### 2011-06-01 tex/coverletter.tex
versions as submitted to Nature

%        File: coverletter.tex
%     Created: Tue May 31 08:00 PM 2011 P
% Last Change: Tue May 31 08:00 PM 2011 P
%
\documentclass[a4paper]{letter}
\begin{document}
\address{Carl Boettiger \\
Department of Ecology and Evolution \\
University of California \\
One Shields Avenue, \\
Davis CA 95616\\
Fax: (530) 752-3350\\
Carl Boettiger: cboettig@ucdavis.edu  \\
Alan Hastings: amhastings@ucdavis.edu}
\telephone{(530) 752-8116} 
\signature{Carl B and Alan H}
\begin{letter}{ }
\opening{To the editor,}
Enclosed is a manuscript we are submitting as a Letter to Nature.  
We believe that this paper is ideally suited for publication in Nature as it attacks a very important problem -
detecting critical transitions or “tipping points” – that ranges across many fields.  
This work represents a fundamental advance in this problem by providing the first analysis of when early detection will and will not be reliable.
We discover that prevailing proposals based on correlations in summary statistics are highly unreliable on typical data,
but propose a model-based approach which can substantially reduce the fraction false alarms and missed detections.
By quantifying error rates, identifying in what data detection may be probable,
and providing a more powerful tool for detection, 
we believe we have made key steps towards making early warning of sudden shifts possible.

For Nature’s readers:
The ability to predict sudden changes in system state prior to their occurrence is of importance in many fields, but particularly in ecology and environmental science where anthropogenic activities at multiple scales are threatening many systems.  The possibility of doing so without a specific model of a system has been emphasized in a series of papers based essentially on ideas from bifurcations in dynamical systems.  Our work provides the first statistical approach to this problem that properly takes into account issues that arise from considering a single realization and for the first time accurately quantifies both the probability of false alarms and missing impending transitions.

For the general public:
The ability to predict sudden changes in system prior to their occurrence is of importance in many fields, but particularly in ecology and environmental science.  Any forecasting is not without uncertainty; effective management decisions must quantify risks in order to weight the cost of missing an early warning signal against the cost of a false alarm.  Our work provides the first treatment of this uncertainty, and also demonstrating a novel approach for dramatically reducing the error rates.

We would like to recommend the following individuals as potential reviewers: Dr. Otso Ovaskainen, Dr. Benjamin Bolker, Dr. Stephen Ellner, Dr. N. Thompson Hobbs, Chris Wikle, Dr. Anthony Ives

\closing{Sincerely,}

\encl{ Text: 1496 words (Microsoft Word format) \\
Figures: 3 items, (jpeg format), layout as single figure, shown in pdf.  \\
Supplementary Information (pdf) Contains: 12 sections, 19 pages, 7 figures. \\
Data and source code: all data and code used in the analysis is provided as an R package, a link to the package is provided in the manuscript and in the supplement.  
}

\end{letter}
\end{document}

### 2011-06-01 appendix/appendix.tex
coverletter should be text, not a file (for Science)

%        File: figures.tex
%     Created: Thu Mar 24 12:00 PM 2011 P
% Last Change: Thu Mar 24 12:00 PM 2011 P
%

\documentclass[authoryear, preprint,review,12pt]{elsarticle}
%\usepackage{endfloat}
%\documentclass[final,5p,twocolumn]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{lineno}
\linenumbers
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

%\graphicspath{{/home/cboettig/Documents/ucdavis/research/phylotrees/images/}}

\journal{Nature} 
\begin{document}
\begin{frontmatter}
\title{Quantifying Limits to Detection of Early Warning for Critical Transitions  }
\author[davis]{Carl Boettiger\corref{cor1}}
\ead{cboettig@ucdavis.edu}
\author[davis]{Alan M Hastings}
%\author[davis]{}
\cortext[cor1]{Corresponding author.}
\address[davis]{Center for Population Biology, University of California, Davis, United States}

%\begin{abstract}
%\end{abstract}

%\begin{keyword}
%\sep 
%\end{keyword}
\end{frontmatter}
%\include{figures.tex}
%input{figures.tex}
%\include{body.tex}
%\input{body.tex}

\appendix

\section{The data sets}
All data sets and analyses performed are included in the accompanying R package.
The simulated data sets in Figure 1a of the text are produced with 100 data points
sampled evenly sampled over a time interval (0,100) using the linearized models directly, see~\ref{modelderivations}.
The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
The constant data is produced using the simulation method for the constant environment model~\eqref{OU},
with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system. 
The Glaciation data comes from~\citet{Petit1999}, accessible from NOAA:
\href{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}.
The data is preprocessed by interpoplation and detrending to be as consistent as possible with the analysis presented in~\citet{Dakos2008},
analyzing the third glaciation event. 
The processed data consists of 121 sample points. 
The match is not exact since~\citet{Dakos2008} estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
The code for processing the data from the original~\citet{Petit1999} file is included in the accompanying R package.
The Algae example comes from the chemostat ``H6'' in the experiments in \emph{Daphnia magna} populations of~\citet{Drake2010}. 
The analysis in that paper focuses on data averaged over the ensembles, though the supplement includes analysis over individual replicates. 
This individual replicates was chosen as an example of a single replicate 
that showed a statistically significant correlation in variance over window of time where critical slowing down was expected. 
Further empirical examples that have been previously investigated for early warning signs are also analyzed under our approach in~\ref{examples}.

\section{Averaging over ensembles vs averaging over time}\label{correlations}
The approach of detecting early warning signals by summary statistics is motivated by the ensemble dynamics of a stochastic system.  If we had an infinite number of identical, replicate systems, we could average across replicates to obtain asymptotic estimates of the (for instance) the variance and autocorrelation of the stochastic process $X_t$ at each time $t$, 
\begin{align}
V(X_t) &= \int X_t^2 P(X_t) \ud X_t - \int X_t P(X_t) \ud X_t \\
\text{Autocorr}(X_{\tau}) &= \int (X_t - X_{t-\tau}) P(X_t) \ud X_t 
\end{align}
and so forth for higher moments.  
All of the integrals are over replicates, $X_t$.  
To apply the theory to a single replicate and instead compute the statistics over time windows one must assert that the system is ergodic
--roughly, independent of the time over which it is measured~\citep{Gardiner2009}.  
This cannot be true if the system is experiencing a gradual loss of stability leading to a critical transition -- but may hold  
approximately if the change is gradual relative to the window, and the window is long relative to the relaxation time of the system.
Then if the moments are calculated across the ensemble at sufficiently wide sampling intervals, one may argue that they are approximately independent.  
Only under these assumptions can we apply the approach of summary statistics to a single replicate.  
Even then, we would need techniques from time-series analysis test, 
since we have not satisfied the assumptions of the standard correlations tests.  
Much of the same data are used to compute sequential values of the moments, hence sequential values are temporally correlated.

This lack of independence violates the assumptions under which the $p$-values associated with typical correlation tests are calculated.
For instance the rank correlations such as Kendall's $\tau$ test a null hypothesis that increases in the $V(X_t)$, etc.
are independent of increases in time, which is not satisfied even when the environment is constant and $X_t$ is stationary.
Calculating $p$-values for Pearson's test needs parametric assumptions such as normal distributed variables as well as independence to define the null distribution.
As both the stationary and the deteriorating environment violate the assumptions of the null, with sufficient data both appear significant.
Consequently the $p$-values seen in Figure 1a and common in the current approaches cannot be interpreted as the probability a warning signal would be observed by chance in a stable system.

It is possible to define a $p$-value for these correlation statistics by estimating a null distribution consistent with the null hypothesis, as we do through the model-based Monte Carlo estimation of these distributions in Figure 1b of the text, (Type I error).

\section{Error Types}\label{errortypes}
The $p$-values given in Figure 1a of the text do not correspond to the probabilities 
that the measured value of the statistical test (Kendall's $\tau$) or greater would be observed under a stable system (previous section, \ref{correlations}). 
The approach illustrated in Figure 1b estimates the appropriate null distribution of $\tau$ 
for each warning statistic (variance, autocorrelation, etc.) under the null hypothesis that the system is stable, 
%using Monte Carlo parametric bootstrap approach, section~\ref{parametricbootstrap} 
following the assumption of the stable model defined in the text.\footnote{
While other stable models are possible, this model corresponds to the underlying bifurcation theory behind the design of early warning signals.
Many systems may have dynamics that fall entirely outside the scope of early warning signals~\cite{Hastings2010}. 
Handling such systems as may have dynamics that correspond poorly to any of the models treated here can be addressed by tests of model adequacy, see~\ref{modelchoice}.}

The appropriate $p$-value for the test statistic can be determined directly from this distribution by integrating to the right of the observed value.
This determines the probability of a false alarm, Type I error.
We are fortunate to be able to estimate the distribution of the test statistic under the alternative hypothesis that the system is stable (red distributions),
and hence can also estimate the power of the test, defined at the accepted false-positive rate.
Having accepted a false-positive rate determines a position beyond which only less than that fraction of the null distribution lies. 
We can ask what fraction of the test distribution lies beyond this fraction as well.
If the false-positive rate is set very low, we lose power. 
The fraction of the test distribution lying to the other side of this accepted rate (1 minus the power) is the probability of a missed event, Type II error.
There is an obvious trade-off between the threshold set for type I error and the resulting Type II error.

Note that while the Type I error is dependent directly on the value observed in the data (black triangles of Figure 1b \& 1c),
the Type II error depends only on the distributions themselves (though it should be remembered that these are also estimated from the data by means of the model fitting).

While it is customary in the scientific literature to accept 5\% false-alarm rate,
we emphasize here that this is properly a choice set by management,
under consideration of the relative costs of false alarms and missed events.
In particular, having an estimate of power is very useful in selecting this rate --
if the probability of a missed event is quite high at a 5\% false alarm rate,
it may make sense to use a less conservative estimate of when to act.
We further note that the best management decisions will be based on the probabilities of both outcomes,
rather than forcing a binary decision~\citep{Brozovic2011}.
For this reason we emphasize reporting the entire distributions in Figure 1b \& 1c, rather than only summary information about the error rates.

\section{Model Derivations}\label{modelderivations}
The sudden collapses we wish to detect are driven by bifurcations when an eigenvalue changes sign in the corresponding deterministic model,
resulting in the sudden loss of a stable state and the onset of a critical transition.
There are many possible models that can contain such a transition.
These may be highly nonlinear models with multiple attractors.
The diversity of possible models can be classified by the type of bifurcation,
such as the saddle-node bifurcation in which a stable point (node) and unstable point (saddle) collide and annihilate,
or a trans-critical bifurcation, in which a stable point becomes unstable.  
(Any dynamical systems textbook, such as \citep{Guckenheimer1983} can give a good overview of such models and bifurcations). 
In general we will be unable to estimate such models accurately if the system has only been observed near the favorable attractor (i.e. before the critical transition occurs),
even if we knew the functional form of the model.  Fortunately the detection of early warning is based on a \emph{local} theory, predicting changes in the dynamics near the current attractor.
Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model, but only the linearized model near the bifurcation.

Near the region of the bifurcation,
we can use the simple normal form equation for the bifurcation rather than the full model,
but these are still non-linear and difficult to estimate accurately from the data.
Critical slowing down is based on taking this one step further -- looking at the eigenvalue around the stable point.
This is equivalent to linearizing the normal forms of the bifurcations around the stable point.
By so doing we obtain a simple but general models we can estimate from data.
We are left with explicit, stochastic models that reflect the behavior of the more complicated systems around the stable point, and we can see how they change as they approach the bifurcation.

\subsection{Saddle-node bifurcation}
The most frequently considered bifurcation model in early warning signals is the saddle-node bifurcation.  
In this bifurcation, a gradual decline is interrupted by a transition when a stable node and saddle node collide and annihilate, 
leaving the system to return to another attractor, if one exists.  
Because this can occur while the system is far from the other attractor, 
this transition appears particularly sudden and has been a focal concern~\citep{Scheffer2001, Scheffer2009}.

Many models with alternative stable states considered in ecological systems can exhibit this kind of bifurcation.  For instance, a common model is one with a saturating birth rate (such as Holling Type III) and linear death rate, 
\begin{equation}
dX_t = \left( \frac{e K x^2}{X^2 + h_t^2} - e X_t - a_t\right) dt + \sigma \sqrt{ \frac{e K x^2}{X^2 + h_t^2} + e X_t + a_t} dB_t \label{ass}
\end{equation}
This model can experience a saddle-node bifurcation through increased mortality by increasing either $e$ or $a$.  This parameterization is only an example, of course many others are possible~\citep{Scheffer2009a, Scheffer2001, Strogatz2001a, Guckenheimer1983}.  
In this case we have specified the noise dependence on $X_t$ explicitly for the influence of demographic (intrinsic) noise.
The saddle-node bifurcation has normal form
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
Linearizing this model differs from the transcritical bifurcation, since the location of stable point moves as the bifurcation parameter changes.
Transforming the canonical form to allow for an arbitrary mean $\theta$,  the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$, which gives our second model, LSN: 
\begin{equation}
\ud X = 2\sqrt{ r_t } (\phi - X_t)\ud t + \sigma\sqrt{\phi } \ud B_t. \label{LSN}
\end{equation}

\subsection{Transcritical Bifurcation}
\citet{Drake2010} induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} for purpose of testing early warning signals~\citep{Drake2010}. The transcritical bifurcation

For instance, consider the stochastic Levin's model~\citep{Levins1969} (a logistic model), where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent) colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).  
\begin{equation}
\ud X_t = \left( c_t X_t (1-X_t/K) - e_t X_t \right) \ud t + \sigma \sqrt{\frac{e_t}{c_t}} \ud B_t \label{levins},
\end{equation}
such a model can be derived from an individual-based description with or without a stochastic environment, i.e.  \citep{Kampen2007a, Nisbet2004a}.  The model contains a transcritical bifurcation when $c_t < e_t$.  The normal form of the bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t x - x^2 
\label{transcritical}
\end{equation}

We can rewrite the deterministic part of the Levins model, Eq~\ref{levins} in normal form without loss of generality by taking $r_t = 1 - e_t/c_t$.  
We can linearize around the stable equilibrium value $\hat x(t) = 1 - e_t/c_t = r_t$, 
\begin{align}
\dot x &=  f(x) \approx f'(x)|_{\hat x} (x - \hat x) + \ldots, \\
 &= (r_t - 2 x_t|_{x=\hat x}) (x - \hat x), \\
 &= r_t(r_t - x).
\end{align}
Note the equilibrium time dependence arises not from the internal dynamics but from the changing parameter values.  We can express the linearized stochastic dynamics then by the time dependent mean-reverting process.  We refer to this model as the linearized transcritical bifurcation.

\begin{equation}
\ud X_t = r_t (r_t - X_t/K) \ud t + \sigma \sqrt{1+r_t} \ud B_t \label{LTC}
\end{equation}

The pattern of the saddle-node differs from that of the transcritical in the square-root dependence on the bifurcation parameter.

\subsection{Modeling limitations}
Many models may differ substantially from~\eqref{ass} and~\eqref{levins} and yet correspond to the linearizations illustrated here.  
The linearizations have thus been constructed with enough free parameters to capture the gross features of the system,
(scaling the equilibrium position, variance, and timescale) while corresponding to the approximations of normal forms of the bifurcations.  
Despite this, there are severe limitations in formulating even a linear approximation without ample knowledge of the system. 
Note that specifying this model such that likelihoods can be calculated still requires a specification of the rate of change of $r_t$.
We assume a linear model,
\begin{equation}
r_t = r_0 - m t
\label{R_t}
\end{equation}
though the approach could easily be extended to an arbitrary model.
A gradual change can be approximately linear over the time interval of interest, and is appropriate to distinguish from a stable model.
Nonlinear rates of change in the environmental conditions will generally be harder to detect (either by summary statistics or the likelihood approach),
thus this assumption is consistent with testing the best case scenario for these methods, illustrating the intrinsic limits of these approaches.

We also draw attention to the fact that noise scales with the mean dynamics.  
For intrinsically stochastic processes, such as demographic stochasticity, this will scale as the square root, 
while for external perturbations to the system dynamics (environmental stochasticity) this would scale linearly.
We have assumed an intrinsic noise process in both models~\eqref{LSN} and~\eqref{LTC}, 
though an alternative model with linear scaling would also be plausible.  
Note that in these models, the variance observed is determined by the ratio of the Brownian terms (in front of $\ud B_t)$
to the stabilizing force (in front of $\ud t$), so that in general larger $r_t$ will correspond to smaller system variances
regardless of this smaller correction due to the noise scaling with the mean dynamics.  
Either should perform better than the stable model if a system is slowly losing stability,
though using the better-matching model will increase the power of the approach.

The models here are neither more nor less general than the seemingly model-free approach of summary statistics.  
We have proposed these models to match the assumptions of the summary statistics -- 
that a system is approaching a bifurcation in which a node loses stability as an eigenvalue passes through zero.

\subsection{Stable Model}
When the system is stable, $r_t$ is constant and both models~\eqref{LSN} and~\eqref{LTC} reduce to a simple Ornstein-Uhlenbeck process, 
\begin{equation}
\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
\end{equation}

\section{Likelihood calculations}\label{likelihood}
The linearization above is not only justified by the nature of and the properties we wish to estimate from the data,
but also convenient for the calculations. 
In particular, the models are specified by linear stochastic differential equations, hence their solutions are Gaussian.
The fundamental challenge is that in modeling a gradual loss of stability, the solutions are also explicitly time-dependent.
Model fitting and comparison requires we can write down the likelihood expression for each model given the time-series data.

The probability $P(M|X)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
\begin{equation}
\log P(M | X) =  \sum_i \log P(x_i | x_{i-1}, t_i)
\end{equation}
As the processes are Gaussian, the probability density $P$ is normally distributed and we need only calculate its first and second moments.  For the OU process (OU process) this is trivial:

\begin{align}
E(x_i) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
V(x_i) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
\label{OUsoln}
\end{align}

For the time dependent models, we have analytic forms only for the dynamical equations of these moments, which we must integrate numerically over each time interval.  This is responsible for most of the computational effort required of this approach.  For the linearized transcritical bifurcation:

\begin{align}
\frac{\ud }{\ud t} E(x_i)&=  r(t)(r(t) - x_i) \\
\frac{\ud}{\ud t} V(x_i) &=  -2 r(t) V(x_i) + (1+r(t))\sigma^2 
\label{LTCsoln}
\end{align}

For the linearized saddle-node bifurcation:
\begin{align}
\frac{\ud }{\ud t} E(x_i)&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
\frac{\ud}{\ud t} V(x_i) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
\label{LSNsoln}
\end{align}

These are numerically integrated using \texttt{lsoda} routine for the likelihood calculation.

\section{The likelihood statistic}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

The hypotheses considered here are more challenging then the original lemma, as they are composite in nature:
they specify two model forms (stable and changing stability)
but with model parameters that must be first estimated from the data.
Comparing models whose parameters have been estimated by maximum likelihood is first treated by~\citet{Cox1961, Cox1962},
and has since been developed in this simulation estimation of the null distribution~\citep{McLachlan1987}, by parametric bootstrap estimate~\citep{Efron1987}.  
Cox's $\delta$ statistic is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.

Let $L_0$ be the likelihood function for model 0, let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; and define $L_1$ , $\theta_1$ , $L_1$ similarly for model 1. The statistic we will use is $\delta$, defined to be twice the difference in log likelihood of observing the data under the two MLE models,
$\delta = -2 (\log L_0 - \log L_1 )$

This approach has since been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996}.  
We have extended the approach by generating the test distribution as well as a null distribution, allowing estimate of power and Type II error.

\section{A note on information criteria}\label{aic}
While it would be possible to compare these models directly by a simple test such as the Akaike Information criterion to compare such models, this would be to attempt to answer the question statistically that should be a management decision -- whether to treat the data as indicative of an impending critical transition -- while leaving unanswered the statistical question of the relative risks of missed events and false alarms.  Using the distributions we compute in Figure 1c, it is easy to assess the performance an AIC-based decision criteria would have, Figure~\ref{aic}.  Without the approach described here, these error rates would be unknown.  Given the distributions, it is clear that the there would be no need to accept the AIC criterion.  For instance, in the empirical data, third panel of Figure~\ref{aic}, it would make sense to use a much more stringent threshold, thereby decreasing the rate of false alarms without any substantial increase in the rate of missed events.  When power of detection is very low such, as in the middle panel, error rates can be particularly high.  Low power could result from only a very gradual loss of power or from inadequate sampling, which can be estimated by comparing parameter distributions, see section~\ref{parameterdistributions}.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{aic.pdf}
\end{center}
\caption{False alarm and missed detection rate that result from an AIC decision criterion.}
\label{fig:aic}
\end{figure}

\section{Parameter Distributions}\label{parameterdistributions}
% Detecting weak signals
The approach taken in Figure 1c provides a powerful way to indicate when a warning signal can be detected, but more care must be taken with the case in which a signal is not detected.  It will always be easier to say a system is deteriorating (rejecting the null hypothesis that a system is stable) then to establish that a system is stable, as we cannot rule out that we have missed the detection for lack of power.  Thus the left and right panels of Figure 1c make for clear cases of destabilizing systems, while more care must be taken in drawing a conclusion about the middle panel.

Can we argue that the middle system is truly stable, or could it be experiencing a very gradual loss of stability? The lack of power could be the result of inadequate amount of data or the result of a very gradual change that is difficult to distinguish from  a stable system.  In such cases, the distribution of the estimated stability loss parameter can help establish a lack of power (but potentially rapid rate of stability-loss) from a gradual loss.  Figure~\ref{fig:par} shows an example of this for the case of the simulated loss of stability and the simulated constant environment.  When estimating a changing-stability model on the constant environment data, the distribution for the rate of change is tightly and symmetrically distributed around zero, suggesting that the lack of a signal from this data in the likelihood test is not simply due to a lack of power to detect a very gradual change.  As this data is simulated under a constant environment, this is exactly what we would expect to find given adequate sampling.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{parameter_dist.pdf}
\end{center}
\caption{.}
\label{fig:par}
\end{figure}

Distributions of all parameter estimates are obtained without extra computational cost by the likelihood approach, and are reported by our package.  Note that this parameter $m$ represents linearized or average rate of loss, and does not preclude the possibility of the rate of deterioration accelerating.

\section{Model choice and model adequacy}\label{modelchoice}
Any approach is only as good as its underlying assumptions.  In this paper we have leveraged the machinery of likelihood-based approaches to both evaluate the performance of existing methods and provide a statistical test that more closely matches the assumptions of the basic theory.  For this approach to be useful, the data must still be well-approximated by the models used.  We have provided two different simple bifurcation models that could each represent the hypothesis of gradual decline in stability: linearized transcritical bifurcation and the linearized saddle-node bifurcation; Equations~\eqref{LTC},~\eqref{LSN}.  As they both represent a gradual loss of stability either can be used, but the test will be most powerful by using the one that best corresponds to the data.  While knowledge of the system alone can suggest this (such as in the case of~\citet{Drake2010}), it is straight forward to compare the models against each other using the same likelihood framework, as illustrated in Figure~\ref{fig:modelchoice} for several empirical examples.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=.25\textwidth]{deut_modelchoice.png}
\includegraphics[width=.25\textwidth]{caco3_modelchoice.png}
\end{center}
\caption{Model }
\label{fig:modelchoice}
\end{figure}

As discussed in the text, many systems may not correspond well to any of the simple dynamics upon which the detection of early warning signals is based~\citep{Hastings2010}.  Such dynamics can be particularly misleading for current methods based on summary statistics.  The Monte Carlo bootstrapping of the likelihood approach can be applied as a test of model adequacy, as illustrated in \citet{Goldman1993} and discussed in \citet{Sullivan2005b} in a phylogenetics context. If the observed likelihood ratio falls far from the distributions estimated under either model, the model is likely an inadequate description and subsequent inference may be misleading.

% Choosing between models

\section{R package tutorial}
We provide an R package with simple implementation of the methods described here.  The package is available from: \href{https://github.com/cboettig/warningsignals/archives/master}{https://github.com/cboettig/warningsignals/archives/master}, where it will be actively maintained and developed.  The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood analysis illustrated in Figure 1c:

\begin{verbatim}
> require(warningsignals)
>
> # Load a sample dataset
> data(glaciationIII)
>
> # Fit constant (OU) model and a stability-loss model (LSN);
> models <- fit_models(glaciationIII, 'LSN')
>
> results <- montecarlotest(models$const, models$timedep, n=2000, cpu=16)
> plot(results)
\end{verbatim}

The package can generate traditional warning signals (as in Figure 1a) and bootstraps of their uncertainty (as in Figure 1b) for a variety of correlation tests (Pearson's, Kendall's, Spearman's) and summary statistics (variance, autocorrelation, skew) as well.  Copies of data used in the paper are provided as examples.  The Monte Carlo methods can take advantage of multiple core processors or clusters by specifying the number of CPUs in the function argument.   Full documentation of functions and data are provided in the package.

\section{Examples}\label{examples}
In this section we illustrate the main results of the paper in three further empirical datasets from the climate record.  
\begin{figure}[hb]
\begin{center}
\includegraphics[width=\textwidth]{appendix_fig1}
\end{center}
\caption{Examples for three more empirical datasets from climate records, analysed for warning signals in~\citet{Dakos2008}.  Calcium carbonate data originally from~\citet{Tripati2005}, Deuterium concentrations in the periods before major ice ages from~\citet{Petit1999}.}
\label{fig:1a}
\end{figure}

\begin{figure}
\begin{center}
%\includegraphics[width=.8\textwidth]{figure2_appendix.pdf}
\includegraphics{appendix_fig2}
\end{center}
\caption{Bootstraps of the correlations observed in Figure~\ref{fig:1a}, presented as in Figure 1b in the text.  The distribution of the test statistic under the null (blue) and test (red) hypotheses are difficult to distinguish reliably, across all indicators for each data set.}
\label{fig:1b}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics{appendix_fig3}
%\includegraphics[width=.8\textwidth]{figure3_appendix.pdf}
\end{center}
\caption{Bootstraps of likelihood ratios for the data observed in Figure~\ref{fig:1a}, presented as in Figure 1c in the text.  The distribution of the test statistic under the null (blue) and test (red) hypotheses are more clearly distinguished in this approach, supporting the evidence for a warning signal.}
\label{fig:1c}
\end{figure}

\section{Comparisons for additional summary statistics}
Higher moments have also been proposed for the detection of warning signals, such as the skewness~\citep{Guttal2008a}, illustrated for the datasets presented in the text in Figure~\ref{fig:1}.  The skewness appears to have even less power than the other indicators discussed in the text.

\begin{figure}[ht]
\begin{center}
\subfigure[ ]{\includegraphics{Boettiger_fig1_appendix.jpg}}
\subfigure[ ] {\includegraphics{Boettiger_fig2_appendix.jpg}}
\end{center}
\caption{Increasing skewness is no more likely under the stable hypothesis than the stability-loss hypothesis for each of these data.}
\label{fig:1}
\end{figure}

\pagebreak

\section*{ }%bibliography
\bibliography{/home/cboettig/Documents/bibliographies/library.bib}

\end{document}

### 2011-06-30 appendix/appendix.tex
a day of writing

@@
 \section{Model Derivations}\label{modelderivations}
+
 The sudden collapses we wish to detect are driven by bifurcations when an eigenvalue changes sign in the corresponding deterministic model,

### 2011-07-05 appendix/appendix.tex
draft sent to Alan, edits on appendix

--- 2011-06-01 appendix/appendix.tex
@@
 
-\section{The data sets}
+\section{Method Summary}
+For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see Section~\ref{modelderivations}, and Section~\ref{likelihood} for the corresponding likelihood equations.)  The models are stochastic equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.
+Distributions are determined by a parametric bootstrap: 500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
+For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation) were computed over a sliding window equal to half the observed time, and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.
+The estimates of $\tau$ under the replicate simulations from the stable system and the estimates under the deteriorating stability system make up the two distributions which are integrated at different thresholds to create the ROC curves.
+
+The likelihood test is performed as in the parametric bootstrap; but instead of calculating summary-statistics, each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores, (Cox's delta statistic, see Section~\ref{Cox}).
+Note that steps such as simulating under either model or estimating each of the indicator statistics are much faster than estimating the model parameters by maximum likelihood, and that estimating the deteriorating stability model is substantially more intensive than estimating the stable model which has a closed form solution for the likelihood expression, see~\citet{likelihood}.
+While in the summary-statistic approach outlined above this is estimated only once, for the likelihood approach this must be estimated 1000 times (once on each of the 500 replicates from the stable simulation and the 500 replicates from the deteriorating stability simulation), making this approach nearly 1000 times slower.
+Refitting the models on each replicate, rather merely computing the likelihood under the original estimates, is essential to the power of this approach~\citep{Huelsenbeck1996}.⁠
+
+\section{The data sets}\label{data}
 All data sets and analyses performed are included in the accompanying R package.
-The simulated data sets in Figure 1a of the text are produced with 100 data points
-sampled evenly sampled over a time interval (0,100) using the linearized models directly, see~\ref{modelderivations}.
+The simulated data sets are produced with 40 data points
+sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, see~\ref{modelderivations}.
 The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
@@
 with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system.
+
 The Glaciation data comes from~\citet{Petit1999}, accessible from NOAA:
@@
 The code for processing the data from the original~\citet{Petit1999} file is included in the accompanying R package.
-The Algae example comes from the chemostat ``H6'' in the experiments in \emph{Daphnia magna} populations of~\citet{Drake2010}. 
+The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments in \emph{Daphnia magna} populations of~\citet{Drake2010}. 
 The analysis in that paper focuses on data averaged over the ensembles, though the supplement includes analysis over individual replicates. 
@@
 Further empirical examples that have been previously investigated for early warning signs are also analyzed under our approach in~\ref{examples}.
-
-\section{Averaging over ensembles vs averaging over time}\label{correlations}
-The approach of detecting early warning signals by summary statistics is motivated by the ensemble dynamics of a stochastic system.
-If we had an infinite number of identical, replicate systems, we could average across replicates to obtain asymptotic estimates of the (for instance) the variance and autocorrelation of the stochastic process $X_t$ at each time $t$,
-\begin{align}
-V(X_t) &= \int X_t^2 P(X_t) \ud X_t - \int X_t P(X_t) \ud X_t \\
-\text{Autocorr}(X_{\tau}) &= \int (X_t - X_{t-\tau}) P(X_t) \ud X_t 
-\end{align}
-and so forth for higher moments.  
-All of the integrals are over replicates, $X_t$.  
-To apply the theory to a single replicate and instead compute the statistics over time windows one must assert that the system is ergodic
---roughly, independent of the time over which it is measured~\citep{Gardiner2009}.  
-This cannot be true if the system is experiencing a gradual loss of stability leading to a critical transition -- but may hold  
-approximately if the change is gradual relative to the window, and the window is long relative to the relaxation time of the system.
-Then if the moments are calculated across the ensemble at sufficiently wide sampling intervals, one may argue that they are approximately independent.  
-Only under these assumptions can we apply the approach of summary statistics to a single replicate.  
-Even then, we would need techniques from time-series analysis test, 
-since we have not satisfied the assumptions of the standard correlations tests.  
-Much of the same data are used to compute sequential values of the moments, hence sequential values are temporally correlated.
-
-This lack of independence violates the assumptions under which the $p$-values associated with typical correlation tests are calculated.
-For instance the rank correlations such as Kendall's $\tau$ test a null hypothesis that increases in the $V(X_t)$, etc.
-are independent of increases in time, which is not satisfied even when the environment is constant and $X_t$ is stationary.
-Calculating $p$-values for Pearson's test needs parametric assumptions such as normal distributed variables as well as independence to define the null distribution.
-As both the stationary and the deteriorating environment violate the assumptions of the null, with sufficient data both appear significant.
-Consequently the $p$-values seen in Figure 1a and common in the current approaches cannot be interpreted as the probability a warning signal would be observed by chance in a stable system.
-
-It is possible to define a $p$-value for these correlation statistics by estimating a null distribution consistent with the null hypothesis, as we do through the model-based Monte Carlo estimation of these distributions in Figure 1b of the text, (Type I error).
-
-\section{Error Types}\label{errortypes}
-The $p$-values given in Figure 1a of the text do not correspond to the probabilities 
-that the measured value of the statistical test (Kendall's $\tau$) or greater would be observed under a stable system (previous section, \ref{correlations}). 
-The approach illustrated in Figure 1b estimates the appropriate null distribution of $\tau$ 
-for each warning statistic (variance, autocorrelation, etc.) under the null hypothesis that the system is stable, 
-%using Monte Carlo parametric bootstrap approach, section~\ref{parametricbootstrap} 
-following the assumption of the stable model defined in the text.\footnote{
-While other stable models are possible, this model corresponds to the underlying bifurcation theory behind the design of early warning signals.
-Many systems may have dynamics that fall entirely outside the scope of early warning signals~\cite{Hastings2010}. 
-Handling such systems as may have dynamics that correspond poorly to any of the models treated here can be addressed by tests of model adequacy, see~\ref{modelchoice}.}
-
-The appropriate $p$-value for the test statistic can be determined directly from this distribution by integrating to the right of the observed value.
-This determines the probability of a false alarm, Type I error.
-We are fortunate to be able to estimate the distribution of the test statistic under the alternative hypothesis that the system is stable (red distributions),
-and hence can also estimate the power of the test, defined at the accepted false-positive rate.
-Having accepted a false-positive rate determines a position beyond which only less than that fraction of the null distribution lies. 
-We can ask what fraction of the test distribution lies beyond this fraction as well.
-If the false-positive rate is set very low, we lose power. 
-The fraction of the test distribution lying to the other side of this accepted rate (1 minus the power) is the probability of a missed event, Type II error.
-There is an obvious trade-off between the threshold set for type I error and the resulting Type II error.
-
-Note that while the Type I error is dependent directly on the value observed in the data (black triangles of Figure 1b \& 1c),
-the Type II error depends only on the distributions themselves (though it should be remembered that these are also estimated from the data by means of the model fitting).
-
-While it is customary in the scientific literature to accept 5\% false-alarm rate,
-we emphasize here that this is properly a choice set by management,
-under consideration of the relative costs of false alarms and missed events.
-In particular, having an estimate of power is very useful in selecting this rate --
-if the probability of a missed event is quite high at a 5\% false alarm rate,
-it may make sense to use a less conservative estimate of when to act.
-We further note that the best management decisions will be based on the probabilities of both outcomes,
-rather than forcing a binary decision~\citep{Brozovic2011}.
-For this reason we emphasize reporting the entire distributions in Figure 1b \& 1c, rather than only summary information about the error rates.
 
@@
 or a trans-critical bifurcation, in which a stable point becomes unstable.  
-(Any dynamical systems textbook, such as \citep{Guckenheimer1983} can give a good overview of such models and bifurcations). 
-In general we will be unable to estimate such models accurately if the system has only been observed near the favorable attractor (i.e. before the critical transition occurs),
-even if we knew the functional form of the model.  Fortunately the detection of early warning is based on a \emph{local} theory, predicting changes in the dynamics near the current attractor.
-Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model, but only the linearized model near the bifurcation.
+(Any dynamical systems textbook, such as \citet{Guckenheimer1983} can give a good overview of such models and bifurcations). 
+In general we will be unable to estimate such models accurately if the system has only been observed near the favorable attractor
+(i.e. before the critical transition occurs),
+even if we knew the functional form of the model.
+Fortunately the detection of early warning is based on a \emph{local} theory,
+predicting changes in the dynamics near the current attractor.
+Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
+but only the linearized model near the bifurcation.
 
@@
 By so doing we obtain a simple but general models we can estimate from data.
-We are left with explicit, stochastic models that reflect the behavior of the more complicated systems around the stable point, and we can see how they change as they approach the bifurcation.
+We are left with explicit, stochastic models that reflect the behavior of the more complicated systems around the stable point,
+and we can see how they change as they approach the bifurcation.
+
+Routines for fitting each model by maximum likelihood and simulating under each model are provided in the accompanying \texttt{warningsignals} \texttt{R} package.
 
@@
 \end{equation}
-Linearizing this model differs from the transcritical bifurcation, since the location of stable point moves as the bifurcation parameter changes.
-Transforming the canonical form to allow for an arbitrary mean $\theta$,  the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$, which gives our second model, LSN:
+Linearizing this model differs from the transcritical bifurcation,
+since the location of stable point moves as the bifurcation parameter changes.
+Transforming the canonical form to allow for an arbitrary mean $\theta$,
+the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$,
+which gives the model we refer to as the Linearized Saddle Node, LSN: 
 \begin{equation}
@@
 \subsection{Transcritical Bifurcation}
-\citet{Drake2010} induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} for purpose of testing early warning signals~\citep{Drake2010}. The transcritical bifurcation
-
-For instance, consider the stochastic Levin's model~\citep{Levins1969} (a logistic model), where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent) colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).
+Another simple bifurcation that may be common in ecological systems with alternative stable states is the transcritical bifurcation; for instance, \citet{Drake2010} induce such a bifurcation in laboratory populations of \emph{Daphnia magna} for purpose of testing early warning signals~\citep{Drake2010}.
+To see how this bifurcation arises, consider for example, the stochastic Levin's model~\citep{Levins1969} (a logistic model), where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent) colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).
 \begin{equation}
@@
 We can express the linearized stochastic dynamics then by the time dependent mean-reverting process.
-We refer to this model as the linearized transcritical bifurcation.
+We refer to this model as the linearized transcritical bifurcation, LTC.
 
@@
 \end{equation}
+This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere (\emph{e.g.} \citet{Dakos2008, Guttal2008a}).
 
@@
 
-The probability $P(M|X)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
+The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
 \begin{equation}
-\log P(M | X) =  \sum_i \log P(x_i | x_{i-1}, t_i)
+\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
 \end{equation}
-As the processes are Gaussian, the probability density $P$ is normally distributed and we need only calculate its first and second moments.  For the OU process (OU process) this is trivial:
+As the processes are Gaussian, the probability density $P$ is normally distributed and we need only calculate its first and second moments (see, \emph{e.g.}~\citep{Gardiner2009}).
+For any process of the form
+
+\begin{equation}
+  \ud X_t = f(X_t) \ud t + g(X_t) \ud B_t 
+  \label{general}
+\end{equation}
+
+where $f(x)$ and $g(x)$ are linear functions of $x$, the solution is Gaussian with moments evolving according the the coupled ordinary differential equations:
 
 \begin{align}
-E(x_i) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
-V(x_i) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
+ \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
+\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
+  \label{general_moments}
+\end{align}
+
+For the OU process, we can solve this in closed form,~\citep{Gardiner2009}.
+
+\begin{align}
+  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
+V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
 \label{OUsoln}
@@
 
-For the time dependent models, we have analytic forms only for the dynamical equations of these moments, which we must integrate numerically over each time interval.
+For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
 This is responsible for most of the computational effort required of this approach.
@@
 \begin{align}
-\frac{\ud }{\ud t} E(x_i)&=  r(t)(r(t) - x_i) \\
-\frac{\ud}{\ud t} V(x_i) &=  -2 r(t) V(x_i) + (1+r(t))\sigma^2 
+\frac{\ud }{\ud t} E(x_i| M = \text{LTC})&=  r(t)(r(t) - x_i) \\
+\frac{\ud}{\ud t} V(x_i| M = \text{LTC}) &=  -2 r(t) V(x_i|M) + (1+r(t))\sigma^2 
 \label{LTCsoln}
@@
 
-These are numerically integrated using \texttt{lsoda} routine for the likelihood calculation.
+These are numerically integrated using \texttt{lsoda} routine avialble in \texttt{R} for the likelihood calculation.
 
-\section{The likelihood statistic}
+\section{The likelihood statistic}\label{Cox}
 Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
@@
 We have extended the approach by generating the test distribution as well as a null distribution, allowing estimate of power and Type II error.
-
-\section{A note on information criteria}\label{aic}
-While it would be possible to compare these models directly by a simple test such as the Akaike Information criterion to compare such models, this would be to attempt to answer the question statistically that should be a management decision -- whether to treat the data as indicative of an impending critical transition -- while leaving unanswered the statistical question of the relative risks of missed events and false alarms.
-Using the distributions we compute in Figure 1c, it is easy to assess the performance an AIC-based decision criteria would have, Figure~\ref{aic}.
-Without the approach described here, these error rates would be unknown.
-Given the distributions, it is clear that the there would be no need to accept the AIC criterion.
-For instance, in the empirical data, third panel of Figure~\ref{aic}, it would make sense to use a much more stringent threshold, thereby decreasing the rate of false alarms without any substantial increase in the rate of missed events.
-When power of detection is very low such, as in the middle panel, error rates can be particularly high.
-Low power could result from only a very gradual loss of power or from inadequate sampling, which can be estimated by comparing parameter distributions, see section~\ref{parameterdistributions}.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{aic.pdf}
-\end{center}
-\caption{False alarm and missed detection rate that result from an AIC decision criterion.}
-\label{fig:aic}
-\end{figure}
-
-\section{Parameter Distributions}\label{parameterdistributions}
-% Detecting weak signals
-The approach taken in Figure 1c provides a powerful way to indicate when a warning signal can be detected, but more care must be taken with the case in which a signal is not detected.
-It will always be easier to say a system is deteriorating (rejecting the null hypothesis that a system is stable) then to establish that a system is stable, as we cannot rule out that we have missed the detection for lack of power.
-Thus the left and right panels of Figure 1c make for clear cases of destabilizing systems, while more care must be taken in drawing a conclusion about the middle panel.
-
-Can we argue that the middle system is truly stable, or could it be experiencing a very gradual loss of stability?
-The lack of power could be the result of inadequate amount of data or the result of a very gradual change that is difficult to distinguish from  a stable system.
-In such cases, the distribution of the estimated stability loss parameter can help establish a lack of power (but potentially rapid rate of stability-loss) from a gradual loss.
-Figure~\ref{fig:par} shows an example of this for the case of the simulated loss of stability and the simulated constant environment.
-When estimating a changing-stability model on the constant environment data, the distribution for the rate of change is tightly and symmetrically distributed around zero, suggesting that the lack of a signal from this data in the likelihood test is not simply due to a lack of power to detect a very gradual change.
-As this data is simulated under a constant environment, this is exactly what we would expect to find given adequate sampling.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\textwidth]{parameter_dist.pdf}
-\end{center}
-\caption{.}
-\label{fig:par}
-\end{figure}
-
-Distributions of all parameter estimates are obtained without extra computational cost by the likelihood approach, and are reported by our package.
-Note that this parameter $m$ represents linearized or average rate of loss, and does not preclude the possibility of the rate of deterioration accelerating.
 
@@
 The package is available from: \href{https://github.com/cboettig/warningsignals/archives/master}{https://github.com/cboettig/warningsignals/archives/master}, where it will be actively maintained and developed.
-The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood analysis illustrated in Figure 1c:
+The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood-based analysis:
 
@@
 
-The package can generate traditional warning signals (as in Figure 1a) and bootstraps of their uncertainty (as in Figure 1b) for a variety of correlation tests (Pearson's, Kendall's, Spearman's) and summary statistics (variance, autocorrelation, skew) as well.
+The package also performs the computationally faster but less powerful summary-statistic tests, computes the resulting distributions under the different hypotheses and the corresponding ROC curves.
+The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coeffient of variation).
 Copies of data used in the paper are provided as examples.
 The Monte Carlo methods can take advantage of multiple core processors or clusters by specifying the number of CPUs in the function argument.
-Full documentation of functions and data are provided in the package.
+Full documentation of functions and data are provided in the package, as well as example scripts which replicate the analyses presented in the paper.
 
@@
 In this section we illustrate the main results of the paper in three further empirical datasets from the climate record.
-\begin{figure}[hb]
-\begin{center}
-\includegraphics[width=\textwidth]{appendix_fig1}
-\end{center}
-\caption{Examples for three more empirical datasets from climate records, analysed for warning signals in~\citet{Dakos2008}.
-Calcium carbonate data originally from~\citet{Tripati2005}, Deuterium concentrations in the periods before major ice ages from~\citet{Petit1999}.}
-\label{fig:1a}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-%\includegraphics[width=.8\textwidth]{figure2_appendix.pdf}
-\includegraphics{appendix_fig2}
-\end{center}
-\caption{Bootstraps of the correlations observed in Figure~\ref{fig:1a}, presented as in Figure 1b in the text.
-The distribution of the test statistic under the null (blue) and test (red) hypotheses are difficult to distinguish reliably, across all indicators for each data set.}
-\label{fig:1b}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{appendix_fig3}
-%\includegraphics[width=.8\textwidth]{figure3_appendix.pdf}
-\end{center}
-\caption{Bootstraps of likelihood ratios for the data observed in Figure~\ref{fig:1a}, presented as in Figure 1c in the text.
-The distribution of the test statistic under the null (blue) and test (red) hypotheses are more clearly distinguished in this approach, supporting the evidence for a warning signal.}
-\label{fig:1c}
-\end{figure}
-
-\section{Comparisons for additional summary statistics}
-Higher moments have also been proposed for the detection of warning signals, such as the skewness~\citep{Guttal2008a}, illustrated for the datasets presented in the text in Figure~\ref{fig:1}.
-The skewness appears to have even less power than the other indicators discussed in the text.
-
-\begin{figure}[ht]
-\begin{center}
-\subfigure[ ]{\includegraphics{Boettiger_fig1_appendix.jpg}}
-\subfigure[ ] {\includegraphics{Boettiger_fig2_appendix.jpg}}
-\end{center}
-\caption{Increasing skewness is no more likely under the stable hypothesis than the stability-loss hypothesis for each of these data.}
-\label{fig:1}
-\end{figure}
 

### 2011-07-08 appendix/appendix.tex
minor updates to appendix

@@
 
-The likelihood test is performed as in the parametric bootstrap; but instead of calculating summary-statistics, each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores, (Cox's delta statistic, see Section~\ref{Cox}).
-Note that steps such as simulating under either model or estimating each of the indicator statistics are much faster than estimating the model parameters by maximum likelihood, and that estimating the deteriorating stability model is substantially more intensive than estimating the stable model which has a closed form solution for the likelihood expression, see~\citet{likelihood}.
+The likelihood test is performed as in the parametric bootstrap; but instead of calculating summary-statistics, each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores, (Cox's $\delta$ statistic, see Section~\ref{Cox}).
+Note that steps such as simulating under either model or estimating each of the indicator statistics are much faster than estimating the model parameters by maximum likelihood, and that estimating the deteriorating stability model is substantially more intensive than estimating the stable model which has a closed form solution for the likelihood expression, see~\ref{likelihood}.
 While in the summary-statistic approach outlined above this is estimated only once, for the likelihood approach this must be estimated 1000 times (once on each of the 500 replicates from the stable simulation and the 500 replicates from the deteriorating stability simulation), making this approach nearly 1000 times slower.
-Refitting the models on each replicate, rather merely computing the likelihood under the original estimates, is essential to the power of this approach~\citep{Huelsenbeck1996}.⁠
+Refitting the models on each replicate, rather merely computing the likelihood under the original estimates, is essential to the power of this approach~\citep{Huelsenbeck1996}.
+
+When replicates are created by simulation under the estimated models, we record data at the same time intervals observed in the original data (as in Fig.
+3, main text).
+It is straight forward to allow this interval to be different, representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve.
+The warning statistics are then estimated on each set of simulated data as before.
 
@@
 sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, see~\ref{modelderivations}.
-The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
-The constant data is produced using the simulation method for the constant environment model~\eqref{OU},
-with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system.
+The data was then fit to the LSN model (see~\ref{modelderivations}) to compute the ROC curves.
+%The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
+%The constant data is produced using the simulation method for the constant environment model~\eqref{OU},
+%with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system.
 
-The Glaciation data comes from~\citet{Petit1999}, accessible from NOAA:
+The Glaciation data comes from~\citet{Petit1999}, accessible from NOAA:\\
 \href{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}.
@@
 The code for processing the data from the original~\citet{Petit1999} file is included in the accompanying R package.
+
 The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments in \emph{Daphnia magna} populations of~\citet{Drake2010}. 
@@
 This approach has since been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996}.  
-We have extended the approach by generating the test distribution as well as a null distribution, allowing estimate of power and Type II error.
+We have extended the approach by generating the test distribution as well as a null distribution in order to compute ROC curves.
 

### 2011-07-17 appendix/appendix.tex
some edits on appendix, finished synthesis of feedback from Marissa, Marcel, Alistair, Alex

@@
 
+\section{On a hypothesis-testing framework}
+It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
+As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.
+In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.
+By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.
+
+The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of its approaches.
+While the $p$-values presented in the main text are based on parametric assumptions not met by the data, the rather different $p$ values presented in appendix table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$ 2 and H$_0$ 3 are very similar to our stable system, though H$_0$ 1 has lost the correlation structure expected in any time-series).
+The plots in Fig.
+3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.
+Our approach replicate these findings for the $\tau$-statistic.
+We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.
+
+The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
+As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.
+Fig.
+S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.
+The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.
+The reality is worse, we find it overlaps almost completely.
+We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.
+
+The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''  %A statistical test is meaningless without at least some notion of power.
+As it is perhaps the linguistics of hypothesis testing as much as the quantitative difficulty of the analysis that has led this step to be entirely ignored (though we admit some surprise at the extent to which the question of significance or false alarms has also been ignored), we hope
+
 \section{Model choice and model adequacy}\label{modelchoice}
@@
 As they both represent a gradual loss of stability either can be used, but the test will be most powerful by using the one that best corresponds to the data.
-While knowledge of the system alone can suggest this (such as in the case of~\citet{Drake2010}), it is straight forward to compare the models against each other using the same likelihood framework, as illustrated in Figure~\ref{fig:modelchoice} for several empirical examples.
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=.25\textwidth]{deut_modelchoice.png}
-\includegraphics[width=.25\textwidth]{caco3_modelchoice.png}
-\end{center}
-\caption{Model }
-\label{fig:modelchoice}
-\end{figure}
+While knowledge of the system alone can suggest this (such as in the case of~\citet{Drake2010}), it is straight forward to compare the models against each other using the same likelihood framework.
 
@@
 If the observed likelihood ratio falls far from the distributions estimated under either model, the model is likely an inadequate description and subsequent inference may be misleading.
-
-% Choosing between models
 

### 2011-07-21 appendix/appendix.tex
minor edits from Alan

@@
 \usepackage{amsmath,amsfonts}
-\usepackage{subfigure}
+\usepackage{subfigure, verbatim}
 \usepackage[pdftex]{color}
@@
 
-\journal{Nature} 
+\journal{Science} 
 \begin{document}
 \begin{frontmatter}
-\title{Quantifying Limits to Detection of Early Warning for Critical Transitions  }
+\title{Quantifying Limits to Detection of Early Warning for Critical Transitions: Supplementary Online Materials}
 \author[davis]{Carl Boettiger\corref{cor1}}
@@
 
-%\begin{abstract}
-%\end{abstract}
-
-%\begin{keyword}
-%\sep 
-%\end{keyword}
 \end{frontmatter}
-%\include{figures.tex}
-%input{figures.tex}
-%\include{body.tex}
-%\input{body.tex}
 
@@
 \section{Method Summary}
-For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see Section~\ref{modelderivations}, and Section~\ref{likelihood} for the corresponding likelihood equations.)  The models are stochastic equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.
+For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see~\ref{modelderivations} and~\ref{likelihood} for the corresponding likelihood equations).
+The models are stochastic equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.
 Distributions are determined by a parametric bootstrap: 500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
@@
 
+In addition to the descriptions provided here, the source code for implementing each of the analyses and generating the plots presented is provided as an R package,~\ref{R}.
+Further details on the implementation can be found in the function documentation.
+
 \section{The data sets}\label{data}
-All data sets and analyses performed are included in the accompanying R package.
+All data sets and analyses performed are included in the accompanying R package, described in~\ref{R}.
 The simulated data sets are produced with 40 data points
 sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, see~\ref{modelderivations}.
-The data was then fit to the LSN model (see~\ref{modelderivations}) to compute the ROC curves.
+The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.
+Code to run this individual-based simulation under a specified parameter values is provided in the \texttt{R} package.
+See \verb|ibm_sims.R| demo for the example used here.
 %The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
@@
 \href{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}.
-The data is preprocessed by interpoplation and detrending to be as consistent as possible with the analysis presented in~\citet{Dakos2008},
+The data is preprocessed by interpolation and de-trending to be as consistent as possible with the analysis presented in~\citet{Dakos2008},
 analyzing the third glaciation event. 
@@
 
-where $f(x)$ and $g(x)$ are linear functions of $x$, the solution is Gaussian with moments evolving according the the coupled ordinary differential equations:
+where $f(x)$ and $g(x)$ are linear functions of $x$, the solution is Gaussian with moments evolving according the coupled ordinary differential equations:
 
@@
 
-These are numerically integrated using \texttt{lsoda} routine avialble in \texttt{R} for the likelihood calculation.
+These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.
 
-\section{The likelihood statistic}\label{Cox}
+\section{Comparing Models}\label{Cox}
 Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
@@
 We have extended the approach by generating the test distribution as well as a null distribution in order to compute ROC curves.
+
+\pagebreak
+\section{The distributions for the ROC curves}
+Figure 1 (main text) shows hypothetical distributions of a test statistic under a stable system and a system approaching a critical transition in order to illustrate how an ROC curve is calculated.
+Figure~\ref{fig:3dist} shows the distributions used to generate the ROC curves in Figure 3 of the main text.
+
+\begin{figure}[!ht]
+  \begin{center}
+    \includegraphics[width=.6\linewidth]{3dist}
+  \end{center}
+  \caption{The distributions underlying the ROC curves presented in Figure 3 of the main text.
+As in Figure 1 (main text), the blue distribution corresponds to the stable model, while the red distribution corresponds to the model of a system approaching a critical transition.
+Both models are estimated from the data by maximum likelihood.
+The top panel shows the likelihood statistic, where the $x$-axis is the likelihood ratio $\delta$.
+The increase in the various summary statistics is characterized by an Kendall's $\tau$.
+The value of the statistic ($\tau$, $\delta$) estimated from the data set is shown by the triangle.
+Highly overlapping distributions correspond to ROC curves near the 1:1 line.}
+  \label{fig:3dist}
+\end{figure}
 
@@
 
-The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''  %A statistical test is meaningless without at least some notion of power.
-As it is perhaps the linguistics of hypothesis testing as much as the quantitative difficulty of the analysis that has led this step to be entirely ignored (though we admit some surprise at the extent to which the question of significance or false alarms has also been ignored), we hope
+The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
 
@@
 
-\section{R package tutorial}
+\section{R package tutorial}\label{R}
 We provide an R package with simple implementation of the methods described here.
@@
 The package also performs the computationally faster but less powerful summary-statistic tests, computes the resulting distributions under the different hypotheses and the corresponding ROC curves.
-The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coeffient of variation).
+The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coefficient of variation).
 Copies of data used in the paper are provided as examples.
@@
 \section{Examples}\label{examples}
-In this section we illustrate the main results of the paper in three further empirical datasets from the climate record.
+In this section we illustrate the main results of the paper in several further empirical datasets from the climate record, Figs~\ref{fig:a3}, \ref{fig:a4}, and~\ref{fig:a3dist}.
 
-\pagebreak
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[width=\linewidth]{a3}
+  \end{center}
+  \caption{ROC curves for two additional data sets also analyzed in~\citet{Dakos2008}.
+The area under each curve (AUC) is given in the legend.
+(a) The end of Greenhouse Earth period 34 million years ago may be detected in the 482 observations of CaCO$_3$ concentrations~\citep{Tripati2005}.
+Despite the strong increase in autocorrelation suggested by the estimate of $\tau$, (see Fig~\ref{fig:a3dist}), the autocorrelation statistic performs little better than chance (AUC = 0.54).
+The variance in this signal actually decreases, causing the warning signal based on variance to perform substantially worse than chance.
+Despite this, the likelihood approach still distinguishes reliably between the estimated deteriorating and stable hypotheses.
+(b) Deuterium levels provide an early warning of the transition into the first ice age (Glaciation I) that appears in the time series from~\citep{Petit1999} (N=591).
+Autocorrelation performs better than the other indicators, but again the likelihood approach is most reliable.
+Figure~\ref{fig:a3dist} shows the corresponding distributions used to calculate the ROC curves.}
+  \label{fig:a3}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[width=\linewidth]{a4}
+  \end{center}
+  \caption{ROC curves generated on sub-sampled data for the additional data sets.
+The signal in the CaCO$_3$ data, Greenhouse Earth transition, is stronger than that in the deuterium data (Glaciation transition), making detection possible with fewer data points.}
+  \label{fig:a4}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics[width=\linewidth]{a3dist}
+  \end{center}
+  \caption{The distributions corresponding to the ROC curves in Fig~\ref{fig:a3}.
+The top panel shows the likelihood statistic, where the $x$-axis is the likelihood ratio $\delta$.
+The increase in the various summary statistics is characterized by an Kendall's $\tau$.
+The value of the statistic ($\tau$, $\delta$) estimated from the data set is shown by the triangle.
+}
+  \label{fig:a3dist}
+\end{figure}
 

### 2011-08-02 appendix/appendix.tex
updated copy sending to alan, para 2 & 3 still in trouble

@@
 \author[davis]{Alan M Hastings}
-%\author[davis]{}
 \cortext[cor1]{Corresponding author.}
 \address[davis]{Center for Population Biology, University of California, Davis, United States}
-
 \end{frontmatter}
@@
 \appendix
+\renewcommand*\thefigure{S\arabic{figure}}
+\renewcommand*\theequation{S\arabic{equation}}
 
@@
 \subsection{Transcritical Bifurcation}
-Another simple bifurcation that may be common in ecological systems with alternative stable states is the transcritical bifurcation; for instance, \citet{Drake2010} induce such a bifurcation in laboratory populations of \emph{Daphnia magna} for purpose of testing early warning signals~\citep{Drake2010}.
+Another simple bifurcation that may be common in ecological systems is the transcritical bifurcation; for instance, \citet{Drake2010} induce such a bifurcation in laboratory populations of \emph{Daphnia magna} for purpose of testing early warning signals~\citep{Drake2010}.
 To see how this bifurcation arises, consider for example, the stochastic Levin's model~\citep{Levins1969} (a logistic model), where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent) colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).
@@
 \begin{align}
-\dot x &=  f(x) \approx f'(x)|_{\hat x} (x - \hat x) + \ldots, \\
- &= (r_t - 2 x_t|_{x=\hat x}) (x - \hat x), \\
- &= r_t(r_t - x).
+\dot x &=  f(x) \approx f'(x)|_{\hat x} (x - \hat x) + \ldots, \nonumber \\
+ &= (r_t - 2 x_t|_{x=\hat x}) (x - \hat x), \nonumber\\
+ &= r_t(r_t - x). \nonumber
 \end{align}
@@
 The pattern of the saddle-node differs from that of the transcritical in the square-root dependence on the bifurcation parameter.
+We note that the transcritical bifurcation involves a gradual decline towards zero, whereas the saddle-node bifurcation involves a sudden transition from a non-trivial equilibrium point.
+Thus saddle-node bifurcations are the more typical focus of early warning efforts, even though both exhibit the phenomenon of critical slowing down.
+We include this example for the \emph{Daphnia} data from~\citet{Drake2010}, showing how our approach could be applied in that scenario.
 
@@
 
-Let $L_0$ be the likelihood function for model 0, let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; and define $L_1$ , $\theta_1$ , $L_1$ similarly for model 1.
-The statistic we will use is $\delta$, defined to be twice the difference in log likelihood of observing the data under the two MLE models,
-$\delta = -2 (\log L_0 - \log L_1 )$
-
-This approach has since been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996}.  
-We have extended the approach by generating the test distribution as well as a null distribution in order to compute ROC curves.
+Let $L_0$ be the likelihood function for model 0, 
+let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
+be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
+and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
+The statistic we will use is $\delta$, 
+defined to be twice the difference in log likelihood of observing the data under the two MLE models,
+$\delta = -2 (\log L_0 - \log L_1 )$.  
+This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
+We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
+Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.
 
@@
 The value of the statistic ($\tau$, $\delta$) estimated from the data set is shown by the triangle.
-Highly overlapping distributions correspond to ROC curves near the 1:1 line.}
+Highly overlapping distributions correspond to ROC curves near the 1:1 line in Figure 3.}
   \label{fig:3dist}
@@
 The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of its approaches.
-While the $p$-values presented in the main text are based on parametric assumptions not met by the data, the rather different $p$ values presented in appendix table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$ 2 and H$_0$ 3 are very similar to our stable system, though H$_0$ 1 has lost the correlation structure expected in any time-series).
+While the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data, the rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
 The plots in Fig.
@@
 The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
-
-\section{Model choice and model adequacy}\label{modelchoice}
-Any approach is only as good as its underlying assumptions.
-In this paper we have leveraged the machinery of likelihood-based approaches to both evaluate the performance of existing methods and provide a statistical test that more closely matches the assumptions of the basic theory.
-For this approach to be useful, the data must still be well-approximated by the models used.
-We have provided two different simple bifurcation models that could each represent the hypothesis of gradual decline in stability: linearized transcritical bifurcation and the linearized saddle-node bifurcation; Equations~\eqref{LTC},~\eqref{LSN}.
-As they both represent a gradual loss of stability either can be used, but the test will be most powerful by using the one that best corresponds to the data.
-While knowledge of the system alone can suggest this (such as in the case of~\citet{Drake2010}), it is straight forward to compare the models against each other using the same likelihood framework.
-
-As discussed in the text, many systems may not correspond well to any of the simple dynamics upon which the detection of early warning signals is based~\citep{Hastings2010}.
-Such dynamics can be particularly misleading for current methods based on summary statistics.
-The Monte Carlo bootstrapping of the likelihood approach can be applied as a test of model adequacy, as illustrated in \citet{Goldman1993} and discussed in \citet{Sullivan2005b} in a phylogenetics context.
-If the observed likelihood ratio falls far from the distributions estimated under either model, the model is likely an inadequate description and subsequent inference may be misleading.
 

### 2011-08-02 appendix/appendix.tex
submitted to SCIENCE

@@
 \section{Method Summary}
-For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see~\ref{modelderivations} and~\ref{likelihood} for the corresponding likelihood equations).
-The models are stochastic equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.
+For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see~\ref{modelderivations} and~\ref{likelihood} for the corresponding likelihood expressions).
+The models are stochastic differential equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.
 Distributions are determined by a parametric bootstrap: 500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
@@
 See \verb|ibm_sims.R| demo for the example used here.
-%The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
-%The constant data is produced using the simulation method for the constant environment model~\eqref{OU},
-%with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system.
 
@@
 such as the saddle-node bifurcation in which a stable point (node) and unstable point (saddle) collide and annihilate,
-or a trans-critical bifurcation, in which a stable point becomes unstable.  
-(Any dynamical systems textbook, such as \citet{Guckenheimer1983} can give a good overview of such models and bifurcations). 
+or a transcritical bifurcation, in which a stable point becomes unstable.  
+(A dynamical systems text such as \citet{Guckenheimer1983} will give a good overview of such bifurcations). 
 In general we will be unable to estimate such models accurately if the system has only been observed near the favorable attractor
@@
 
-Near the region of the bifurcation,
-we can use the simple normal form equation for the bifurcation rather than the full model,
-but these are still non-linear and difficult to estimate accurately from the data.
-Critical slowing down is based on taking this one step further -- looking at the eigenvalue around the stable point.
-This is equivalent to linearizing the normal forms of the bifurcations around the stable point.
+The theory of critical slowing down (\emph{e.g.} \citet{Scheffer2009}) is based on taking this one step further -- 
+looking at the eigenvalue around the stable point.
+This is equivalent to linearizing the bifurcations around the stable point.
 By so doing we obtain a simple but general models we can estimate from data.
@@
 
-\section{On a hypothesis-testing framework}
+\section{On a hypothesis-testing framework}\label{Dakos}
 It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
@@
 The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of its approaches.
-While the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data, the rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
+We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.
+For this reason we generate the null distribution directly from the model as outlined above.
+The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
 The plots in Fig.

### 2011-10-11 boettiger.tex
PNAS format in tex format

%        File: boettiger.tex
%     Created: Mon Oct 10 04:00 PM 2011 P
% Last Change: Mon Oct 10 04:00 PM 2011 P
%%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BASIC CLASS FILE 
%% PNAStwo for two column articles is called by default.
%% Uncomment PNASone for single column articles. One column class
%% and style files are available upon request from pnas@nas.edu.
%% (uncomment means get rid of the '%' in front of the command)

%\documentclass{pnasone}
\documentclass{pnastwo}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Changing position of text on physical page:
%% Since not all printers position
%% the printed page in the same place on the physical page,
%% you can change the position yourself here, if you need to:

% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
                         %  physical page; positive dimension will lower it.

                         %% You may set the dimension to the size that you need.

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% OPTIONAL GRAPHICS STYLE FILE

                         %% Requires graphics style file (graphicx.sty), used for inserting
                         %% .eps files into LaTeX articles.
                         %% Note that inclusion of .eps files is for your reference only;
                         %% when submitting to PNAS please submit figures separately.

                         %% Type into the square brackets the name of the driver program 
                         %% that you are using. If you don't know, try dvips, which is the
                         %% most common PC driver, or textures for the Mac. These are the options:

                         % [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
                         % [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
                         % [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]

                         \usepackage[pdftex]{graphicx}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% OPTIONAL POSTSCRIPT FONT FILES

                         %% PostScript font files: You may need to edit the PNASoneF.sty
                         %% or PNAStwoF.sty file to make the font names match those on your system. 
                         %% Alternatively, you can leave the font style file commands commented out
                         %% and typeset your article using the default Computer Modern 
                         %% fonts (recommended). If accepted, your article will be typeset
                         %% at PNAS using PostScript fonts.

                         % Choose PNASoneF for one column; PNAStwoF for two column:
                         %\usepackage{PNASoneF}
                         %\usepackage{pnastwof}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% ADDITIONAL OPTIONAL STYLE FILES

                         %% The AMS math files are commonly used to gain access to useful features
                         %% like extended math fonts and math commands.

                         \usepackage{amssymb,amsfonts,amsmath}
                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% OPTIONAL MACRO FILES
                         %% Insert self-defined macros here.
                         %% \newcommand definitions are recommended; \def definitions are supported

                         %\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
                         %\def\s{\sigma}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         %% Don't type in anything in the following section:
                         %%%%%%%%%%%%
                         %% For PNAS Only:
                         \contributor{Submitted to Proceedings
                         of the National Academy of Sciences of the United States of America}
                         \url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
                         \copyrightyear{2008}
                         \issuedate{Issue Date}
                         \volume{Volume}
                         \issuenumber{Issue Number}
                         %%%%%%%%%%%%

                         \begin{document}

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

                         %% For titles, only capitalize the first letter
                         %% \title{Almost sharp fronts for the surface quasi-geostrophic equation}

                         \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}

                         %% Enter authors via the \author command.  
                         %% Use \affil to define affiliations.
                         %% (Leave no spaces between author name and \affil command)

                         %% Note that the \thanks{} command has been disabled in favor of
                         %% a generic, reserved space for PNAS publication footnotes.

                         %% \author{<author name>
                         %% \affil{<number>}{<Institution>}} One number for each institution.
                         %% The same number should be used for authors that
                         %% are affiliated with the same institution, after the first time
                         %% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
                         %% Then, before last author ...
                         %% \and
                         %% \author{<author name>
                         %% \affil{<number>}{}}

                         %% For example, assuming Garcia and Sonnery are both affiliated with
                         %% Universidad de Murcia:
                         %% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
                         %% United Kingdom},
                         %% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
                         %% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}

                         \author{Carl Boettiger\affil{1}{Center for Population Biology, University of California, Davis}, 
                         Alan Hastings\affil{2}{Department of Environmental Science and Policy, University of California, Davis} }

                         \contributor{Submitted to Proceedings of the National Academy of Sciences
                         of the United States of America}

                         %% The \maketitle command is necessary to build the title page.
                         \maketitle

                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
                         \begin{article}

                           \begin{abstract}

Catastrophic regime shifts in complex natural systems may be averted through advanced detection. 
Recent work has provided a proof-of-principle that many systems approaching a catastrophic transition look different 
through the lens of early warning indicators such as rising variance or increased return times.  
Without the benefit of hindsight of a collapse, 
applications of these approaches must quantify how reliable different indicators are in avoiding false alarms, 
and how sensitive they are to missing subtle warning signs.  
We propose an approach which quantifies this trade-off between reliability and sensitivity, 
allows comparisons between different indicators, 
and estimates what data are necessary to achieve a desired error rate.  
We also introduce a likelihood-based indicator that is more sensitive than previously proposed summary statistics.

                           \end{abstract}

                             %% When adding keywords, separate each term with a straight line: |
                             \keywords{ early warning signals | tipping point | alternative stable states | likelihood methods }

                             %% Optional for entering abbreviations, separate the abbreviation from
                             %% its definition with a comma, separate each pair with a semicolon:
                             %% for example:
                             %% \abbreviations{SAM, self-assembled monolayer; OTS,
                             %% octadecyltrichlorosilane}

                             % \abbreviations{}

                             %% The first letter of the article should be drop cap: \dropcap{}
                             %\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts

                             %% Enter the text of your article beginning here and ending before
                             %% \begin{acknowledgements}
                             %% Section head commands for your reference:
                             %% \section{}
                             %% \subsection{}
                             %% \subsubsection{}

\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠. 
Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011} , and semi-arid ecosystems~\cite{Kefi2007}⁠. 
Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
it is important to develop the ability to predict impending regime shifts based on early warning signs. 
With a good model of the system, system-specific approaches could be useful~\cite{Lenton2009}.
but in cases where good models are not available general approaches are needed~\cite{Scheffer2009}⁠.
We focus on the common scenario of gradual change yielding a sudden shift or bifurcation~\cite{Scheffer2001, Scheffer2009}⁠
We describe a novel statistical approach that is appropriate for the goal of informing management.

Though previous research has identified patterns such as increasing variance or autocorrelation in time-series approaching a critical transition,
there has been no method for forecasting the probability of a collapse from a natural system.
As real-world applications of early warning signals will not be able to point to the same differences between controlled and
manipulated systems that have been demonstrated in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
and fewer still that determine the distribution of that quantity in the event that the system is not approaching a transition --
the reliability or false alarm rate~\cite{Dakos2008}⁠. 
None have considered the distribution of such a statistic in the event that a system is approaching a transition -- 
the sensitivity to avoid missed detection.  
We need to quantify such uncertainty/error rates to compare between possible methods~\cite{Contamin2009}
and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.

Detection of impending transitions is not just a scientific question and the goal should not solely be to detect with a 5\% chance of a false positive,
and a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
with which it will fail to provide any advanced warning. 
Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
In this paper, we
(1) present the first measurements of reliability and sensitivity for early-warning indicators, 
(2) we use this method to estimate the amount of data required to achieve a given level of confidence,  
(3) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of an insensitive method 
(4) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.

We illustrate the trade-off between false alarms and failed detection using receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠. 
The curves represent the corresponding false alarm rate at any detection sensitivity,  Fig. 1.
The closer these distributions are to one-another, the more severe the trade-off.  
If the distributions overlap exactly, the ROC curve has a constant slope of unity.  
The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
Different early-warning indicators will vary in their sensitivity to detect differences between stable systems and those approaching a critical transition, making the ROC curves a natural way to compare their performance.  
Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
The area under the ROC curve provides a summary of how well the method compares to a random guess --
an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance, while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.

To generate ROC curves for common early warning indicators such as 
increases in autocorrelation, variance, and skew, 
we must quantify the increase and generate the expected distributions of the focal statistic.  
While the existing literature frequently presents only a 
visual increase in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
we rely on the few examples that quantify an increasing trend using Kendall's tau correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
Values of tau near unity indicate a strongly increasing trend in the warning indicator – suggestive of an approaching transition, while values near zero suggest a lack of a trend – characteristic of stable systems.  In principle, other quantifications of increasing variance, autocorrelation, etc, may perform better than tau and could be analyzed just as well in this framework.  While we use this established metric of increase, we also introduce a novel warning indicator that is not based on the gradual increase of a summary statistic but on likelihood comparisons.

Having identified an indicator statistic, we must be able to generate its distribution under both the scenario of a stable system and one approaching a transition before we can compute the ROC curve.  We fit general stochastic models representing common bifurcations (e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010} ) for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each\footnote[1]{See methods.}⁠.  By estimating the statistic (e.g. tau) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves\footnote[1]{See methods.}⁠.  Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of tau, which provides a more direct and more powerful test than detecting an increase in summary statistics.  As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.

\section{Examples}

We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
We explore two empirical examples:
deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\cite{Dakos2008},
and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\cite{Drake2010}.  
For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
and the same system in a constant environment. 
The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
we need a way to identify such cases where we lack the data to trust the pattern.

The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system.  
The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems.  
The best-performing summary statistic would, at a 5\% false positive rate, have a 10\% chance of catching a true positive 
in both the Daphnia and Glaciation data set.  
Meanwhile, our likelihood-based approach performs substantially better, 
having respectively a 30\% in Daphnia and 100\% chance in Glaciation set of catching a true positive at a 5\% false positive rate. 
Because we have the entire ROC curve, we are not restricted to an evaluation at a fixed false-positive rate such as 5\%, 
but can accept more false-positives to avoid missing as many impending catastrophes.  
This is particularly valuable in the context of critical transitions, 
where data are often limited and the consequences of missed detection can be far more severe than false alarms.

\section{Discussion}
A pressing question in the application of early warning signals is quantifying when the data are sufficient for detection to be reliable~\cite{Scheffer2009, Scheffer2010, Inman2011}. 
Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
for each of the indicators (Fig 4). 
Using our generic models estimated from the original data as in Fig. 3, 
we generate the corresponding distributions by simulating under the models as before, 
but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
would be hampered or improved with more or less data.  
While all methods improve as the number of data points increase, 
the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
as expected~\cite{Carpenter2011};⁠
skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4).

``A best-case scenario.''
Although not all sudden collapses correspond to our model assumptions~\cite{Hastings2010}⁠,
including dynamical systems where the variance actually decreases as a sudden shift approaches \cite{Schreiber2003, Schreiber2008},⁠
using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
but would be amenable to the approach used here.

The conclusion is not simply that likelihood approaches are more reliable, 
but rather more broadly that warning signals should consider the inherent trade-off between sensitivity and accuracy, and must quantify how this trade-off depends on both the indicators used and the data available.  
Using the methods we have outlined when designing early warning strategies for natural systems can ensure that data collection has adequate power to offer a reasonable chance of detection. 
The approach developed here estimates the risk of both missed events and false alarms; concepts which are critical to prediction-based management.

                             %% == end of paper:

                             %% Optional Materials and Methods Section
                             %% The Materials and Methods section header will be added automatically.

                             %% Enter any subheads and the Materials and Methods text below.
                             \begin{materials}
                               \subsection{Models}
                               \subsection{Data sets}
                             \end{materials}

                             %% Optional Appendix or Appendices
                             \appendix

                             %Appendix text...
                             %% or, for appendix with title, use square brackets:
                             %% \appendix[Appendix Title]

                             \begin{acknowledgments}

Author Contributions
C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper.

Acknowledgements
S. Schreiber, M. Holyoak, M. Baskett, and A. Perkins provided comments on the manuscript.  This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.

Data and code available at: https://github.com/cboettig/warningsignals/archives/master. 
                             \end{acknowledgments}

                             %% PNAS does not support submission of supporting .tex files such as BibTeX.
                             %% Instead all references must be included in the article .tex document. 
                             %% If you currently use BibTeX, your bibliography is formed because the 
                             %% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
                             %% .tex document. To conform to PNAS requirements, copy the reference listings
                             %% from your .bbl file and add them to the article .tex file, using the
                             %% bibliography environment described above.

                             %%  Contact pnas@nas.edu if you need assistance with your
                             %%  bibliography.

                             % Sample bibliography item in PNAS format:
                             %% \bibitem{in-text reference} comma-separated author names up to 5,
                             %% for more than 5 authors use first author last name et al. (year published)
                             %% article title  {\it Journal Name} volume #: start page-end page.
                             %% ie,
                             % \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
                             % A short C-terminal sequence is necessary and sufficient for the
                             % targeting of chitinases to the plant vacuole. 
                             % {\it Proc Natl Acad Sci USA} 88:10362-10366.

                             %% Enter the largest bibliography number in the facing curly brackets
                             %% following \begin{thebibliography}

                             \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
                             \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
                           \end{article}
                           %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

                           %% Adding Figure and Table References
                           %% Be sure to add figures and tables after \end{article}
                           %% and before \end{document}

                           %% For figures, put the caption below the illustration.
                           %%
                           %% \begin{figure}
                           %% \caption{Almost Sharp Front}\label{afoto}
                           %% \end{figure}

                           %% For Tables, put caption above table
                           %%
                           %% Table caption should start with a capital letter, continue with lower case
                           %% and not have a period at the end
                           %% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
                           %% keep that much space between every line in the table.

                           %% \begin{table}
                           %% \caption{Repeat length of longer allele by age of onset class}
                           %% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
                           %% table text
                           %% \end{tabular}
                           %% \end{table}

                           %% For two column figures and tables, use the following:

                           %% \begin{figure*}
                           %% \caption{Almost Sharp Front}\label{afoto}
                           %% \end{figure*}

                           %% \begin{table*}
                           %% \caption{Repeat length of longer allele by age of onset class}
                           %% \begin{tabular}{ccc}
                           %% table text
                           %% \end{tabular}
                           %% \end{table*}

                           \end{document}

### 2011-10-11 boettiger.tex
PNAS format doesn't seem to reference figures

@@
 In this paper, we
-(1) present the first measurements of reliability and sensitivity for early-warning indicators, 
-(2) we use this method to estimate the amount of data required to achieve a given level of confidence,  
-(3) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of an insensitive method 
-(4) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.
+(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
+(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
+(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of an insensitive method 
+(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.
 
 We illustrate the trade-off between false alarms and failed detection using receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠.
-The curves represent the corresponding false alarm rate at any detection sensitivity,  Fig. 1.
+The curves represent the corresponding false alarm rate at any detection sensitivity, Fig 1.
 The closer these distributions are to one-another, the more severe the trade-off.  
@@
 we rely on the few examples that quantify an increasing trend using Kendall's tau correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
-Values of tau near unity indicate a strongly increasing trend in the warning indicator – suggestive of an approaching transition, while values near zero suggest a lack of a trend – characteristic of stable systems.
-In principle, other quantifications of increasing variance, autocorrelation, etc, may perform better than tau and could be analyzed just as well in this framework.
-While we use this established metric of increase, we also introduce a novel warning indicator that is not based on the gradual increase of a summary statistic but on likelihood comparisons.
+Values of tau near unity indicate a strongly increasing trend in the warning indicator --
+suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
+In principle, other quantifications of increasing variance, autocorrelation, etc,
+may perform better than tau and could be analyzed just as well in this framework. 
+While we use this established metric of increase, we also introduce a novel warning indicator
+that is not based on the gradual increase of a summary statistic but on likelihood comparisons.
 
-Having identified an indicator statistic, we must be able to generate its distribution under both the scenario of a stable system and one approaching a transition before we can compute the ROC curve.
-We fit general stochastic models representing common bifurcations (e.g.
-saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010} ) for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each\footnote[1]{See methods.}⁠.
-By estimating the statistic (e.g.
-tau) on each replicate we generate distributions such as depicted in Fig.
-1 and compute ROC curves\footnote[1]{See methods.}⁠.
-Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of tau, which provides a more direct and more powerful test than detecting an increase in summary statistics.
+Having identified an indicator statistic,
+we must be able to generate its distribution under both the scenario of a stable system and 
+one approaching a transition before we can compute the ROC curve.
+We fit general stochastic models representing common bifurcations
+(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010} )
+for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each\footnote[1]{See methods.}⁠. 
+By estimating the statistic (e.g. tau) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves\footnote[1]{See methods.}⁠.
+Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of tau,
+which provides a more direct and more powerful test than detecting an increase in summary statistics.
 As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.
@@
                            %% \end{figure}
+ \begin{figure*}
+   \begin{center}
+     \includegraphics[width=\linewidth]{Fig1}
+     \label{fig1}
+     \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
+Bottom row: Points along the ROC curve are calculated for each possible threshold indicated in the top row.
+The false positive rate is the integral of the distribution of the test statistic under the stable system right of the threshold (blue shaded area, corresponding to blue vertical line).
+The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).
+Successive columns show the threshold increasing, tracing out the ROC curve (successively gray lines show the threshold used for the matching gray point added to the ROC curve).}
+  \end{center}
+ \end{figure*}
+
+ \begin{figure*}
+   \begin{center}
+     \includegraphics[width=\linewidth]{Fig2}
+     \label{fig2}
+     \caption{Early warning signals in simulated and empirical data sets.
+The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation due to a gradually deteriorating environment (Deteriorating).
+Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated in a chemostat towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a critical transition in glaciation (Glaciation).
+Increases in summary statistics such as variance, autocorrelation, or skew, computed over a moving window, are used to indicate if a system is moving towards a critical transition (see supplementary online text).
+The increase is measured by the correlation coefficient $\tau$.
+Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
+  \end{center}
+ \end{figure*}
+
+ \begin{figure*}
+   \begin{center}
+     \includegraphics[width=\linewidth]{Fig3}
+     \label{fig3}
+     \caption{ROC curves for different early warning indicators on simulated and empirical data.
+The area under the curve, inset, indicates how the method compares to a random guess (0.5).
+The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data.
+The simulated dataset has 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121.
+See corresponding distributions, Fig.
+S1, and additional examples, Fig.
+S2.}
+  \end{center}
+ \end{figure*}
+
+ \begin{figure*}
+   \begin{center}
+     \includegraphics[width=\linewidth]{Fig4}
+     \label{fig4}
+     \caption{ROC curves (from Fig.
+3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.
+Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.}
+  \end{center}
+ \end{figure*}
 

### 2011-10-12 appendix/appendix.tex
pnas edits

@@
 
-The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of its approaches.
+The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
 We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.

### 2011-10-12 boettiger.tex
pnas edits

@@
                          %\def\s{\sigma}
+
+                         \newcommand{\ud}{\mathrm{d}}
 
@@
 but in cases where good models are not available general approaches are needed~\cite{Scheffer2009}⁠.
-We focus on the common scenario of gradual change yielding a sudden shift or bifurcation~\cite{Scheffer2001, Scheffer2009}⁠
+We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system~\cite{Scheffer2001, Scheffer2009}.
+% hmm, not environmental examples\dots ⁠
 We describe a novel statistical approach that is appropriate for the goal of informing management.
@@
 We illustrate the trade-off between false alarms and failed detection using receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠.
-The curves represent the corresponding false alarm rate at any detection sensitivity, Fig 1.
+The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
 The closer these distributions are to one-another, the more severe the trade-off.  
@@
 While the existing literature frequently presents only a 
-visual increase in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006}, 
-we rely on the few examples that quantify an increasing trend using Kendall's tau correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
-Values of tau near unity indicate a strongly increasing trend in the warning indicator --
+% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
+% so this is not a reasonable statistical characterization
+visual or heuristic measure of increase\footnote{That is, do not provide $p$-values or a statistic for which they could be generated.} in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006},
+we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
+Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
 suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
 In principle, other quantifications of increasing variance, autocorrelation, etc,
-may perform better than tau and could be analyzed just as well in this framework. 
+may perform better than $\tau$ and could be analyzed just as well in this framework. 
 While we use this established metric of increase, we also introduce a novel warning indicator
@@
 (e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010} )
-for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each\footnote[1]{See methods.}⁠. 
-By estimating the statistic (e.g. tau) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves\footnote[1]{See methods.}⁠.
-Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of tau,
+for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
+By estimating the statistic (e.g. $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
+Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of $\tau$,
 which provides a more direct and more powerful test than detecting an increase in summary statistics.
@@
 
+%%%%%%% Converting from power to detection %%%%%%%%%%%%
+
 \section{Discussion}
@@
 
-``A best-case scenario.''
-Although not all sudden collapses correspond to our model assumptions~\cite{Hastings2010}⁠,
-including dynamical systems where the variance actually decreases as a sudden shift approaches \cite{Schreiber2003, Schreiber2008},⁠
+%% Threshold
+%% Imagine a missed detection of no more than 10% is acceptable.
+
+Moreover, this illustrates that the summary statistics are very unreliable indicators. 
+A low false alarm rate is only achieved by poor sensitivity; 
+or conversely sensitivity it only achieved with a high rate of false alarms.  
+This is under assumptions that are highly favorable to these statistics, corresponding directly to the process assumptions.  
+The reason for this comes largely from the difficulty in quantifying an increase without a baseline -- 
+the distribution of tau under the null model covers almost the entire range.  %Ref FIGURE
+
+Our approach of estimating uncertainty and using the likelihood indicator is neither more nor less general than the other indicators we discuss.  
+All rely on the assumption of an underlying bifurcation driven by the principle eigenvalue passing through zero.\footnote{
+Skew, a higher order term, has additional assumption about the asymmetry,
+and is thus generically less powerful than signals on the lower order moments.} 
+These methods are robust to the assumptions of Gaussian noise and linearity of the model,
+as illustrated by the fit to the simulated data, 
+which instead comes from a nonlinear, discrete-state Markov process (see Appendix).  
+The likelihood approach still performs the best on this data, hence it is not less general.
+
+When the underlying bifurcation model does not apply, neither summary statistics nor our model-based description will be useful.
+
+We stress that all of the early warning signals approaches must only be applied when the underlying assumptions can be justified.  
+The summary statistics such as variance or autocorrelation are no more general than our explicit models, % wow
+and systems that do not correspond to these bifurcations can exhibit sudden transitions without any warning~\cite{Hastings2010}.
+For instance, bifurcations resulting in sudden transitions have been observed in ecological models in which 
+the variance actually deceases as a sudden shift approaches~\cite{Schreiber2003, Schreiber2008}.
+
+Meanwhile, if our knowledge of the system gives reason to be concerned that it may experience a saddle-node or transcritical bifurcation,
+
+If we can use increasing variance as an indicator, than we can use the approach described here to quantify uncertainty.
+
+Although not all sudden collapses correspond to our model assumptions,
+including dynamical systems where the variance actually decreases as a sudden shift approaches,⁠
 using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
@@
 
+The approach presented here offer a best-case scenario for these warning signals,
+illustrating that substantial uncertainty can remain despite the appearance of a positive increase.
+Our results highlight the need to quantify uncertainty in the application of early warning signals,
+and suggest that a hypothesis testing approach may be too restrictive a framework for that purpose.
+Using the ROC curve, we obtain a more general picture of the trade-off between false alarms and failed detections
+than can be provided by a significance test at a fixed false-alarm rate.  
+The curve can guide the choice of threshold used to indicate a warning signal on a case-by-case basis.
+
 The conclusion is not simply that likelihood approaches are more reliable, 
-but rather more broadly that warning signals should consider the inherent trade-off between sensitivity and accuracy, and must quantify how this trade-off depends on both the indicators used and the data available.
-Using the methods we have outlined when designing early warning strategies for natural systems can ensure that data collection has adequate power to offer a reasonable chance of detection. 
-The approach developed here estimates the risk of both missed events and false alarms; concepts which are critical to prediction-based management.
+but rather more broadly that warning signals should consider
+the inherent trade-off between sensitivity and accuracy,
+and must quantify how this trade-off depends on both the indicators used and the data available.  
+The approach developed here estimates the risk of both failed detection and false alarms;
+concepts which are critical to prediction-based management.  
+Using the methods we have outlined when designing early warning strategies for natural systems
+can ensure that data collection has adequate power to offer a reasonable chance of detection.
 
@@
                              \begin{materials}
-                               \subsection{Models}
+                               \subsection{Methods}\label{methods}
+For each dataset, parameters are estimated by maximum likelihood for a stable model
+and a model of a system approaching a critical transition (Appendix).
+The models are stochastic differential equations (SDEs)
+that can be used to simulate data under the same sampling regime as the original data, 
+or at any specified set of time points.
+Distributions are determined by a parametric bootstrap:
+500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
+For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation)
+were computed over a sliding window equal to half the observed time (other window sizes had worse performance, see R package), 
+and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.
+The estimates of $\tau$ under the replicate simulations from the stable system
+and the estimates under the deteriorating stability system
+make up the two distributions which are integrated at a sliding threshold to create the ROC curves, as illustrated in Fig 1.
+
+The likelihood test is performed as in the parametric bootstrap; 
+but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\cite{Cox1961}.
+Refitting the models on each replicate, 
+rather merely computing the likelihood under the original estimates,
+makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\cite{Huelsenbeck1996}.
+
+When replicates are created by simulation under the estimated models,
+we record data at the same time intervals observed in the original data (as in Fig. 3, main text). 
+It is straight forward to allow this interval to be different,
+representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve, Fig 4.
+
+The source code for implementing each of the analyses and generating the plots presented is provided as an R package, \verb|warningsignals|.
+
                                \subsection{Data sets}
+The simulated data sets are produced with 40 data points
+sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, Eq~\eqref{master}.
+The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.
+
+The Glaciation data is accessible from NOAA~\cite{Petit1999}.
+The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
+to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\cite{Dakos2008},
+analyzing the third glaciation event, consisting of 121 sample points. 
+The match is not exact since estimates the de-trending window size manually,
+but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
+De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\cite{Dakos2008}, 
+which did not apply to the other data sets~\cite{Drake2010}.
+
+The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\cite{Drake2010}. 
+This individual replicates was chosen as an example of a single replicate 
+that showed a statistically significant correlation in variance over window of time where critical slowing down was expected. 
+Further empirical examples that have been previously investigated for early warning signs are also analyzed under our approach in SI Text.
+
+All data and code for simulations and data processing are found in the R package.
+
                              \end{materials}
@@
                              \appendix 
+                             \subsection{Models}
+The detection of early warning is based on a \emph{local} theory,
+predicting changes in the dynamics near the current attractor.
+Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
+but only the linearized model near the bifurcation.
+
+\subsubsection{Saddle-node bifurcation}
+Many models with alternative stable states considered in ecological systems can exhibit this kind of bifurcation.
+The individual-based simulation follows the master equation  
+\begin{align}
+  P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
+  b_n &= \frac{e K x^2}{X^2 + h_t^2} \nonumber \\ 
+  d_n &= e X_t - a_t   \nonumber
+\end{align}
+This model can experience a saddle-node bifurcation through increased mortality by increasing either $e$ or $a$. 
+The SDE approximation holds for reasonable population sizes:
+\begin{equation}
+dX_t = \left( \frac{e K x^2}{X^2 + h_t^2} - e X_t - a_t\right) dt + \sigma \sqrt{ \frac{e K x^2}{X^2 + h_t^2} + e X_t + a_t} dB_t \label{ass}
+\end{equation}
+This parameterization is only an example, of course many others are possible~\cite{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
+Under appropriate transformation, all can be written in the same normal form for he saddle-node bifurcation,
+\begin{equation}
+\frac{\ud x}{\ud t} = r_t- x^2.
+\label{saddle-node}
+\end{equation}
+While in principle this model could be estimated directly from the data,
+if all observations have been prior to the crash (\emph{i.e.} near the desirable equilibrium)
+we have no data which is informative about the nonlinearity and such estimates will be unreliable.  
+Fortunately the theory of early warning signals is a \emph{local} one, 
+and we can linearize around the equilibrium to obtain a model we can estimate both more easily and more accurately.
+
+Transforming the canonical form to allow for an arbitrary mean $\theta$,
+the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$,
+which gives the model we refer to as the Linearized Saddle Node, LSN: 
+\begin{equation}
+\ud X = 2\sqrt{ r_t } (\phi - X_t)\ud t + \sigma\sqrt{\phi } \ud B_t. \label{LSN}
+\end{equation}
+
+\subsubsection{Transcritical Bifurcation}
+Drake \& Griffen induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} to test early warning signals~\cite{Drake2010}.  
+This bifurcation also arises in many common models.  For example, the stochastic Levin's model~\cite{Levins1969} (a logistic model),
+\begin{equation}
+\ud X_t = \left( c_t X_t (1-X_t/K) - e_t X_t \right) \ud t + \sigma \sqrt{\frac{e_t}{c_t}} \ud B_t \label{levins},
+\end{equation}
+where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent)
+colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).  
+The model contains a transcritical bifurcation when $c_t < e_t$.  The normal form of the bifurcation is
+\begin{equation}
+\frac{\ud x}{\ud t} = r_t x - x^2 
+\label{transcritical}
+\end{equation}
+%We can rewrite the deterministic part of the Levins model, Eq~\eqref{levins} in normal form without loss of generality by taking $r_t = 1 - e_t/c_t$.  
+%We can linearize around the stable equilibrium value $\hat x(t) = 1 - e_t/c_t = r_t$, 
+%\begin{align}
+%\dot x &=  f(x) \approx f'(x)|_{\hat x} (x - \hat x) + \ldots, \nonumber \\
+% &= (r_t - 2 x_t|_{x=\hat x}) (x - \hat x), \nonumber\\
+% &= r_t(r_t - x). \nonumber
+%\end{align}
+%Note the equilibrium time dependence arises not from the internal dynamics but from the changing parameter values.
+We can express the linearized stochastic dynamics then by the time dependent mean-reverting process. 
+We refer to this model as the linearized transcritical bifurcation, LTC.  
+\begin{equation}
+\ud X_t = r_t (r_t - X_t/K) \ud t + \sigma \sqrt{1+r_t} \ud B_t \label{LTC}
+\end{equation}
+
+%The pattern of the saddle-node differs from that of the transcritical in the square-root dependence on the bifurcation parameter.
+Note that the transcritical bifurcation involves a gradual decline towards zero,
+whereas the saddle-node bifurcation involves a sudden transition from a non-trivial equilibrium point.
+Thus saddle-node bifurcations are the more typical focus of early warning efforts, even though both exhibit the phenomenon of critical slowing down.
+%We include this example for the \emph{Daphnia} data~\cite{Drake2010}, showing how our approach could be applied in that scenario.
+
+%Many models may differ substantially from~\eqref{ass} and~\eqref{levins} and yet correspond to the linearizations illustrated here.  
+%The linearizations have thus been constructed with enough free parameters to capture the gross features of the system,
+%(scaling the equilibrium position, variance, and timescale) while corresponding to the approximations of normal forms of the bifurcations.  
+%Despite this, there are severe limitations in formulating even a linear approximation without ample knowledge of the system. 
+Likelihood calculations also require specifying the rate of change of $r_t$, 
+which we take to first order as
+\begin{equation}
+r_t = r_0 - m t.
+\label{R_t}
+\end{equation}
+Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power. 
+%\subsubsection{Stable Model}
+When the system is stable, $r_t$ is constant and both models~\eqref{LSN} and~\eqref{LTC} reduce to a simple Ornstein-Uhlenbeck process, 
+\begin{equation}
+\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
+\end{equation}
+This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere (\emph{e.g.} \cite{Dakos2008, Guttal2008a}).
+
+\subsection{Likelihood calculations}\label{likelihood}
+The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
+\begin{equation}
+\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
+\end{equation}
+For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\cite{Gardiner2009} to solve the moment equations for mean and variance respectively:
+\begin{align}
+ \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
+\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
+  \label{general_moments}
+\end{align}
+
+For the OU process, we can solve this in closed form, 
+\begin{align}
+  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
+V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
+\label{OUsoln}
+\end{align}
+
+For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
+For the linearized transcritical:
+
+\begin{align}
+\frac{\ud }{\ud t} E(x_i| M = \text{LTC})&=  r(t)(r(t) - x_i) \\
+\frac{\ud}{\ud t} V(x_i| M = \text{LTC}) &=  -2 r(t) V(x_i|M) + (1+r(t))\sigma^2 
+\label{LTCsoln}
+\end{align}
+
+For the linearized saddle-node bifurcation:
+\begin{align}
+\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
+\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
+\label{LSNsoln}
+\end{align}
+
+These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.
 

### 2011-10-14 boettiger.tex
adding figures to new appendix

@@
 The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
-The best-performing summary statistic would, at a 5\% false positive rate, have a 10\% chance of catching a true positive 
-in both the Daphnia and Glaciation data set.  
-Meanwhile, our likelihood-based approach performs substantially better, 
-having respectively a 30\% in Daphnia and 100\% chance in Glaciation set of catching a true positive at a 5\% false positive rate. 
-Because we have the entire ROC curve, we are not restricted to an evaluation at a fixed false-positive rate such as 5\%, 
-but can accept more false-positives to avoid missing as many impending catastrophes.  
-This is particularly valuable in the context of critical transitions, 
+For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
+set to a 5\% false positive rate: 25\% in the simulated set, 5.0\% in the \emph{Daphnia} set, and 5.4\% in the Glaciation data set.
+Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives would be correctly identified from the respective systems.
+The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%, as we would in classical hypothesis testing.  
+We could instead decide that we wanted no to catch no less than 90\% of the true-positives, and read off the corresponding false positives.  
+For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate, while for likelihood a 55\%, 35\%, 0\% false positive.\footnote{
+Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic, as illustrated in Fig. 1.  
+Comparing this threshold with the observed value classifies the signal as detected or undetected.  See SI text.}
+This flexibility is particularly valuable in the context of critical transitions, 
 where data are often limited and the consequences of missed detection can be far more severe than false alarms.
@@
 \section{Discussion}
-A pressing question in the application of early warning signals is quantifying when the data are sufficient for detection to be reliable~\cite{Scheffer2009, Scheffer2010, Inman2011}. 
+A pressing question in the application of early warning signals is quantifying
+when the data are sufficient for detection to be reliable~\cite{Scheffer2009, Scheffer2010, Inman2011}. 
 Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
@@
 %% Imagine a missed detection of no more than 10% is acceptable.
+
+%%% Model adequacy
+In any application of early warning signals, it is essential to address the question of model-adequacy.  
+This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
+as it is for the approach taken here: it has been shown that these statistics correspond to particular assumptions about 
+an underlying process, and systems that violate these assumptions exhibit sudden transitions without any warning~\cite{Hastings2010}.
+Elsewhere sudden transitions have been shown in ecological models where the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
+Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
+All are based around a bifurcation process in which the principle eigenvalue passes through zero.
+
+% remove skew, CV
+
+It may appear that our approach is more limited as the models are explcitly linear with Gaussian noise processes.
+This is consistent
+
+While
 
@@
 The summary statistics such as variance or autocorrelation are no more general than our explicit models, % wow
-and systems that do not correspond to these bifurcations can exhibit sudden transitions without any warning~\cite{Hastings2010}.
-For instance, bifurcations resulting in sudden transitions have been observed in ecological models in which 
-the variance actually deceases as a sudden shift approaches~\cite{Schreiber2003, Schreiber2008}.
-
+and 
 Meanwhile, if our knowledge of the system gives reason to be concerned that it may experience a saddle-node or transcritical bifurcation,

### 2011-10-14 pnas_appendix.tex
adding figures to new appendix

%        File: pnas_appendix.tex
%     Created: Wed Oct 12 05:00 PM 2011 P
% Last Change: Wed Oct 12 05:00 PM 2011 P
%

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,review,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

%\graphicspath{{/home/cboettig/Documents/ucdavis/research/phylotrees/images/}}

\journal{Proceedings of the National Academy of Sciences } 
\begin{document}
%\begin{frontmatter}
%  \title{  }
%  \author[cpb]{Carl Boettiger\corref{cor1}}
%  \author[alan]{Alan Hastings}
%  \ead{cboettig@ucdavis.edu}
%  \cortext[cor1]{Corresponding author.}
%  \address[cpb]{Center for Population Biology, University of California, Davis, United States}
%  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, United States}
%\end{frontmatter}
%

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS1.pdf}
  \end{center}
  \caption{Distribution of the indicator statistic}
  \label{fig:S1}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS2.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data.}
  \label{fig:S2}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS3.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
  \label{fig:S3}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics{FigS4.pdf}
  \end{center}
  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation I data.}
  \label{fig:S4}
\end{figure}

The ROC curve indicates the trade off between false alarms and failed detections, 
from which we may choose a threshold value that will allow us to classify the data set as having detected a warning or not.  
For instance, in the case of the \emph{Daphnia} data, the 90\% true-positive, 35\% false-positive rate corresponds with a threshold of 
$\delta = 2.5$.  The observed $\delta=6.0$, clearly above that threshold.  
We can of ask what is the probability of a false alarm or missed event at the observed threshold of 6 as well: \%9.5 false alarm rate.  
While it is possible to classify

which can guide the classification of a particular system as 
When the curve lies very close to the 1:1 line, there is essentially no information provided by the given warning indicator,
and classifying

To determine whether the data should be considered

 \section*{ }%bibliography
 \bibliography{/home/cboettig/Documents/bibliographies/library.bib}

 \end{document}

### 2011-10-17 boettiger.tex
a few edits

@@
 For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
-set to a 5\% false positive rate: 25\% in the simulated set, 5.0\% in the \emph{Daphnia} set, and 5.4\% in the Glaciation data set.
-Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives would be correctly identified from the respective systems.
-The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%, as we would in classical hypothesis testing.  
-We could instead decide that we wanted no to catch no less than 90\% of the true-positives, and read off the corresponding false positives.  
-For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate, while for likelihood a 55\%, 35\%, 0\% false positive.\footnote{
-Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic, as illustrated in Fig. 1.  
+set to a 5\% false positive rate:
+25\% in the simulated set,
+5.0\% in the \emph{Daphnia} set,
+and 5.4\% in the Glaciation data set.
+Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
+would be correctly identified from the respective systems.
+The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
+as we would in classical hypothesis testing.  
+We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
+and read off the corresponding false positives.  
+For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
+while for likelihood a 55\%, 35\%, 0\% false positive.\footnote{
+Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
+as illustrated in Fig. 1.  
 Comparing this threshold with the observed value classifies the signal as detected or undetected.  See SI text.}
@@
 
+By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
+our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
+Other literature has already begun to address the additional challenges posed when the underlying 
+dynamics do not correspond to these models~\cite{Hastings2010}.
+Our results illustrate that even in this best-case scenario, 
+reliable identification of warning signals from summary statistics can be difficult.  
+The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
+or even the distribution of that statstic under the null model of a stable system, 
+but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
+By estimating the ROC curve for a given set of data, 
+we can better avoid applying warning signals in cases of inadequate power.
+By taking advantage of the assumptions being made to write down a specific likelihood function,
+We can develop indicators that get the most from this data.
+
 %%% Model adequacy
@@
 This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
-as it is for the approach taken here: it has been shown that these statistics correspond to particular assumptions about 
-an underlying process, and systems that violate these assumptions exhibit sudden transitions without any warning~\cite{Hastings2010}.
-Elsewhere sudden transitions have been shown in ecological models where the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
+as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
+and systems that violate these assumptions can exhibit sudden transitions without any warning,
+such as chaotic systems or those with non-smooth potential functions~\cite{Hastings2010}.
+Sudden transitions have also been observed in ecological models in which 
+the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
 Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
-All are based around a bifurcation process in which the principle eigenvalue passes through zero.
+As the bifurcation results from the principle eigenvalue passing through zero, 
+the warning signal is expected in linear-order dynamics;
+estimation of the nonlinear model is less powerful and less accurate.  
+The preformance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
+(introduced by the Poisson proccess demographic events), 
+demonstrates the accuracy under violation of these assumptions.
 
-% remove skew, CV
-
-It may appear that our approach is more limited as the models are explcitly linear with Gaussian noise processes.
-This is consistent
-
-While
-
-Moreover, this illustrates that the summary statistics are very unreliable indicators. 
-A low false alarm rate is only achieved by poor sensitivity; 
-or conversely sensitivity it only achieved with a high rate of false alarms.  
-This is under assumptions that are highly favorable to these statistics, corresponding directly to the process assumptions.  
-The reason for this comes largely from the difficulty in quantifying an increase without a baseline -- 
-the distribution of tau under the null model covers almost the entire range.  %Ref FIGURE
-
-Our approach of estimating uncertainty and using the likelihood indicator is neither more nor less general than the other indicators we discuss.  
-All rely on the assumption of an underlying bifurcation driven by the principle eigenvalue passing through zero.\footnote{
-Skew, a higher order term, has additional assumption about the asymmetry,
-and is thus generically less powerful than signals on the lower order moments.} 
-These methods are robust to the assumptions of Gaussian noise and linearity of the model,
-as illustrated by the fit to the simulated data, 
-which instead comes from a nonlinear, discrete-state Markov process (see Appendix).  
-The likelihood approach still performs the best on this data, hence it is not less general.
-
-When the underlying bifurcation model does not apply, neither summary statistics nor our model-based description will be useful.
-
-We stress that all of the early warning signals approaches must only be applied when the underlying assumptions can be justified.  
-The summary statistics such as variance or autocorrelation are no more general than our explicit models, % wow
-and 
-Meanwhile, if our knowledge of the system gives reason to be concerned that it may experience a saddle-node or transcritical bifurcation,
-
-If we can use increasing variance as an indicator, than we can use the approach described here to quantify uncertainty.
-
-Although not all sudden collapses correspond to our model assumptions,
-including dynamical systems where the variance actually decreases as a sudden shift approaches,⁠
-using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
-Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
-but would be amenable to the approach used here.
+%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
+%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
+%but would be amenable to the approach used here.
 
@@
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig3}
+     \includegraphics[width=\linewidth]{Fig3.pdf}
      \label{fig3}
@@
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig4}
+     \includegraphics[width=\linewidth]{Fig4.pdf}
      \label{fig4}

### 2011-10-17 pnas_appendix.tex
a few edits

@@
 %\documentclass[authoryear,5p]{elsarticle}
-\documentclass[authoryear,review,11pt]{elsarticle}
+%\documentclass[authoryear,review,11pt]{elsarticle}
+\documentclass[11pt]{elsarticle}
 \bibliographystyle{elsarticle-harv}
@@
 
+\section{Distributions of the indicator statistic}
+The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.
+The value of the statistic observed in the original data is indicated by a black triangle.
+Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.
+Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.
+The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.
+
 \begin{figure}[ht]
@@
   \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data.}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a stable system.}
   \label{fig:S2}
@@
   \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a system under deteriorating stability.}
   \label{fig:S3}
@@
   \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation I data.}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
   \label{fig:S4}
@@
 
-The ROC curve indicates the trade off between false alarms and failed detections, 
-from which we may choose a threshold value that will allow us to classify the data set as having detected a warning or not.  
-For instance, in the case of the \emph{Daphnia} data, the 90\% true-positive, 35\% false-positive rate corresponds with a threshold of 
-$\delta = 2.5$.  The observed $\delta=6.0$, clearly above that threshold.  
-We can of ask what is the probability of a false alarm or missed event at the observed threshold of 6 as well: \%9.5 false alarm rate.  
-While it is possible to classify
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics{FigS5.pdf}
+  \end{center}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation data.}
+  \label{fig:S5}
+\end{figure}
 
-which can guide the classification of a particular system as 
-When the curve lies very close to the 1:1 line, there is essentially no information provided by the given warning indicator,
-and classifying
+\section{Comparing Models}\label{Cox}
+Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
+The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
+The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
+which essentially asserts that comparing likelihoods is the most powerful test
+of a choice between two hypotheses~\citep{Neyman1933}, which motivates
+tests from the simple likelihood ratio test up through modern model adequacy methods.
 
-To determine whether the data should be considered
+The hypotheses considered here are more challenging then the original lemma, as they are composite in nature:
+they specify two model forms (stable and changing stability)
+but with model parameters that must be first estimated from the data.
+Comparing models whose parameters have been estimated by maximum likelihood is first treated by~\citet{Cox1961, Cox1962},
+and has since been developed in this simulation estimation of the null distribution~\citep{McLachlan1987}, by parametric bootstrap estimate~\citep{Efron1987}.  
+Cox's $\delta$ statistic is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.
+
+Let $L_0$ be the likelihood function for model 0, 
+let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
+be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
+and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
+The statistic we will use is $\delta$, 
+defined to be twice the difference in log likelihood of observing the data under the two MLE models,
+$\delta = -2 (\log L_0 - \log L_1 )$.  
+This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
+We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
+Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.
+
+\pagebreak
+
+\section{On a hypothesis-testing framework}\label{Dakos}
+It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
+As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.
+In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.
+By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.
+
+The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
+We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.
+For this reason we generate the null distribution directly from the model as outlined above.
+The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
+The plots in Fig.
+3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.
+Our approach replicate these findings for the $\tau$-statistic.
+We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.
+
+The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
+As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.
+Fig.
+S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.
+The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.
+The reality is worse, we find it overlaps almost completely.
+We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.
+
+The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
+
+\section{R package tutorial}\label{R}
+We provide an R package with simple implementation of the methods described here.
+The package is available from: \href{https://github.com/cboettig/warningsignals/archives/master}{https://github.com/cboettig/warningsignals/archives/master}, where it will be actively maintained and developed.
+The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood-based analysis:
+
+\begin{verbatim}
+> require(warningsignals)
+>
+> # Load a sample dataset
+> data(glaciationIII)
+>
+> # Fit constant (OU) model and a stability-loss model (LSN);
+> models <- fit_models(glaciationIII, 'LSN')
+>
+> results <- montecarlotest(models$const, models$timedep, n=2000, cpu=16)
+> plot(results)
+\end{verbatim}
+
+The package also performs the computationally faster but less powerful summary-statistic tests, computes the resulting distributions under the different hypotheses and the corresponding ROC curves.
+The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coefficient of variation).
+Copies of data used in the paper are provided as examples.
+The Monte Carlo methods can take advantage of multiple core processors or clusters by specifying the number of CPUs in the function argument.
+Full documentation of functions and data are provided in the package, as well as example scripts which replicate the analyses presented in the paper.
 

### 2011-10-17 boettiger.tex
best-case phrase added to intro

@@
 it is important to develop the ability to predict impending regime shifts based on early warning signs. 
-With a good model of the system, system-specific approaches could be useful~\cite{Lenton2009}.
+With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009}.
 but in cases where good models are not available general approaches are needed~\cite{Scheffer2009}⁠.
-We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system~\cite{Scheffer2001, Scheffer2009}.
-% hmm, not environmental examples\dots ⁠
-We describe a novel statistical approach that is appropriate for the goal of informing management.
+We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system~\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}}.
+This meets the assumptions and thus provides a ``best-case'' test of detection approaches,
+yet we will demonstrate the resulting uncertainty can remain very high.
+We describe a novel statistical approach for handling this uncertainty that is appropriate for the goal of informing management.
 
@@
 suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
-In principle, other quantifications of increasing variance, autocorrelation, etc,
+In principle, other quantifications of increasing variance, autocorrelation, etc.,
 may perform better than $\tau$ and could be analyzed just as well in this framework. 
@@
 The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
-or even the distribution of that statstic under the null model of a stable system, 
+or even the distribution of that statistic under the null model of a stable system, 
 but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
@@
 estimation of the nonlinear model is less powerful and less accurate.  
-The preformance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
-(introduced by the Poisson proccess demographic events), 
+The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
+(introduced by the Poisson process demographic events), 
 demonstrates the accuracy under violation of these assumptions.
@@
 
-Data and code available at: https://github.com/cboettig/warningsignals/archives/master. 
+Data and code available at: https://github.com/cboettig/warningsignals. 
                              \end{acknowledgments}

### 2011-10-17 boettiger.tex
PNAS submission ready

@@
                          \newcommand{\ud}{\mathrm{d}}
-
+                         \renewcommand{\thefootnote}{\fnsymbol{footnote}}
                          %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Catastrophic regime shifts in complex natural systems may be averted through advanced detection. 
-Recent work has provided a proof-of-principle that many systems approaching a catastrophic transition look different 
+Recent work has provided a proof-of-principle that many systems approaching a catastrophic transition may be identified 
 through the lens of early warning indicators such as rising variance or increased return times.  
@@
 and estimates what data are necessary to achieve a desired error rate. 
-We also introduce a likelihood-based indicator that is more sensitive than previously proposed summary statistics.
+We show these error rates can be quite severe for common indicators even under favorable assumptions, 
+and illustrate how a likelihood-based indicator can improve this performance.
 
@@
 \dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
-Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011} , and semi-arid ecosystems~\cite{Kefi2007}⁠.
+Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠.
 Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
@@
 it is important to develop the ability to predict impending regime shifts based on early warning signs. 
-With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009}.
-but in cases where good models are not available general approaches are needed~\cite{Scheffer2009}⁠.
-We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system~\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}}.
-This meets the assumptions and thus provides a ``best-case'' test of detection approaches,
-yet we will demonstrate the resulting uncertainty can remain very high.
-We describe a novel statistical approach for handling this uncertainty that is appropriate for the goal of informing management.
+With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
+but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
+We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}
+Making this assumption provides a ``best-case'' test of detection approaches;
+we will demonstrate the resulting uncertainty can be very high.
+We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.
 
-Though previous research has identified patterns such as increasing variance or autocorrelation in time-series approaching a critical transition,
-there has been no method for forecasting the probability of a collapse from a natural system.
-As real-world applications of early warning signals will not be able to point to the same differences between controlled and
-manipulated systems that have been demonstrated in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
+Though previous research has identified patterns such as increasing variance or autocorrelation
+in time-series approaching a critical transition,
+there has been no method for forecasting the \emph{probability} of a collapse from a natural system.
+Real-world applications of early warning signals will not be able to point to such differences between controlled and
+manipulated systems that have been utilized in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
 While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
 there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
-and fewer still that determine the distribution of that quantity in the event that the system is not approaching a transition --
-the reliability or false alarm rate~\cite{Dakos2008}⁠. 
-None have considered the distribution of such a statistic in the event that a system is approaching a transition -- 
-the sensitivity to avoid missed detection.  
-We need to quantify such uncertainty/error rates to compare between possible methods~\cite{Contamin2009}
+and fewer still that statistically estimate the false alarm rate~\cite{Dakos2008} or 
+the probability of missed detection⁠. 
+We need to quantify such uncertainty and error rates to compare between possible methods~\cite{Contamin2009}
 and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
 
-Detection of impending transitions is not just a scientific question and the goal should not solely be to detect with a 5\% chance of a false positive,
-and a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
+Detection of impending transitions is not just a scientific question 
+and the goal should not solely be to detect with a 5\% chance of a false positive:
+a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
 There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
@@
 (b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
-(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of an insensitive method 
+(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of insensitive indicators 
 (d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.
 
-We illustrate the trade-off between false alarms and failed detection using receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠.
+We illustrate the trade-off between false alarms and failed detection using 
+receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠. 
 The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
@@
 While we use this established metric of increase, we also introduce a novel warning indicator
-that is not based on the gradual increase of a summary statistic but on likelihood comparisons.
+that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).
 
@@
 We fit general stochastic models representing common bifurcations
-(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010} )
+(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010})
 for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
@@
 By taking advantage of the assumptions being made to write down a specific likelihood function,
-We can develop indicators that get the most from this data.
+we can develop indicators that get the most from this data.
 
 %%% Model adequacy
-In any application of early warning signals, it is essential to address the question of model-adequacy.  
+In any application of early warning signals, it is essential to address the question of model adequacy.  
 This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
@@
 
-Author Contributions
-C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper.
-
-Acknowledgements
 S.
@@
 This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
+
+Author Contributions
+C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper.
 
@@
 The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data.
-The simulated dataset has 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121.
+The simulated data sets have 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121.
 See corresponding distributions, Fig.
-S1, and additional examples, Fig.
-S2.}
+S1.}
   \end{center}
@@
 3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.
-Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.}
+Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.
+A warning sighal is not detected in stable sytem simulation.
+Corresponding distributions shown in Fig S2-5.}
   \end{center}

### 2011-10-17 pnas_appendix.tex
PNAS submission ready

@@
 %
-
-\section{Distributions of the indicator statistic}
-The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.
-The value of the statistic observed in the original data is indicated by a black triangle.
-Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.
-Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.
-The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics{FigS1.pdf}
-  \end{center}
-  \caption{Distribution of the indicator statistic}
-  \label{fig:S1}
-\end{figure}
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics{FigS2.pdf}
-  \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a stable system.}
-  \label{fig:S2}
-\end{figure}
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics{FigS3.pdf}
-  \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a system under deteriorating stability.}
-  \label{fig:S3}
-\end{figure}
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics{FigS4.pdf}
-  \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data.}
-  \label{fig:S4}
-\end{figure}
-
-\begin{figure}[ht]
-  \begin{center}
-    \includegraphics{FigS5.pdf}
-  \end{center}
-  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation data.}
-  \label{fig:S5}
-\end{figure}
-
+\appendix
+\renewcommand*\thefigure{S\arabic{figure}}
+\renewcommand*\theequation{S\arabic{equation}}
 \section{Comparing Models}\label{Cox}
@@
 
+ \section{Distributions of the indicator statistic}
+The ROC curves are generated from distributions of the measure of increase, $\tau$, in a warning statistic (variance, autocorrelation, etc.) or the likelihood ratio $\delta$ from the likelihood approach.
+The value of the statistic observed in the original data is indicated by a black triangle.
+Deciding whether or not this observed value indicates an early warning or not depends on the location of the threshold.
+Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.
+The large overlap between the distribution under the stable model (blue) and the deteriorating stability system (red) indicates how difficult it can be for such statistics to descriminate between these alternatives.
+The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics{FigS1.pdf}
+  \end{center}
+  \caption{Distribution of the indicator statistics corresponding to the ROC curves in Fig.
+3, main text.
+As in Fig.
+1, the expected distribution of the statistic under the null hypothesis of a stable system is shown in blue, while the distribution under the estimated deteriorating stability system is shown in red.
+The value of the test statistic ($\delta$ in Likelihood, top row, otherwise $\tau$), observed in the original timeseries is indicated by the black triangle.
+Distributions are produced from a Gaussian kernel density estimate from 500 replicates, as described in Methods.}
+  \label{fig:S1}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics{FigS2.pdf}
+  \end{center}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a stable system, corresponding to the first column of Fig.
+4, main text.
+Distributions as described in caption of Fig~\ref{fig:S1}.}
+  \label{fig:S2}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics{FigS3.pdf}
+  \end{center}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the simulated data of a system under deteriorating stability, corresponding to the second column of Fig.
+4, main text.
+Distributions as described in caption of Fig~\ref{fig:S2}.}
+  \label{fig:S3}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics{FigS4.pdf}
+  \end{center}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the \emph{Daphnia} data, corresponding to the third column of Fig.
+4, main text.
+Distributions as described in caption of Fig~\ref{fig:S1}.}
+  \label{fig:S4}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{center}
+    \includegraphics{FigS5.pdf}
+  \end{center}
+  \caption{Distributions of the indicator statistic as the sampling effort increases in the Glaciation data, corresponding to the fourth column of Fig.
+4, main text.
+Distributions as described in caption of Fig~\ref{fig:S1}.}
+  \label{fig:S5}
+\end{figure}
+
+\pagebreak
+
 \section*{ }%bibliography

### 2011-10-18 boettiger.tex
bib changes to follow editorial constraints

@@
                              %% \subsubsection{}
-
+                             \section{Introduction}
 \dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
@@
 
-\section{Examples}
+\section{Results}
 
@@
                              %% following \begin{thebibliography}
+\begin{thebibliography}{10}
 
-                             \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
-                             \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
+\bibitem{Holling1973}
+Holling CS
+\newblock (1973) {Resilience and Stability of Ecological Systems}.
+\newblock \emph{Annual Review of Ecology and Systematics} 4:1--23.
+
+\bibitem{Wissel1984}
+Wissel C
+\newblock (1984) {A universal law of the characteristic return time near
+  thresholds}.
+\newblock \emph{Oecologia} 65:101--107.
+
+\bibitem{Scheffer2001}
+Scheffer M, Carpenter SR, Foley JA, Folke C, Walker B
+\newblock (2001) {Catastrophic shifts in ecosystems.}
+\newblock \emph{Nature} 413:591--6.
+
+\bibitem{Scheffer2009}
+Scheffer M, {et~al.}
+\newblock (2009) {Early-warning signals for critical transitions.}
+\newblock \emph{Nature} 461:53--9.
+
+\bibitem{Drake2010}
+Drake JM, Griffen BD
+\newblock (2010) {Early warning signals of extinction in deteriorating
+  environments}.
+\newblock \emph{Nature} 467:456--459.
+
+\bibitem{Carpenter2011}
+Carpenter SR, {et~al.}
+\newblock (2011) {Early Warnings of Regime Shifts: A Whole-Ecosystem
+  Experiment}.
+\newblock \emph{Science} 1079.
+
+\bibitem{Bellwood2004}
+Bellwood DR, Hughes TP, Folke C, Nystr\"{o}m M
+\newblock (2004) {Confronting the coral reef crisis.}
+\newblock \emph{Nature} 429:827--33.
+
+\bibitem{Berkes2006}
+Berkes F, {et~al.}
+\newblock (2006) {Ecology. Globalization, roving bandits, and marine
+  resources.}
+\newblock \emph{Science (New York, N.Y.)} 311:1557--8.
+
+\bibitem{Kefi2007}
+K\'{e}fi S, {et~al.}
+\newblock (2007) {Spatial vegetation patterns and imminent desertification in
+  Mediterranean arid ecosystems.}
+\newblock \emph{Nature} 449:213--7.
+
+\bibitem{Lenton2009}
+Lenton TM, {et~al.}
+\newblock (2009) {Using GENIE to study a tipping point in the climate system.}
+\newblock \emph{Philosophical transactions. Series A, Mathematical, physical,
+  and engineering sciences} 367:871--84.
+
+\bibitem{Carpenter2006}
+Carpenter SR, Brock WA
+\newblock (2006) {Rising variance: a leading indicator of ecological
+  transition.}
+\newblock \emph{Ecology letters} 9:311--8.
+
+\bibitem{Dakos2008}
+Dakos V, {et~al.}
+\newblock (2008) {Slowing down as an early warning signal for abrupt climate
+  change.}
+\newblock \emph{Proceedings of the National Academy of Sciences of the United
+  States of America} 105:14308--12.
+
+\bibitem{Guttal2008}
+Guttal V, Jayaprakash C
+\newblock (2008) {Spatial variance and spatial skewness: leading indicators of
+  regime shifts in spatial ecological systems}.
+\newblock \emph{Theoretical Ecology} 2:3--12.
+
+\bibitem{Guttal2008a}
+Guttal V, Jayaprakash C
+\newblock (2008) {Changing skewness: an early warning signal of regime shifts
+  in ecosystems.}
+\newblock \emph{Ecology letters} 11:450--60.
+
+\bibitem{Dakos2011}
+Dakos V, K\'{e}fi S, Rietkerk M, Nes EHV, Scheffer M
+\newblock (2011) {Slowing Down in Spatially Patterned Ecosystems at the Brink
+  of Collapse}.
+\newblock \emph{The American Naturalist}.
+
+\bibitem{Contamin2009}
+Contamin R, Ellison AM
+\newblock (2009) {Indicators of regime shifts in ecological systems: what do we
+  need to know and when do we need to know it?}
+\newblock \emph{Ecological applications : a publication of the Ecological
+  Society of America} 19:799--816.
+
+\bibitem{Scheffer2010}
+Scheffer M
+\newblock (2010) {Complex systems: Foreseeing tipping points.}
+\newblock \emph{Nature} 467:411--2.
+
+\bibitem{Inman2011}
+Inman M
+\newblock (2011) {Sending out an SOS}.
+\newblock \emph{Nature Climate Change} pp 1--4.
+
+\bibitem{Green1989}
+Green DM, Swets JA
+\newblock (1989) \emph{{Signal Detection Theory and Psychophysics}}
+\newblock (Peninsula Pub), p 521.
+
+\bibitem{Keller2009}
+Keller RP, Lodge DM, Lewis MA, Shogren JF, eds
+\newblock (2009) \emph{{Bioeconomics of Invasive Species: Integrating Ecology,
+  Economics, Policy, and Management}}
+\newblock (Oxford University Press, USA), p 320.
+
+\bibitem{Dakos2009}
+Dakos V, Nes EH, Donangelo R, Fort H, Scheffer M
+\newblock (2009) {Spatial correlation as leading indicator of catastrophic
+  shifts}.
+\newblock \emph{Theoretical Ecology} pp 163--174.
+
+\bibitem{VanNes2007}
+van Nes EH, Scheffer M
+\newblock (2007) {Slow recovery from perturbations as a generic indicator of a
+  nearby catastrophic shift.}
+\newblock \emph{The American Naturalist} 169:738--47.
+
+\bibitem{Biggs2009}
+Biggs R, Carpenter SR, Brock Wa
+\newblock (2009) {Turning back from the brink: detecting an impending regime
+  shift in time to avert it.}
+\newblock \emph{Proceedings of the National Academy of Sciences of the United
+  States of America} 106:826--31.
+
+\bibitem{Cox1961}
+Cox DR
+\newblock (1961) \emph{{Tests of Seperate Families of Hypotheses}}
+\newblock No.{}~2, pp 105 -- 123.
+
+\bibitem{Hastings2010}
+Hastings A, Wysham DB
+\newblock (2010) {Regime shifts in ecological systems can occur with no
+  warning.}
+\newblock \emph{Ecology letters}.
+
+\bibitem{Schreiber2003}
+Schreiber SJ
+\newblock (2003) {Allee effects, extinctions, and chaotic transients in simple
+  population models}.
+\newblock \emph{Theoretical population biology} 64:201--209.
+
+\bibitem{Schreiber2008}
+Schreiber SJ, Rudolf VHW
+\newblock (2008) {Crossing habitat boundaries: coupling dynamics of ecosystems
+  through complex life cycles.}
+\newblock \emph{Ecology letters} 11:576--87.
+
+\bibitem{Huelsenbeck1996}
+Huelsenbeck JP, Bull JJ
+\newblock (1996) {A Likelihood Ratio Test to Detect Conflicting Phylogenetic
+  Signal}.
+\newblock \emph{Systematic Biology} 45:92--98.
+
+\bibitem{Petit1999}
+Petit JR, {et~al.}
+\newblock (1999) {Climate and atmospheric history of the past 420,000 years
+  from the Vostok ice core, Antarctica}.
+\newblock \emph{Nature} 399:429--436.
+
+\bibitem{Strogatz2001a}
+Strogatz SH
+\newblock (2001) \emph{{Nonlinear Dynamics And Chaos: With Applications To
+  Physics, Biology, Chemistry, And Engineering (Studies in Nonlinearity)}}
+\newblock (Westview Press), p 512.
+
+\bibitem{Guckenheimer1983}
+Guckenheimer J, Holmes P
+\newblock (1983) \emph{{Nonlinear Oscillations, Dynamical Systems, and
+  Bifurcations of Vector Fields (Applied Mathematical Sciences Vol. 42)}}
+\newblock (Springer), p 480.
+
+\bibitem{Levins1969}
+Levins R
+\newblock (1969) {Some demographic and genetic consequences of environmental
+  heterogeneity for biological control}.
+\newblock \emph{Bulletin of the Entomological Society of America} 15:237--240.
+
+\bibitem{Gardiner2009}
+Gardiner C
+\newblock (2009) \emph{{Stochastic Methods: A Handbook for the Natural and
+  Social Sciences (Springer Series in Synergetics)}}
+\newblock (Springer), p 447.
+
+\end{thebibliography}
+                           %  \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
+                            % \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
                            \end{article}
@@
                            %% \end{figure}
- \begin{figure*}
+                           \section{Figure Legends}
+\begin{figure}[hb]
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig1}
+%     \includegraphics[width=\linewidth]{Fig1}
      \label{fig1}
@@
   \end{center}
- \end{figure*}
+ \end{figure}
 
- \begin{figure*}
+ \begin{figure}[h]
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig2}
+%     \includegraphics[width=\linewidth]{Fig2}
      \label{fig2}
@@
   \end{center}
- \end{figure*}
+ \end{figure}
 
- \begin{figure*}
+ \begin{figure}[h]
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig3.pdf}
+%     \includegraphics[width=\linewidth]{Fig3.pdf}
      \label{fig3}
@@
   \end{center}
- \end{figure*}
+ \end{figure}
 
- \begin{figure*}
+ \begin{figure}[h!]
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig4.pdf}
+%     \includegraphics[width=\linewidth]{Fig4.pdf}
      \label{fig4}
@@
   \end{center}
- \end{figure*}
+ \end{figure}
 

### 2011-11-04 appendix.tex
ecology letters copy submitted

--- 2011-10-17 pnas_appendix.tex
@@
-%        File: pnas_appendix.tex
-%     Created: Wed Oct 12 05:00 PM 2011 P
-% Last Change: Wed Oct 12 05:00 PM 2011 P
-%
-
 %\documentclass[authoryear,5p]{elsarticle}
@@
 
-\journal{Proceedings of the National Academy of Sciences } 
+\journal{a Letter to \emph{Ecology Letters}, Supplementary Information} 
 \begin{document}
@@
 Thresholds involving larger values of $\tau$ or $\delta$ will generally involve fewer false alarms, but also catch fewer of the true positives.
-The large overlap between the distribution under the stable model (blue) and the deteriorating stability system (red) indicates how difficult it can be for such statistics to descriminate between these alternatives.
+The large overlap between the distribution under the stable model (blue) and the deteriorating stability system (red) indicates how difficult it can be for such statistics to discriminate between these alternatives.
 The nature of this trade-off is illustrated by the ROC curve, which can therefore be helpful in selecting the threshold for a particular statistic.
@@
 1, the expected distribution of the statistic under the null hypothesis of a stable system is shown in blue, while the distribution under the estimated deteriorating stability system is shown in red.
-The value of the test statistic ($\delta$ in Likelihood, top row, otherwise $\tau$), observed in the original timeseries is indicated by the black triangle.
+The value of the test statistic ($\delta$ in Likelihood, top row, otherwise $\tau$), observed in the original time-series is indicated by the black triangle.
 Distributions are produced from a Gaussian kernel density estimate from 500 replicates, as described in Methods.}

### 2011-11-04 ecologyletters.tex
ecology letters copy submitted

--- 2011-10-17 boettiger.tex
@@
-%        File: boettiger.tex
-%     Created: Mon Oct 10 04:00 PM 2011 P
-% Last Change: Mon Oct 10 04:00 PM 2011 P
-%%% PNAStmpl.tex
-%% Template file to use for PNAS articles prepared in LaTeX
-%% Version: Apr 14, 2008
+%        File: ecologyletters.tex
+%     Created: Sat Oct 29 02:00 PM 2011 P
+% Last Change: Sat Oct 29 02:00 PM 2011 P
+%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% BASIC CLASS FILE 
-%% PNAStwo for two column articles is called by default.
-%% Uncomment PNASone for single column articles. One column class
-%% and style files are available upon request from pnas@nas.edu.
-%% (uncomment means get rid of the '%' in front of the command)
+% Title page
+% - author contributions
+%- the article title
+%- the full name(s) of all author(s), affiliation(s), e-mail address(es) of all author(s)
+%- a short running title (abbreviated form of title) of less than 45 characters including spaces
+%- up to 10 keywords for indexing purposes. It is very important that the keywords be chosen carefully.
+%- the type of article (Ideas and Perspectives, Letter, or Reviews and Syntheses)
+%- the number of words in the abstract, in the manuscript as a whole, and in the main text (excluding abstract, acknowledgements, references, table and figure legends)
+%- the number of references (or for papers with extensive meta-analyses, the number of data-source references AND the number of general references)
+%- the number of figures and tables
+%- the name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent.
+%
 
-%\documentclass{pnasone}
-\documentclass{pnastwo}
+%\documentclass[authoryear,5p]{elsarticle}
+\documentclass[authoryear,review,11pt]{my_elsarticle}
+\bibliographystyle{elsarticle-harv}
+\usepackage{graphicx}
+\usepackage{amsmath,amsfonts}
+\usepackage{lineno}
+\linenumbers
+\usepackage{subfigure}
+\usepackage[pdftex]{color}
+\definecolor{darkblue}{rgb}{0,0,0.5}
+\definecolor{darkgreen}{rgb}{0,0.5,0}
+%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
+\usepackage[pdftex, colorlinks]{hyperref}
+\textwidth 6.75in
+\oddsidemargin -0.15in
+\evensidemargin -0.15in
+\textheight 9in
+\topmargin -0.5in
+\newcommand{\ud}{\mathrm{d}}
+\newcommand{\E}{\mathrm{E}}
+\newcommand{\C}{\mathrm{Cov}}
+\newcommand{\V}{\mathrm{Var}}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Changing position of text on physical page:
-%% Since not all printers position
-%% the printed page in the same place on the physical page,
-%% you can change the position yourself here, if you need to:
+%\journal{A Letter to Ecology Letters } 
+\begin{document}
+\begin{frontmatter}
+  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
+  \author[cpb]{Carl Boettiger\corref{cor1}}
+  \ead{cboettig@ucdavis.edu, 610-389-6087}
+  \author[esp]{Alan Hastings}
+  %\author[info]{}
+  %\author[davis]{}
+  \cortext[cor1]{Corresponding author.}
+  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
+  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}
 
-% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
-                         %  physical page; positive dimension will lower it.
-
-                         %% You may set the dimension to the size that you need.
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% OPTIONAL GRAPHICS STYLE FILE
-
-                         %% Requires graphics style file (graphicx.sty), used for inserting
-                         %% .eps files into LaTeX articles.
-                         %% Note that inclusion of .eps files is for your reference only;
-                         %% when submitting to PNAS please submit figures separately.
-
-                         %% Type into the square brackets the name of the driver program 
-                         %% that you are using. If you don't know, try dvips, which is the
-                         %% most common PC driver, or textures for the Mac. These are the options:
-
-                         % [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
-                         % [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
-                         % [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]
-
-                         \usepackage[pdftex]{graphicx}
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% OPTIONAL POSTSCRIPT FONT FILES
-
-                         %% PostScript font files: You may need to edit the PNASoneF.sty
-                         %% or PNAStwoF.sty file to make the font names match those on your system. 
-                         %% Alternatively, you can leave the font style file commands commented out
-                         %% and typeset your article using the default Computer Modern 
-                         %% fonts (recommended). If accepted, your article will be typeset
-                         %% at PNAS using PostScript fonts.
-
-                         % Choose PNASoneF for one column; PNAStwoF for two column:
-                         %\usepackage{PNASoneF}
-                         %\usepackage{pnastwof}
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% ADDITIONAL OPTIONAL STYLE FILES
-
-                         %% The AMS math files are commonly used to gain access to useful features
-                         %% like extended math fonts and math commands.
-
-                         \usepackage{amssymb,amsfonts,amsmath}
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% OPTIONAL MACRO FILES
-                         %% Insert self-defined macros here.
-                         %% \newcommand definitions are recommended; \def definitions are supported
-
-                         %\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
-                         %\def\s{\sigma}
-
-                         \newcommand{\ud}{\mathrm{d}}
-                         \renewcommand{\thefootnote}{\fnsymbol{footnote}}
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% Don't type in anything in the following section:
-                         %%%%%%%%%%%%
-                         %% For PNAS Only:
-                         \contributor{Submitted to Proceedings
-                         of the National Academy of Sciences of the United States of America}
-                         \url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
-                         \copyrightyear{2008}
-                         \issuedate{Issue Date}
-                         \volume{Volume}
-                         \issuenumber{Issue Number}
-                         %%%%%%%%%%%%
-
-                         \begin{document}
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-                         %% For titles, only capitalize the first letter
-                         %% \title{Almost sharp fronts for the surface quasi-geostrophic equation}
-
-                         \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
-
-                         %% Enter authors via the \author command.  
-                         %% Use \affil to define affiliations.
-                         %% (Leave no spaces between author name and \affil command)
-
-                         %% Note that the \thanks{} command has been disabled in favor of
-                         %% a generic, reserved space for PNAS publication footnotes.
-
-                         %% \author{<author name>
-                         %% \affil{<number>}{<Institution>}} One number for each institution.
-                         %% The same number should be used for authors that
-                         %% are affiliated with the same institution, after the first time
-                         %% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
-                         %% Then, before last author ...
-                         %% \and
-                         %% \author{<author name>
-                         %% \affil{<number>}{}}
-
-                         %% For example, assuming Garcia and Sonnery are both affiliated with
-                         %% Universidad de Murcia:
-                         %% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
-                         %% United Kingdom},
-                         %% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
-                         %% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}
-
-                         \author{Carl Boettiger\affil{1}{Center for Population Biology, University of California, Davis}, 
-                         Alan Hastings\affil{2}{Department of Environmental Science and Policy, University of California, Davis} }
-
-                         \contributor{Submitted to Proceedings of the National Academy of Sciences
-                         of the United States of America}
-
-                         %% The \maketitle command is necessary to build the title page.
-                         \maketitle
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         \begin{article}
+ % \address[info]{ \\ 
+  %              }
 
@@
 
-                             %% When adding keywords, separate each term with a straight line: |
-                             \keywords{ early warning signals | tipping point | alternative stable states | likelihood methods }
+  \begin{keyword}
+early warning signals \sep tipping point \sep alternative stable states \sep likelihood methods 
+   \end{keyword}
+ \end{frontmatter}
+ \emph{
+  \textbf{Running title}: Limits to Detection of Early Warning \\ 
+  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 33 references, 4 Figures.
+  }
 
-                             %% Optional for entering abbreviations, separate the abbreviation from
-                             %% its definition with a comma, separate each pair with a semicolon:
-                             %% for example:
-                             %% \abbreviations{SAM, self-assembled monolayer; OTS,
-                             %% octadecyltrichlorosilane}
-
-                             % \abbreviations{}
-
-                             %% The first letter of the article should be drop cap: \dropcap{}
-                             %\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts
-
-                             %% Enter the text of your article beginning here and ending before
-                             %% \begin{acknowledgements}
-                             %% Section head commands for your reference:
-                             %% \section{}
-                             %% \subsection{}
-                             %% \subsubsection{}
-
-\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
-Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠.
+ \section{Introduction}
+There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
+Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠.
 Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
@@
 it is important to develop the ability to predict impending regime shifts based on early warning signs. 
-With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
-but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
-We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}
+With a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009},
+but in cases where specific models are not available general approaches are needed~\citep{Scheffer2009}⁠.
+We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
+(\emph{i.e.} where the dominant eigenvalue passes through zero,
+as discussed in~\citet{Scheffer2001, Scheffer2009}). 
 Making this assumption provides a ``best-case'' test of detection approaches;
@@
 Real-world applications of early warning signals will not be able to point to such differences between controlled and
-manipulated systems that have been utilized in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
-While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
-there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
-and fewer still that statistically estimate the false alarm rate~\cite{Dakos2008} or 
+manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠.
+While the number of potential patterns grows steadily~\citep{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
+there are few examples that statistically quantify the patterns~\citep{Dakos2008, Dakos2011},⁠
+and fewer still that statistically estimate the false alarm rate~\citep{Dakos2008} or 
 the probability of missed detection⁠. 
-We need to quantify such uncertainty and error rates to compare between possible methods~\cite{Contamin2009}
-and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
+We need to quantify such uncertainty and error rates to compare between possible methods~\citep{Contamin2009}
+and address the continual concern about how much data is necessary⁠ for robust detection~\citep{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
 
@@
 We illustrate the trade-off between false alarms and failed detection using 
-receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\cite{Green1989, Keller2009}⁠. 
+receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
 The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
@@
 % so this is not a reasonable statistical characterization
-visual or heuristic measure of increase\footnote{That is, do not provide $p$-values or a statistic for which they could be generated.} in early-warning indicators~\cite{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006},
-we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\cite{Dakos2008, Dakos2011, Dakos2009}.
+visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006},
+we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\citep{Dakos2008, Dakos2011, Dakos2009}.
 Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
@@
 We fit general stochastic models representing common bifurcations
-(e.g. saddle-node~\cite{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\cite{Drake2010})
+(\emph{e.g.} saddle-node~\citep{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\citep{Drake2010})
 for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
-By estimating the statistic (e.g. $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
-Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\cite{Cox1961}⁠ instead of $\tau$,
+By estimating the statistic (\emph{e.g.} $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
+Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\citep{Cox1961}⁠ instead of $\tau$,
 which provides a more direct and more powerful test than detecting an increase in summary statistics.
@@
 
-\section{Examples}
+\section{Materials \& Methods}\label{methods}
 
-We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
-We explore two empirical examples:
-deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\cite{Dakos2008},
-and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\cite{Drake2010}.  
-For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
-and the same system in a constant environment. 
-The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
-against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
-While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
-we need a way to identify such cases where we lack the data to trust the pattern.
-
-The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
-frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
-The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
-For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
-set to a 5\% false positive rate:
-25\% in the simulated set,
-5.0\% in the \emph{Daphnia} set,
-and 5.4\% in the Glaciation data set.
-Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
-would be correctly identified from the respective systems.
-The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
-as we would in classical hypothesis testing.  
-We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
-and read off the corresponding false positives.  
-For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
-while for likelihood a 55\%, 35\%, 0\% false positive.\footnote{
-Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
-as illustrated in Fig. 1.  
-Comparing this threshold with the observed value classifies the signal as detected or undetected.  See SI text.}
-This flexibility is particularly valuable in the context of critical transitions, 
-where data are often limited and the consequences of missed detection can be far more severe than false alarms.
-
-%%%%%%% Converting from power to detection %%%%%%%%%%%%
-
-\section{Discussion}
-A pressing question in the application of early warning signals is quantifying
-when the data are sufficient for detection to be reliable~\cite{Scheffer2009, Scheffer2010, Inman2011}. 
-Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
-for each of the indicators (Fig 4). 
-Using our generic models estimated from the original data as in Fig. 3, 
-we generate the corresponding distributions by simulating under the models as before, 
-but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
-would be hampered or improved with more or less data.  
-While all methods improve as the number of data points increase, 
-the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
-as expected~\cite{Carpenter2011};⁠
-skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4).
-
-%% Threshold
-%% Imagine a missed detection of no more than 10% is acceptable.
-
-By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
-our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
-Other literature has already begun to address the additional challenges posed when the underlying 
-dynamics do not correspond to these models~\cite{Hastings2010}.
-Our results illustrate that even in this best-case scenario, 
-reliable identification of warning signals from summary statistics can be difficult.  
-The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
-or even the distribution of that statistic under the null model of a stable system, 
-but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
-By estimating the ROC curve for a given set of data, 
-we can better avoid applying warning signals in cases of inadequate power.
-By taking advantage of the assumptions being made to write down a specific likelihood function,
-we can develop indicators that get the most from this data.
-
-%%% Model adequacy
-In any application of early warning signals, it is essential to address the question of model adequacy.  
-This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
-as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
-and systems that violate these assumptions can exhibit sudden transitions without any warning,
-such as chaotic systems or those with non-smooth potential functions~\cite{Hastings2010}.
-Sudden transitions have also been observed in ecological models in which 
-the variance deceases rather then increases as the system nears a bifurcation point~\cite{Schreiber2003, Schreiber2008}.  
-Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
-As the bifurcation results from the principle eigenvalue passing through zero, 
-the warning signal is expected in linear-order dynamics;
-estimation of the nonlinear model is less powerful and less accurate.  
-The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
-(introduced by the Poisson process demographic events), 
-demonstrates the accuracy under violation of these assumptions.
-
-%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
-%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
-%but would be amenable to the approach used here.
-
-The approach presented here offer a best-case scenario for these warning signals,
-illustrating that substantial uncertainty can remain despite the appearance of a positive increase.
-Our results highlight the need to quantify uncertainty in the application of early warning signals,
-and suggest that a hypothesis testing approach may be too restrictive a framework for that purpose.
-Using the ROC curve, we obtain a more general picture of the trade-off between false alarms and failed detections
-than can be provided by a significance test at a fixed false-alarm rate.  
-The curve can guide the choice of threshold used to indicate a warning signal on a case-by-case basis.
-
-The conclusion is not simply that likelihood approaches are more reliable, 
-but rather more broadly that warning signals should consider
-the inherent trade-off between sensitivity and accuracy,
-and must quantify how this trade-off depends on both the indicators used and the data available.  
-The approach developed here estimates the risk of both failed detection and false alarms;
-concepts which are critical to prediction-based management.  
-Using the methods we have outlined when designing early warning strategies for natural systems
-can ensure that data collection has adequate power to offer a reasonable chance of detection.
-
-                             %% == end of paper:
-
-                             %% Optional Materials and Methods Section
-                             %% The Materials and Methods section header will be added automatically.
-
-                             %% Enter any subheads and the Materials and Methods text below.
-                             \begin{materials}
-                               \subsection{Methods}\label{methods}
-For each dataset, parameters are estimated by maximum likelihood for a stable model
-and a model of a system approaching a critical transition (Appendix).
+\subsection{Statistical methodology}
+For each given time-series, parameters are estimated by maximum likelihood 
+for a stable model and a model of a system approaching a critical transition (see Models).
 The models are stochastic differential equations (SDEs)
@@
 The likelihood test is performed as in the parametric bootstrap; 
-but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\cite{Cox1961}.
+but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\citep{Cox1961}.
 Refitting the models on each replicate, 
 rather merely computing the likelihood under the original estimates,
-makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\cite{Huelsenbeck1996}.
-
+makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\citep{Huelsenbeck1996}.
 When replicates are created by simulation under the estimated models,
@@
 representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve, Fig 4.
-
 The source code for implementing each of the analyses and generating the plots presented is provided as an R package, \verb|warningsignals|.
 
-                               \subsection{Data sets}
+\subsection{Example Data}
 The simulated data sets are produced with 40 data points
@@
 
-The Glaciation data is accessible from NOAA~\cite{Petit1999}.
+The Glaciation data is accessible from NOAA~\citep{Petit1999}.
 The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
-to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\cite{Dakos2008},
+to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\citep{Dakos2008},
 analyzing the third glaciation event, consisting of 121 sample points. 
@@
 but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
-De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\cite{Dakos2008}, 
-which did not apply to the other data sets~\cite{Drake2010}.
+De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008}, 
+which did not apply to the other data sets~\citep{Drake2010}.
 
-The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\cite{Drake2010}. 
+The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
 This individual replicates was chosen as an example of a single replicate 
 that showed a statistically significant correlation in variance over window of time where critical slowing down was expected.
-Further empirical examples that have been previously investigated for early warning signs are also analyzed under our approach in SI Text.
-
 All data and code for simulations and data processing are found in the R package.
 
-                             \end{materials}
-
-                             %% Optional Appendix or Appendices
-                             \appendix 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Supp. materials or main text?  %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Models}
@@
 Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
-but only the linearized model near the bifurcation.
+but we can approximate the model near the bifurcation.  
+Many possible complex models will correspond to this approximation.
 
-\subsubsection{Saddle-node bifurcation}
-Many models with alternative stable states considered in ecological systems can exhibit this kind of bifurcation.
-The individual-based simulation follows the master equation  
+%\textbf{Saddle-node bifurcation.}
+One of the most common critical transitions modeled in ecological systems with alternative stable states is a saddle-node bifurcation~\citep[\{emph{e.g.}[]{Scheffer2001, Scheffer2009, Guttal2008a, VanNes2007, Biggs2009}.
+The individual-based simulation follows the master equation, 
 \begin{align}
   P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
-  b_n &= \frac{e K x^2}{X^2 + h_t^2} \nonumber \\ 
-  d_n &= e X_t - a_t   \nonumber
+  b_n &= \frac{e K x^2}{X^2 + h_t^2} \\ 
+  d_n &= e X_t - a_t   
 \end{align}
@@
 \end{equation}
-This parameterization is only an example, of course many others are possible~\cite{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
+This parameterization is only an example, of course many others are possible~\citep{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
 Under appropriate transformation, all can be written in the same normal form for he saddle-node bifurcation,
@@
 
-\subsubsection{Transcritical Bifurcation}
-Drake \& Griffen induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} to test early warning signals~\cite{Drake2010}.  
-This bifurcation also arises in many common models.  For example, the stochastic Levin's model~\cite{Levins1969} (a logistic model),
+\citet{Drake2010} induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} to test early warning signals.  
+This bifurcation also arises in many common models.  For example, the stochastic version of~\citet{Levins1969}'s model (a logistic model),
 \begin{equation}
@@
 \end{equation}
-%We can rewrite the deterministic part of the Levins model, Eq~\eqref{levins} in normal form without loss of generality by taking $r_t = 1 - e_t/c_t$.  
-%We can linearize around the stable equilibrium value $\hat x(t) = 1 - e_t/c_t = r_t$, 
-%\begin{align}
-%\dot x &=  f(x) \approx f'(x)|_{\hat x} (x - \hat x) + \ldots, \nonumber \\
-% &= (r_t - 2 x_t|_{x=\hat x}) (x - \hat x), \nonumber\\
-% &= r_t(r_t - x). \nonumber
-%\end{align}
-%Note the equilibrium time dependence arises not from the internal dynamics but from the changing parameter values.
 We can express the linearized stochastic dynamics then by the time dependent mean-reverting process. 
@@
 Thus saddle-node bifurcations are the more typical focus of early warning efforts, even though both exhibit the phenomenon of critical slowing down.
-%We include this example for the \emph{Daphnia} data~\cite{Drake2010}, showing how our approach could be applied in that scenario.
 
-%Many models may differ substantially from~\eqref{ass} and~\eqref{levins} and yet correspond to the linearizations illustrated here.  
-%The linearizations have thus been constructed with enough free parameters to capture the gross features of the system,
-%(scaling the equilibrium position, variance, and timescale) while corresponding to the approximations of normal forms of the bifurcations.  
-%Despite this, there are severe limitations in formulating even a linear approximation without ample knowledge of the system. 
 Likelihood calculations also require specifying the rate of change of $r_t$, 
@@
 Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
-%\subsubsection{Stable Model}
 When the system is stable, $r_t$ is constant and both models~\eqref{LSN} and~\eqref{LTC} reduce to a simple Ornstein-Uhlenbeck process,
@@
 \end{equation}
-This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere (\emph{e.g.} \cite{Dakos2008, Guttal2008a}).
+This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.
 
@@
 \end{equation}
-For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\cite{Gardiner2009} to solve the moment equations for mean and variance respectively:
+For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
 \begin{align}
@@
 These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Supp. materials or main text?  %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-                             %Appendix text...
-                             %% or, for appendix with title, use square brackets:
-                             %% \appendix[Appendix Title]
+\section{Results}
 
-                             \begin{acknowledgments}
+We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
+We explore two empirical examples:
+deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\citep{Dakos2008},
+and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\citep{Drake2010}.  
+For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
+and the same system in a constant environment. 
+The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
+against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
+While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
+we need a way to identify such cases where we lack the data to trust the pattern.
 
+The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
+frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
+The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
+For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
+set to a 5\% false positive rate:
+25\% in the simulated set,
+5.0\% in the \emph{Daphnia} set,
+and 5.4\% in the Glaciation data set.
+Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
+would be correctly identified from the respective systems.
+The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
+as we would in classical hypothesis testing.  
+We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
+and read off the corresponding false positives.  
+For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
+while for likelihood a 55\%, 35\%, 0\% false positive.
+Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
+as illustrated in Fig. 1.  
+Comparing this threshold with the observed value classifies the signal as detected or undetected, (see Supporting Information).
+This flexibility is particularly valuable in the context of critical transitions, 
+where data are often limited and the consequences of missed detection can be far more severe than false alarms.
+
+%%%%%%% Converting from power to detection %%%%%%%%%%%%
+
+\section{Discussion}
+A pressing question in the application of early warning signals is quantifying
+when the data are sufficient for detection to be reliable~\citep{Scheffer2009, Scheffer2010, Inman2011}. 
+Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
+for each of the indicators (Fig 4). 
+Using our generic models estimated from the original data as in Fig. 3, 
+we generate the corresponding distributions by simulating under the models as before, 
+but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
+would be hampered or improved with more or less data.  
+While all methods improve as the number of data points increase, 
+the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
+as expected~\citep{Carpenter2011};⁠
+skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4). 
+%\citet{Dakos2011a} demonstrate that changes to environmental coupling decrease.  also use replicate simulations to assess expected response, but not the probability of accuracy on single replicates.
+%% Threshold
+%% Imagine a missed detection of no more than 10% is acceptable.
+
+By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
+our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
+Other literature has already begun to address the additional challenges posed when the underlying 
+dynamics do not correspond to these models~\citep{Hastings2010}.
+Our results illustrate that even in this best-case scenario, 
+reliable identification of warning signals from summary statistics can be difficult.  
+The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
+or even the distribution of that statistic under the null model of a stable system, 
+but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
+By estimating the ROC curve for a given set of data, 
+we can better avoid applying warning signals in cases of inadequate power.
+By taking advantage of the assumptions being made to write down a specific likelihood function,
+we can develop indicators that get the most from this data.
+
+%%% Model adequacy
+In any application of early warning signals, it is essential to address the question of model adequacy.  
+This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
+as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
+and systems that violate these assumptions can exhibit sudden transitions without any warning,
+such as chaotic systems or those with non-smooth potential functions~\citep{Hastings2010}.
+Sudden transitions have also been observed in ecological models in which 
+the variance deceases rather then increases as the system nears a bifurcation point~\citep{Schreiber2003, Schreiber2008}.  
+Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
+As the bifurcation results from the principle eigenvalue passing through zero, 
+the warning signal is expected in linear-order dynamics;
+estimation of the nonlinear model is less powerful and less accurate.  
+The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
+(introduced by the Poisson process demographic events), 
+demonstrates the accuracy under violation of these assumptions.
+
+%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
+%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
+%but would be amenable to the approach used here.
+
+The approach presented here offer a best-case scenario for these warning signals,
+illustrating that substantial uncertainty can remain despite the appearance of a positive increase.
+Our results highlight the need to quantify uncertainty in the application of early warning signals,
+and suggest that a hypothesis testing approach may be too restrictive a framework for that purpose.
+Using the ROC curve, we obtain a more general picture of the trade-off between false alarms and failed detections
+than can be provided by a significance test at a fixed false-alarm rate.  
+The curve can guide the choice of threshold used to indicate a warning signal on a case-by-case basis.
+
+The conclusion is not simply that likelihood approaches are more reliable, 
+but rather more broadly that warning signals should consider
+the inherent trade-off between sensitivity and accuracy,
+and must quantify how this trade-off depends on both the indicators used and the data available.  
+The approach developed here estimates the risk of both failed detection and false alarms;
+concepts which are critical to prediction-based management.  
+Using the methods we have outlined when designing early warning strategies for natural systems
+can ensure that data collection has adequate power to offer a reasonable chance of detection.
+
+\section{Acknowledgments}
 S.
@@
 This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
+C.B.
+wrote the software and performed the analysis with advice from A.H.
+C.B.
+and A.H.
+designed the study and wrote the paper.
+Data and code are available at: https://github.com/cboettig/warningsignals.
 
-Author Contributions
-C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper.
+ % Computational Sciences Graduate Fellowship from the Department of Energy under grant number DE-FG02-97ER25308. 
+ \section{References}%bibliography
+ \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
 
-Data and code available at: https://github.com/cboettig/warningsignals. 
-                             \end{acknowledgments}
-
-                             %% PNAS does not support submission of supporting .tex files such as BibTeX.
-                             %% Instead all references must be included in the article .tex document. 
-                             %% If you currently use BibTeX, your bibliography is formed because the 
-                             %% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-                             %% .tex document. To conform to PNAS requirements, copy the reference listings
-                             %% from your .bbl file and add them to the article .tex file, using the
-                             %% bibliography environment described above.
-
-                             %%  Contact pnas@nas.edu if you need assistance with your
-                             %%  bibliography.
-
-                             % Sample bibliography item in PNAS format:
-                             %% \bibitem{in-text reference} comma-separated author names up to 5,
-                             %% for more than 5 authors use first author last name et al. (year published)
-                             %% article title  {\it Journal Name} volume #: start page-end page.
-                             %% ie,
-                             % \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
-                             % A short C-terminal sequence is necessary and sufficient for the
-                             % targeting of chitinases to the plant vacuole. 
-                             % {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-                             %% Enter the largest bibliography number in the facing curly brackets
-                             %% following \begin{thebibliography}
-
-                             \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
-                             \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
-                           \end{article}
-                           %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-                           %% Adding Figure and Table References
-                           %% Be sure to add figures and tables after \end{article}
-                           %% and before \end{document}
-
-                           %% For figures, put the caption below the illustration.
-                           %%
-                           %% \begin{figure}
-                           %% \caption{Almost Sharp Front}\label{afoto}
-                           %% \end{figure}
- \begin{figure*}
+\begin{figure}[hb]
    \begin{center}
@@
   \end{center}
- \end{figure*}
+ \end{figure}
 
- \begin{figure*}
+ \begin{figure}[h]
    \begin{center}
@@
   \end{center}
- \end{figure*}
+ \end{figure}
 
- \begin{figure*}
+ \begin{figure}[h!]
    \begin{center}
@@
 Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.
-A warning sighal is not detected in stable sytem simulation.
+A warning signal is not detected in stable system simulation.
 Corresponding distributions shown in Fig S2-5.}
   \end{center}
- \end{figure*}
-
-                           %% For Tables, put caption above table
-                           %%
-                           %% Table caption should start with a capital letter, continue with lower case
-                           %% and not have a period at the end
-                           %% Using @{\vrule height ?? depth ?? width0pt} in the tabular preamble will
-                           %% keep that much space between every line in the table.
-
-                           %% \begin{table}
-                           %% \caption{Repeat length of longer allele by age of onset class}
-                           %% \begin{tabular}{@{\vrule height 10.5pt depth4pt  width0pt}lrcccc}
-                           %% table text
-                           %% \end{tabular}
-                           %% \end{table}
-
-                           %% For two column figures and tables, use the following:
-
-                           %% \begin{figure*}
-                           %% \caption{Almost Sharp Front}\label{afoto}
-                           %% \end{figure*}
-
-                           %% \begin{table*}
-                           %% \caption{Repeat length of longer allele by age of onset class}
-                           %% \begin{tabular}{ccc}
-                           %% table text
-                           %% \end{tabular}
-                           %% \end{table*}
+ \end{figure}
 

### 2011-11-09 ecologyletters.tex
preprint mode, copy sent to John Drake

@@
 %\documentclass[authoryear,5p]{elsarticle}
-\documentclass[authoryear,review,11pt]{my_elsarticle}
+\documentclass[authoryear,preprint,11pt]{elsarticle}
 \bibliographystyle{elsarticle-harv}
@@
 
-%\journal{A Letter to Ecology Letters } 
+\journal{\tiny Ecology Letters} 
 \begin{document}
@@
  \end{frontmatter}
- \emph{
-  \textbf{Running title}: Limits to Detection of Early Warning \\ 
-  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 33 references, 4 Figures.
+\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
+  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
+  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
 }
@@
 Sudden transitions have also been observed in ecological models in which 
-the variance deceases rather then increases as the system nears a bifurcation point~\citep{Schreiber2003, Schreiber2008}.  
+the variance deceases rather then increases as the system nears a bifurcation point~\citep{Schreiber2003, Schreiber2008, Dakos2011a}.  
 Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
@@
 This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
-C.B.
-wrote the software and performed the analysis with advice from A.H.
-C.B.
-and A.H.
-designed the study and wrote the paper.
 Data and code are available at: https://github.com/cboettig/warningsignals.
@@
  \section{References}%bibliography
- \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
+ \bibliography{/home/cboettig/Documents/Mendeley/bib/library.bib}
 
@@
 The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).
-Successive columns show the threshold increasing, tracing out the ROC curve (successively gray lines show the threshold used for the matching gray point added to the ROC curve).}
+Successive columns show the threshold increasing, tracing out the ROC curve.}
   \end{center}
@@
      \caption{Early warning signals in simulated and empirical data sets.
-The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation due to a gradually deteriorating environment (Deteriorating).
-Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated in a chemostat towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a critical transition in glaciation (Glaciation).
-Increases in summary statistics such as variance, autocorrelation, or skew, computed over a moving window, are used to indicate if a system is moving towards a critical transition (see supplementary online text).
+The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation (Deteriorating).
+Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a glaciation period (Glaciation).
+Increases in summary statistics, computed over a moving window, have often been used to indicate if a system is moving towards a critical transition.
 The increase is measured by the correlation coefficient $\tau$.

### 2012-01-13 boettiger.tex
rewrites..

--- 2011-10-17 boettiger.tex
@@
-%        File: boettiger.tex
-%     Created: Mon Oct 10 04:00 PM 2011 P
-% Last Change: Mon Oct 10 04:00 PM 2011 P
-%%% PNAStmpl.tex
-%% Template file to use for PNAS articles prepared in LaTeX
-%% Version: Apr 14, 2008
+%        File: ecologyletters.tex
+%     Created: Sat Oct 29 02:00 PM 2011 P
+% Last Change: Sat Oct 29 02:00 PM 2011 P
+%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% BASIC CLASS FILE 
-%% PNAStwo for two column articles is called by default.
-%% Uncomment PNASone for single column articles. One column class
-%% and style files are available upon request from pnas@nas.edu.
-%% (uncomment means get rid of the '%' in front of the command)
+% Title page
+% - author contributions
+%- the article title
+%- the full name(s) of all author(s), affiliation(s), e-mail address(es) of all author(s)
+%- a short running title (abbreviated form of title) of less than 45 characters including spaces
+%- up to 10 keywords for indexing purposes. It is very important that the keywords be chosen carefully.
+%- the type of article (Ideas and Perspectives, Letter, or Reviews and Syntheses)
+%- the number of words in the abstract, in the manuscript as a whole, and in the main text (excluding abstract, acknowledgements, references, table and figure legends)
+%- the number of references (or for papers with extensive meta-analyses, the number of data-source references AND the number of general references)
+%- the number of figures and tables
+%- the name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent.
+%
 
-%\documentclass{pnasone}
-\documentclass{pnastwo}
+%\documentclass[authoryear,5p]{elsarticle}
+\documentclass[authoryear,preprint,11pt]{elsarticle}
+\bibliographystyle{elsarticle-harv}
+\usepackage{graphicx}
+\usepackage{amsmath,amsfonts}
+\usepackage{lineno}
+\linenumbers
+\usepackage{subfigure}
+\usepackage[pdftex]{color}
+\definecolor{darkblue}{rgb}{0,0,0.5}
+\definecolor{darkgreen}{rgb}{0,0.5,0}
+%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
+\usepackage[pdftex, colorlinks]{hyperref}
+\textwidth 6.75in
+\oddsidemargin -0.15in
+\evensidemargin -0.15in
+\textheight 9in
+\topmargin -0.5in
+\newcommand{\ud}{\mathrm{d}}
+\newcommand{\E}{\mathrm{E}}
+\newcommand{\C}{\mathrm{Cov}}
+\newcommand{\V}{\mathrm{Var}}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Changing position of text on physical page:
-%% Since not all printers position
-%% the printed page in the same place on the physical page,
-%% you can change the position yourself here, if you need to:
+\journal{\tiny Ecology Letters} 
+\begin{document}
+\begin{frontmatter}
+  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
+  \author[cpb]{Carl Boettiger\corref{cor1}}
+  \ead{cboettig@ucdavis.edu, 610-389-6087}
+  \author[esp]{Alan Hastings}
+  %\author[info]{}
+  %\author[davis]{}
+  \cortext[cor1]{Corresponding author.}
+  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
+  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}
 
-% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
-                         %  physical page; positive dimension will lower it.
-
-                         %% You may set the dimension to the size that you need.
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% OPTIONAL GRAPHICS STYLE FILE
-
-                         %% Requires graphics style file (graphicx.sty), used for inserting
-                         %% .eps files into LaTeX articles.
-                         %% Note that inclusion of .eps files is for your reference only;
-                         %% when submitting to PNAS please submit figures separately.
-
-                         %% Type into the square brackets the name of the driver program 
-                         %% that you are using. If you don't know, try dvips, which is the
-                         %% most common PC driver, or textures for the Mac. These are the options:
-
-                         % [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
-                         % [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
-                         % [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]
-
-                         \usepackage[pdftex]{graphicx}
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% OPTIONAL POSTSCRIPT FONT FILES
-
-                         %% PostScript font files: You may need to edit the PNASoneF.sty
-                         %% or PNAStwoF.sty file to make the font names match those on your system. 
-                         %% Alternatively, you can leave the font style file commands commented out
-                         %% and typeset your article using the default Computer Modern 
-                         %% fonts (recommended). If accepted, your article will be typeset
-                         %% at PNAS using PostScript fonts.
-
-                         % Choose PNASoneF for one column; PNAStwoF for two column:
-                         %\usepackage{PNASoneF}
-                         %\usepackage{pnastwof}
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% ADDITIONAL OPTIONAL STYLE FILES
-
-                         %% The AMS math files are commonly used to gain access to useful features
-                         %% like extended math fonts and math commands.
-
-                         \usepackage{amssymb,amsfonts,amsmath}
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% OPTIONAL MACRO FILES
-                         %% Insert self-defined macros here.
-                         %% \newcommand definitions are recommended; \def definitions are supported
-
-                         %\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
-                         %\def\s{\sigma}
-
-                         \newcommand{\ud}{\mathrm{d}}
-                         \renewcommand{\thefootnote}{\fnsymbol{footnote}}
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         %% Don't type in anything in the following section:
-                         %%%%%%%%%%%%
-                         %% For PNAS Only:
-                         \contributor{Submitted to Proceedings
-                         of the National Academy of Sciences of the United States of America}
-                         \url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
-                         \copyrightyear{2008}
-                         \issuedate{Issue Date}
-                         \volume{Volume}
-                         \issuenumber{Issue Number}
-                         %%%%%%%%%%%%
-
-                         \begin{document}
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-                         %% For titles, only capitalize the first letter
-                         %% \title{Almost sharp fronts for the surface quasi-geostrophic equation}
-
-                         \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
-
-                         %% Enter authors via the \author command.  
-                         %% Use \affil to define affiliations.
-                         %% (Leave no spaces between author name and \affil command)
-
-                         %% Note that the \thanks{} command has been disabled in favor of
-                         %% a generic, reserved space for PNAS publication footnotes.
-
-                         %% \author{<author name>
-                         %% \affil{<number>}{<Institution>}} One number for each institution.
-                         %% The same number should be used for authors that
-                         %% are affiliated with the same institution, after the first time
-                         %% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
-                         %% Then, before last author ...
-                         %% \and
-                         %% \author{<author name>
-                         %% \affil{<number>}{}}
-
-                         %% For example, assuming Garcia and Sonnery are both affiliated with
-                         %% Universidad de Murcia:
-                         %% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
-                         %% United Kingdom},
-                         %% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
-                         %% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}
-
-                         \author{Carl Boettiger\affil{1}{Center for Population Biology, University of California, Davis}, 
-                         Alan Hastings\affil{2}{Department of Environmental Science and Policy, University of California, Davis} }
-
-                         \contributor{Submitted to Proceedings of the National Academy of Sciences
-                         of the United States of America}
-
-                         %% The \maketitle command is necessary to build the title page.
-                         \maketitle
-
-                         %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-                         \begin{article}
+ % \address[info]{ \\ 
+  %              }
 
@@
 
-                             %% When adding keywords, separate each term with a straight line: |
-                             \keywords{ early warning signals | tipping point | alternative stable states | likelihood methods }
+  \begin{keyword}
+early warning signals \sep tipping point \sep alternative stable states \sep likelihood methods 
+   \end{keyword}
+ \end{frontmatter}
+\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
+  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
+  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
+}
 
-                             %% Optional for entering abbreviations, separate the abbreviation from
-                             %% its definition with a comma, separate each pair with a semicolon:
-                             %% for example:
-                             %% \abbreviations{SAM, self-assembled monolayer; OTS,
-                             %% octadecyltrichlorosilane}
-
-                             % \abbreviations{}
-
-                             %% The first letter of the article should be drop cap: \dropcap{}
-                             %\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts
-
-                             %% Enter the text of your article beginning here and ending before
-                             %% \begin{acknowledgements}
-                             %% Section head commands for your reference:
-                             %% \section{}
-                             %% \subsection{}
-                             %% \subsubsection{}
-
-\dropcap{T}here is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\cite{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
-Many important ecosystems may currently be threatened with collapse, including corals~\cite{Bellwood2004}, fisheries~\cite{Berkes2006}⁠, lakes~\cite{Carpenter2011}, and semi-arid ecosystems~\cite{Kefi2007}⁠.
+\section{Introduction}
+There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
+Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠.
 Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
@@
 it is important to develop the ability to predict impending regime shifts based on early warning signs. 
-With a good model of the system, detail-oriented approaches could be useful~\cite{Lenton2009},
-but in cases where specific models are not available general approaches are needed~\cite{Scheffer2009}⁠.
-We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system.\footnote{A system involving a local bifurcation where the eigenvalue approaches zero \cite{Scheffer2001, Scheffer2009}.}
+With a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009},
+but in cases where specific models are not available general approaches are needed~\citep{Scheffer2009}⁠.
+
+We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
+(\emph{i.e.} where the dominant eigenvalue passes through zero,
+as discussed in~\citet{Scheffer2001, Scheffer2009}). 
 Making this assumption provides a ``best-case'' test of detection approaches;
@@
 
+\subsection{The implicit model}
+Foundational work on early warning signals has operated under the often-implicit assumption that the system dynamics contain a saddle-node bifurcation and looks for patterns that would be associated with approaching this bifurcation point.
+Such methods
+
+\subsection{Explicit use of the model}
+Detection of warning signals as model choice (``non-parametric'' Carpenter \& Brock, unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)
+
+\subsection{Why information criteria will not serve.}
+
+\subsection{Beyond hypothesis testing}
+Hypothesis testing is built upon the objective of
+
+\citet{Seekell2011}
+
+\subsection{Visualizing the informativeness through ROC curves}
+
+%% EDIT -- 
 Though previous research has identified patterns such as increasing variance or autocorrelation
@@
 Real-world applications of early warning signals will not be able to point to such differences between controlled and
-manipulated systems that have been utilized in experimental studies of early warning~\cite{Drake2010, Carpenter2011}⁠.
-While the number of potential patterns grows steadily~\cite{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
-there are few examples that statistically quantify the patterns~\cite{Dakos2008, Dakos2011},⁠
-and fewer still that statistically estimate the false alarm rate~\cite{Dakos2008} or 
+manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠.
+While the number of potential patterns grows steadily~\citep{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
+there are few examples that statistically quantify the patterns~\citep{Dakos2008, Dakos2011},⁠
+and fewer still that statistically estimate the false alarm rate~\citep{Dakos2008} or 
 the probability of missed detection⁠. 
-We need to quantify such uncertainty and error rates to compare between possible methods~\cite{Contamin2009}
-and address the continual concern about how much data is necessary⁠ for robust detection~\cite{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
+We need to quantify such uncertainty and error rates to compare between possible methods~\citep{Contamin2009}
+and address the continual concern about how much data is necessary⁠ for robust detection~\citep{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
 
+%% EDIT
 Detection of impending transitions is not just a scientific question 
@@
 These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Supp. materials or main text?  %
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-                             %Appendix text...
-                             %% or, for appendix with title, use square brackets:
-                             %% \appendix[Appendix Title]
+\section{Results}
 
-                             \begin{acknowledgments}
+We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
+We explore two empirical examples:
+deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\citep{Dakos2008},
+and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\citep{Drake2010}.  
+For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
+and the same system in a constant environment. 
+The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
+against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
+While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
+we need a way to identify such cases where we lack the data to trust the pattern.
 
+The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
+frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
+The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
+For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
+set to a 5\% false positive rate:
+25\% in the simulated set,
+5.0\% in the \emph{Daphnia} set,
+and 5.4\% in the Glaciation data set.
+Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
+would be correctly identified from the respective systems.
+The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
+as we would in classical hypothesis testing.  
+We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
+and read off the corresponding false positives.  
+For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
+while for likelihood a 55\%, 35\%, 0\% false positive.
+Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
+as illustrated in Fig. 1.  
+Comparing this threshold with the observed value classifies the signal as detected or undetected, (see Supporting Information).
+This flexibility is particularly valuable in the context of critical transitions, 
+where data are often limited and the consequences of missed detection can be far more severe than false alarms.
+
+%%%%%%% Converting from power to detection %%%%%%%%%%%%
+
+\section{Discussion}
+A pressing question in the application of early warning signals is quantifying
+when the data are sufficient for detection to be reliable~\citep{Scheffer2009, Scheffer2010, Inman2011}. 
+Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
+for each of the indicators (Fig 4). 
+Using our generic models estimated from the original data as in Fig. 3, 
+we generate the corresponding distributions by simulating under the models as before, 
+but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
+would be hampered or improved with more or less data.  
+While all methods improve as the number of data points increase, 
+the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
+as expected~\citep{Carpenter2011};⁠
+skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4). 
+%\citet{Dakos2011a} demonstrate that changes to environmental coupling decrease.  also use replicate simulations to assess expected response, but not the probability of accuracy on single replicates.
+%% Threshold
+%% Imagine a missed detection of no more than 10% is acceptable.
+
+By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
+our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
+Other literature has already begun to address the additional challenges posed when the underlying 
+dynamics do not correspond to these models~\citep{Hastings2010}.
+Our results illustrate that even in this best-case scenario, 
+reliable identification of warning signals from summary statistics can be difficult.  
+The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
+or even the distribution of that statistic under the null model of a stable system, 
+but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
+By estimating the ROC curve for a given set of data, 
+we can better avoid applying warning signals in cases of inadequate power.
+By taking advantage of the assumptions being made to write down a specific likelihood function,
+we can develop indicators that get the most from this data.
+
+%%% Model adequacy
+In any application of early warning signals, it is essential to address the question of model adequacy.  
+This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
+as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
+and systems that violate these assumptions can exhibit sudden transitions without any warning,
+such as chaotic systems or those with non-smooth potential functions~\citep{Hastings2010}.
+Sudden transitions have also been observed in ecological models in which 
+the variance deceases rather then increases as the system nears a bifurcation point~\citep{Schreiber2003, Schreiber2008, Dakos2011a}.  
+Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
+As the bifurcation results from the principle eigenvalue passing through zero, 
+the warning signal is expected in linear-order dynamics;
+estimation of the nonlinear model is less powerful and less accurate.  
+The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
+(introduced by the Poisson process demographic events), 
+demonstrates the accuracy under violation of these assumptions.
+
+%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
+%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
+%but would be amenable to the approach used here.
+
+The approach presented here offer a best-case scenario for these warning signals,
+illustrating that substantial uncertainty can remain despite the appearance of a positive increase.
+Our results highlight the need to quantify uncertainty in the application of early warning signals,
+and suggest that a hypothesis testing approach may be too restrictive a framework for that purpose.
+Using the ROC curve, we obtain a more general picture of the trade-off between false alarms and failed detections
+than can be provided by a significance test at a fixed false-alarm rate.  
+The curve can guide the choice of threshold used to indicate a warning signal on a case-by-case basis.
+
+The conclusion is not simply that likelihood approaches are more reliable, 
+but rather more broadly that warning signals should consider
+the inherent trade-off between sensitivity and accuracy,
+and must quantify how this trade-off depends on both the indicators used and the data available.  
+The approach developed here estimates the risk of both failed detection and false alarms;
+concepts which are critical to prediction-based management.  
+Using the methods we have outlined when designing early warning strategies for natural systems
+can ensure that data collection has adequate power to offer a reasonable chance of detection.
+
+\section{Acknowledgments}
 S.
@@
 This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
+Data and code are available at: https://github.com/cboettig/warningsignals.
 
-Author Contributions
-C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper.
+ % Computational Sciences Graduate Fellowship from the Department of Energy under grant number DE-FG02-97ER25308. 
+ \section{References}%bibliography
+ \bibliography{boettiger}
 
-Data and code available at: https://github.com/cboettig/warningsignals. 
-                             \end{acknowledgments}
-
-                             %% PNAS does not support submission of supporting .tex files such as BibTeX.
-                             %% Instead all references must be included in the article .tex document. 
-                             %% If you currently use BibTeX, your bibliography is formed because the 
-                             %% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-                             %% .tex document. To conform to PNAS requirements, copy the reference listings
-                             %% from your .bbl file and add them to the article .tex file, using the
-                             %% bibliography environment described above.
-
-                             %%  Contact pnas@nas.edu if you need assistance with your
-                             %%  bibliography.
-
-                             % Sample bibliography item in PNAS format:
-                             %% \bibitem{in-text reference} comma-separated author names up to 5,
-                             %% for more than 5 authors use first author last name et al. (year published)
-                             %% article title  {\it Journal Name} volume #: start page-end page.
-                             %% ie,
-                             % \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
-                             % A short C-terminal sequence is necessary and sufficient for the
-                             % targeting of chitinases to the plant vacuole. 
-                             % {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-                             %% Enter the largest bibliography number in the facing curly brackets
-                             %% following \begin{thebibliography}
-
-                             \bibliographystyle{/home/cboettig/Documents/tex/pnas.bst}
-                             \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
-                           \end{article}
-                           %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-                           %% Adding Figure and Table References
-                           %% Be sure to add figures and tables after \end{article}
-                           %% and before \end{document}
-
-                           %% For figures, put the caption below the illustration.
-                           %%
-                           %% \begin{figure}
-                           %% \caption{Almost Sharp Front}\label{afoto}
-                           %% \end{figure}
- \begin{figure*}
+\begin{figure}[hb]
    \begin{center}
@@
 The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).
-Successive columns show the threshold increasing, tracing out the ROC curve (successively gray lines show the threshold used for the matching gray point added to the ROC curve).}
+Successive columns show the threshold increasing, tracing out the ROC curve.}
   \end{center}
- \end{figure*}
+ \end{figure}
 
- \begin{figure*}
+ \begin{figure}[h]
    \begin{center}

### 2012-01-13 boettiger.tex
rewrites..

@@
 Foundational work on early warning signals has operated under the often-implicit assumption that the system dynamics contain a saddle-node bifurcation and looks for patterns that would be associated with approaching this bifurcation point.
-Such methods
+The typical pattern has been a trend in a summary statistic such as variance, autocorrelation, skew, spectral ratio~\citep{Biggs2009}.
+While attractive for their simplicity, such approaches are subject to numerous challenges.
 
-\subsection{Explicit use of the model}
-Detection of warning signals as model choice (``non-parametric'' Carpenter \& Brock, unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)
+1) While superficially simple, 
+calculation of these statistics is subject to 
+several somewhat arbitrary decisions and additional assumptions,
+such as the choice of window size over which the statistic is computed,
+the choice of detrending,
+whether the consecutive averaging windows should overlap~\citep[\emph{e.g.}][]{Lenton2012}. % discusses these challenges, as does most of the others to lesser extent 
+% ergodic problem
+
+2) The warning signal, ``an increase in statistic $x$,'' is not quantitative,
+making it difficult to attribute statistical significance to this statement.
+While much of the literature employing these statistics simply avoids 
+any attempt to quantify the significance of the detection, 
+\citet{Dakos2008} relies on Kendall's correlation coefficient to quantify an increase,
+and then goes through some pains to consider potential null models
+that could be used to calculate statistical significance.  
+Other measures of increase, such as Pearson's, have also been proposed.
+
+3) Specifying an apprpriate null model for such a calculation is difficult.  
+A supposed strength of the summary statistic approach 
+the non-parametric significance test for Kendalls' tau tests against the null model that the sequence is independent,
+something that is clearly violated when the statistic is computed in a sliding window.
+Not surprisingly, \citet{Dakos2008} find this method overestimates the number of significant events relative to other nulls.  
+Other non-parametric methods are tempting, 
+such as shuffling the order of the data-points in the time-series,
+but this breaks even the natural autocorrelation expected in a stable system that is not approaching any transition,
+and thus should not be considered for a null.  
+Relying on a more explicit but flexible model %or class of models
+such as autogregressive processes is more consistent.  
+% Does this go here?
+We will see that such generic model-based approaches are becoming more common, 
+and argue how they can be used to maximum advantage
+% by simulation, likelihood fitting, and ROC curves.
+
+% See Seekell et al for a summary stat with a good null
+
+4) The underlying assumption that the system contains a saddle-node bifurcation
+can be easilyoverlooked in summary-statistic based approach.  
+For instance, variance may increase for a wide variety of reasons that 
+do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008} 
+or may not increase as a bifurcation is approached~\citep{Livnia2012, Dakos2011a}.
+
+% Does this section go here?
+A system may experience a critical transition due to a rare fluctuation,
+or a perturbation to the system state, %(Does this need a figure)?
+rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.  
+When applying detection approaches to data that have already experienced a critical transition,
+we must condition on the 
+and attempting to ascertain the cause of the transition
+These are alternative null hypotheses 
+ %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
+\citep{Ditlevsen2010} % telling these two cases apart
+\citep{Livina2011} noise induced transitions cannot be forecast.
+
+5) Summary-statistic approaches inherently lack power. 
+Methods for the detection of early warning signals are continually challenged by 
+inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
+After all, the challenge of extrapolating into the future across a dramatic shift is immense,
+there is essentially no historical president for successfully \& reliably predicting regime shifts. % \cite Tetlock?
+The signals one attempts to detect with these approaches are extremely weak.
+Worse still, beyond this intuitive notion, there is little quantification of just how powerful our methods are,
+even in artificial scenarios where uncertainy conforms to our parametric assumptions.  
+% Seekell Am Nat, Contamin Eco Apps
+We characterize the power
+
+\subsection{Explicit use of the models}
+Detection of warning signals as model choice (``non-parametric'' \citet{Carpenter2011e},  unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)
 

### 2012-01-18 boettiger.tex
edits

@@
  \end{frontmatter}
-\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
-  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
-  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
-}
+%\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
+%  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
+%  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
+% }
 
@@
 \subsection{The implicit model}
-Foundational work on early warning signals has operated under the often-implicit assumption that the system dynamics contain a saddle-node bifurcation and looks for patterns that would be associated with approaching this bifurcation point.
-The typical pattern has been a trend in a summary statistic such as variance, autocorrelation, skew, spectral ratio~\citep{Biggs2009}.
+Foundational work on early warning signals has operated under the often-implicit assumption
+that the system dynamics contain a saddle-node bifurcation and
+looks for patterns that would be associated with approaching this bifurcation point.
+The typical pattern has been a trend in a summary statistic such as
+variance, autocorrelation, skew, spectral ratio~\citep{Biggs2009}.
 While attractive for their simplicity, such approaches are subject to numerous challenges.
 
-1) While superficially simple, 
-calculation of these statistics is subject to 
+1) While superficially simple, calculation of these statistics is subject to 
 several somewhat arbitrary decisions and additional assumptions,
@@
 \subsection{Explicit use of the models}
+Several recent approaches to early warning signals have considered fitting models, though have not considered the model-choice problem.  
 Detection of warning signals as model choice (``non-parametric'' \citet{Carpenter2011e},  unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)

### 2012-01-18 boettiger.tex
updates

@@
 
-\journal{\tiny Ecology Letters} 
+\journal{\tiny Journal of the Royal Society Interface} 
 \begin{document}
@@
   \author[cpb]{Carl Boettiger\corref{cor1}}
-  \ead{cboettig@ucdavis.edu, 610-389-6087}
+  \ead{cboettig@ucdavis.edu}
   \author[esp]{Alan Hastings}
@@
 
-We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
-(\emph{i.e.} where the dominant eigenvalue passes through zero,
-as discussed in~\citet{Scheffer2001, Scheffer2009}). 
-Making this assumption provides a ``best-case'' test of detection approaches;
-we will demonstrate the resulting uncertainty can be very high.
-We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.
+%We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
+%(\emph{i.e.} where the dominant eigenvalue passes through zero,
+%as discussed in~\citet{Scheffer2001, Scheffer2009}). 
+%Making this assumption provides a ``best-case'' test of detection approaches;
+%we will demonstrate the resulting uncertainty can be very high.
+%We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.
 
-\subsection{The implicit model}
-Foundational work on early warning signals has operated under the often-implicit assumption
+\subsection{Summary statistics and the implicit model}
+Foundational work on early warning signals has operated under the often-implicit assumption %often unstated?
 that the system dynamics contain a saddle-node bifurcation and
@@
 While attractive for their simplicity, such approaches are subject to numerous challenges.
+In this paper we argue for a model-based approach, 
+and describe how this can be done in a way that best addresses these difficulties.
 
-1) While superficially simple, calculation of these statistics is subject to 
+\textbf{arbitrary windows}
+While superficially simple to use, the calculation of statistics is subject to 
 several somewhat arbitrary decisions and additional assumptions,
@@
 the choice of detrending,
-whether the consecutive averaging windows should overlap~\citep[\emph{e.g.}][]{Lenton2012}. % discusses these challenges, as does most of the others to lesser extent 
-% ergodic problem
+whether the consecutive averaging windows should overlap.  
+\citet{Lenton2012} compare the performance of different approaches to these issues,
+showing that choice of window and method used for detrending indeed influence the results,
+and each approach considered has pros and cons.  
+Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
+it is difficult to evaluate and compare between such methods.  
+A model-based approach side-steps these issue entirely.  
+% discusses these challenges, as does most of the others to lesser extent 
+% Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
+% and that it assumes stationarity, which is precisely the thing we're looking to change?  
+% Similarly the detrending approaches can remove the signal?
+% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points, and interpolating is incorrect, introducing fictitious correlations?
 
 2) The warning signal, ``an increase in statistic $x$,'' is not quantitative,
-making it difficult to attribute statistical significance to this statement.
+making it difficult to compare between signals or to
+attribute a statistical significance to the detection.
 While much of the literature employing these statistics simply avoids 
@@
 
-3) Specifying an apprpriate null model for such a calculation is difficult.  
-A supposed strength of the summary statistic approach 
-the non-parametric significance test for Kendalls' tau tests against the null model that the sequence is independent,
+3) Specifying an appropriate null model is difficult.  
+% simulation of replicates under the original model
+As the summary statistic approach 
+the non-parametric significance test for Kendall's tau tests against the null model that the sequence is independent,
 something that is clearly violated when the statistic is computed in a sliding window.
@@
 do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008} 
-or may not increase as a bifurcation is approached~\citep{Livnia2012, Dakos2011a}.
+or may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
 
@@
 \citep{Ditlevsen2010} % telling these two cases apart
-\citep{Livina2011} noise induced transitions cannot be forecast.
+\citep{Livina2012} noise induced transitions cannot be forecast.
 

### 2012-01-19 boettiger.tex
writing

@@
 
-\textbf{arbitrary windows}
+\textbf{arbitrary windows}\\
 While superficially simple to use, the calculation of statistics is subject to 
@@
 % Similarly the detrending approaches can remove the signal?
-% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points, and interpolating is incorrect, introducing fictitious correlations?
+% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points,
+% and interpolating is incorrect, introducing fictitious correlations?
 
-2) The warning signal, ``an increase in statistic $x$,'' is not quantitative,
+\textbf{no quantitative measures}\\
+The warning signal, ``an increase in statistic $x$,'' is not quantitative,
 making it difficult to compare between signals or to
@@
 
-3) Specifying an appropriate null model is difficult.  
+\textbf{problem of null models}\\
+Specifying an appropriate null model is difficult.  
 % simulation of replicates under the original model
@@
 
-4) The underlying assumption that the system contains a saddle-node bifurcation
-can be easilyoverlooked in summary-statistic based approach.  
-For instance, variance may increase for a wide variety of reasons that 
-do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008} 
-or may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
+\textbf{hidden assumptions}\\
+The underlying assumption that the system contains a saddle-node bifurcation
+can be easily overlooked in summary-statistic based approach.  
+For instance, variance may increase for reasons that 
+do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
+Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
+Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
+
+Even when the system is characterized by 
+simple alternative stable states in the saddle-node model,
+transitions between states may be due to other mechanisms than a bifurcation, such as a purely noise-induced transition,
+or a
 
@@
 rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.
+
 When applying detection approaches to data that have already experienced a critical transition,
 we must condition on the
+
 and attempting to ascertain the cause of the transition

### 2012-01-20 boettiger.tex
writing

@@
-%        File: ecologyletters.tex
-%     Created: Sat Oct 29 02:00 PM 2011 P
-% Last Change: Sat Oct 29 02:00 PM 2011 P
-%
-
-% Title page
-% - author contributions
-%- the article title
-%- the full name(s) of all author(s), affiliation(s), e-mail address(es) of all author(s)
-%- a short running title (abbreviated form of title) of less than 45 characters including spaces
-%- up to 10 keywords for indexing purposes. It is very important that the keywords be chosen carefully.
-%- the type of article (Ideas and Perspectives, Letter, or Reviews and Syntheses)
-%- the number of words in the abstract, in the manuscript as a whole, and in the main text (excluding abstract, acknowledgements, references, table and figure legends)
-%- the number of references (or for papers with extensive meta-analyses, the number of data-source references AND the number of general references)
-%- the number of figures and tables
-%- the name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent.
-%
-
 %\documentclass[authoryear,5p]{elsarticle}
@@
 \newcommand{\V}{\mathrm{Var}}
+
+\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}
 
@@
 The typical pattern has been a trend in a summary statistic such as
-variance, autocorrelation, skew, spectral ratio~\citep{Biggs2009}.
+variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
+skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
+% consider additional citations
 While attractive for their simplicity, such approaches are subject to numerous challenges.
@@
 
-\textbf{arbitrary windows}\\
-While superficially simple to use, the calculation of statistics is subject to 
-several somewhat arbitrary decisions and additional assumptions,
+\subsubsection*{Arbitrary windows}\\
+While superficially simple to use, the calculation of these statistics is subject to 
+several arbitrary decisions and additional assumptions,
 such as the choice of window size over which the statistic is computed,
 the choice of detrending,
-whether the consecutive averaging windows should overlap.  
+and whether and by how much the consecutive time-windows used in the calculation should overlap.  
 \citet{Lenton2012} compare the performance of different approaches to these issues,
@@
 A model-based approach side-steps these issue entirely.  
-% discusses these challenges, as does most of the others to lesser extent 
-% Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
-% and that it assumes stationarity, which is precisely the thing we're looking to change?  
-% Similarly the detrending approaches can remove the signal?
-% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points,
-% and interpolating is incorrect, introducing fictitious correlations?
+\cb{
+ Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
+ and that it assumes stationarity, which is precisely the thing we're looking to change?  
+ Similarly could mention that the detrending approaches can remove the signal?
+ We could also mention that some popular approaches such as autocorrelation require evenly-spaced points,
+ and interpolating is incorrect, introducing fictitious correlations?
+}
 
-\textbf{no quantitative measures}\\
+\subsubsection*{No quantitative measures}\\
 The warning signal, ``an increase in statistic $x$,'' is not quantitative,
@@
 attribute a statistical significance to the detection.
-While much of the literature employing these statistics simply avoids 
-any attempt to quantify the significance of the detection, 
-\citet{Dakos2008} relies on Kendall's correlation coefficient to quantify an increase,
-and then goes through some pains to consider potential null models
-that could be used to calculate statistical significance.  
-Other measures of increase, such as Pearson's, have also been proposed.
+Some authors have suggested Kendall's correlation coefficient to quantify an increase~\citep{Dakos2008, Dakos20011} 
+in autocorrelation or variance.  
+Other measures of increase, such as Pearson's, have also been proposed~\citep{Drake2010}.  
+Much of the literature using summary statistics simply forgoes quantifying
+the increase or estimating significance.  
+While this may be satisfactory in experimental systems with controls and replicates to compare to,~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
+any real-world application of early warning methods will not have the luxury of controls and replicates.
+In such cases a quantitative definition of detection is essential.  
+\cb{ We will see that correlation tests in summary statistics are an ineffective way to do this.  }
 
-\textbf{problem of null models}\\
-Specifying an appropriate null model is difficult.  
-% simulation of replicates under the original model
-As the summary statistic approach 
-the non-parametric significance test for Kendall's tau tests against the null model that the sequence is independent,
-something that is clearly violated when the statistic is computed in a sliding window.
-Not surprisingly, \citet{Dakos2008} find this method overestimates the number of significant events relative to other nulls.  
-Other non-parametric methods are tempting, 
-such as shuffling the order of the data-points in the time-series,
-but this breaks even the natural autocorrelation expected in a stable system that is not approaching any transition,
-and thus should not be considered for a null.  
-Relying on a more explicit but flexible model %or class of models
-such as autogregressive processes is more consistent.  
-% Does this go here?
-We will see that such generic model-based approaches are becoming more common, 
-and argue how they can be used to maximum advantage
-% by simulation, likelihood fitting, and ROC curves.
+\subsubsection*{Problematic null models}\\
+Specifying an appropriate null model is also difficult.  
+Non-parametric null hypotheses seem to require the fewest assumptions but in fact can be the most problematic.  
+For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
+assumes only that the two variables are independent, but this is
+an assumption that is violated by the very experimental design:
+since one variable is time and the other a moving time-averaged window.  
+Under such a test any adequately large data set will find a significant result, regardless of whether a warning signal exists.
+A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
+this destroys the natural autocorrelation in the time series.  
+\cb{I feel citing \citet{Dakos2011} and \citet{Dakos2008} for these two examples would be impolite\dots} 
+Parametric null models have been more promising, and have brought us closer to a model-based approach where assumptions are explicit.  
+For instance, autoregressive models~\citet{Dakos2008} have been proposed, which are first parameterized from the data.  
+\cb{is this next line okay to leave in?}
+Estimating the null models directly from the data using a maximum likelihood framework, will provide the most powerful basis for this test.  
+\citet{Seekell2011} proposes a conditional heterscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}
 
-% See Seekell et al for a summary stat with a good null
+\cb{This is a particular concern given the tendency for warning signals to be found only in the largest data sets, 
+which has been taken as evidence that large data sets are required for successful detection,
+not that large data sets have higher false-positive rates under certain metrics.}
+\cb{A similar problem exists with the proposal that large deviations from the running variance,
+i.e. in excess of 2 standard deviations, be taken as evidence of an increase.
+This likewise guarantees the false positive rate goes to unity as the number of observations increases.}
 
-\textbf{hidden assumptions}\\
+\subsubsection*{Hidden assumptions}\\
 The underlying assumption that the system contains a saddle-node bifurcation
@@
 Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.   
+While no approach will be applicable to all classes of sudden transitions,
+it is certainly still useful to have an approach that detect transitions driven by 
+saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.
+
+% do we need the energy diagram figure to emphasize this is our assumption?
+Even restricting ourselves to this classic alternate stable state model,
+we must still consider at least three possible scenarios that could explain a sudden transition.  
+The first is a large perturbation of the system state, moving the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
+This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
+(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
+The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.  
+The third scenario is that a slowly varying parameter of the system leads to a bifurcation in which the current stable state is destroyed and the system drifts to the alternative stable state~\citep{Scheffer2009}.
+So while it is not the only explanation for a sudden transition between alternative stable states, it would still be valuable to know
+
+this ``blue sky catastrophe'' or saddle node bifurcation is the focus of
+
+System specific knowledge 
+Any theory of early warning signals must restrict itself to the subset of transitions that exhibit warning signals
+% a warning signal general enough to catch
 
@@
 
+%% TOPIC FOR ANOTHER NOTE
 When applying detection approaches to data that have already experienced a critical transition,
-we must condition on the
+we have conditioned on the fact the transition has occurred.  This is a very different problem than 
+forecasting the probabilities of systems that have not collapsed!  Prosecutor's fallacy!
 
@@
 Real-world applications of early warning signals will not be able to point to such differences between controlled and
-manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠.
+manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠
 While the number of potential patterns grows steadily~\citep{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods

### 2012-01-20 boettiger.tex
fallacy note added

@@
 Estimating the null models directly from the data using a maximum likelihood framework, will provide the most powerful basis for this test.  
-\citet{Seekell2011} proposes a conditional heterscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}
+\citet{Seekell2011} proposes a conditional heteroscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}
 
@@
 \subsubsection*{Hidden assumptions}\\
+Assumes that the system sits effectively at equilibrium around a stable state  
+and tests whether the system may be experiencing the slow change of a parameter that is weakening the stability of that state,
+moving the system towards a saddle-node bifurcation.  
+As such processes may be common in natural systems this is an important scenario to consider,
+but without the clarity of assumptions made in a modeling approach there is a risk that
+early warning signs will be used in situations where they do not apply.
+
+Differing system dynamics
 The underlying assumption that the system contains a saddle-node bifurcation
@@
 
-% do we need the energy diagram figure to emphasize this is our assumption?
+% this will be made precise later
+
+\cb{ do we need the energy diagram figure to emphasize this is our assumption?}
 Even restricting ourselves to this classic alternate stable state model,
 we must still consider at least three possible scenarios that could explain a sudden transition.
-The first is a large perturbation of the system state, moving the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
+
+%Differing causes of the transition.
+Even when we can exclude or ignore such dynamics and 
+restrict ourselves to systems that can support a saddle-node bifurcation,
+any early warning signals approach must further assume that a changing parameter 
+has brought the system closer to the bifurcation.  This excludes at least three alternative
+explanations for the transition.  
+The first is that a large perturbation of the system state 
+has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.  
 This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast. 
@@
 The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.  
-The third scenario is that a slowly varying parameter of the system leads to a bifurcation in which the current stable state is destroyed and the system drifts to the alternative stable state~\citep{Scheffer2009}.
-So while it is not the only explanation for a sudden transition between alternative stable states, it would still be valuable to know
+As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.  
+The third scenario is that the system does pass through a saddle-node bifurcation, 
+but rather than gradually and monotonically approaching the critical point, the 
+bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.
 
-this ``blue sky catastrophe'' or saddle node bifurcation is the focus of
+% Of course the literature has always stuck carefully to these assumptions, but there has sometimes been the suggestion that the summary statistics themselves were somehow more general than this, and that framing these assumptions in explicit and precise formulae, in models, would be misleading.
 
-System specific knowledge 
-Any theory of early warning signals must restrict itself to the subset of transitions that exhibit warning signals
-% a warning signal general enough to catch
-
-Even when the system is characterized by 
-simple alternative stable states in the saddle-node model,
-transitions between states may be due to other mechanisms than a bifurcation, such as a purely noise-induced transition,
-or a
-
-% Does this section go here?
-A system may experience a critical transition due to a rare fluctuation,
-or a perturbation to the system state, %(Does this need a figure)?
-rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.
-
-%% TOPIC FOR ANOTHER NOTE
-When applying detection approaches to data that have already experienced a critical transition,
-we have conditioned on the fact the transition has occurred.  This is a very different problem than 
-forecasting the probabilities of systems that have not collapsed!  Prosecutor's fallacy!
+%%%% TOPIC FOR ANOTHER NOTE: Early warning signals and the Prosecutor's Fallacy %%%%%%%
+%When applying detection approaches to data that have already experienced a critical transition,
+%we have conditioned on the fact the transition has occurred.  This is a very different problem than 
+%forecasting the probabilities of systems that have not collapsed!
 
@@
 \citep{Ditlevsen2010} % telling these two cases apart
-\citep{Livina2012} noise induced transitions cannot be forecast.
 

### 2012-01-20 fallacy.tex
fallacy note added

--- 2012-01-19 boettiger.tex
@@
 
+\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}
+
 \journal{\tiny Journal of the Royal Society Interface} 
@@
 \begin{frontmatter}
-  \title{Quantifying Limits to Detection of Early Warning for Critical Transitions}
+  \title{Early Warning Signals and the Prosecutor's Fallacy}
   \author[cpb]{Carl Boettiger\corref{cor1}}
@@
 
-Catastrophic regime shifts in complex natural systems may be averted through advanced detection. 
-Recent work has provided a proof-of-principle that many systems approaching a catastrophic transition may be identified 
-through the lens of early warning indicators such as rising variance or increased return times.  
-Without the benefit of hindsight of a collapse, 
-applications of these approaches must quantify how reliable different indicators are in avoiding false alarms, 
-and how sensitive they are to missing subtle warning signs.  
-We propose an approach which quantifies this trade-off between reliability and sensitivity, 
-allows comparisons between different indicators, 
-and estimates what data are necessary to achieve a desired error rate. 
-We show these error rates can be quite severe for common indicators even under favorable assumptions, 
-and illustrate how a likelihood-based indicator can improve this performance.
+  Early warning signals have been proposed to forecast the possibility of a 
+  critical transition, such as the eutriphication of a lake, desertification
+  of savanna, the collapse of a coral reef, the end of a glacial period, or 
+  the onset of epilepsy in the brain.  While such transitions may be difficult
+  to study in the laboratory setting and unethical to introduce on the a larger
+  natural scales where they may occur, it has often been observed that history
+  has already performed the experiment for us, time and again.  Here we point 
+  out a critical difference between selecting systems for study based on the 
+  fact that we have observed a critical transition and those systems for which
+  we wish to forecast the approach of a transition. This difference arising
+  from conditional probabilities is often known as the the Prosecutor's 
+  Fallacy.  As we pursue this important question of early warning signals, 
+  we must be careful to avoid this trap.
 
@@
  \end{frontmatter}
-%\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
-%  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
-%  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
-% }
-
-\section{Introduction}
-There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
-Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠.
-Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
-and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
-it is important to develop the ability to predict impending regime shifts based on early warning signs. 
-With a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009},
-but in cases where specific models are not available general approaches are needed~\citep{Scheffer2009}⁠.
-
-%We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
-%(\emph{i.e.} where the dominant eigenvalue passes through zero,
-%as discussed in~\citet{Scheffer2001, Scheffer2009}). 
-%Making this assumption provides a ``best-case'' test of detection approaches;
-%we will demonstrate the resulting uncertainty can be very high.
-%We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.
-
-\subsection{Summary statistics and the implicit model}
-Foundational work on early warning signals has operated under the often-implicit assumption %often unstated?
-that the system dynamics contain a saddle-node bifurcation and
-looks for patterns that would be associated with approaching this bifurcation point.
-The typical pattern has been a trend in a summary statistic such as
-variance, autocorrelation, skew, spectral ratio~\citep{Biggs2009}.
-While attractive for their simplicity, such approaches are subject to numerous challenges.
-In this paper we argue for a model-based approach, 
-and describe how this can be done in a way that best addresses these difficulties.
-
-\textbf{arbitrary windows}\\
-While superficially simple to use, the calculation of statistics is subject to 
-several somewhat arbitrary decisions and additional assumptions,
-such as the choice of window size over which the statistic is computed,
-the choice of detrending,
-whether the consecutive averaging windows should overlap.  
-\citet{Lenton2012} compare the performance of different approaches to these issues,
-showing that choice of window and method used for detrending indeed influence the results,
-and each approach considered has pros and cons.  
-Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
-it is difficult to evaluate and compare between such methods.  
-A model-based approach side-steps these issue entirely.  
-% discusses these challenges, as does most of the others to lesser extent 
-% Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
-% and that it assumes stationarity, which is precisely the thing we're looking to change?  
-% Similarly the detrending approaches can remove the signal?
-% Should we also mention that some popular approaches such as autocorrelation require evenly-spaced points,
-% and interpolating is incorrect, introducing fictitious correlations?
-
-\textbf{no quantitative measures}\\
-The warning signal, ``an increase in statistic $x$,'' is not quantitative,
-making it difficult to compare between signals or to
-attribute a statistical significance to the detection.
-While much of the literature employing these statistics simply avoids 
-any attempt to quantify the significance of the detection, 
-\citet{Dakos2008} relies on Kendall's correlation coefficient to quantify an increase,
-and then goes through some pains to consider potential null models
-that could be used to calculate statistical significance.  
-Other measures of increase, such as Pearson's, have also been proposed.
-
-\textbf{problem of null models}\\
-Specifying an appropriate null model is difficult.  
-% simulation of replicates under the original model
-As the summary statistic approach 
-the non-parametric significance test for Kendall's tau tests against the null model that the sequence is independent,
-something that is clearly violated when the statistic is computed in a sliding window.
-Not surprisingly, \citet{Dakos2008} find this method overestimates the number of significant events relative to other nulls.  
-Other non-parametric methods are tempting, 
-such as shuffling the order of the data-points in the time-series,
-but this breaks even the natural autocorrelation expected in a stable system that is not approaching any transition,
-and thus should not be considered for a null.  
-Relying on a more explicit but flexible model %or class of models
-such as autogregressive processes is more consistent.  
-% Does this go here?
-We will see that such generic model-based approaches are becoming more common, 
-and argue how they can be used to maximum advantage
-% by simulation, likelihood fitting, and ROC curves.
-
-% See Seekell et al for a summary stat with a good null
-
-\textbf{hidden assumptions}\\
-The underlying assumption that the system contains a saddle-node bifurcation
-can be easily overlooked in summary-statistic based approach.  
-For instance, variance may increase for reasons that 
-do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
-Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
-Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
-
-Even when the system is characterized by 
-simple alternative stable states in the saddle-node model,
-transitions between states may be due to other mechanisms than a bifurcation, such as a purely noise-induced transition,
-or a
-
-% Does this section go here?
-A system may experience a critical transition due to a rare fluctuation,
-or a perturbation to the system state, %(Does this need a figure)?
-rather than a bifurcation~\citep{Scheffer2001, Scheffer2009}.
-
-When applying detection approaches to data that have already experienced a critical transition,
-we must condition on the
-
-and attempting to ascertain the cause of the transition
-These are alternative null hypotheses 
- %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
-\citep{Ditlevsen2010} % telling these two cases apart
-\citep{Livina2012} noise induced transitions cannot be forecast.
-
-5) Summary-statistic approaches inherently lack power. 
-Methods for the detection of early warning signals are continually challenged by 
-inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
-After all, the challenge of extrapolating into the future across a dramatic shift is immense,
-there is essentially no historical president for successfully \& reliably predicting regime shifts. % \cite Tetlock?
-The signals one attempts to detect with these approaches are extremely weak.
-Worse still, beyond this intuitive notion, there is little quantification of just how powerful our methods are,
-even in artificial scenarios where uncertainy conforms to our parametric assumptions.  
-% Seekell Am Nat, Contamin Eco Apps
-We characterize the power
-
-\subsection{Explicit use of the models}
-Several recent approaches to early warning signals have considered fitting models, though have not considered the model-choice problem.  
-Detection of warning signals as model choice (``non-parametric'' \citet{Carpenter2011e},  unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)
-
-\subsection{Why information criteria will not serve.}
-
-\subsection{Beyond hypothesis testing}
-Hypothesis testing is built upon the objective of
-
-\citet{Seekell2011}
-
-\subsection{Visualizing the informativeness through ROC curves}
-
-%% EDIT -- 
-Though previous research has identified patterns such as increasing variance or autocorrelation
-in time-series approaching a critical transition,
-there has been no method for forecasting the \emph{probability} of a collapse from a natural system.
-Real-world applications of early warning signals will not be able to point to such differences between controlled and
-manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠.
-While the number of potential patterns grows steadily~\citep{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
-there are few examples that statistically quantify the patterns~\citep{Dakos2008, Dakos2011},⁠
-and fewer still that statistically estimate the false alarm rate~\citep{Dakos2008} or 
-the probability of missed detection⁠. 
-We need to quantify such uncertainty and error rates to compare between possible methods~\citep{Contamin2009}
-and address the continual concern about how much data is necessary⁠ for robust detection~\citep{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
-
-%% EDIT
-Detection of impending transitions is not just a scientific question 
-and the goal should not solely be to detect with a 5\% chance of a false positive:
-a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
-There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
-with which it will fail to provide any advanced warning. 
-Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
-In this paper, we
-(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
-(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
-(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of insensitive indicators 
-(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.
-
-We illustrate the trade-off between false alarms and failed detection using 
-receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
-The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
-The closer these distributions are to one-another, the more severe the trade-off.  
-If the distributions overlap exactly, the ROC curve has a constant slope of unity.  
-The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
-Different early-warning indicators will vary in their sensitivity to detect differences between stable systems and those approaching a critical transition, making the ROC curves a natural way to compare their performance.
-Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
-we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
-The area under the ROC curve provides a summary of how well the method compares to a random guess --
-an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance, while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.
-
-To generate ROC curves for common early warning indicators such as 
-increases in autocorrelation, variance, and skew, 
-we must quantify the increase and generate the expected distributions of the focal statistic.  
-While the existing literature frequently presents only a 
-% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
-% so this is not a reasonable statistical characterization
-visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006},
-we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\citep{Dakos2008, Dakos2011, Dakos2009}.
-Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
-suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
-In principle, other quantifications of increasing variance, autocorrelation, etc.,
-may perform better than $\tau$ and could be analyzed just as well in this framework. 
-While we use this established metric of increase, we also introduce a novel warning indicator
-that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).
-
-Having identified an indicator statistic,
-we must be able to generate its distribution under both the scenario of a stable system and 
-one approaching a transition before we can compute the ROC curve.
-We fit general stochastic models representing common bifurcations
-(\emph{e.g.} saddle-node~\citep{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\citep{Drake2010})
-for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
-By estimating the statistic (\emph{e.g.} $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
-Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\citep{Cox1961}⁠ instead of $\tau$,
-which provides a more direct and more powerful test than detecting an increase in summary statistics.
-As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.
-
-\section{Materials \& Methods}\label{methods}
-
-\subsection{Statistical methodology}
-For each given time-series, parameters are estimated by maximum likelihood 
-for a stable model and a model of a system approaching a critical transition (see Models).
-The models are stochastic differential equations (SDEs)
-that can be used to simulate data under the same sampling regime as the original data, 
-or at any specified set of time points.
-Distributions are determined by a parametric bootstrap:
-500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
-For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation)
-were computed over a sliding window equal to half the observed time (other window sizes had worse performance, see R package), 
-and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.
-The estimates of $\tau$ under the replicate simulations from the stable system
-and the estimates under the deteriorating stability system
-make up the two distributions which are integrated at a sliding threshold to create the ROC curves, as illustrated in Fig 1.
-
-The likelihood test is performed as in the parametric bootstrap; 
-but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\citep{Cox1961}.
-Refitting the models on each replicate, 
-rather merely computing the likelihood under the original estimates,
-makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\citep{Huelsenbeck1996}.
-When replicates are created by simulation under the estimated models,
-we record data at the same time intervals observed in the original data (as in Fig. 3, main text). 
-It is straight forward to allow this interval to be different,
-representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve, Fig 4.
-The source code for implementing each of the analyses and generating the plots presented is provided as an R package, \verb|warningsignals|.
-
-\subsection{Example Data}
-The simulated data sets are produced with 40 data points
-sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, Eq~\eqref{master}.
-The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.
-
-The Glaciation data is accessible from NOAA~\citep{Petit1999}.
-The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
-to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\citep{Dakos2008},
-analyzing the third glaciation event, consisting of 121 sample points. 
-The match is not exact since estimates the de-trending window size manually,
-but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
-De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008}, 
-which did not apply to the other data sets~\citep{Drake2010}.
-
-The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
-This individual replicates was chosen as an example of a single replicate 
-that showed a statistically significant correlation in variance over window of time where critical slowing down was expected.
-All data and code for simulations and data processing are found in the R package.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Supp. materials or main text?  %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Models}
-The detection of early warning is based on a \emph{local} theory,
-predicting changes in the dynamics near the current attractor.
-Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
-but we can approximate the model near the bifurcation.  
-Many possible complex models will correspond to this approximation.
-
-%\textbf{Saddle-node bifurcation.}
-One of the most common critical transitions modeled in ecological systems with alternative stable states is a saddle-node bifurcation~\citep[\{emph{e.g.}[]{Scheffer2001, Scheffer2009, Guttal2008a, VanNes2007, Biggs2009}.
-The individual-based simulation follows the master equation, 
-\begin{align}
-  P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
-  b_n &= \frac{e K x^2}{X^2 + h_t^2} \\ 
-  d_n &= e X_t - a_t   
-\end{align}
-This model can experience a saddle-node bifurcation through increased mortality by increasing either $e$ or $a$. 
-The SDE approximation holds for reasonable population sizes:
-\begin{equation}
-dX_t = \left( \frac{e K x^2}{X^2 + h_t^2} - e X_t - a_t\right) dt + \sigma \sqrt{ \frac{e K x^2}{X^2 + h_t^2} + e X_t + a_t} dB_t \label{ass}
-\end{equation}
-This parameterization is only an example, of course many others are possible~\citep{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
-Under appropriate transformation, all can be written in the same normal form for he saddle-node bifurcation,
-\begin{equation}
-\frac{\ud x}{\ud t} = r_t- x^2.
-\label{saddle-node}
-\end{equation}
-While in principle this model could be estimated directly from the data,
-if all observations have been prior to the crash (\emph{i.e.} near the desirable equilibrium)
-we have no data which is informative about the nonlinearity and such estimates will be unreliable.  
-Fortunately the theory of early warning signals is a \emph{local} one, 
-and we can linearize around the equilibrium to obtain a model we can estimate both more easily and more accurately.
-
-Transforming the canonical form to allow for an arbitrary mean $\theta$,
-the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$,
-which gives the model we refer to as the Linearized Saddle Node, LSN: 
-\begin{equation}
-\ud X = 2\sqrt{ r_t } (\phi - X_t)\ud t + \sigma\sqrt{\phi } \ud B_t. \label{LSN}
-\end{equation}
-
-\citet{Drake2010} induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} to test early warning signals.  
-This bifurcation also arises in many common models.  For example, the stochastic version of~\citet{Levins1969}'s model (a logistic model),
-\begin{equation}
-\ud X_t = \left( c_t X_t (1-X_t/K) - e_t X_t \right) \ud t + \sigma \sqrt{\frac{e_t}{c_t}} \ud B_t \label{levins},
-\end{equation}
-where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent)
-colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).  
-The model contains a transcritical bifurcation when $c_t < e_t$.  The normal form of the bifurcation is
-\begin{equation}
-\frac{\ud x}{\ud t} = r_t x - x^2 
-\label{transcritical}
-\end{equation}
-We can express the linearized stochastic dynamics then by the time dependent mean-reverting process. 
-We refer to this model as the linearized transcritical bifurcation, LTC.  
-\begin{equation}
-\ud X_t = r_t (r_t - X_t/K) \ud t + \sigma \sqrt{1+r_t} \ud B_t \label{LTC}
-\end{equation}
-
-%The pattern of the saddle-node differs from that of the transcritical in the square-root dependence on the bifurcation parameter.
-Note that the transcritical bifurcation involves a gradual decline towards zero,
-whereas the saddle-node bifurcation involves a sudden transition from a non-trivial equilibrium point.
-Thus saddle-node bifurcations are the more typical focus of early warning efforts, even though both exhibit the phenomenon of critical slowing down.
-
-Likelihood calculations also require specifying the rate of change of $r_t$, 
-which we take to first order as
-\begin{equation}
-r_t = r_0 - m t.
-\label{R_t}
-\end{equation}
-Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
-When the system is stable, $r_t$ is constant and both models~\eqref{LSN} and~\eqref{LTC} reduce to a simple Ornstein-Uhlenbeck process,
-\begin{equation}
-\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
-\end{equation}
-This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.
-
-\subsection{Likelihood calculations}\label{likelihood}
-The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
-\begin{equation}
-\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
-\end{equation}
-For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
-\begin{align}
- \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
-\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
-  \label{general_moments}
-\end{align}
-
-For the OU process, we can solve this in closed form, 
-\begin{align}
-  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
-V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
-\label{OUsoln}
-\end{align}
-
-For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
-For the linearized transcritical:
-
-\begin{align}
-\frac{\ud }{\ud t} E(x_i| M = \text{LTC})&=  r(t)(r(t) - x_i) \\
-\frac{\ud}{\ud t} V(x_i| M = \text{LTC}) &=  -2 r(t) V(x_i|M) + (1+r(t))\sigma^2 
-\label{LTCsoln}
-\end{align}
-
-For the linearized saddle-node bifurcation:
-\begin{align}
-\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
-\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
-\label{LSNsoln}
-\end{align}
-
-These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.  
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Supp. materials or main text?  %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
-
-We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
-We explore two empirical examples:
-deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\citep{Dakos2008},
-and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\citep{Drake2010}.  
-For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
-and the same system in a constant environment. 
-The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
-against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
-While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
-we need a way to identify such cases where we lack the data to trust the pattern.
-
-The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
-frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
-The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
-For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
-set to a 5\% false positive rate:
-25\% in the simulated set,
-5.0\% in the \emph{Daphnia} set,
-and 5.4\% in the Glaciation data set.
-Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
-would be correctly identified from the respective systems.
-The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
-as we would in classical hypothesis testing.  
-We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
-and read off the corresponding false positives.  
-For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
-while for likelihood a 55\%, 35\%, 0\% false positive.
-Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
-as illustrated in Fig. 1.  
-Comparing this threshold with the observed value classifies the signal as detected or undetected, (see Supporting Information).
-This flexibility is particularly valuable in the context of critical transitions, 
-where data are often limited and the consequences of missed detection can be far more severe than false alarms.
-
-%%%%%%% Converting from power to detection %%%%%%%%%%%%
-
-\section{Discussion}
-A pressing question in the application of early warning signals is quantifying
-when the data are sufficient for detection to be reliable~\citep{Scheffer2009, Scheffer2010, Inman2011}. 
-Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
-for each of the indicators (Fig 4). 
-Using our generic models estimated from the original data as in Fig. 3, 
-we generate the corresponding distributions by simulating under the models as before, 
-but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
-would be hampered or improved with more or less data.  
-While all methods improve as the number of data points increase, 
-the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
-as expected~\citep{Carpenter2011};⁠
-skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4). 
-%\citet{Dakos2011a} demonstrate that changes to environmental coupling decrease.  also use replicate simulations to assess expected response, but not the probability of accuracy on single replicates.
-%% Threshold
-%% Imagine a missed detection of no more than 10% is acceptable.
-
-By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
-our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
-Other literature has already begun to address the additional challenges posed when the underlying 
-dynamics do not correspond to these models~\citep{Hastings2010}.
-Our results illustrate that even in this best-case scenario, 
-reliable identification of warning signals from summary statistics can be difficult.  
-The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
-or even the distribution of that statistic under the null model of a stable system, 
-but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
-By estimating the ROC curve for a given set of data, 
-we can better avoid applying warning signals in cases of inadequate power.
-By taking advantage of the assumptions being made to write down a specific likelihood function,
-we can develop indicators that get the most from this data.
-
-%%% Model adequacy
-In any application of early warning signals, it is essential to address the question of model adequacy.  
-This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
-as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
-and systems that violate these assumptions can exhibit sudden transitions without any warning,
-such as chaotic systems or those with non-smooth potential functions~\citep{Hastings2010}.
-Sudden transitions have also been observed in ecological models in which 
-the variance deceases rather then increases as the system nears a bifurcation point~\citep{Schreiber2003, Schreiber2008, Dakos2011a}.  
-Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
-As the bifurcation results from the principle eigenvalue passing through zero, 
-the warning signal is expected in linear-order dynamics;
-estimation of the nonlinear model is less powerful and less accurate.  
-The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
-(introduced by the Poisson process demographic events), 
-demonstrates the accuracy under violation of these assumptions.
-
-%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
-%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
-%but would be amenable to the approach used here.
-
-The approach presented here offer a best-case scenario for these warning signals,
-illustrating that substantial uncertainty can remain despite the appearance of a positive increase.
-Our results highlight the need to quantify uncertainty in the application of early warning signals,
-and suggest that a hypothesis testing approach may be too restrictive a framework for that purpose.
-Using the ROC curve, we obtain a more general picture of the trade-off between false alarms and failed detections
-than can be provided by a significance test at a fixed false-alarm rate.  
-The curve can guide the choice of threshold used to indicate a warning signal on a case-by-case basis.
-
-The conclusion is not simply that likelihood approaches are more reliable, 
-but rather more broadly that warning signals should consider
-the inherent trade-off between sensitivity and accuracy,
-and must quantify how this trade-off depends on both the indicators used and the data available.  
-The approach developed here estimates the risk of both failed detection and false alarms;
-concepts which are critical to prediction-based management.  
-Using the methods we have outlined when designing early warning strategies for natural systems
-can ensure that data collection has adequate power to offer a reasonable chance of detection.
-
-\section{Acknowledgments}
-S.
-Schreiber, M.
-Holyoak, M.
-Baskett, and A.
-Perkins provided comments on the manuscript.
-This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
-Data and code are available at: https://github.com/cboettig/warningsignals.
-
- % Computational Sciences Graduate Fellowship from the Department of Energy under grant number DE-FG02-97ER25308. 
  \section{References}%bibliography
  \bibliography{boettiger}
-
-\begin{figure}[hb]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig1}
-     \label{fig1}
-     \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
-Bottom row: Points along the ROC curve are calculated for each possible threshold indicated in the top row.
-The false positive rate is the integral of the distribution of the test statistic under the stable system right of the threshold (blue shaded area, corresponding to blue vertical line).
-The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).
-Successive columns show the threshold increasing, tracing out the ROC curve.}
-  \end{center}
- \end{figure}
-
- \begin{figure}[h]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig2}
-     \label{fig2}
-     \caption{Early warning signals in simulated and empirical data sets.
-The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation (Deteriorating).
-Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a glaciation period (Glaciation).
-Increases in summary statistics, computed over a moving window, have often been used to indicate if a system is moving towards a critical transition.
-The increase is measured by the correlation coefficient $\tau$.
-Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
-  \end{center}
- \end{figure}
-
- \begin{figure}[h]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig3.pdf}
-     \label{fig3}
-     \caption{ROC curves for different early warning indicators on simulated and empirical data.
-The area under the curve, inset, indicates how the method compares to a random guess (0.5).
-The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data.
-The simulated data sets have 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121.
-See corresponding distributions, Fig.
-S1.}
-  \end{center}
- \end{figure}
-
- \begin{figure}[h!]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig4.pdf}
-     \label{fig4}
-     \caption{ROC curves (from Fig.
-3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.
-Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.
-A warning signal is not detected in stable system simulation.
-Corresponding distributions shown in Fig S2-5.}
-  \end{center}
- \end{figure}
-
  \end{document}

### 2012-01-21 boettiger.tex
writing

@@
 
-\subsubsection*{Arbitrary windows}\\
+\subsubsection*{Arbitrary windows}
 While superficially simple to use, the calculation of these statistics is subject to 
@@
 
-\subsubsection*{No quantitative measures}\\
+\subsubsection*{No quantitative measures}
 The warning signal, ``an increase in statistic $x$,'' is not quantitative,
@@
 
-\subsubsection*{Problematic null models}\\
+\subsubsection*{Problematic null models}
 Specifying an appropriate null model is also difficult.  
@@
 
-\subsubsection*{Hidden assumptions}\\
+\subsubsection*{Hidden assumptions}
 Assumes that the system sits effectively at equilibrium around a stable state  
@@
 
-Differing system dynamics
+%Differing system dynamics
 The underlying assumption that the system contains a saddle-node bifurcation
@@
 \cb{ do we need the energy diagram figure to emphasize this is our assumption?}
-Even restricting ourselves to this classic alternate stable state model,
-we must still consider at least three possible scenarios that could explain a sudden transition.
-
 %Differing causes of the transition.
@@
 (An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
-The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.  
+The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}  
 As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.  
@@
 
-%%%% TOPIC FOR ANOTHER NOTE: Early warning signals and the Prosecutor's Fallacy %%%%%%%
-%When applying detection approaches to data that have already experienced a critical transition,
-%we have conditioned on the fact the transition has occurred.  This is a very different problem than 
-%forecasting the probabilities of systems that have not collapsed!
-
-and attempting to ascertain the cause of the transition
-These are alternative null hypotheses 
- %(Dakos 2008, Lenton,  see Lenton 2009 transition, see 
-\citep{Ditlevsen2010} % telling these two cases apart
-
-5) Summary-statistic approaches inherently lack power. 
+\subsubsection{Summary-statistic approaches have less statistical power.}
 Methods for the detection of early warning signals are continually challenged by 
 inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
-After all, the challenge of extrapolating into the future across a dramatic shift is immense,
-there is essentially no historical president for successfully \& reliably predicting regime shifts. % \cite Tetlock?
+The challenge of extrapolating into the future across a dramatic shift is immense;
+there is essentially no historical president for successfully \& reliably predicting regime shifts
+outside of controlled settings. % \cite Tetlock?
 The signals one attempts to detect with these approaches are extremely weak.
-Worse still, beyond this intuitive notion, there is little quantification of just how powerful our methods are,
-even in artificial scenarios where uncertainy conforms to our parametric assumptions.  
-% Seekell Am Nat, Contamin Eco Apps
-We characterize the power
+Despite the widespread recognition of the this need for large data sets, 
+there has been very few studies quantitative studies of power looking at how much data is actually required~\citep{Contamin2009}.
+\cb{this is the only example that claims to measure statistical power.
+I find it incredible that Ellison, contrary to his own ecological statistics textbook, decides to just invent an arbitrary statistic (log ratio of the maximum value in the statistic seen before and after onset of the perturbation) and call this a measure of ``statistical power.''}
+The Neyman-Pearson Lemma guarantees us that the most powerful test between hypotheses compares the 
+likelihood that the data was produced under each. This requires a model-based approach, which,
+as we will see, can have substantially more power to detect early warning signals than summary-statistic based approaches.
 
 \subsection{Explicit use of the models}
-Several recent approaches to early warning signals have considered fitting models, though have not considered the model-choice problem.  
-Detection of warning signals as model choice (``non-parametric'' \citet{Carpenter2011e},  unpublished Dakos, Kuehn arxiv, Dakos 2011?, Gross 2011)
+Model-based approaches are beginning to play a larger role in early warning signals, 
+though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.  
+While choosing appropriate models without system-specific knowledge is challenging, 
+much can be accomplished by framing the implicit assumptions above into more precise equations.
+\citet{Lade2011} introduces the idea of generalized models for early warning signals, and 
+\citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.  
+While presented as a ``non-parametric'' approach,~\citet{Carpenter2011e} is essentially similar, where the dynamics are approximated 
+by an explicit stochastic differential equation (SDE) whose coefficients are estimated from statistical moments of the data.
+
+\subsection{Model Choice}
+Representing
+
+We introduce a similar set of SDE models to these, 
+\cb{should this next part come later?}
+and go on to illustrate how these models can be fit to data by maximum likelihood and compared 
+by a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
+characterizes the rate of missed detections and false alarms expected in the estimate 
+(or equivalently, estimates both the power and statistical significance of the test).
+
+\subsection{Models}
+The normal form for the saddle-node bifurcation is
+\begin{equation}
+\frac{\ud x}{\ud t} = r_t- x^2.
+\label{saddle-node}
+\end{equation}
+where $x$ is the state variable and $r_t$ our bifurcation parameter, which may be slowly varying in time.  
+Transforming this canonical form to allow for an arbitrary mean $\theta$,
+the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi(r_t)$.
+We expand around the fixed point and express in terms of the stochastic dynamics:
+
+\begin{equation}
+\ud X = \sqrt{ r_t } (\phi(r_t) - X_t)\ud t + \sigma\sqrt{\phi(r_t) } \ud B_t. \label{LSN}
+\end{equation}
+
+It is also possible to derive this same expression from the master equation of various stochastic birth/death (or gains \& losses) models,
+that contain saddle-node bifurcations, as we show in the appendix.  
+Alternate dynamical forms are possible but will not differ substantially.
+
+As we discuss above, the early warning signal paradigm must also include some assumptions on how the bifurcation parameter, $r_t$, is changing.
+We assume a gradual, monotonic change which we approximate to first order:
+\begin{equation}
+r_t = r_0 - m t.
+\label{R_t}
+\end{equation}
+Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
+When the system is stable, $r_t$ is constant and the linearized saddle-node model~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
+\begin{equation}
+\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
+\end{equation}
+This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.
+
+\subsection{Likelihood calculations}\label{likelihood}
+The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
+\begin{equation}
+\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
+\end{equation}
+For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
+\begin{align}
+ \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
+\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
+  \label{general_moments}
+\end{align}
+
+For the OU process, we can solve this in closed form, 
+\begin{align}
+  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
+V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
+\label{OUsoln}
+\end{align}
+
+For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
+For the linearized saddle-node bifurcation:
+\begin{align}
+\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
+\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
+\label{LSNsoln}
+\end{align}
+
+These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.
 
@@
 While the existing literature frequently presents only a 
-% e.g. Drake \& Griffen use any two-sigma deviation as indicative, yet any sufficiently sampled null process will produce a positive warning,  
-% so this is not a reasonable statistical characterization
 visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006},
@@
 All data and code for simulations and data processing are found in the R package.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Supp. materials or main text?  %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Models}
-The detection of early warning is based on a \emph{local} theory,
-predicting changes in the dynamics near the current attractor.
-Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model,
-but we can approximate the model near the bifurcation.  
-Many possible complex models will correspond to this approximation.
-
-%\textbf{Saddle-node bifurcation.}
-One of the most common critical transitions modeled in ecological systems with alternative stable states is a saddle-node bifurcation~\citep[\{emph{e.g.}[]{Scheffer2001, Scheffer2009, Guttal2008a, VanNes2007, Biggs2009}.
-The individual-based simulation follows the master equation, 
-\begin{align}
-  P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
-  b_n &= \frac{e K x^2}{X^2 + h_t^2} \\ 
-  d_n &= e X_t - a_t   
-\end{align}
-This model can experience a saddle-node bifurcation through increased mortality by increasing either $e$ or $a$. 
-The SDE approximation holds for reasonable population sizes:
-\begin{equation}
-dX_t = \left( \frac{e K x^2}{X^2 + h_t^2} - e X_t - a_t\right) dt + \sigma \sqrt{ \frac{e K x^2}{X^2 + h_t^2} + e X_t + a_t} dB_t \label{ass}
-\end{equation}
-This parameterization is only an example, of course many others are possible~\citep{Scheffer2009, Scheffer2001, Strogatz2001a, Guckenheimer1983}. 
-Under appropriate transformation, all can be written in the same normal form for he saddle-node bifurcation,
-\begin{equation}
-\frac{\ud x}{\ud t} = r_t- x^2.
-\label{saddle-node}
-\end{equation}
-While in principle this model could be estimated directly from the data,
-if all observations have been prior to the crash (\emph{i.e.} near the desirable equilibrium)
-we have no data which is informative about the nonlinearity and such estimates will be unreliable.  
-Fortunately the theory of early warning signals is a \emph{local} one, 
-and we can linearize around the equilibrium to obtain a model we can estimate both more easily and more accurately.
-
-Transforming the canonical form to allow for an arbitrary mean $\theta$,
-the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi$,
-which gives the model we refer to as the Linearized Saddle Node, LSN: 
-\begin{equation}
-\ud X = 2\sqrt{ r_t } (\phi - X_t)\ud t + \sigma\sqrt{\phi } \ud B_t. \label{LSN}
-\end{equation}
-
-\citet{Drake2010} induce a transcritical bifurcation in laboratory populations of \emph{Daphnia magna} to test early warning signals.  
-This bifurcation also arises in many common models.  For example, the stochastic version of~\citet{Levins1969}'s model (a logistic model),
-\begin{equation}
-\ud X_t = \left( c_t X_t (1-X_t/K) - e_t X_t \right) \ud t + \sigma \sqrt{\frac{e_t}{c_t}} \ud B_t \label{levins},
-\end{equation}
-where $X_t$ is the number of occupied patches of some total number $K$, $c_t$ is a (time-dependent)
-colonization rate and $e_t$ an extinction rate (both scaled by the number of patches $K$).  
-The model contains a transcritical bifurcation when $c_t < e_t$.  The normal form of the bifurcation is
-\begin{equation}
-\frac{\ud x}{\ud t} = r_t x - x^2 
-\label{transcritical}
-\end{equation}
-We can express the linearized stochastic dynamics then by the time dependent mean-reverting process. 
-We refer to this model as the linearized transcritical bifurcation, LTC.  
-\begin{equation}
-\ud X_t = r_t (r_t - X_t/K) \ud t + \sigma \sqrt{1+r_t} \ud B_t \label{LTC}
-\end{equation}
-
-%The pattern of the saddle-node differs from that of the transcritical in the square-root dependence on the bifurcation parameter.
-Note that the transcritical bifurcation involves a gradual decline towards zero,
-whereas the saddle-node bifurcation involves a sudden transition from a non-trivial equilibrium point.
-Thus saddle-node bifurcations are the more typical focus of early warning efforts, even though both exhibit the phenomenon of critical slowing down.
-
-Likelihood calculations also require specifying the rate of change of $r_t$, 
-which we take to first order as
-\begin{equation}
-r_t = r_0 - m t.
-\label{R_t}
-\end{equation}
-Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
-When the system is stable, $r_t$ is constant and both models~\eqref{LSN} and~\eqref{LTC} reduce to a simple Ornstein-Uhlenbeck process,
-\begin{equation}
-\ud X_t = r (\theta - X_t) \ud t + \sigma \ud B_t \label{OU}
-\end{equation}
-This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.
-
-\subsection{Likelihood calculations}\label{likelihood}
-The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
-\begin{equation}
-\log P(X | M)=  \sum_i \log P(x_i | x_{i-1}, t_i)
-\end{equation}
-For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
-\begin{align}
- \frac{\ud }{\ud t} E(x| M)&=  f(x) \\
-\frac{\ud}{\ud t} V(x| M) &=  -\partial_x f(x) V(x|M) + g(x)^2 
-  \label{general_moments}
-\end{align}
-
-For the OU process, we can solve this in closed form, 
-\begin{align}
-  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
-V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
-\label{OUsoln}
-\end{align}
-
-For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
-For the linearized transcritical:
-
-\begin{align}
-\frac{\ud }{\ud t} E(x_i| M = \text{LTC})&=  r(t)(r(t) - x_i) \\
-\frac{\ud}{\ud t} V(x_i| M = \text{LTC}) &=  -2 r(t) V(x_i|M) + (1+r(t))\sigma^2 
-\label{LTCsoln}
-\end{align}
-
-For the linearized saddle-node bifurcation:
-\begin{align}
-\frac{\ud }{\ud t} E(x_i| M = \text{LSN})&=  2\sqrt{r(t)}(\sqrt{r(t)}+\theta - x_i) \\
-\frac{\ud}{\ud t} V(x_i| M = \text{LSN}) &=  -2 \sqrt{r(t)} V(x_i) + \sigma^2 ( \sqrt{r(t)}+\theta )
-\label{LSNsoln}
-\end{align}
-
-These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.  
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Supp. materials or main text?  %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2012-01-23 appendix.tex
preprint mode, copy sent to John Drake
rewrites..
revisions continue

@@
 \renewcommand*\theequation{S\arabic{equation}}
-\section{Comparing Models}\label{Cox}
-Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
-The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
-The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
-which essentially asserts that comparing likelihoods is the most powerful test
-of a choice between two hypotheses~\citep{Neyman1933}, which motivates
-tests from the simple likelihood ratio test up through modern model adequacy methods.
-
-The hypotheses considered here are more challenging then the original lemma, as they are composite in nature:
-they specify two model forms (stable and changing stability)
-but with model parameters that must be first estimated from the data.
-Comparing models whose parameters have been estimated by maximum likelihood is first treated by~\citet{Cox1961, Cox1962},
-and has since been developed in this simulation estimation of the null distribution~\citep{McLachlan1987}, by parametric bootstrap estimate~\citep{Efron1987}.  
-Cox's $\delta$ statistic is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.
-
-Let $L_0$ be the likelihood function for model 0, 
-let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
-be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
-and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
-The statistic we will use is $\delta$, 
-defined to be twice the difference in log likelihood of observing the data under the two MLE models,
-$\delta = -2 (\log L_0 - \log L_1 )$.  
-This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
-We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
-Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.
-
-\pagebreak
-
-\section{On a hypothesis-testing framework}\label{Dakos}
-It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
-As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.
-In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.
-By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.
-
-The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
-We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.
-For this reason we generate the null distribution directly from the model as outlined above.
-The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
-The plots in Fig.
-3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.
-Our approach replicate these findings for the $\tau$-statistic.
-We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.
-
-The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
-As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.
-Fig.
-S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.
-The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.
-The reality is worse, we find it overlaps almost completely.
-We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.
-
-The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
 
@@
 \section*{ }%bibliography
- \bibliography{/home/cboettig/Documents/bibliographies/library.bib}
+ \bibliography{boettiger.bib}
 

### 2012-01-23 boettiger.tex
revisions continue

@@
 
- % \address[info]{ \\ 
-  %              }
-
   \begin{abstract}
@@
  \end{frontmatter}
-%\emph{\noindent \textbf{Running title}: Limits to Detection of Early Warning \\ 
-%  \textbf{Article Type}: Letter.  Abstract: 132 words, Text: 3,472 words, 34 references, 4 Figures. \\
-%  \textbf{Author Contributions}: C.B. wrote the software and performed the analysis with advice from A.H.  C.B. and A.H. designed the study and wrote the paper. 
-% }
 
@@
 
-%We focus on the common scenario of a gradual environmental change yielding a sudden shift or bifurcation in a system 
-%(\emph{i.e.} where the dominant eigenvalue passes through zero,
-%as discussed in~\citet{Scheffer2001, Scheffer2009}). 
-%Making this assumption provides a ``best-case'' test of detection approaches;
-%we will demonstrate the resulting uncertainty can be very high.
-%We describe a novel statistical approach for handling uncertainty that is appropriate for the goal of informing management.
+\cb{Need more introduction defining critical transitions, early warning signals etc?}
 
-\subsection{Summary statistics and the implicit model}
+\section{The summary statistics approach}
 Foundational work on early warning signals has operated under the often-implicit assumption %often unstated?
@@
 
-\subsection{Explicit use of the models}
+\section{A model based approach}
 Model-based approaches are beginning to play a larger role in early warning signals, 
@@
 
-\subsection{Model Choice}
-Representing
+\cb{We probably need to give some sense where this section is headed, so here's my minimal summary}
 
-We introduce a similar set of SDE models to these, 
-\cb{should this next part come later?}
-and go on to illustrate how these models can be fit to data by maximum likelihood and compared 
-by a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
-characterizes the rate of missed detections and false alarms expected in the estimate 
-(or equivalently, estimates both the power and statistical significance of the test).
+We introduce a set of SDE models that is closely related to these examples,
+and go on to illustrate how these models can be fit to data by maximum likelihood
+and compared by a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
+characterizes the rate of missed detections and false alarms expected in the estimate. 
+\cb{or equivalently, could say: estimates both the power and statistical significance of the test.}  
+\cb{Is it clear that the maximum likelihood fitting and the simulation/bootstrapping are novel here, as is the power? }
+
+\subsection{Early warning signals as model choice}
+It may be useful to think of the detection of early warning signals as a problem of model choice, rather than one of pattern recognition.  
+The model-choice based approach attempts to frame each of the possible scenarios as structurally different equations,
+making the assumptions explicit.  
+In any model choice problem, it is important to identify the goal of the exercise -- such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
+In this case generality is more important than realism or predictive capability: 
+we will write down a general model that is capable of approximating the wide class of models that experience a critical transition through a saddle node bifurcation,
+and another general model that is capable of representing these same systems when they are not approaching a bifurcation.  
+These may be thought of as the hypothesis and null hypothesis,
+though they are in fact compound hypotheses: in each case, we must first estimate the model parameters from the data before we can compare the models.  
+In this approach it is not assumed that ``reality'' is included in the models being tested per se, 
+but that one of the models is a better approximation of the true dynamics than the other.  
+System whose dynamics violate the assumptions common to both models, 
+such as in the examples of ~\citet{Hastings2009} where systems exhibit sudden transitions without warning, 
+fall outside the set of cases where this approach would be valid; 
+though the inability of either model to match the system dynamics could be an indication of such a violation.
 
 \subsection{Models}
-The normal form for the saddle-node bifurcation is
+The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
 \begin{equation}
@@
 \end{equation}
-For~\eqref{LSN},~\eqref{LTC} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
+For~\eqref{LSN} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
 \begin{align}
@@
 
-For the OU process, we can solve this in closed form, 
+For the OU process, we can solve this in closed form over an interval of time $t_i$ between subsequent observations: 
 \begin{align}
-  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} \theta \left(1 - e^{-rt_i} \right) \\
+  E(x_i| M = \text{OU}) &= X_{i-1} e^{-r t_i} + \theta \left(1 - e^{-rt_i} \right) \\
 V(x_i| M = \text{OU}) &= \frac{\sigma^2}{2 r} \left(1 - e^{-2 r t_i} \right)
@@
 
-\subsection{Why information criteria will not serve.}
+\subsection{Comparing Models}\label{Cox}
+Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
+The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
+The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
+which essentially asserts that comparing likelihoods is the most powerful test
+of a choice between two hypotheses~\citep{Neyman1933}, which motivates
+tests from the simple likelihood ratio test up through modern model adequacy methods.
+
+The hypotheses considered here are more challenging then the original lemma, as they are composite in nature:
+they specify two model forms (stable and changing stability)
+but with model parameters that must be first estimated from the data.
+Comparing models whose parameters have been estimated by maximum likelihood is first treated by~\citet{Cox1961, Cox1962},
+and has since been developed in this simulation estimation of the null distribution~\citep{McLachlan1987}, by parametric bootstrap estimate~\citep{Efron1987}.  
+Cox's $\delta$ statistic is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.
+
+Let $L_0$ be the likelihood function for model 0, 
+let $\theta_0 = \arg \max \theta_0 \in \Omega_0$, ($L_0 (\theta_0 |X)$) 
+be the maximum likelihood estimator for $\theta_0$ given $X$, and let $L_0 = L_0 (\theta_0 |X)$; 
+and define $L_1$, $\theta_1$, $L_1$ similarly for model 1. 
+The statistic we will use is $\delta$, 
+defined to be twice the difference in log likelihood of observing the data under the two MLE models,
+$\delta = -2 (\log L_0 - \log L_1 )$.  
+This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
+We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
+Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.
+
+\section{On a hypothesis-testing framework}\label{Dakos}
+\cb{Here's a longer section on hypothesis testing that was originally in the appendix.
+One of the Ecology Letters reviewers seemed very upset that this information was in the appendix and not in the main text, so here it is.
+This section is quite specific in references to ~\citet{Dakos2008}, and also quite long.
+My alternative proposal is to omit this section, and include the section I have on ''Beyond Hypothesis testing'' that follows after this}
+It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
+As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.
+In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.
+By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.
+
+The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
+We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.
+For this reason we generate the null distribution directly from the model as outlined above.
+The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
+The plots in Fig.
+3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.
+Our approach replicate these findings for the $\tau$-statistic.
+We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.
+
+The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
+As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.
+Fig.
+S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.
+The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.
+The reality is worse, we find it overlaps almost completely.
+We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.
+
+The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
+
+\subsection{Information criteria will not serve.}
+One will commonly observe models representing alternative processes being compared in this way through 
+the use of various information criteria such as the Akaike information criterion.  
+While tempting to apply in this situation, such approaches are not suited to this problem for several reasons.  
+The first is that information criteria are not concerned with the model choice objective we have in mind,
+as they are typically applied to find an adequate model description without too many parameters that the system may be overfit.  
+More pointedly, information criteria have no inherent notion of uncertainty.
+Information criteria tests alone will not tell us our chances of a false alarm, of missing a real signal, 
+or how much data we need to be confident in our ability to detect transitions.
 
 \subsection{Beyond hypothesis testing}
-Hypothesis testing is built upon the objective of
+\cb{If we remove the ``On a hypothesis-testing framework'' narrative above, I would then insert this.  I'm not sure if the wording of this next section is too casual.}
+%Hypothesis testing only runs into trouble when we make the mistake of substituting uncertainty for certainty, that we imagine the phrase ``statistically significant'' is meaningful.  
+The language of hypothesis testing is built around a strong bias that false positives are worse than false negatives, 
+and consequently an a emphasis on $p$-values rather than power.  
+In the context of early warning signals this is perilous -- 
+it suggests that we would rather fail to predict a catastrophe than to sound a false alarm, 
+while the question of when we have enough data to adequately detect an approaching transition
+should be at least as important as the chance of a false alarm.
 
-\citet{Seekell2011}
-
-\subsection{Visualizing the informativeness through ROC curves}
-
-%% EDIT -- 
-Though previous research has identified patterns such as increasing variance or autocorrelation
-in time-series approaching a critical transition,
-there has been no method for forecasting the \emph{probability} of a collapse from a natural system.
-Real-world applications of early warning signals will not be able to point to such differences between controlled and
-manipulated systems that have been utilized in experimental studies of early warning~\citep{Drake2010, Carpenter2011}⁠
-While the number of potential patterns grows steadily~\citep{Carpenter2006, Dakos2008, Guttal2008, Guttal2008a, Dakos2011}, % add spectral methods
-there are few examples that statistically quantify the patterns~\citep{Dakos2008, Dakos2011},⁠
-and fewer still that statistically estimate the false alarm rate~\citep{Dakos2008} or 
-the probability of missed detection⁠. 
-We need to quantify such uncertainty and error rates to compare between possible methods~\citep{Contamin2009}
-and address the continual concern about how much data is necessary⁠ for robust detection~\citep{Scheffer2001, Dakos2008, Carpenter2011, Scheffer2010, Inman2011}.
-
-%% EDIT
-Detection of impending transitions is not just a scientific question 
-and the goal should not solely be to detect with a 5\% chance of a false positive:
-a higher false alarm rate may well be  worth a smaller chance of  missing a catastrophic shift. 
-There is an obvious trade-off between the frequency with which a given approach will experience a false alarm and
-with which it will fail to provide any advanced warning. 
-Choosing the appropriate trade off requires knowledge of the likelihood of both kinds of error.
-In this paper, we
-(a) present the first measurements of reliability and sensitivity for early-warning indicators, 
-(b) we use this method to estimate the amount of data required to achieve a given level of confidence,  
-(c) we show where existing methods have performed little better than chance due to a lack of sufficient data or the use of insensitive indicators 
-(d) we introduce a new indicator based on likelihood which substantially out-performs the existing methods.
-
+\subsection{ROC Curves}
 We illustrate the trade-off between false alarms and failed detection using 
 receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
-The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig 1.
+The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig~\ref{fig:fig1}.
 The closer these distributions are to one-another, the more severe the trade-off.  
@@
 The ROC curve demonstrates this trade-off between accuracy and sensitivity.  
-Different early-warning indicators will vary in their sensitivity to detect differences between stable systems and those approaching a critical transition, making the ROC curves a natural way to compare their performance.
+Different early-warning indicators will vary in their sensitivity to detect differences
+between stable systems and those approaching a critical transition,
+making the ROC curves a natural way to compare their performance.  
 Since the shape of the curve will also depend on the duration and frequency of the time-series observations,
@@
 The area under the ROC curve provides a summary of how well the method compares to a random guess --
-an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance, while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.
+an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance,
+while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.
 
-To generate ROC curves for common early warning indicators such as 
-increases in autocorrelation, variance, and skew, 
-we must quantify the increase and generate the expected distributions of the focal statistic.  
-While the existing literature frequently presents only a 
-visual or heuristic measure of increase (\emph{i.e.}, do not provide $p$-values or a statistic for which they could be generated) in early-warning indicators~\citep{Scheffer2009, Drake2010, Carpenter2011, Carpenter2006},
-we rely on the few examples that quantify an increasing trend using Kendall's $\tau$ correlation statistic~\citep{Dakos2008, Dakos2011, Dakos2009}.
-Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator --
-suggestive of an approaching transition, while values near zero suggest a lack of a trend -- characteristic of stable systems.
-In principle, other quantifications of increasing variance, autocorrelation, etc.,
-may perform better than $\tau$ and could be analyzed just as well in this framework. 
-While we use this established metric of increase, we also introduce a novel warning indicator
-that is not based on the gradual increase of a summary statistic but on likelihood comparisons (see Methods).
+\section{Examples}
 
-Having identified an indicator statistic,
-we must be able to generate its distribution under both the scenario of a stable system and 
-one approaching a transition before we can compute the ROC curve.
-We fit general stochastic models representing common bifurcations
-(\emph{e.g.} saddle-node~\citep{Scheffer2009, Guttal2008a, VanNes2007, Biggs2009} and transcritical bifurcations~\citep{Drake2010})
-for both scenarios to the data by maximum likelihood, and then simulate 500 replicate data sets under each⁠. 
-By estimating the statistic (\emph{e.g.} $\tau$) on each replicate we generate distributions such as depicted in Fig. 1 and compute ROC curves (see Methods).
-Our approach provides a novel method for comparing alternatives by their likelihood ratio statistic~\citep{Cox1961}⁠ instead of $\tau$,
-which provides a more direct and more powerful test than detecting an increase in summary statistics.
-As these methods are computationally demanding, we provide a multi-processor enabled R package to assist with the analysis.
-
-\section{Materials \& Methods}\label{methods}
-
-\subsection{Statistical methodology}
-For each given time-series, parameters are estimated by maximum likelihood 
-for a stable model and a model of a system approaching a critical transition (see Models).
-The models are stochastic differential equations (SDEs)
-that can be used to simulate data under the same sampling regime as the original data, 
-or at any specified set of time points.
-Distributions are determined by a parametric bootstrap:
-500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).
-For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation)
-were computed over a sliding window equal to half the observed time (other window sizes had worse performance, see R package), 
-and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.
-The estimates of $\tau$ under the replicate simulations from the stable system
-and the estimates under the deteriorating stability system
-make up the two distributions which are integrated at a sliding threshold to create the ROC curves, as illustrated in Fig 1.
-
-The likelihood test is performed as in the parametric bootstrap; 
-but each simulated replicate is refit to both stable and deteriorating models to generate a difference in likelihood scores~\citep{Cox1961}.
-Refitting the models on each replicate, 
-rather merely computing the likelihood under the original estimates,
-makes this much more computationally intensive than the summary statistic calculation but is essential to the power of this approach~\citep{Huelsenbeck1996}.
-When replicates are created by simulation under the estimated models,
-we record data at the same time intervals observed in the original data (as in Fig. 3, main text). 
-It is straight forward to allow this interval to be different,
-representing the effect of increased or decreased sampling effort on the reliability-sensitivity trade-off in the ROC curve, Fig 4.
-The source code for implementing each of the analyses and generating the plots presented is provided as an R package, \verb|warningsignals|.
-
-\subsection{Example Data}
+\subsection*{Data}
 The simulated data sets are produced with 40 data points
-sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, Eq~\eqref{master}.
+sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation. 
 The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.
@@
 The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
-This individual replicates was chosen as an example of a single replicate 
-that showed a statistically significant correlation in variance over window of time where critical slowing down was expected.
-All data and code for simulations and data processing are found in the R package.
+This individual replicate was chosen as an example that showed 
+a pattern of increasing variance over the 16 data points where the system was being manipulated towards a crash.
+All data and code for simulations and analysis are found in the accompanying R package, \verb|warningsignals|.
+
+%%%%%%%%%%%%%% 
+\section{Comparing the performance of summary statistics and model-based approaches}  
+Due to the variety of ways in which early warning signals based on summary statistics are implemented and evaluated
+it is difficult to give a straight-forward comparison between them and the performance of this model-based approach.
+However, by adopting one of one of the quantitative measures of a warning signal pattern, such as Kendall's $\tau$~\citep{Dakos2008, Dakos2011, Dakos2009}, 
+we are able to make a side-by-side comparison of the different summary statistics and the model based approach in the
+context of false alarms and failed detections shown by the ROC curve.  
+Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator, 
+which is supposed to be inidicative of an approaching transition.  
+Values near zero suggest a lack of a trend, as expected in stable systems.
+
+Having estimated the stable and critical transition models from the data, 
+we can simulate 500 replicate data sets under each⁠. 
+We then calculated the warning signals statistic over a sliding window of size equal to one-half the length of the time series,
+and compute the correlation coefficient $\tau$ for the statistic over each replicate from both stable and impending transition models.  
+This results in a distribution of $\tau$ values coming from a model of a stable system, 
+and a corresponding distribution of $\tau$ values coming from the model with an impending transition.
 
@@
 We explore two empirical examples:
-deuterium concentrations (Fig. 2c) provide an observational dataset previously cited as an early warning signal of a critical transition in glaciation~\citep{Dakos2008},
+deuterium concentrations (Fig. 2c) provide an observational dataset
+previously cited as an early warning signal of a critical transition in glaciation~\citep{Dakos2008},
 and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\citep{Drake2010}.  
@@
 %\citet{Dakos2011a} demonstrate that changes to environmental coupling decrease.  also use replicate simulations to assess expected response, but not the probability of accuracy on single replicates.
-%% Threshold
-%% Imagine a missed detection of no more than 10% is acceptable.
 
-By assuming the underlying model corresponds to a saddle-node or transcritical bifurcation,
+By assuming the underlying model corresponds to a saddle-node bifurcation,
 our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach. 
@@
 
-%Using our model-driven approach may make it easier to identify dynamics that do not correspond with the underlying assumptions.  
-%Including measurement error and focusing on such different classes of dynamics are other complications we do not address at this time, 
-%but would be amenable to the approach used here.
-
-The approach presented here offer a best-case scenario for these warning signals,
-illustrating that substantial uncertainty can remain despite the appearance of a positive increase.
-Our results highlight the need to quantify uncertainty in the application of early warning signals,
-and suggest that a hypothesis testing approach may be too restrictive a framework for that purpose.
-Using the ROC curve, we obtain a more general picture of the trade-off between false alarms and failed detections
-than can be provided by a significance test at a fixed false-alarm rate.  
-The curve can guide the choice of threshold used to indicate a warning signal on a case-by-case basis.
-
 The conclusion is not simply that likelihood approaches are more reliable, 
@@
      \includegraphics[width=\linewidth]{Fig1}
-     \label{fig1}
+     \label{fig:fig1}
      \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
@@
      \includegraphics[width=\linewidth]{Fig2}
-     \label{fig2}
+     \label{fig:fig2}
      \caption{Early warning signals in simulated and empirical data sets.

### 2012-01-23 boettiger.tex
minor edits

@@
 by an explicit stochastic differential equation (SDE) whose coefficients are estimated from statistical moments of the data.
+\citet{Dakos2011a} starts by explicitly assuming these same SDE models but uses them only to define the summary statistics.
+% If we are to build are methods on the assumptions
 
 \cb{We probably need to give some sense where this section is headed, so here's my minimal summary}
-
 We introduce a set of SDE models that is closely related to these examples,

### 2012-01-31 boettiger.tex
edits

@@
 receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
-The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig~\ref{fig:fig1}.
+The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig~\ref{fig:roc_intro}.
 The closer these distributions are to one-another, the more severe the trade-off.  
@@
 \section{Examples}
+We illustrate this approach on simulated data as well as several natural time-series that have been previously analyzed for early warning signals.
 
 \subsection*{Data}
-The simulated data sets are produced with 40 data points
-sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation. 
-The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves.
+The simulation implements an individual, continuous-time stochastic birth-death process with rates given by the master equation,
 
-The Glaciation data is accessible from NOAA~\citep{Petit1999}.
+\begin{align}
+    P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\ 
+    b_n &= \frac{e K x^2}{X^2 + h_t^2} \\ 
+    d_n &= e X_t + a_t   
+\end{align}
+
+This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977}, 
+with stochasticity introduced from directly through the demographic process.   
+This model is forced through a bifurcation by gradually increasing the $a$ parameter,
+which increases can be thought of as an increasing toxicity of the environmnent
+(from $a_0 = 100$ increasing at constant rate of 0.09 units/unit time).  
+Other parameters are: $Xo = 730$, $e = 0.5$, $K = 1000$, $h = 200$.  
+We run this model over a time interval from 0 to 500.  
+and sample at 40 evenly spaced time points, which were used for subsequent analysis.
+We refer to this as the simulation under a degrading environment.
+We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.  
+For the convenience of the inquisitive reader, we have also provided a simple function in the 
+associated R package where the user can vary the sampling scheme and parameter values and rerun this analysis.
+
+Our first empirical data set comes from the glaciation record seen in deuterium levels in Antartic ice cores~\citep{Petit1999}, 
+as analyzed by~\citet{Dakos2008}.
 The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing 
-to be as consistent as possible with the original analysis from Dakos \emph{et al.}~\citep{Dakos2008},
-analyzing the third glaciation event, consisting of 121 sample points. 
+to be as consistent as possible with the original analysis.
+We focus on the third glaciation event, consisting of 121 sample points. 
 The match is not exact since estimates the de-trending window size manually,
@@
 
-The \emph{Daphnia} example comes from the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
+The second empirical data set comes from the population dynamics of 
+\emph{Daphnia} living in the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
 This individual replicate was chosen as an example that showed 
@@
 
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{figures/Simulation_warningsignal.pdf}
+     \label{fig:simulation}
+     \caption{A model-based calculation of warning signals for the simulated data example.}
+  \end{center}
+ \end{figure}
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{figures/GlaciationIII_warningsignal.pdf}
+     \label{fig:glaciation}
+     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).}
+  \end{center}
+ \end{figure}
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics{figures/Chemostat_warningsignal.pdf}
+     \label{fig:chemostat}
+     \caption{A model-based calculation of warning signals for the Daphnia data analyzed in ~\citet{Drake2010} (Chemostat H6).}
+  \end{center}
+ \end{figure}
+
 \begin{figure}[hb]
@@
      \includegraphics[width=\linewidth]{Fig1}
-     \label{fig:fig1}
+     \label{fig:roc_intro}
      \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
@@
      \includegraphics[width=\linewidth]{Fig2}
-     \label{fig:fig2}
+     \label{fig:summary_stats}
      \caption{Early warning signals in simulated and empirical data sets.
@@
 Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
-  \end{center}
- \end{figure}
-
- \begin{figure}[h]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig3.pdf}
-     \label{fig3}
-     \caption{ROC curves for different early warning indicators on simulated and empirical data.
-The area under the curve, inset, indicates how the method compares to a random guess (0.5).
-The likelihood method performs substantially better than other metrics, particularly on the weaker trends seen in the empirical data.
-The simulated data sets have 40 points, ``Daphnia'' has 16, the ``Glaciation'' has 121.
-See corresponding distributions, Fig.
-S1.}
   \end{center}

### 2012-02-06 boettiger.tex
ironing out text

@@
 %\documentclass[authoryear,5p]{elsarticle}
-\documentclass[authoryear,preprint,11pt]{elsarticle}
+%\documentclass[authoryear,preprint,11pt]{elsarticle}
+\documentclass[authoryear,review,11pt]{elsarticle}
 \bibliographystyle{elsarticle-harv}
@@
 \linenumbers
-\usepackage{subfigure}
-\usepackage[pdftex]{color}
-\definecolor{darkblue}{rgb}{0,0,0.5}
-\definecolor{darkgreen}{rgb}{0,0.5,0}
-%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
 \usepackage[pdftex, colorlinks]{hyperref}
+\usepackage{float}
+
 \textwidth 6.75in
@@
 \topmargin -0.5in
+
 \newcommand{\ud}{\mathrm{d}}
@@
 
-\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}
+\newcommand{\cb}[1]{{\it (#1)}}
 
@@
   \author[esp]{Alan Hastings}
-  %\author[info]{}
-  %\author[davis]{}
   \cortext[cor1]{Corresponding author.}
@@
 
+% rewrite
+
   \end{abstract}
@@
 
-\cb{Need more introduction defining critical transitions, early warning signals etc?}
+% erg, a bit mor transition needed?
+
+We discuss the limitations of current approaches that rely on summary statistics
+and introduce a model based approach to detect early warning signals.
+We illustrate how stochastic differential equation (SDE) models can be used 
+to reflect the uncertainty inherent in the detection of early warning signals.
+We caution against paradigms that are not useful for capturing this uncertainty in 
+a model-selection based approach, such as information criteria,
+and suggest reciever-operating characteristics~\citep{Green1989, Keller2009}
+as a powerful way to illustrate the sensitivity of different data sets 
+and different early warning indicators have in detecting early warning signals.
 
 \section{The summary statistics approach}
-Foundational work on early warning signals has operated under the often-implicit assumption %often unstated?
-that the system dynamics contain a saddle-node bifurcation and
-looks for patterns that would be associated with approaching this bifurcation point.
-The typical pattern has been a trend in a summary statistic such as
+Foundational work on early warning signals has operated under the often-implicit assumption 
+that the system dynamics contain a saddle-node bifurcation by
+looking for patterns that are associated with kind of transition.
+Typical patterns include an increasing trend in a summary statistic such as
 variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
 skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
-% consider additional citations
 While attractive for their simplicity, such approaches are subject to numerous challenges.
@@
 and describe how this can be done in a way that best addresses these difficulties.
+We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.
+
+%%% hmm, what about all the red-herring topics, like change point analysis?
+Add a discussion such as Contamin & Ellison ~\citep{Contamin2009} have about forecasting shifts, not post-hoc identification of shifts.
 
 \subsubsection*{Arbitrary windows}
-While superficially simple to use, the calculation of these statistics is subject to 
-several arbitrary decisions and additional assumptions,
-such as the choice of window size over which the statistic is computed,
-the choice of detrending,
-and whether and by how much the consecutive time-windows used in the calculation should overlap.  
-\citet{Lenton2012} compare the performance of different approaches to these issues,
-showing that choice of window and method used for detrending indeed influence the results,
-and each approach considered has pros and cons.  
-Without a consistent choice or a clear mechanism to determine one, % Should we say this part?
-it is difficult to evaluate and compare between such methods.  
-A model-based approach side-steps these issue entirely.  
-\cb{
- Should we mention the ergodic problem here - that averaging over windows isn't averaging across time series, 
- and that it assumes stationarity, which is precisely the thing we're looking to change?  
- Similarly could mention that the detrending approaches can remove the signal?
- We could also mention that some popular approaches such as autocorrelation require evenly-spaced points,
- and interpolating is incorrect, introducing fictitious correlations?
-}
+While superficially simple to use,
+the calculation of these statistics is subject to several arbitrary decisions and assumptions.
+
+A basic difficulty arises from the need to assume a time-series is ergodic 
+while trying to test if it is not.  
+Theoretically, the increasing trend in variance, autocorrelation, or other statistics
+is something that would be measured across an ensemble -- across replicates. %% citation? 
+As true replicates are seldom avialable in systems for which developing warning signals would be most desirable,
+these methods average across a single replicate using a moving window in time.  
+The selection of the size of this window and whether and by how much to overlap consecutive windows
+varies across the literature.  
+\citet{Lenton2012} demonstrates that these differences can influence the results, 
+and that the different choices each carry advantages and disadvantages.
+
+In addition to introducing the challenge of selecting a window size, 
+this ergodic assumption, 
+that averaging over time is equivalent to averaging over replicate realizations,  
+raises further difficulties.  
+While appropriate for a system that is stationary, 
+or changing slowly enough in the window that it may appear stationary,
+the assumption is at odds with the original hypothesis 
+that the system is approaching a saddle-node bifurcation.
+
+Further, certain statistics such as the critical slowing down measured in autocorrelation 
+require data that is evenly sampled in time.  
+Interpolating from existing data to create evenly spaced points is particularly problematic,
+as this introduces an artificial autocorrelation into the data.
 
 \subsubsection*{No quantitative measures}
-The warning signal, ``an increase in statistic $x$,'' is not quantitative,
-making it difficult to compare between signals or to
+Summary statistics typically invoke qualitative patterns such as an increase in statistic $x$, 
+rather than a quantitative measure of the early warning pattern.  
+This makes it difficult to compare between signals or to
 attribute a statistical significance to the detection.
-Some authors have suggested Kendall's correlation coefficient to quantify an increase~\citep{Dakos2008, Dakos20011} 
+Some authors have suggested Kendall's correlation coefficient, 
+$\tau$, could be used to quantify an increase~\citep{Dakos2008, Dakos20011} 
 in autocorrelation or variance.  
-Other measures of increase, such as Pearson's, have also been proposed~\citep{Drake2010}.  
-Much of the literature using summary statistics simply forgoes quantifying
-the increase or estimating significance.  
-While this may be satisfactory in experimental systems with controls and replicates to compare to,~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
-any real-world application of early warning methods will not have the luxury of controls and replicates.
-In such cases a quantitative definition of detection is essential.  
-\cb{ We will see that correlation tests in summary statistics are an ineffective way to do this.  }
+Other measures of increase, such as Pearson's correlation coefficient have also been proposed~\citep{Drake2010},
+while most of the literature simply forgoes quantifying the increase or estimating significance.  
+While adequate in experimental systems that can compare patterns between controls and replicates~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
+any real-world application of these approaches must be useful on a single time-series of observations.
+In these cases a quantitative definition of a statistically significant detection is essential. 
+Using Kendall's $\tau$ we will illustrate how such a quantitative measure 
+can be used to better capture the uncertainty in identifying the early warning pattern.
 
@@
 For instance, the standard non-parametric hypothesis test with Kendall's tau rank correlation coefficient
-assumes only that the two variables are independent, but this is
-an assumption that is violated by the very experimental design:
-since one variable is time and the other a moving time-averaged window.  
-Under such a test any adequately large data set will find a significant result, regardless of whether a warning signal exists.
+assumes only that the two variables are independent,
+but this is an assumption that is violated by the very experimental design:
+temporal correlations will exist in any finely-enough samled time series, 
+and moving windows introduce temporal correlations in the statistics.
+Under such a test any adequately large data set will find a significant result,
+regardless of whether a warning signal exists.
 A similar problem arises when the points in the time series are reordered to create a null hypothesis -- 
 this destroys the natural autocorrelation in the time series.  
-\cb{I feel citing \citet{Dakos2011} and \citet{Dakos2008} for these two examples would be impolite\dots} 
-Parametric null models have been more promising, and have brought us closer to a model-based approach where assumptions are explicit.  
-For instance, autoregressive models~\citet{Dakos2008} have been proposed, which are first parameterized from the data.  
-\cb{is this next line okay to leave in?}
-Estimating the null models directly from the data using a maximum likelihood framework, will provide the most powerful basis for this test.  
-\citet{Seekell2011} proposes a conditional heteroscedaticity as a summary statistic particularly because probability models for the null distribution are readily available.\cb{Though as a measure of changing variance, it suffers from the weaknesses mentioned above, but no need to mention that.}
+More promising parametric null models have been proposed, 
+such as autoregressive models in~\citet{Dakos2008}, bringing us  
+closer to a model-based approach with explicit assumptions.  
+Others have looked for alternative summary statistics where 
+reasonable null models are more readily available, 
+such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.
 
-\cb{This is a particular concern given the tendency for warning signals to be found only in the largest data sets, 
-which has been taken as evidence that large data sets are required for successful detection,
-not that large data sets have higher false-positive rates under certain metrics.}
-\cb{A similar problem exists with the proposal that large deviations from the running variance,
-i.e. in excess of 2 standard deviations, be taken as evidence of an increase.
-This likewise guarantees the false positive rate goes to unity as the number of observations increases.}
+%To characterize the power of an potential early warning signal we need more than the 
+%distribution of some leading indicator under the null hypothesis of a stable system.  
+%Very little work has addressed the distribution of any proposed indicator under the 
+%actual test hypothesis, that is, direclty measured statistical power. 
+%(~\citep{Contamin2009} use a measure of effect size, the log ratio of maximum value of differences between adjacent indicator values in the time-series, as a measure of power
 
+%This is a particular concern given the tendency for warning signals to be found more often in large data sets, 
+%which could be evidence that large data sets are required for successful detection~\citep{Inman2011, Scheffer2010,Held2004, Dakos2008, Scheffer2009, Guttal2008, Carpenter2011 Bestelmeyer2011},
+%rathern than that large data sets have higher false-positive rates under certain metrics.
+% Ellison Eco App paper -- calls it power but it is not
+%
 \subsubsection*{Hidden assumptions}
-Assumes that the system sits effectively at equilibrium around a stable state  
-and tests whether the system may be experiencing the slow change of a parameter that is weakening the stability of that state,
-moving the system towards a saddle-node bifurcation.  
-As such processes may be common in natural systems this is an important scenario to consider,
-but without the clarity of assumptions made in a modeling approach there is a risk that
-early warning signs will be used in situations where they do not apply.
-
 %Differing system dynamics
@@
 
-\cb{ do we need the energy diagram figure to emphasize this is our assumption?}
 %Differing causes of the transition.
@@
 
-\subsubsection{Summary-statistic approaches have less statistical power.}
+\subsubsection*{Summary-statistic approaches have less statistical power.}
 Methods for the detection of early warning signals are continually challenged by 
@@
 
-\subsection{Early warning signals as model choice}
+\subsection*{Early warning signals as model choice}
 It may be useful to think of the detection of early warning signals as a problem of model choice, rather than one of pattern recognition.  
@@
 
-\subsection{Models}
+\subsection*{Models}
 The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
@@
 
-\subsection{Likelihood calculations}\label{likelihood}
+\subsection*{Likelihood calculations}
 The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
@@
 
-\subsection{Comparing Models}\label{Cox}
+\subsection*{Comparing Models}
 Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
@@
 defined to be twice the difference in log likelihood of observing the data under the two MLE models,
-$\delta = -2 (\log L_0 - \log L_1 )$.  
+\begin{equation}\label{delta}
+\delta = -2 (\log L_0 - \log L_1 ).
+\end{equation}
 This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
@@
 
-\section{On a hypothesis-testing framework}\label{Dakos}
-\cb{Here's a longer section on hypothesis testing that was originally in the appendix.
+%\section{On a hypothesis-testing framework}
+%\cb{Here's a longer section on hypothesis testing that was originally in the appendix.
 One of the Ecology Letters reviewers seemed very upset that this information was in the appendix and not in the main text, so here it is.
@@
 My alternative proposal is to omit this section, and include the section I have on ''Beyond Hypothesis testing'' that follows after this}
-It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
+%It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
 As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.
@@
 By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.
-
-The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
+%
+%The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
 We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.
@@
 We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.
-
-The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
+%
+%The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
 As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.
@@
 We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.
+%
+%The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
+%
+%
+%
 
-The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
-
-\subsection{Information criteria will not serve.}
+\subsection*{Information criteria will not serve.}
 One will commonly observe models representing alternative processes being compared in this way through 
@@
 
-\subsection{Beyond hypothesis testing}
+\subsection*{Beyond hypothesis testing}
 \cb{If we remove the ``On a hypothesis-testing framework'' narrative above, I would then insert this.  I'm not sure if the wording of this next section is too casual.}
@@
 
-\subsection{ROC Curves}
+\subsection*{ROC Curves}
 We illustrate the trade-off between false alarms and failed detection using 
@@
 while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.
+
+\begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{Fig1}
+     \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
+Bottom row: Points along the ROC curve are calculated for each possible threshold indicated in the top row.
+The false positive rate is the integral of the distribution of the test statistic under the stable system right of the threshold (blue shaded area, corresponding to blue vertical line).
+The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).
+Successive columns show the threshold increasing, tracing out the ROC curve.}
+     \label{fig:roc_intro}
+  \end{center}
+ \end{figure}
 
@@
 
-Our first empirical data set comes from the glaciation record seen in deuterium levels in Antartic ice cores~\citep{Petit1999}, 
+The first empirical data set comes from the population dynamics of 
+\emph{Daphnia} living in the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
+This individual replicate was chosen as an example that showed 
+a pattern of increasing variance over the 16 data points where the system was being manipulated towards a crash.
+All data and code for simulations and analysis are found in the accompanying R package, \verb|warningsignals|.
+
+Our second empirical data set comes from the glaciation record seen in deuterium levels in Antartic ice cores~\citep{Petit1999}, 
 as analyzed by~\citet{Dakos2008}.
@@
 
-The second empirical data set comes from the population dynamics of 
-\emph{Daphnia} living in the chemostat ``H6'' in the experiments of Drake \& Griffen~\citep{Drake2010}. 
-This individual replicate was chosen as an example that showed 
-a pattern of increasing variance over the 16 data points where the system was being manipulated towards a crash.
-All data and code for simulations and analysis are found in the accompanying R package, \verb|warningsignals|.
+\subsection*{Analysis}
+We perform the identical analysis proceedure described above on each of these three data sets.
+First, we estimate parameters for the null and test model to each data set by maximum likelihood.
+Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better --
+the deviances $\delta$, Eqn~\eqref{delta} -- are 5.1, 6.0, 83.9 respectively for the simulation,
+the chemostat data and the glaciation data. 
+The size of these differences reflects not only the magnitude of the difference in fit
+between the models but also the arbitrary units of the raw likelihoods,
+which are smaller for larger data-sets.  
+Consequently the glaciation score reflects as much the greater length of its time series as it does anything else.
+
+To identify if these differences are significant,
+and by what probability they could arise as a false alarm or a missed event,
+we simulate 500 replicate time series from each estimated model,
+as visualized in the middle panel of each data set,
+Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}.
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=.85\linewidth]{figures/Simulation_warningsignal.pdf}
+     \caption{A model-based calculation of warning signals for the simulated data example.}
+     \label{fig:simulation}
+  \end{center}
+ \end{figure}
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=.85\linewidth]{figures/Chemostat_warningsignal.pdf}
+     \caption{A model-based calculation of warning signals for the Daphnia data analyzed in ~\citet{Drake2010} (Chemostat H6).}
+     \label{fig:chemostat}
+  \end{center}
+ \end{figure}
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=.85\linewidth]{figures/GlaciationIII_warningsignal.pdf}
+     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).}
+     \label{fig:glaciation}
+  \end{center}
+ \end{figure}
+
+The model parameters of both models are re-estimated on both families of replicates 
+(the null and test, \emph{i.e.} $2 \times 2 \times 500$ fits). 
+The differences in the likelihood values between the model estimates produced from the first set of simulations 
+determines the null distribution for the deviance statistic $\delta$.  
+As the constant OU process model is nested within the time-heterogenous model, these values are always positive,
+but tend to be not as large as those produced when the models are fit to the second family of data. 
+These two distributions are shown in the lower panel of 
+Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}, 
+along with the deviance observed in the original data (vertical line).
+
+The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.  
+The tendancy of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other 
+indicates the probability our observed data corresponds best with that model  -- 
+either approaching a critical transition or remaining stable.  
+While it trivial to assign a statistical significance to this observation based 
+on how far into the tail of the (red) null distribution it falls, 
+for the reasons we discussed we prefer the more symmetric comparison of the probability that 
+this value was observed in either distribution.  
+We visualize the trade-off between false alarms and failed detection using the ROC curves introduced above.
+
+Because ROC curves compare areas under the probability distributions they can always be presented on a uniform 0-1 axis
+for both true positives and false positives, allowing us to overlay the curves from different data on the same plot,
+Figure~\ref{fig:rocdata}
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{figures/rocplot.pdf}
+     \caption{ }
+     \label{fig:rocdata}
+  \end{center}
+ \end{figure}
 
@@
 it is difficult to give a straight-forward comparison between them and the performance of this model-based approach.
-However, by adopting one of one of the quantitative measures of a warning signal pattern, such as Kendall's $\tau$~\citep{Dakos2008, Dakos2011, Dakos2009}, 
+However, by adopting one of one of the quantitative measures of a warning signal pattern,
+such as Kendall's $\tau$~\citep{Dakos2008, Dakos2011, Dakos2009}, 
 we are able to make a side-by-side comparison of the different summary statistics and the model based approach in the
@@
 Values of $\tau$ near unity indicate a strongly increasing trend in the warning indicator, 
-which is supposed to be inidicative of an approaching transition.  
+which is supposed to be indicative of an approaching transition.  
 Values near zero suggest a lack of a trend, as expected in stable systems.
 
-Having estimated the stable and critical transition models from the data, 
-we can simulate 500 replicate data sets under each⁠. 
-We then calculated the warning signals statistic over a sliding window of size equal to one-half the length of the time series,
-and compute the correlation coefficient $\tau$ for the statistic over each replicate from both stable and impending transition models.  
+\begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{figures/Fig2}
+     \caption{Early warning signals in simulated and empirical data sets.
+The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation (Deteriorating).
+Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a glaciation period (Glaciation).
+Increases in summary statistics, computed over a moving window, have often been used to indicate if a system is moving towards a critical transition.
+The increase is measured by the correlation coefficient $\tau$.
+Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
+     \label{fig:summary_stats}
+  \end{center}
+ \end{figure}
+
+Figure~\ref{fig:summary_stats} shows the time series for each data set in columns 
+and the early warning indicators  of variance and autocorrelation 
+computed over a sliding window for each.  
+Kendall's correlation coefficient $\tau$ is calculated for each warning indicator and displayed on the graphs, inset.  
+For comparison, the left-most column includes data simulated under a stable system, 
+which nevertheless shows a chance increasing autocorrelation with a $\tau=0.7$
+We can adapt the approach we have described above to determine how often such a strong increase would appear 
+by chance in a stable system as follows.
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{figures/summary_box.pdf}
+     \caption{}
+     \label{fig:summary_box}
+  \end{center}
+ \end{figure}
+
+By estimating the stable and critical transition models from the data,
+and simulate 500 replicate data sets under each as in the analysis above,
+we can then calculate the warning signals statistic over a sliding window 
+of size equal to one-half the length of the time series,
+and compute the correlation coefficient $\tau$ measuring the degree to which the statistic shows an increasing trend.   
 This results in a distribution of $\tau$ values coming from a model of a stable system, 
 and a corresponding distribution of $\tau$ values coming from the model with an impending transition.  
+These distributions are shown in Figure~\ref{fig:summary_box}.
 
-\section{Results}
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=\linewidth]{figures/summary_roc.pdf}
+     \caption{}
+     \label{fig:summary_roc}
+  \end{center}
+ \end{figure}
 
-We illustrate this analysis of uncertainty in early warning indicators in simulated and empirical data sets (Fig. 2). 
-We explore two empirical examples:
-deuterium concentrations (Fig. 2c) provide an observational dataset
-previously cited as an early warning signal of a critical transition in glaciation~\citep{Dakos2008},
-and \emph{Daphnia magna} concentrations (Fig. 2d) manipulated in a chemostat towards a critical transition~\citep{Drake2010}.  
-For comparison, we include an individual-based simulation of a system approaching a saddle-node bifurcation due to environmental deterioration,
-and the same system in a constant environment. 
-The challenge of accurate detection is clearly illustrated by comparing the patterns for a system approaching transition (b-d) 
-against the simulation of a stable system, in which chance fluctuations give the appearance of steadily rising autocorrelation. 
-While such a pattern will fade in a longer or more frequently sampled time-series of the same system, 
-we need a way to identify such cases where we lack the data to trust the pattern.
-
-The ROC curves for these data (Fig. 3) show that the summary-statistic based indicators 
+ The ROC curves for these data (Fig.~\ref{fig:summary_roc}) show that the summary-statistic based indicators 
 frequently lack the sensitivity to distinguish reliably between observed patterns from a stable or unstable system. 
-The large correlations observed in the empirical examples (Fig. 2) are not uncommon in stable systems. 
+The large correlations observed in the empirical examples (Fig.~\ref{fig:summary_stats}) are not uncommon in stable systems. 
 For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
@@
 as we would in classical hypothesis testing.  
-We could instead decide that we wanted no to catch no less than 90\% of the true-positives,
+We could instead decide that we wanted not to catch no less than 90\% of the true-positives,
 and read off the corresponding false positives.  
@@
 while for likelihood a 55\%, 35\%, 0\% false positive.
-Each false-positive, true-positive coordinate on the ROC curve corresponds to a threshold of the indicator statistic,
-as illustrated in Fig. 1.  
-Comparing this threshold with the observed value classifies the signal as detected or undetected, (see Supporting Information).
-This flexibility is particularly valuable in the context of critical transitions, 
-where data are often limited and the consequences of missed detection can be far more severe than false alarms.
-
-%%%%%%% Converting from power to detection %%%%%%%%%%%%
 
@@
 
- \begin{figure}
-   \begin{center}
-     \includegraphics[width=\linewidth]{figures/Simulation_warningsignal.pdf}
-     \label{fig:simulation}
-     \caption{A model-based calculation of warning signals for the simulated data example.}
-  \end{center}
- \end{figure}
-
- \begin{figure}
-   \begin{center}
-     \includegraphics[width=\linewidth]{figures/GlaciationIII_warningsignal.pdf}
-     \label{fig:glaciation}
-     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).}
-  \end{center}
- \end{figure}
-
- \begin{figure}
-   \begin{center}
-     \includegraphics{figures/Chemostat_warningsignal.pdf}
-     \label{fig:chemostat}
-     \caption{A model-based calculation of warning signals for the Daphnia data analyzed in ~\citet{Drake2010} (Chemostat H6).}
-  \end{center}
- \end{figure}
-
-\begin{figure}[hb]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig1}
-     \label{fig:roc_intro}
-     \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
-Bottom row: Points along the ROC curve are calculated for each possible threshold indicated in the top row.
-The false positive rate is the integral of the distribution of the test statistic under the stable system right of the threshold (blue shaded area, corresponding to blue vertical line).
-The true positive rate is the integral of the system approaching a transition left of the threshold (red shaded area, corresponds to the red line).
-Successive columns show the threshold increasing, tracing out the ROC curve.}
-  \end{center}
- \end{figure}
-
- \begin{figure}[h]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig2}
-     \label{fig:summary_stats}
-     \caption{Early warning signals in simulated and empirical data sets.
-The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation (Deteriorating).
-Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a glaciation period (Glaciation).
-Increases in summary statistics, computed over a moving window, have often been used to indicate if a system is moving towards a critical transition.
-The increase is measured by the correlation coefficient $\tau$.
-Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
-  \end{center}
- \end{figure}
-
- \begin{figure}[h!]
-   \begin{center}
-     \includegraphics[width=\linewidth]{Fig4.pdf}
-     \label{fig4}
-     \caption{ROC curves (from Fig.
-3) are recalculated under different sampling frequencies in the simulation step (see supplementary online text), using a total of 25, 50, or 200 points.
-Increasing the sampling improves the trade-off between sensitivity and reliability for all indicators, though the likelihood approach is most effective.
-A warning signal is not detected in stable system simulation.
-Corresponding distributions shown in Fig S2-5.}
-  \end{center}
- \end{figure}
-
  \end{document}

### 2012-02-07 boettiger.tex
draft to Alan, Noam

@@
 Some authors have suggested Kendall's correlation coefficient, 
-$\tau$, could be used to quantify an increase~\citep{Dakos2008, Dakos20011} 
+$\tau$, could be used to quantify an increase~\citep{Dakos2008, Dakos2011} 
 in autocorrelation or variance.  
@@
 
-%To characterize the power of an potential early warning signal we need more than the 
-%distribution of some leading indicator under the null hypothesis of a stable system.  
-%Very little work has addressed the distribution of any proposed indicator under the 
-%actual test hypothesis, that is, direclty measured statistical power. 
-%(~\citep{Contamin2009} use a measure of effect size, the log ratio of maximum value of differences between adjacent indicator values in the time-series, as a measure of power
-
-%This is a particular concern given the tendency for warning signals to be found more often in large data sets, 
-%which could be evidence that large data sets are required for successful detection~\citep{Inman2011, Scheffer2010,Held2004, Dakos2008, Scheffer2009, Guttal2008, Carpenter2011 Bestelmeyer2011},
-%rathern than that large data sets have higher false-positive rates under certain metrics.
-% Ellison Eco App paper -- calls it power but it is not
-%
 \subsubsection*{Hidden assumptions}
@@
 The underlying assumption that the system contains a saddle-node bifurcation
-can be easily overlooked in summary-statistic based approach.  
+can be easily overlooked in common summary-statistics based approaches.  
 For instance, variance may increase for reasons that 
@@
 
-% this will be made precise later
-
 %Differing causes of the transition.
@@
 restrict ourselves to systems that can support a saddle-node bifurcation,
-any early warning signals approach must further assume that a changing parameter 
-has brought the system closer to the bifurcation.  This excludes at least three alternative
-explanations for the transition.  
+these approaches must further assume that a changing parameter 
+has brought the system closer to the bifurcation.  
+This excludes at least three alternative explanations for the transition.  
 The first is that a large perturbation of the system state 
@@
 
+%% Not sure if this paragraph above is necessary.
+
 % Of course the literature has always stuck carefully to these assumptions, but there has sometimes been the suggestion that the summary statistics themselves were somehow more general than this, and that framing these assumptions in explicit and precise formulae, in models, would be misleading.
@@
 \subsubsection*{Summary-statistic approaches have less statistical power.}
-Methods for the detection of early warning signals are continually challenged by 
-inadequate data  %\cite\cite\cite %often worse than we think -- as we will illustrate
-The challenge of extrapolating into the future across a dramatic shift is immense;
-there is essentially no historical president for successfully \& reliably predicting regime shifts
-outside of controlled settings. % \cite Tetlock?
-The signals one attempts to detect with these approaches are extremely weak.
+Methods for the detection of early warning signals are continually challenged by inadequate data~\citep{Inman2011, Scheffer2010,Held2004, Dakos2008, Scheffer2009, Guttal2008, Carpenter2011, Bestelmeyer2011}.
 Despite the widespread recognition of the this need for large data sets, 
-there has been very few studies quantitative studies of power looking at how much data is actually required~\citep{Contamin2009}.
-\cb{this is the only example that claims to measure statistical power.
-I find it incredible that Ellison, contrary to his own ecological statistics textbook, decides to just invent an arbitrary statistic (log ratio of the maximum value in the statistic seen before and after onset of the perturbation) and call this a measure of ``statistical power.''}
-The Neyman-Pearson Lemma guarantees us that the most powerful test between hypotheses compares the 
-likelihood that the data was produced under each. This requires a model-based approach, which,
-as we will see, can have substantially more power to detect early warning signals than summary-statistic based approaches.
+there has been very few studies quantitative studies of power to determine at how much data is required~\citep{Contamin2009},
+how often a particular method would produce a false alarm or fail to detect a signal,
+and which tests will be the most powerful or sensitive.  
+The Neyman-Pearson Lemma demonstrates that the most powerful test between hypotheses 
+compares the likelihood that the data was produced under each~\citep{Neyman1933}. 
+Such likelihood calculations require a model-based approach.
 
@@
 While choosing appropriate models without system-specific knowledge is challenging, 
-much can be accomplished by framing the implicit assumptions above into more precise equations.
+much can be accomplished by framing the implicit assumptions into equations.
 \citet{Lade2011} introduces the idea of generalized models for early warning signals, and 
 \citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.  
-While presented as a ``non-parametric'' approach,~\citet{Carpenter2011e} is essentially similar, where the dynamics are approximated 
-by an explicit stochastic differential equation (SDE) whose coefficients are estimated from statistical moments of the data.
-\citet{Dakos2011a} starts by explicitly assuming these same SDE models but uses them only to define the summary statistics.
-% If we are to build are methods on the assumptions
+\citet{Carpenter2011e} and \citet{Dakos2011a} start by assuming the dynamics obey a generic
+stochastic differential equation (SDE), but use this only to derive or define the summary statistics of interest.
 
-\cb{We probably need to give some sense where this section is headed, so here's my minimal summary}
-We introduce a set of SDE models that is closely related to these examples,
-and go on to illustrate how these models can be fit to data by maximum likelihood
-and compared by a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
+In this section we outline how the detection of early warning signals may be thought of
+as a problem of model choice.
+We next show generic models can be constructed under the assumptions discussed above
+and estimated from the data in a maximum likelihood framework.
+We highlight the disadvantages of comparing these estimates by information criterea, 
+and insead introduce a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
 characterizes the rate of missed detections and false alarms expected in the estimate.
-\cb{or equivalently, could say: estimates both the power and statistical significance of the test.}  
-\cb{Is it clear that the maximum likelihood fitting and the simulation/bootstrapping are novel here, as is the power? }
 
 \subsection*{Early warning signals as model choice}
-It may be useful to think of the detection of early warning signals as a problem of model choice, rather than one of pattern recognition.  
-The model-choice based approach attempts to frame each of the possible scenarios as structurally different equations,
-making the assumptions explicit.  
-In any model choice problem, it is important to identify the goal of the exercise -- such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
+It may be useful to think of the detection of early warning signals 
+as a problem of model choice rather than one of pattern recognition.  
+The model choice approach attempts to frame each of the possible scenarios as structurally different equations,
+each with unknown parameters that must be estimated from the data.
+In any model choice problem, it is important to identify the goal of the exercise --
+such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
 In this case generality is more important than realism or predictive capability: 
-we will write down a general model that is capable of approximating the wide class of models that experience a critical transition through a saddle node bifurcation,
-and another general model that is capable of representing these same systems when they are not approaching a bifurcation.  
+we will write down a general model that is capable of approximating
+a wide class of models in which regime shifts are characterized by a saddle node bifurcation,
+and a second generic model that is capable of representing the behavior of such systems
+when they are not approaching a bifurcation.  
 These may be thought of as the hypothesis and null hypothesis,
-though they are in fact compound hypotheses: in each case, we must first estimate the model parameters from the data before we can compare the models.  
+though they are in fact compound hypotheses,
+as we must first estimate the model parameters from the data.  
 In this approach it is not assumed that ``reality'' is included in the models being tested per se, 
@@
 System whose dynamics violate the assumptions common to both models, 
-such as in the examples of ~\citet{Hastings2009} where systems exhibit sudden transitions without warning, 
+such as in the examples of ~\citet{Hastings2010} where systems exhibit sudden transitions without warning, 
 fall outside the set of cases where this approach would be valid; 
@@
 \end{equation}
-where $x$ is the state variable and $r_t$ our bifurcation parameter, which may be slowly varying in time.  
+where $x$ is the state variable and $r_t$ our bifurcation parameter, 
+which may be slowly varying in time and 
+consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.  
 Transforming this canonical form to allow for an arbitrary mean $\theta$,
@@
 
-It is also possible to derive this same expression from the master equation of various stochastic birth/death (or gains \& losses) models,
-that contain saddle-node bifurcations, as we show in the appendix.  
-Alternate dynamical forms are possible but will not differ substantially.
+%It is also possible to derive this same expression from the master equation 
+%of various stochastic birth/death (or gains \& losses) models
+%that contain saddle-node bifurcations, as we show in the appendix.  
+%Alternate dynamical forms are possible but will not differ substantially.
 
-As we discuss above, the early warning signal paradigm must also include some assumptions on how the bifurcation parameter, $r_t$, is changing.
+As we discuss above, in this paradigme we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
 We assume a gradual, monotonic change which we approximate to first order: 
@@
 Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
-When the system is stable, $r_t$ is constant and the linearized saddle-node model~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
+When the system is stable, $r_t$ is constant and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process, 
 \begin{equation}
@@
 
-For the time dependent models, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
-For the linearized saddle-node bifurcation:
+For the time dependent model, we have analytic forms only for the dynamical equations of these moments from equation~\eqref{general_moments}, which we must integrate numerically over each time interval.
+The moments of Equation~\eqref{LSN} are given by
 \begin{align}
@@
 \subsection*{Comparing Models}
-Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
-The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
+Likelihood methods form the basis of much of modern statistics
+in both Frequentist and Bayesian paradigms.  
+The ability evaluate likelihoods directly by computation has made it
+possible to treat cases that do not conform to traditional assumptions more directly.
 The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
 which essentially asserts that comparing likelihoods is the most powerful test
-of a choice between two hypotheses~\citep{Neyman1933}, which motivates
+of a choice between two hypotheses~\citep{Neyman1933}, and motivates
 tests from the simple likelihood ratio test up through modern model adequacy methods.
 
-The hypotheses considered here are more challenging then the original lemma, as they are composite in nature:
+The hypotheses considered here are more challenging then the original lemma provides for, 
+as they are composite in nature:
 they specify two model forms (stable and changing stability)
@@
 and has since been developed in this simulation estimation of the null distribution~\citep{McLachlan1987}, by parametric bootstrap estimate~\citep{Efron1987}.  
-Cox's $\delta$ statistic is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.
+Cox's $\delta$ statistic (often called the deviance between models) 
+is simply the difference between the log likelihoods of these maximum likelihood estimates, defined as follows.
 
@@
 \end{equation}
-This approach has been applied to the problem of model adequacy~\citep{Goldman1993} and model choice~\citep{Huelsenbeck1996} in other contexts.  
-We have extended the approach by generating the test distribution as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
-Our results have shown that $\delta$ tends to result in lower error rates than $\tau$, the measure of increase in a summary statistic.
+This approach has been applied to the problem of model adequacy~\citep{Goldman1993} 
+and model choice~\citep{Huelsenbeck1996} in other contexts.  
+We have extended the approach by generating the test distribution 
+as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.  
+%Our results show that $\delta$ tends to result in lower error rates than $\tau$,
+%a measure of increase in a summary statistic.
 
-%\section{On a hypothesis-testing framework}
-%\cb{Here's a longer section on hypothesis testing that was originally in the appendix.
-One of the Ecology Letters reviewers seemed very upset that this information was in the appendix and not in the main text, so here it is.
-This section is quite specific in references to ~\citet{Dakos2008}, and also quite long.
-My alternative proposal is to omit this section, and include the section I have on ''Beyond Hypothesis testing'' that follows after this}
-%It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
-As this introduces the unnecessary nuisance parameter, the statistical significance criterion, it is not presented in the text.
-In the hypothesis testing framework, a false positive is a Type I error, which is defined relative to some arbitrary statistical significance criterion, most commonly 0.05.
-By changing this criterion, one can increase or decrease the probability of the Type I error at the cost of decreasing or increasing the Type II error, also defined relative to this criterion.
-%
-%The work by~\citet{Dakos2008} is one of the few commendable examples that attempts a careful quantification of the statistical significance of the increase observed in the indicator statistic (autocorrelation in this case).
-We note that the $p$-values presented in the main text of that work are based on parametric assumptions not met by the data -- the autocorrelation is computed by a sliding window, and the points are part of a time series, both conditions that do not satisfy the null hypothesis that $x$ and $y$ are independent, which is used in approximating a normal distribution of $\tau$ values.
-For this reason we generate the null distribution directly from the model as outlined above.
-The rather different $p$ values presented in appendix of~\citet{Dakos2008}, Table S3 are constructed with more reasonable null hypotheses (in particular, H$_0$-2 and H$_0$-3 are very similar to our stable system, though H$_0$-1 has lost the correlation structure expected in any time-series).
-The plots in Fig.
-3 confirm our finding that almost any value of Kendall's $\tau$ is possible under the null hypothesis, and the $p$-values in table S3 show that the detection method is not significant in any of the time series that had fewer than 400 points.
-Our approach replicate these findings for the $\tau$-statistic.
-We were glad to find that our likelihood-based approach does not have this difficulty, easily identifying the warning signal in these shorter time series, while also reliably avoiding false alarms.
-%
-%The approach taken by~\citet{Dakos2008} does not present the corresponding analysis of power or Type II error.
-As statistical significance could be achieved with a method that selects less than one in every twenty data sets at random, it is impossible to know whether the approach is actually detecting a signal or simply a random pattern in the noise without a test of power.
-Fig.
-S3, highlights the importance of so doing -- the null distribution is almost uniform across all possible values of $\tau$.
-The distribution of $\tau$ under the test distribution will not be able to avoid at least some overlap on this bounded domain.
-The reality is worse, we find it overlaps almost completely.
-We were glad to see that in estimating the distribution of $\tau$ for very large data sets, it does tighten around zero, while the distribution expected under a true positive condenses around more positive values, indicating that there is at least some discriminative ability in the test.
-%
-%The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''
-%
-%
 %
@@
 \subsection*{Information criteria will not serve.}
-One will commonly observe models representing alternative processes being compared in this way through 
+One will commonly observe models representing alternative processes being compared through 
 the use of various information criteria such as the Akaike information criterion.  
@@
 \subsection*{Beyond hypothesis testing}
-\cb{If we remove the ``On a hypothesis-testing framework'' narrative above, I would then insert this.  I'm not sure if the wording of this next section is too casual.}
-%Hypothesis testing only runs into trouble when we make the mistake of substituting uncertainty for certainty, that we imagine the phrase ``statistically significant'' is meaningful.  
-The language of hypothesis testing is built around a strong bias that false positives are worse than false negatives, 
+It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. 
+This introduces the a nuisance parameter, known as the statistical significance criterion.  
+In the hypothesis testing framework, a false positive is a Type I error, 
+which is defined relative to this arbitrary statistical significance criterion, most commonly 0.05.  
+By changing the criterion, one can increase or decrease the probability of the Type I error 
+at the cost of decreasing or increasing false negative or Type II error, which must also be defined relative to this criterion.
+
+The language of hypothesis testing is built around a bias that false positives are worse than false negatives, 
 and consequently an a emphasis on $p$-values rather than power.  
 In the context of early warning signals this is perilous -- 
-it suggests that we would rather fail to predict a catastrophe than to sound a false alarm, 
-while the question of when we have enough data to adequately detect an approaching transition
-should be at least as important as the chance of a false alarm.
+it suggests that we would rather fail to predict a catastrophe than to sound a false alarm.
+
+To avoid this linguistic bias and the introduction of an nuisance parameter on which to define statistical significance, 
+we propose the use of Reciever Operating characteristic (ROC) curves.
 
@@
 we can use these curves to illustrate by how much a given increase in sampling effort can decrease the rate of false alarms or failed detections.  
-The area under the ROC curve provides a summary of how well the method compares to a random guess --
-an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance,
-while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.
+%The area under the ROC curve provides a summary of how well the method compares to a random guess --
+%an area of 0.5 corresponding to perfectly overlapping distributions is performing no better than chance,
+%while a statistic with area of unity always distinguishes correctly between stable systems and those approaching a transition.
 
@@
    \begin{center}
-     \includegraphics[width=\linewidth]{Fig1}
+     \includegraphics[width=\linewidth]{figures/Fig1.pdf}
      \caption{Top row: The distributions of a hypothetical warning indicator are shown under the case of a stable system (blue) and a system approaching a critical transition (red).
@@
 
-\section{Examples}
+\section{Example Results}
 We illustrate this approach on simulated data as well as several natural time-series that have been previously analyzed for early warning signals.
@@
 for both true positives and false positives, allowing us to overlay the curves from different data on the same plot,
-Figure~\ref{fig:rocdata}
+Figure~\ref{fig:rocdata}.
 
@@
    \begin{center}
-     \includegraphics[width=\linewidth]{figures/Fig2}
+     \includegraphics[width=\linewidth]{figures/Fig2.pdf}
      \caption{Early warning signals in simulated and empirical data sets.
@@
 \section{Discussion}
-A pressing question in the application of early warning signals is quantifying
-when the data are sufficient for detection to be reliable~\citep{Scheffer2009, Scheffer2010, Inman2011}. 
-Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
-for each of the indicators (Fig 4). 
-Using our generic models estimated from the original data as in Fig. 3, 
-we generate the corresponding distributions by simulating under the models as before, 
-but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
-would be hampered or improved with more or less data.  
-While all methods improve as the number of data points increase, 
-the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
-as expected~\citep{Carpenter2011};⁠
-skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4). 
-%\citet{Dakos2011a} demonstrate that changes to environmental coupling decrease.  also use replicate simulations to assess expected response, but not the probability of accuracy on single replicates.
+
+% \cb{This section used to discuss the increased data sampling version, now removed}
+%A pressing question in the application of early warning signals is quantifying
+%when the data are sufficient for detection to be reliable~\citep{Scheffer2009, Scheffer2010, Inman2011}. 
+%Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
+%for each of the indicators (Fig 4). 
+%Using our generic models estimated from the original data as in Fig. 3, 
+%we generate the corresponding distributions by simulating under the models as before, 
+%but with more or less frequent sampling to illustrate how our ability to detect an approaching transition 
+%would be hampered or improved with more or less data.  
+%While all methods improve as the number of data points increase, 
+%the coefficient of variation and the variance show more substantial improvement than autocorrelation, 
+%as expected~\citep{Carpenter2011};⁠
+%skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4).
 
@@
 reliable identification of warning signals from summary statistics can be difficult.  
-The extent of this difficulty is not fully apparent from the warning statistic itself, Fig 2,
-or even the distribution of that statistic under the null model of a stable system, 
-but becomes clear from the ROC curve, where a 5\% false positive rate often corresponds to only a 5\% true positive rate as well.  
+The extent of this difficulty becomes clear from the ROC curves of these statistics, 
+where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
+performing no better than the flip of a coin. %%%% EEEK are you really going to say that?!?! 
 By estimating the ROC curve for a given set of data, 
@@
 By taking advantage of the assumptions being made to write down a specific likelihood function,
-we can develop indicators that get the most from this data.
+we can develop approaches that get the most information from the data available.
 
@@
 In any application of early warning signals, it is essential to address the question of model adequacy.  
-This is every bit as important for the summary-statistic based indicators such as variance and autocorrelation
-as it is for the approach taken here; these statistics correspond to particular assumptions about the underlying process,
-and systems that violate these assumptions can exhibit sudden transitions without any warning,
-such as chaotic systems or those with non-smooth potential functions~\citep{Hastings2010}.
-Sudden transitions have also been observed in ecological models in which 
-the variance deceases rather then increases as the system nears a bifurcation point~\citep{Schreiber2003, Schreiber2008, Dakos2011a}.  
 Our approach formalizes the assumptions about the underlying process to match the assumptions of the other warning signals.  
@@
 estimation of the nonlinear model is less powerful and less accurate.  
-The performance of this approach in the simulated data, which is nonlinear with non-Gaussian noise 
-(introduced by the Poisson process demographic events), 
+The performance of this approach in the simulated data -- which is nonlinear in its dynamics 
+and driven with non-Gaussian noise introduced by the Poisson demographic events -- 
 demonstrates the accuracy under violation of these assumptions.
@@
 \section{Acknowledgments}
-S.
+We would like to thank S.
 Schreiber, M.
@@
 Baskett, and A.
-Perkins provided comments on the manuscript.
+Perkins for comments on the manuscript.
 This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
@@
 \bibliography{boettiger}
-
+%\bibliography{~/Documents/Mendeley/bib/library}
  \end{document}

### 2012-02-07 boettiger.tex
spelling, formatting, linenumbers working with align environments

@@
 
-% erg, a bit mor transition needed?
-
 We discuss the limitations of current approaches that rely on summary statistics
@@
 a model-selection based approach, such as information criteria,
-and suggest reciever-operating characteristics~\citep{Green1989, Keller2009}
+and suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
 as a powerful way to illustrate the sensitivity of different data sets 
@@
 is something that would be measured across an ensemble -- across replicates. %% citation? 
-As true replicates are seldom avialable in systems for which developing warning signals would be most desirable,
+As true replicates are seldom available in systems for which developing warning signals would be most desirable,
 these methods average across a single replicate using a moving window in time.  
@@
 but this is an assumption that is violated by the very experimental design:
-temporal correlations will exist in any finely-enough samled time series, 
+temporal correlations will exist in any finely-enough sampled time series, 
 and moving windows introduce temporal correlations in the statistics.
@@
 and estimated from the data in a maximum likelihood framework.
-We highlight the disadvantages of comparing these estimates by information criterea, 
-and insead introduce a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
+We highlight the disadvantages of comparing these estimates by information criteria, 
+and instead introduce a simulation or bootstrapping approach rooted in~\citet{Cox1961} and~\citet{McLachlan1987} that
 characterizes the rate of missed detections and false alarms expected in the estimate.
@@
 In any model choice problem, it is important to identify the goal of the exercise --
-such as the ability to generalize, to immitate reality, or to predict~\citep{Levins1966}.  
+such as the ability to generalize, to imitate reality, or to predict~\citep{Levins1966}.  
 In this case generality is more important than realism or predictive capability: 
@@
 as we must first estimate the model parameters from the data.  
-In this approach it is not assumed that ``reality'' is included in the models being tested per se, 
+In this approach it is not assumed that ``reality'' is included in the models being tested, 
 but that one of the models is a better approximation of the true dynamics than the other.  
@@
 Transforming this canonical form to allow for an arbitrary mean $\theta$,
-the bifurcation looks like $ dx/dt = r_t- (\theta-x)^2 $, with fixed point $\hat x = \sqrt{r_t} +\theta =: \phi(r_t)$.
+the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
 We expand around the fixed point and express in terms of the stochastic dynamics:
@@
 
-As we discuss above, in this paradigme we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
+As we discuss above, in this paradigm we must include an assumption on how the bifurcation parameter, $r_t$, is changing.
 We assume a gradual, monotonic change which we approximate to first order: 
@@
 \subsection*{Likelihood calculations}
-The probability $P(X|M)$ of the data $X$ given the model $M$ is the product of the probability of observing each point in the time series given the previous point and the length of the interval,  
+The probability \(P(X|M) \) of the data \(X\) given the model \(M\) is the product of the probability of observing each point in the time series given the previous point and the length of the interval,
 \begin{equation}
@@
 \end{equation}
+
 For~\eqref{LSN} or~\eqref{OU} it is sufficient~\citep{Gardiner2009} to solve the moment equations for mean and variance respectively:
+
 \begin{align}
@@
 For the OU process, we can solve this in closed form over an interval of time $t_i$ between subsequent observations:
+
 \begin{align}
@@
 The moments of Equation~\eqref{LSN} are given by
+
 \begin{align}
@@
 The first is that information criteria are not concerned with the model choice objective we have in mind,
-as they are typically applied to find an adequate model description without too many parameters that the system may be overfit.  
+as they are typically applied to find an adequate model description without too many parameters that the system may be over-fit.  
 More pointedly, information criteria have no inherent notion of uncertainty.
@@
 It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing. 
-This introduces the a nuisance parameter, known as the statistical significance criterion.  
+This introduces a nuisance parameter known as the statistical significance criterion.  
 In the hypothesis testing framework, a false positive is a Type I error, 
@@
 The language of hypothesis testing is built around a bias that false positives are worse than false negatives, 
-and consequently an a emphasis on $p$-values rather than power.  
+and consequently an emphasis on $p$-values rather than power.  
 In the context of early warning signals this is perilous -- 
@@
 To avoid this linguistic bias and the introduction of an nuisance parameter on which to define statistical significance, 
-we propose the use of Reciever Operating characteristic (ROC) curves.
+we propose the use of receiver operating characteristic (ROC) curves.
 
@@
 We illustrate the trade-off between false alarms and failed detection using 
-receiver-operating characteristic (ROC) curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
+receiver-operating characteristic curves first developed in signal-processing literature~\citep{Green1989, Keller2009}⁠. 
 The curves represent the corresponding false alarm rate at any detection sensitivity (true positive rate), Fig~\ref{fig:roc_intro}.
@@
 This model is forced through a bifurcation by gradually increasing the $a$ parameter,
-which increases can be thought of as an increasing toxicity of the environmnent
+which increases can be thought of as an increasing toxicity of the environment
 (from $a_0 = 100$ increasing at constant rate of 0.09 units/unit time).  
@@
 
-Our second empirical data set comes from the glaciation record seen in deuterium levels in Antartic ice cores~\citep{Petit1999}, 
+Our second empirical data set comes from the glaciation record seen in deuterium levels in Antarctic ice cores~\citep{Petit1999}, 
 as analyzed by~\citet{Dakos2008}.
@@
 \subsection*{Analysis}
-We perform the identical analysis proceedure described above on each of these three data sets.
+We perform the identical analysis procedure described above on each of these three data sets.
 First, we estimate parameters for the null and test model to each data set by maximum likelihood.
@@
 determines the null distribution for the deviance statistic $\delta$.  
-As the constant OU process model is nested within the time-heterogenous model, these values are always positive,
+As the constant OU process model is nested within the time-heterogeneous model, these values are always positive,
 but tend to be not as large as those produced when the models are fit to the second family of data. 
@@
 The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.  
-The tendancy of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other 
+The tendency of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other 
 indicates the probability our observed data corresponds best with that model  -- 

### 2012-02-09 boettiger.tex
minor edit

@@
 as we would in classical hypothesis testing.  
-We could instead decide that we wanted not to catch no less than 90\% of the true-positives,
+We could instead decide that we wanted no less than 90\% of the true-positives,
 and read off the corresponding false positives.  
-For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
-while for likelihood a 55\%, 35\%, 0\% false positive.
+For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate, respectively,
+while for likelihood this gives us respectivel 55\%, 35\%, and 0\% false positives.
 

### 2012-02-09 boettiger.tex
alan hastings edits to intro

--- 2012-02-07 boettiger.tex
@@
 it is important to develop the ability to predict impending regime shifts based on early warning signs.
-With a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009},
-but in cases where specific models are not available general approaches are needed~\citep{Scheffer2009}⁠.
 
-We discuss the limitations of current approaches that rely on summary statistics
-and introduce a model based approach to detect early warning signals.
-We illustrate how stochastic differential equation (SDE) models can be used 
+A number of particular systems have demonstrated the kinds of relationships that would produce regime shifts, including dynamics of coral reefs (MUMBY ET AL, NATURE), and simple models of metapopulations with differing local population sizes (HASTINGS Hastings, A.
+(1991).
+Structured models of metapopulation dynamics.
+Biological Journal of the Linnean Society, 42, 57-71.).
+In cases like these one sensible approach to understanding whether a regime shift would be likely would be to fit the model using either a time series or else independent estimates of spratters.
+More generally, with a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009}.
+
+However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009} that do not depend on estimating the parameters of a known model of a specific system.
+This has lead to a variety of approaches based on summary statistics (REFS -- ) that look for generic signs of impending regime shifts.
+Our goal here is to extend earlier work by providing much more precise estimates of the ability of different statistics to signal impending regime shifts and to develop new approaches that both are more efficient and also lay bare some of the important assumptions underlying attempts to find general warning signs of regime shifts.
+We caution against paradigms that are not useful for capturing  uncertainty in
+a model-selection based approach, such as information criteria.
+
+We begin our approach by first discussing the limitations of current approaches that rely on summary statistics
+and provide a clear description of assumptions through the introduction of a model based approach to detect early warning signals.
+We then illustrate how stochastic differential equation (SDE) models can be used
 to reflect the uncertainty inherent in the detection of early warning signals.
-We caution against paradigms that are not useful for capturing this uncertainty in 
-a model-selection based approach, such as information criteria,
-and suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
+Finally we suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
 as a powerful way to illustrate the sensitivity of different data sets
-and different early warning indicators have in detecting early warning signals.
+and different early warning indicators have in detecting early warning signals and use this to explore a number of examples.
 

### 2012-02-10 boettiger.tex
working through noam edits

@@
 and illustrate how a likelihood-based indicator can improve this performance.
-
 % rewrite
@@
 There is an increasing recognition of the importance of regime shifts or critical transitions at a variety of scales in ecological systems~\citep{Holling1973, Wissel1984, Scheffer2001, Scheffer2009, Drake2010, Carpenter2011}⁠.
+% broader citations -- see how this is handled by others.  
 Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠.
 Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
-and the idea that management responses would be needed either to avoid an undesirable shift or else to adapt to novel conditions,
+and the need for management to either avoid an undesirable shift or else to adapt to novel conditions,
 it is important to develop the ability to predict impending regime shifts based on early warning signs.
 
-A number of particular systems have demonstrated the kinds of relationships that would produce regime shifts, including dynamics of coral reefs (MUMBY ET AL, NATURE), and simple models of metapopulations with differing local population sizes (HASTINGS Hastings, A.
-(1991).
-Structured models of metapopulation dynamics.
-Biological Journal of the Linnean Society, 42, 57-71.).
-In cases like these one sensible approach to understanding whether a regime shift would be likely would be to fit the model using either a time series or else independent estimates of spratters.
+A number of particular systems have demonstrated the kinds of relationships that would produce regime shifts,
+including dynamics of coral reefs~\citep{Mumby2007},
+and simple models of metapopulations with differing local population sizes~\citep{Hastings1991a}.
+In cases like these one sensible approach to understanding 
+whether a regime shift would be likely would be 
+to fit the model using either a time series or else independent estimates of parameters. 
 More generally, with a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009}.
 
-However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009} that do not depend on estimating the parameters of a known model of a specific system.
-This has lead to a variety of approaches based on summary statistics (REFS -- ) that look for generic signs of impending regime shifts.
-Our goal here is to extend earlier work by providing much more precise estimates of the ability of different statistics to signal impending regime shifts and to develop new approaches that both are more efficient and also lay bare some of the important assumptions underlying attempts to find general warning signs of regime shifts.
+However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009}
+that do not depend on estimating the parameters of a known model of a specific system. 
+This has lead to a variety of approaches based on summary statistics
+~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
+that look for generic signs of impending regime shifts.  
+Our goal here is to extend earlier work by providing much more precise estimates 
+of the ability of different statistics to signal impending regime shifts 
+and to develop new approaches that both are more efficient 
+and also lay bare some of the important assumptions underlying attempts 
+to find general warning signs of regime shifts.  
 We caution against paradigms that are not useful for capturing uncertainty in
@@
 as a powerful way to illustrate the sensitivity of different data sets
-and different early warning indicators have in detecting early warning signals and use this to explore a number of examples.
+and different indicators have in detecting early warning signals and use this to explore a number of examples.
 
@@
 that the system dynamics contain a saddle-node bifurcation by
-looking for patterns that are associated with kind of transition.
+looking for patterns that are associated with this kind of transition.
 Typical patterns include an increasing trend in a summary statistic such as
@@
 
+\subsubsection*{Hidden assumptions}
+%Differing system dynamics
+%% OKAY, so this is a bit of a stretch.
+
+The underlying assumption that the system contains a saddle-node bifurcation,
+can be easily overlooked in common summary-statistics based approaches.
+For instance, variance may increase for reasons that
+do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
+Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
+Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
+Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
+and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
+However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
+they are characterized by a steady approach towards the alternative stable state, 
+such as a population size gradually approaching zero.  
+While no approach will be applicable to all classes of sudden transitions,
+it is certainly still useful to have an approach that detect transitions driven by
+saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.
+
+%Differing causes of the transition.
+Even when we can exclude or ignore such dynamics and
+restrict ourselves to systems that can support a saddle-node bifurcation,
+these approaches must further assume that a changing parameter
+has brought the system closer to the bifurcation.
+This excludes at least three alternative explanations for the transition.
+The first is that a large perturbation of the system state
+has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
+This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
+(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
+The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}
+As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
+The third scenario is that the system does pass through a saddle-node bifurcation,
+but rather than gradually and monotonically approaching the critical point, the
+bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.
+
+%% Not sure if this paragraph above is necessary.
+
+% Of course the literature has always stuck carefully to these assumptions, but there has sometimes been the suggestion that the summary statistics themselves were somehow more general than this, and that framing these assumptions in explicit and precise formulae, in models, would be misleading.
+
 \subsubsection*{Arbitrary windows}
-While superficially simple to use,
-the calculation of these statistics is subject to several arbitrary decisions and assumptions.
-
-A basic difficulty arises from the need to assume a time-series is ergodic
+In addition to the assumption of a saddle node bifurcation, 
+the calculation of these statistics is subject to several additional arbitrary choices.
+A basic difficulty arises from the need to assume a time-series is ergodic: 
+that averaging over time is equivalent to averaging over replicate realizations,
 while trying to test if it is not.
@@
 In addition to introducing the challenge of selecting a window size,
-this ergodic assumption,
-that averaging over time is equivalent to averaging over replicate realizations,
-raises further difficulties.
+this ergodic assumption raises further difficulties.
 While appropriate for a system that is stationary,
@@
 In these cases a quantitative definition of a statistically significant detection is essential.
-Using Kendall's $\tau$ we will illustrate how such a quantitative measure
-can be used to better capture the uncertainty in identifying the early warning pattern.
 
@@
 such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.
-
-\subsubsection*{Hidden assumptions}
-%Differing system dynamics
-The underlying assumption that the system contains a saddle-node bifurcation
-can be easily overlooked in common summary-statistics based approaches.
-For instance, variance may increase for reasons that
-do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
-Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
-Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
-While no approach will be applicable to all classes of sudden transitions,
-it is certainly still useful to have an approach that detect transitions driven by
-saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.
-
-%Differing causes of the transition.
-Even when we can exclude or ignore such dynamics and
-restrict ourselves to systems that can support a saddle-node bifurcation,
-these approaches must further assume that a changing parameter
-has brought the system closer to the bifurcation.
-This excludes at least three alternative explanations for the transition.
-The first is that a large perturbation of the system state
-has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
-This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
-(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
-The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}
-As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
-The third scenario is that the system does pass through a saddle-node bifurcation,
-but rather than gradually and monotonically approaching the critical point, the
-bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.
-
-%% Not sure if this paragraph above is necessary.
-
-% Of course the literature has always stuck carefully to these assumptions, but there has sometimes been the suggestion that the summary statistics themselves were somehow more general than this, and that framing these assumptions in explicit and precise formulae, in models, would be misleading.
 
@@
 the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
-We expand around the fixed point and express in terms of the stochastic dynamics:
+We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:
 
 \begin{equation}
-\ud X = \sqrt{ r_t } (\phi(r_t) - X_t)\ud t + \sigma\sqrt{\phi(r_t) } \ud B_t. \label{LSN}
+\ud X = \sqrt{ r_t } (\phi(r_t) - X_t)\ud t + \sigma\sqrt{\phi(r_t) } \ud B_t \label{LSN}
 \end{equation}
 
+where $B_t$ is the standard Brownian motion. 
+This expression captures the behavior of the system near the stable point as it approaches 
 %It is also possible to derive this same expression from the master equation
@@
 \section{Acknowledgments}
-We would like to thank S.
-Schreiber, M.
-Holyoak, M.
-Baskett, and A.
-Perkins for comments on the manuscript.
-This research was supported by funding from NSF Grant EF 0742674 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308.
-Data and code are available at: https://github.com/cboettig/warningsignals.
+We would like to thank S. Schreiber, M. Holyoak, M. Baskett, and A. Perkins for comments on the manuscript. 
+This research was supported by funding from NSF Grant EF 0742674 
+and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308. 
+Data and code are available at \href{https://github.com/cboettig/warningsignals}{ https://github.com/cboettig/warningsignals}.
 

### 2012-02-10 boettiger.tex
a few updates

@@
 \subsubsection*{Hidden assumptions}
-%Differing system dynamics
-%% OKAY, so this is a bit of a stretch.
-
 The underlying assumption that the system contains a saddle-node bifurcation,
@@
 Other parameters are: $Xo = 730$, $e = 0.5$, $K = 1000$, $h = 200$.
-We run this model over a time interval from 0 to 500.
+We run this model over a time interval from 0 to 500
 and sample at 40 evenly spaced time points, which were used for subsequent analysis.
@@
 
-Because ROC curves compare areas under the probability distributions they can always be presented on a uniform 0-1 axis
-for both true positives and false positives, allowing us to overlay the curves from different data on the same plot,
-Figure~\ref{fig:rocdata}.
+%Because ROC curves compare areas under the probability distributions they can always be presented on a uniform 0-1 axis
+%for both true positives and false positives, allowing us to overlay the curves from different data on the same plot,
+%Figure~\ref{fig:rocdata}.
 

### 2012-02-12 boettiger.tex
added Noam's edits, captions

@@
 We have extended the approach by generating the test distribution
-as well as a null distribution of the statistic $\delta$ in order to compute ROC curves.
-%Our results show that $\delta$ tends to result in lower error rates than $\tau$,
-%a measure of increase in a summary statistic.
+as well as a null distribution of the statistic $\delta$.
+
+\subsection{A simulation-based approach estimates the distributions}
+We perform the identical analysis procedure described above on each of these three data sets.
+First, we estimate parameters for the null and test model to each data set by maximum likelihood.
+Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better --
+
+To identify if these differences are significant,
+and by what probability they could arise as a false alarm or a missed event,
+we simulate 500 replicate time series from each estimated model,
+as visualized in the middle panel of each data set,
+Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}.
+
+The model parameters of both models are re-estimated on both families of replicates
+(the null and test, \emph{i.e.} $2 \times 2 \times 500$ fits).
+The differences in the likelihood values between the model estimates produced from the first set of simulations
+determines the null distribution for the deviance statistic $\delta$.
+As the constant OU process model is nested within the time-heterogeneous model, these values are always positive,
+but tend to be not as large as those produced when the models are fit to the second family of data.
+
+The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.
+The tendency of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other
+indicates the probability our observed data corresponds best with that model  --
+either approaching a critical transition or remaining stable.
+While it trivial to assign a statistical significance to this observation based
+on how far into the tail of the (red) null distribution it falls,
+for the reasons we discussed we prefer the more symmetric comparison of the probability that
+this value was observed in either distribution.
+We visualize the trade-off between false alarms and failed detection using the ROC curves introduced above.
 
@@
 it suggests that we would rather fail to predict a catastrophe than to sound a false alarm.
-
 To avoid this linguistic bias and the introduction of an nuisance parameter on which to define statistical significance,
@@
 We illustrate this approach on simulated data as well as several natural time-series that have been previously analyzed for early warning signals.
+All data and code for simulations and analysis are found in the accompanying R package, \verb|warningsignals|.
 
 \subsection*{Data}
-The simulation implements an individual, continuous-time stochastic birth-death process with rates given by the master equation,
+The simulation implements an individual, continuous-time stochastic birth-death process with rates given by the master equation~\citep{Gardiner2009},
 
 \begin{align}
-    P(n,t) &= b_{n-1} p_{n-1} + d_{n+1}p_{n+1} - (b_n+d_n)p_n  \label{master} \\
-    b_n &= \frac{e K x^2}{X^2 + h_t^2} \\
-    d_n &= e X_t + a_t
+  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
+    b_n &= \frac{e K n^2}{n^2 + h^2} \\
+    d_n &= e n + a_t
 \end{align}
 
+where $P(n,t)$ is the probability of having $n$ individuals at time $t$, $b_n$ is the probability of a birth event occuring in a population of $n$ individuals, $d_n$ the probability of a death.
+$e, K, h$ and $a_t$ are parameters.
 This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977},
-with stochasticity introduced from directly through the demographic process.
+with stochasticity introduced directly through the demographic process.
+We select this model since it is has discrete numbers of individuals, 
+nonlinear processes, and the noise is driven by Poisson process of births and deaths instead of a Gaussian, 
+and thus provides an illustration that our approach is robust to the violations of those
+assumptions in model~\eqref{LSN}.
+
 This model is forced through a bifurcation by gradually increasing the $a$ parameter,
@@
 and sample at 40 evenly spaced time points, which were used for subsequent analysis.
-We refer to this as the simulation under a degrading environment.
-We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.
+This sampling frequency was chosen to be representative of reasonable sampling in biological time-series,
+and provides enough points to detect a signal while not too many that errors can be avoided entirely.  
 For the convenience of the inquisitive reader, we have also provided a simple function in the
 associated R package where the user can vary the sampling scheme and parameter values and rerun this analysis.
+This time series is shown in the top panel of Figure~\ref{fig:simulation}.
+
+%We refer to this as the simulation under a degrading environment.
+%We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.
+
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=.73\linewidth]{figures/Simulation_warningsignal.pdf}
+     \caption{A model-based calculation of warning signals for the simulated data example.
+Top panel: The original time series data on which model parameters for Equations~\eqref{OU}~and~\eqref{LSN} are estimated.
+Middle panel: replicate simulations under the maximum likelihood estimated (MLE) parameters of the null model, Equation~\eqref{OU} and test model, Equation~\eqref{LSN}.
+Bottom panel: The distribution of deviances (differences in log likelihood, Equation~\eqref{delta}), when both null and test models are fit to each of the replicates from the null model, ``null,'' in red, and these differences when estimating for each of the replicates from the test model, in blue.
+The overlap of distributions indicate replicates that will be difficult to tell apart.
+The observed differences in the original data are indicated by the vertical line.}
+     \label{fig:simulation}
+  \end{center}
+ \end{figure}
 
@@
 a pattern of increasing variance over the 16 data points where the system was being manipulated towards a crash.
-All data and code for simulations and analysis are found in the accompanying R package, \verb|warningsignals|.
+This time series is shown in the top panel of Figure~\ref{fig:chemostat}.
 
-Our second empirical data set comes from the glaciation record seen in deuterium levels in Antarctic ice cores~\citep{Petit1999},
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=.85\linewidth]{figures/Chemostat_warningsignal.pdf}
+     \caption{A model-based calculation of warning signals for the Daphnia data analyzed in ~\citet{Drake2010} (Chemostat H6).  
+     Panels as in Figure~\ref{fig:simulation}. }
+     \label{fig:chemostat}
+  \end{center}
+ \end{figure}
+
+Our second empirical data set comes from the glaciation record seen
+in deuterium levels in Antarctic ice cores~\citep{Petit1999},
 as analyzed by~\citet{Dakos2008}.
@@
 which did not apply to the other data sets~\citep{Drake2010}.
+This time series is shown in the top panel of Figure~\ref{fig:glaciation}.
+ \begin{figure}
+   \begin{center}
+     \includegraphics[width=.85\linewidth]{figures/GlaciationIII_warningsignal.pdf}
+     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).  Panels as in Figure~\ref{fig:simulation}.}
+     \label{fig:glaciation}
+  \end{center}
+ \end{figure}
 
 \subsection*{Analysis}
-We perform the identical analysis procedure described above on each of these three data sets.
-First, we estimate parameters for the null and test model to each data set by maximum likelihood.
-Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better --
-the deviances $\delta$, Eqn~\eqref{delta} -- are 5.1, 6.0, 83.9 respectively for the simulation,
+The deviances $\delta$, Eqn~\eqref{delta} -- are 5.1, 6.0, 83.9 respectively for the simulation,
 the chemostat data and the glaciation data.
@@
 
-To identify if these differences are significant,
-and by what probability they could arise as a false alarm or a missed event,
-we simulate 500 replicate time series from each estimated model,
-as visualized in the middle panel of each data set,
-Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}.
+As described above, we simulate 500 replicates under each model, 
+shown in the middle panels of Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation},
+and determine the distributions in likelihood ratio under each, shown in the lower panels.  
+The observed deviance from the original data is also indicated (vertical line).
 
- \begin{figure}
-   \begin{center}
-     \includegraphics[width=.85\linewidth]{figures/Simulation_warningsignal.pdf}
-     \caption{A model-based calculation of warning signals for the simulated data example.}
-     \label{fig:simulation}
-  \end{center}
- \end{figure}
-
- \begin{figure}
-   \begin{center}
-     \includegraphics[width=.85\linewidth]{figures/Chemostat_warningsignal.pdf}
-     \caption{A model-based calculation of warning signals for the Daphnia data analyzed in ~\citet{Drake2010} (Chemostat H6).}
-     \label{fig:chemostat}
-  \end{center}
- \end{figure}
-
- \begin{figure}
-   \begin{center}
-     \includegraphics[width=.85\linewidth]{figures/GlaciationIII_warningsignal.pdf}
-     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).}
-     \label{fig:glaciation}
-  \end{center}
- \end{figure}
-
-The model parameters of both models are re-estimated on both families of replicates
-(the null and test, \emph{i.e.} $2 \times 2 \times 500$ fits).
-The differences in the likelihood values between the model estimates produced from the first set of simulations
-determines the null distribution for the deviance statistic $\delta$.
-As the constant OU process model is nested within the time-heterogeneous model, these values are always positive,
-but tend to be not as large as those produced when the models are fit to the second family of data.
-These two distributions are shown in the lower panel of
-Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation},
-along with the deviance observed in the original data (vertical line).
-
-The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.
-The tendency of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other
-indicates the probability our observed data corresponds best with that model  --
-either approaching a critical transition or remaining stable.
-While it trivial to assign a statistical significance to this observation based
-on how far into the tail of the (red) null distribution it falls,
-for the reasons we discussed we prefer the more symmetric comparison of the probability that
-this value was observed in either distribution.
-We visualize the trade-off between false alarms and failed detection using the ROC curves introduced above.
+The ROC curves for each of these data sets is plotted in Figure~\ref{fig:rocdata}.  
+While differences in the rate at which the system approaches a transition will also 
+improve the ratio of true positives to false positives, 
+here we see the best-sampled data set, Glaciation, with 121 points, 
+also has the clearest signal with no observed errors in the 500 replicates of each type.   
+Comparing the chemostat and simulation curves illustrate how the trade-off between
+false positives and true positives can vary between data.  The chemostat signal, 
+which estimates a relatively rapid rate of change but has less data, captures a higher
+rate of true positives for a given rate of false positives than the simulation data
+set with a weaker rate of change but more data, for false positive rates above 20\%.  
+However, the simulated set with more data performs better if lower false-positive rates
+are desired.
 
@@
      \includegraphics[width=\linewidth]{figures/rocplot.pdf}
-     \caption{ }
+     \caption{ROC curves for the Simulation, Chemostat, and Glaciation data, computed from the distributions shown in Figures~\ref{fig:simulation}, \ref{fig:chemostat} and \ref{fig:glaciation}, bottom panel.}
      \label{fig:rocdata}
@@
      \caption{Early warning signals in simulated and empirical data sets.  
-The first two columns are simulated data from (a) a stable system (Stable), and (b) the same system approaching a saddle-node bifurcation (Deteriorating).
-Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated towards a critical transition (Daphnia), and (d) deuterium concentrations previously cited as an early warning signal of a glaciation period (Glaciation).
-Increases in summary statistics, computed over a moving window, have often been used to indicate if a system is moving towards a critical transition.
+     The first two columns are simulated data from (a) a stable system (Stable), 
+     and (b) the same system approaching a saddle-node bifurcation (Deteriorating).  
+     Empirical examples are from (c) \emph{Daphnia magna} concentrations manipulated towards a critical transition (Daphnia), 
+     and (d) deuterium concentrations previously cited as an early warning signal of a glaciation period (Glaciation). 
+     Increases in summary statistics, computed over a moving window, 
+     have often been used to indicate if a system is moving towards a critical transition.  
      The increase is measured by the correlation coefficient $\tau$.  
-Note that positive correlation does not guarantee the system is moving towards a transition, as seen in the stable system, first column.}
+     Note that positive correlation does not guarantee the system is moving towards a transition, 
+     as seen in the stable system, first column.}
      \label{fig:summary_stats}
@@
      \includegraphics[width=\linewidth]{figures/summary_box.pdf}
-     \caption{}
+     \caption{Box-plots of the distributions of Kendall's~$\tau$ observed for the summary statistic methods variance and autocorrelation, applied to three different data sets (from Figures~\ref{fig:simulation},~\ref{fig:chemostat}, \ref{fig:glaciation}).
+The distributions show extensive overlap, suggesting that it will be difficult to distinguish early warning signals by the correlation coefficient in these summary statistics.}
      \label{fig:summary_box}
@@
 These distributions are shown in Figure~\ref{fig:summary_box}.
+Contrary to the expectation that replicates of the null model (stable system, Equation~\eqref{OU}) would cluster around zero, 
+while the test model, Equation~\eqref{LSN}, would cluster around larger positive $\tau$ values, 
+the observed $\tau$ values on the replicates extend evenly across the range.  
+This results in dramatic overlap and offers little ability to distinguish between the stable replicates 
+and the replicates approaching a transition.
+
+The use of box plots in Figure~\ref{fig:summary_box} provide a convient and familiar way 
+to visualize the overlap between more than two distributions,
+though they lack the resolution of the overlapping density distributions in 
+Figures~\ref{fig:simulation},~\ref{fig:chemostat}, \ref{fig:glaciation}.  
+The overlapping distributions are the natural representation from which to introduce the ROC curve, as in Figure~\ref{fig:roc_intro}.
 
@@
      \includegraphics[width=\linewidth]{figures/summary_roc.pdf}
-     \caption{}
+     \caption{ROC curves compare the performance of the summary statistics variance and autocorrelation against the likelihood-based approach from Figure~\ref{fig:rocdata} on each of three example data sets (Figures~\ref{fig:simulation},~\ref{fig:chemostat}, \ref{fig:glaciation}).}
      \label{fig:summary_roc}
@@
 The large correlations observed in the empirical examples (Fig.~\ref{fig:summary_stats}) are not uncommon in stable systems.
-For instance, we can read from the curve what fraction of true positives would be detected by a variance-based indicator
-set to a 5\% false positive rate:
-25\% in the simulated set,
-5.0\% in the \emph{Daphnia} set,
-and 5.4\% in the Glaciation data set.
-Compare this to likelihood: 61\%, 34\%, and 100\% of the replicate true positives
-would be correctly identified from the respective systems.
-The value of the ROC curve is that we need not restrict the analysis to a pre-specified error such as 5\%,
-as we would in classical hypothesis testing.
-We could instead decide that we wanted not to catch no less than 90\% of the true-positives,
-and read off the corresponding false positives.
-For variance this corresponds respectively to a 49\%, 81\%, and 93\% false positive rate,
-while for likelihood a 55\%, 35\%, 0\% false positive.
+It is notable that in both empirical examples the summary statistics approach does little better than chance in distinguishing 
+replicates that have been simulated from models~\eqref{LSN} and ~\eqref{OU}, 
+despite the fact that these models correspond 
+to the assumptions of the summary statistics approaches.  
+On the simulated data, the varianced based method approaches the 
+true-positive rate of our likelihood method at higher levels of false positives, 
+but performs worse when the desired level of false postives is low.  
+The ROC curve helps us compare the performance of the different approaches at different tolerances.  
+For instance, Table~\ref{tab:falsepos} shows the fraction of true crashes caught at a 5\% false positive rate.
+We can instead set a desired True positive rate and read off the resulting number of false alarms, Table~\ref{tab:truepos}.
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{l|l|l}
+                  &  Variance & Likelihood \\ 
+      \hline
+      Simulation  & 25 \%     & 61\% \\
+      Chemostat   & 5.0\%     & 34\% \\
+      Glaciation  & 5.4\%     & 100\% 
+    \end{tabular}
+    \caption{Fraction of \emph{crashes detected} when the desired false alarm rate is fixed to 5\%}
+    \label{tab:falsepos}
+  \end{center}
+\end{table}
+
+\begin{table}
+  \begin{center}
+    \begin{tabular}{l|l|l}
+                  &  Variance & Likelihood \\ 
+      \hline
+      Simulation  & 49 \%     & 55\% \\
+      Chemostat   & 81 \%     & 35\% \\
+      Glaciation  & 93 \%     & 0\% 
+    \end{tabular}
+    \caption{Fraction of \emph{false alarms} when the desired detection rate is fixed to 90\%}
+    \label{tab:truepos}
+  \end{center}
+\end{table}
 
 \section{Discussion}
-
-% \cb{This section used to discuss the increased data sampling version, now removed}
-%A pressing question in the application of early warning signals is quantifying
-%when the data are sufficient for detection to be reliable~\citep{Scheffer2009, Scheffer2010, Inman2011}.
-%Our approach provides a natural way to address this challenge by estimating the ROC curve under differing data sampling efforts
-%for each of the indicators (Fig 4).
-%Using our generic models estimated from the original data as in Fig. 3,
-%we generate the corresponding distributions by simulating under the models as before,
-%but with more or less frequent sampling to illustrate how our ability to detect an approaching transition
-%would be hampered or improved with more or less data.
-%While all methods improve as the number of data points increase,
-%the coefficient of variation and the variance show more substantial improvement than autocorrelation,
-%as expected~\citep{Carpenter2011};⁠
-%skew performs little better than chance, and likelihood is consistently the most reliable and sensitive (Fig. 4).
-
 By assuming the underlying model corresponds to a saddle-node bifurcation,

### 2012-02-14 boettiger.tex
Merged Alan's edits

@@
 to find general warning signs of regime shifts.
-We caution against paradigms that are not useful for capturing uncertainty in
-a model-selection based approach, such as information criteria.
 
@@
 We then illustrate how stochastic differential equation (SDE) models can be used
-to reflect the uncertainty inherent in the detection of early warning signals.
-Finally we suggest receiver-operating characteristics~\citep{Green1989, Keller2009}
+to reflect the uncertainty inherent in the detection of early warning signals. We caution against paradigms that are not useful for capturing uncertainty in
+a model-selection based approach, such as information criteria.
+Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
 as a powerful way to illustrate the sensitivity of different data sets
@@
 skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
-While attractive for their simplicity, such approaches are subject to numerous challenges.
+While attractive for their simplicity, such approaches must confront numerous challenges.
 In this paper we argue for a model-based approach,
@@
 \subsubsection*{Hidden assumptions}
-The underlying assumption that the system contains a saddle-node bifurcation,
+The underlying assumption that the system contains a saddle-node bifurcation
 can be easily overlooked in common summary-statistics based approaches.
@@
 %Differing causes of the transition.
-Even when we can exclude or ignore such dynamics and
-restrict ourselves to systems that can support a saddle-node bifurcation,
-these approaches must further assume that a changing parameter
+Even when we can exclude or ignore other dynamics and
+restrict ourselves to systems that can produce a saddle-node bifurcation,
+ approaches based on critical slowing down or rising variance (REFS) must further assume that a changing parameter
 has brought the system closer to the bifurcation.
-This excludes at least three alternative explanations for the transition.
-The first is that a large perturbation of the system state
+This assumption excludes at least three alternative explanations for the transition.
+The first possibility is that a large perturbation of the system state
 has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
@@
 (An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
-The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary.~\citep{Ditlevsen2010}
+The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
 As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
@@
 In addition to the assumption of a saddle node bifurcation, 
-the calculation of these statistics is subject to several additional arbitrary choices.
+the calculation of statistics that would be used to detect an impending transition is subject to several additional arbitrary choices.
 A basic difficulty arises from the need to assume a time-series is ergodic: 
@@
 
-Further, certain statistics such as the critical slowing down measured in autocorrelation
+Further, certain statistics such as the critical slowing down measured by autocorrelation
 require data that is evenly sampled in time.
@@
 \section{A model based approach}
-Model-based approaches are beginning to play a larger role in early warning signals,
+Model-based approaches are beginning to play a larger role in early warning signal detection,
 though we have not as yet seen the direct fitting and simulation of models to compare hypotheses.
@@
 consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
-Transforming this canonical form to allow for an arbitrary mean $\theta$,
+Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
 the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
@@
 where $B_t$ is the standard Brownian motion. 
-This expression captures the behavior of the system near the stable point as it approaches 
+This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
 %It is also possible to derive this same expression from the master equation
@@
 Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
-When the system is stable, $r_t$ is constant and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
+When the system is stable, $r_t$ is constant ($m=0$) and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
 \begin{equation}
@@
 in both Frequentist and Bayesian paradigms.
-The ability evaluate likelihoods directly by computation has made it
+The ability to evaluate likelihoods directly by computation has made it
 possible to treat cases that do not conform to traditional assumptions more directly.
@@
 
-The ROC curves for each of these data sets is plotted in Figure~\ref{fig:rocdata}.  
+The ROC curves for each of these data sets are plotted in Figure~\ref{fig:rocdata}.  
 While differences in the rate at which the system approaches a transition will also 
@@
 \section{Discussion}
+
+The challenge of determining early warning signs for impending possible regime shifts requires real attention to the underlying statistical issues and other assumptions.
+Doing this, does, however, open up new possibilities for asking what the goal of detection should be, and for clearly identifying underlying assumptions.
+We consider alternative approaches based either on summary statistics or a likelihood based model choice.
 By assuming the underlying model corresponds to a saddle-node bifurcation,
-our analysis presents a ``best-case scenario'' for both the summary statistics and the likelihood-based approach.
+our analysis presents a ``best-case scenario'' for both  summary statistic and  likelihood-based approaches.
 Other literature has already begun to address the additional challenges posed when the underlying
@@
 reliable identification of warning signals from summary statistics can be difficult.
-The extent of this difficulty becomes clear from the ROC curves of these statistics,
+
+Asa way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms, we introduce an approach based on receiver operator curves.
+These curves XXX.
+The extent of the difficulty in finding reliable indicators of impending regime shifts based on summary statistics.
+becomes clear from the ROC curves of these statistics,
 where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
@@
 we can develop approaches that get the most information from the data available.
+
+DISUCSS EXAMPLES HERE
 

### 2012-02-16 appendix/appendices_knit_.Rnw
adding knitted appendices

%\VignetteIndexEntry{PMC Tutorial}
\documentclass{elsarticle}

%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage[pdftex, colorlinks]{hyperref}
\usepackage{amsmath, amsfonts}  % extended mathematics
\usepackage{booktabs} % book-quality tables
\usepackage{float}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}

%% Looks like a comment but it isn't! This is setting the default behavior for the Sweave chunk options, <<options >>= 
% \SweaveOpts{fig.width=7, fig.height=6, fig.align=center, message=FALSE, warning=FALSE}

\begin{document}
\begin{frontmatter}
  \title{Limits to detection of early warning signals: Appendices}
  \author[davis]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  %\author[davis]{}
  \cortext[cor1]{Corresponding author.}
  \address[davis]{Center for Population Biology, University of California, Davis, United States}

  \begin{abstract}
  These appendices explore several concepts that do not get fully developed in the paper, while also providing an introduction to the early warning signals package we provide to support this paper.  This document is written in Sweave, so the code you see is the code that generates the figures and results shown.   
  \end{abstract}
 \end{frontmatter}

\section{Introduction}
\setkeys{Gin}{width=0.85\textwidth}

<<libraries, echo=FALSE>>=
require(populationdynamics) 
require(earlywarning)
require(plyr)
require(ggplot2)
require(reshape2)
nreps <- 64
@

Early warning signals are most often presented in the literature as graph showing the monotonic increase of some statistic over time.  It is not necessarily obvious how best to quantify this pattern, nor does it seem common practice to do so.  Though these patterns often look pretty montonic, we must be careful not to forget that monotonic increases in a statistic will appear by chance at some rate.  In small data sets, the probabilty of a chance pattern is higher.  These basic statistical concerns seem largely absent from this literature, but if visual patterns speak more strongly than statistics, here's the problem in visual form:

Simulate from a system approaching a collapse~\Sexpr{nreps} times,

<<crash>>=
require(populationdynamics) 
pars = c(Xo = 730, e = 0.5, a = 100, K = 1000, h = 200, i = 0, Da = .09, Dt = 0, p = 2)
time=seq(0, 900, length=200)
sn <- saddle_node_ibm(pars,time, reps=nreps)
@

We do a little reformatting of the data, 
<<reformat>>=
X <- data.frame(time=time, value=sn$x1)
dat <- melt(X, id="time")
names(dat)[2] <- "reps"
@
and compute the autocorrelation over a window in time equal half the length if the timeseries
<<autocorr>>=
acorr <- summary_statistic(dat, window_autocorr, windowsize=length(time)/2)
@

\begin{figure}[H]
\begin{center}
<<crashacor>>=
ggplot(acorr) + geom_line(aes(time, value)) + facet_wrap(~reps) + 
  opts(title="Autocorrelation on replicates from a system approaching a crash")
@
\caption{The pattern of autocorrelation in replicate simulations from a simulation of a system which is approaching a saddle node bifurcation.}\label{crashacor}
\end{center}
\end{figure}

While we can see an average increasing trend, idenifying a particular squiggle as an early warning sign looks a lot more like judgement call then science.  Reality is unlikely to be any less murky, so the purpose this paper is to provide statistical guidance about how to start approaching this uncertainty.

We provide the code in this document to both illustrate the steps required and also that the commands can be copy-pasted into R using different choices for the warning indicator, the sampling of the simulation, etc.

\section{Model-based warning signals}

We load the package and the data set from the individual-based simulations.  To this data we fit both the LSN and OU models, 
<<ibmmodels>>=
require(earlywarning)
data(ibms)
A <- stability_model(ibm_critical, "OU")
B <- stability_model(ibm_critical, "LSN")
observed <- -2 * (logLik(A) - logLik(B))
@

Comparing the models by the bootstrapping procedure is intensive.  `compare` simulates under each model, and then estimates both models on both simulated sets of data (for a total of 4 model fits).  As only the LSN fit is computationally intensive, this takes about twice the time as the above fitting step of model B.  We will need to repeat this many times to build up a bootstrap distribution. Consequently, it is convient to run this step in parallel over multiple processors or clusters of computers.

\subsection{Parallelization of the comparison}
Fortunately R has a diverse array of parallel tools.  Here we set up the flexible \texttt{snowfall} parallel envirnoment, which can run in serial, on a multicore chip, or over a cluster of machines.

<<snowfall, include=FALSE, eval=FALSE>>=
require(snowfall)
sfInit(par=T, cpu=16) 
sfLibrary(earlywarning)
sfExportAll()
@

Rather than build this choice into the earlywarning signals package, the user can choose whatever looping implementation they desire to compute replicates of the comparison.  The actual analysis takes just one line:

<<compare, eval=FALSE>>=
reps <- sfLapply(1:500, function(i) compare(A,B))
@

Our results are cached in the package, so anyone seeking to explore this example can run
<<cache>>=
data(manuscript_data)
@

\subsection{Visualizing the distributions}

Once this step is finished, the remaining analysis is just a matter of shuffling and plotting the data.   We extract the likelihood ratios (deviances) from the null and test simulations with a simple function.  
<<lr>>=
lr <- lik_ratios(reps)
@

We can visualize this data in several ways: as two overlapping distributions,

\begin{figure}[H]
<<dists>>=
require(ggplot2)
ggplot(lr) + geom_density(aes(value, fill=simulation), alpha=.7) + 
  geom_vline(xintercept=observed, lty=2)
@
\end{figure}

or as beanplot, a modification of a boxplot showing the continuous distributions side-by-side:

\begin{figure}[H]
<<beanplot>>=
require(beanplot)
beanplot(value ~ simulation, lr, what=c(0,1,0,0))
abline(h=observed, lty=2)
@
\end{figure}

\subsection{Generating the ROC Curve}

We can make the ROC curve by calculating the true and false positive rates from the overlap of these distributions using the function \verb|roc_data|.  This gives us a data frame of threshold values for our statistic, and the corresponding false positive rate and true positive rate.  We can simply plot those rates against each other to create the ROC curve.

\begin{figure}[H]
<<rocplot>>=
roc <- roc_data(lr)
ggplot(roc) + geom_line(aes(False.positives, True.positives), lwd=1)
@
\end{figure}

\subsection{ Parameter distributions}

We can also look at the bootstraps of the parameters.  Another helper function will reformat this data from reps list.  The fit column uses a two-letter code to indicate first what model was used to simulate the data, and then what model was fit to the data.  For instance, AB means the data was simulated from model A (the null, OU model) but fit to B.

<<getpars>>=
pars <- parameter_bootstraps(reps)
head(pars)
@

There are lots of options for visualizing this relatively high-dimensional data, which we can easily explore with a few commands from the \texttt{ggplot2} package. For instance, we can look at average and range of parameters estimated in the bootstraps of each model:

\begin{figure}[H]
<<parplot>>=
require(Hmisc)
ggplot(subset(pars, fit %in% c("AA", "BB")), aes(fit, value)) +
  stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
  stat_summary(fun.data = median_hilow, geom = "pointrange", 
               position = position_dodge(width = 0.90), conf.int = 0.95) +
  facet_wrap(~parameter, scales="free_y")
@
\end{figure}

\subsection{Visualizing the simulation replicates}

The \texttt{reps} list also contains the simulation data from each replicate

\begin{figure}[H]
<<<repsplot, fig.width=10>>=
sims <- get_replicates(reps)
ggplot2(sims) + geom_line(aes(time, value, group=rep), alpha=0.1) + facet_wrap(~model)
@
\end{figure}

\section{Computing summary statistics with improved sampling}

Autocorrelation performs particularly poorly on the examples in the text, falling on approximately the 1:1 line in the ROC curves.  This example goes through the steps to demonstrate that in a sufficiently-frequently sampled timeseries, autocorrelation does contain *some* signal of early warning, it is simply a weaker signal than the corresponding analysis with variance, which in turn is weaker than the model-based approach. This example also illustrates the use of some of the summary statistics tools.

We begin by running the individual based simulation, supplied in our helper package, populationdynamics, with 500 sample points, instead of the 40 shown in the text.

<<resim>>=
require(populationdynamics) 
pars = c(Xo = 730, e = 0.5, a = 100, K = 1000, h = 200, i = 0, Da = .09, Dt = 0, p = 2)
time=seq(0, 500, length=500)
sn <- saddle_node_ibm(pars,time)
X <- data.frame(time=time, value=sn$x1)
@

We compute the observed value of Kendall's tau for the autocorrelation computed in a moving window over the data.  (The warningtrend function is just a wrapper for the base function cor.test, which handles different window sizes.  The default uses half the length of the timeseries.)

<<observed_tau>>=
observed <- warningtrend(X, window_autocorr) 
observed
@

While this gives us single value which is useful for statistical comparisons that follow, recall that it is more common to simply plot the autocorrelation computed in this manner, and the increase is just the visual pattern.

\begin{figure}[H]
<<plot_autocorr>>=
npts <- length(X[["value"]])
autocorrelation <- window_autocorr(X[["value"]], windowsize=npts/2)
autocorrelation <- c(rep(NA, npts/2), autocorrelation)
plot(X[[time]], autocorrelation, type="l")
@
\end{figure}

To bootstrap the estimate of tau, we need to be able to simulate under a null and test model.  We use our models of a stable system and a system approaching a collapse to do this.  We first estimate the model parameters from the data,

<<morefits>>=
A <- stability_model(X, "OU")
B <- stability_model(X, "LSN")
@

and then we can simulate some replicates

<<moresims>>=
reps <- 100
Asim <- simulate(A, reps)
Bsim <- simulate(B, reps)
@

We tidy up the data a bit; columns should be variables, not replicates.   
<<moretidy>>=
require(reshape2)
Asim <- melt(Asim, id="time")
Bsim <- melt(Bsim, id="time")
names(Asim)[2] <- "rep"
names(Bsim)[2] <- "rep"
@

Now that we have replicates from each process, we can apply the windowed autocorrelation function to each replicate
<<plyrstuff>>=
require(plyr)
wsA <- ddply(Asim, "rep", warningtrend, window_autocorr)
wsB <- ddply(Bsim, "rep", warningtrend, window_autocorr)
@

And gather and plot the distributions,
\begin{figure}[H]
<<tidybean>>=
tidy <- melt(data.frame(null=wsA$tau, test=wsB$tau))
names(tidy) <- c("simulation", "value")
beanplot(value ~ simulation, tidy, what=c(0,1,0,0))
abline(h=observed, lty=2)
@
\end{figure}

and corrsponding ROC plot,
\begin{figure}[H]
<<anotherroc>>=
roc <- roc_data(tidy)
ggplot(roc) + geom_line(aes(False.positives, True.positives), lwd=1)
@
\end{figure}

%\section{Package Installation}

\section*{ }%bibliography
\bibliographystyle{elsarticle-harv}
%\bibliography{/home/cboettig/Documents/Mendeley/bib/library}
%\bibliography{../boettiger}

\end{document}

### 2012-02-16 boettiger.tex
adding knitted appendices

@@
 restrict ourselves to systems that can produce a saddle-node bifurcation,
- approaches based on critical slowing down or rising variance (REFS) must further assume that a changing parameter
-has brought the system closer to the bifurcation.
+approaches based on critical slowing down or rising variance~\citep[\emph{e.g.}][]{Held2004, Scheffer2009, Carpenter2011}
+must further assume that a changing parameter has brought the system closer to the bifurcation.
 This assumption excludes at least three alternative explanations for the transition.
@@
 much can be accomplished by framing the implicit assumptions into equations.
-\citet{Lade2011} introduces the idea of generalized models for early warning signals, and
+\citet{Lade2012} introduces the idea of generalized models for early warning signals, and
 \citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
@@
 
-The challenge of determining early warning signs for impending possible regime shifts requires real attention to the underlying statistical issues and other assumptions.
-Doing this, does, however, open up new possibilities for asking what the goal of detection should be, and for clearly identifying underlying assumptions.
+The challenge of determining early warning signs for impending possible regime shifts 
+requires real attention to the underlying statistical issues and other assumptions.  
+Doing this, does, however, open up new possibilities for asking what the goal of detection should be, 
+and for clearly identifying underlying assumptions. 
 We consider alternative approaches based either on summary statistics or a likelihood based model choice.  
@@
 
-Asa way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms, we introduce an approach based on receiver operator curves.
-These curves XXX.
-The extent of the difficulty in finding reliable indicators of impending regime shifts based on summary statistics.
-becomes clear from the ROC curves of these statistics,
+ % past tense or not?
+We have used three examples to illustrate the performance of this approach in data from 
+simulation, a chemostat experiment, and paleo-atmospheric record; examples differing in 
+sampling intensity and strength of signal of an approaching collapse.  
+While the well-sampled geological data shows an unmistakable signal in this model-based approach,
+the uncertainty in the smaller simulated and experimental data forces a trade-off between errors.  
+%Because such uncertainty may be more typical of early warning signal detection 
+%than the monotonically increasing patterns often used to illustrate the approach,
+
+As a way to clearly illustrate the choices involved in looking for warning signals while avoiding false alarms,
+we introduce an approach based on receiver operator curves. 
+These curves illustrate the extent to which an potential warning signal mitigates the trade-off 
+between missed events and false alarms.  
+The extent of the difficulty in finding reliable indicators of impending regime shifts
+based on summary statistics becomes clear from the ROC curves of these statistics,
 where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
-performing no better than the flip of a coin. %%%% EEEK are you really going to say that?!?!
+performing no better than the flip of a coin. %%%% EEEK can we really say it that strongly?
 By estimating the ROC curve for a given set of data,
@@
 we can develop approaches that get the most information from the data available.
-
-DISUCSS EXAMPLES HERE
 
@@
 demonstrates the accuracy under violation of these assumptions.
+% this robustness to assumptions can also be seen in the performance of the empirical systems
 

### 2012-02-16 appendix/appendices.tex
updating appendices examples

%\VignetteIndexEntry{PMC Tutorial}
\documentclass{elsarticle}
\usepackage{graphicx, color}
\newcommand{\hlnumber}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlfunctioncall}[1]{\textcolor[rgb]{.5,0,.33}{\textbf{#1}}}%
\newcommand{\hlstring}[1]{\textcolor[rgb]{.6,.6,1}{#1}}%
\newcommand{\hlkeyword}[1]{\textbf{#1}}%
\newcommand{\hlargument}[1]{\textcolor[rgb]{.69,.25,.02}{#1}}%
\newcommand{\hlcomment}[1]{\textcolor[rgb]{.18,.6,.34}{#1}}%
\newcommand{\hlroxygencomment}[1]{\textcolor[rgb]{.44,.48,.7}{#1}}%
\newcommand{\hlformalargs}[1]{\hlargument{#1}}%
\newcommand{\hleqformalargs}[1]{\hlargument{#1}}%
\newcommand{\hlassignement}[1]{\textbf{#1}}%
\newcommand{\hlpackage}[1]{\textcolor[rgb]{.59,.71,.145}{#1}}%
\newcommand{\hlslot}[1]{\textit{#1}}%
\newcommand{\hlsymbol}[1]{#1}%
\newcommand{\hlprompt}[1]{\textcolor[rgb]{.5,.5,.5}{#1}}%

\usepackage{color}%

\newsavebox{\hlnormalsizeboxclosebrace}%
\newsavebox{\hlnormalsizeboxopenbrace}%
\newsavebox{\hlnormalsizeboxbackslash}%
\newsavebox{\hlnormalsizeboxlessthan}%
\newsavebox{\hlnormalsizeboxgreaterthan}%
\newsavebox{\hlnormalsizeboxdollar}%
\newsavebox{\hlnormalsizeboxunderscore}%
\newsavebox{\hlnormalsizeboxand}%
\newsavebox{\hlnormalsizeboxhash}%
\newsavebox{\hlnormalsizeboxat}%
\newsavebox{\hlnormalsizeboxpercent}% 
\newsavebox{\hlnormalsizeboxhat}%
\newsavebox{\hlnormalsizeboxsinglequote}%
\newsavebox{\hlnormalsizeboxbacktick}%

\setbox\hlnormalsizeboxopenbrace=\hbox{\begin{normalsize}\verb.{.\end{normalsize}}%
\setbox\hlnormalsizeboxclosebrace=\hbox{\begin{normalsize}\verb.}.\end{normalsize}}%
\setbox\hlnormalsizeboxlessthan=\hbox{\begin{normalsize}\verb.<.\end{normalsize}}%
\setbox\hlnormalsizeboxdollar=\hbox{\begin{normalsize}\verb.$.\end{normalsize}}%
\setbox\hlnormalsizeboxunderscore=\hbox{\begin{normalsize}\verb._.\end{normalsize}}%
\setbox\hlnormalsizeboxand=\hbox{\begin{normalsize}\verb.&.\end{normalsize}}%
\setbox\hlnormalsizeboxhash=\hbox{\begin{normalsize}\verb.#.\end{normalsize}}%
\setbox\hlnormalsizeboxat=\hbox{\begin{normalsize}\verb.@.\end{normalsize}}%
\setbox\hlnormalsizeboxbackslash=\hbox{\begin{normalsize}\verb.\.\end{normalsize}}%
\setbox\hlnormalsizeboxgreaterthan=\hbox{\begin{normalsize}\verb.>.\end{normalsize}}%
\setbox\hlnormalsizeboxpercent=\hbox{\begin{normalsize}\verb.%.\end{normalsize}}%
\setbox\hlnormalsizeboxhat=\hbox{\begin{normalsize}\verb.^.\end{normalsize}}%
\setbox\hlnormalsizeboxsinglequote=\hbox{\begin{normalsize}\verb.'.\end{normalsize}}%
\setbox\hlnormalsizeboxbacktick=\hbox{\begin{normalsize}\verb.`.\end{normalsize}}%
\setbox\hlnormalsizeboxhat=\hbox{\begin{normalsize}\verb.^.\end{normalsize}}%

\newsavebox{\hltinyboxclosebrace}%
\newsavebox{\hltinyboxopenbrace}%
\newsavebox{\hltinyboxbackslash}%
\newsavebox{\hltinyboxlessthan}%
\newsavebox{\hltinyboxgreaterthan}%
\newsavebox{\hltinyboxdollar}%
\newsavebox{\hltinyboxunderscore}%
\newsavebox{\hltinyboxand}%
\newsavebox{\hltinyboxhash}%
\newsavebox{\hltinyboxat}%
\newsavebox{\hltinyboxpercent}% 
\newsavebox{\hltinyboxhat}%
\newsavebox{\hltinyboxsinglequote}%
\newsavebox{\hltinyboxbacktick}%

\setbox\hltinyboxopenbrace=\hbox{\begin{tiny}\verb.{.\end{tiny}}%
\setbox\hltinyboxclosebrace=\hbox{\begin{tiny}\verb.}.\end{tiny}}%
\setbox\hltinyboxlessthan=\hbox{\begin{tiny}\verb.<.\end{tiny}}%
\setbox\hltinyboxdollar=\hbox{\begin{tiny}\verb.$.\end{tiny}}%
\setbox\hltinyboxunderscore=\hbox{\begin{tiny}\verb._.\end{tiny}}%
\setbox\hltinyboxand=\hbox{\begin{tiny}\verb.&.\end{tiny}}%
\setbox\hltinyboxhash=\hbox{\begin{tiny}\verb.#.\end{tiny}}%
\setbox\hltinyboxat=\hbox{\begin{tiny}\verb.@.\end{tiny}}%
\setbox\hltinyboxbackslash=\hbox{\begin{tiny}\verb.\.\end{tiny}}%
\setbox\hltinyboxgreaterthan=\hbox{\begin{tiny}\verb.>.\end{tiny}}%
\setbox\hltinyboxpercent=\hbox{\begin{tiny}\verb.%.\end{tiny}}%
\setbox\hltinyboxhat=\hbox{\begin{tiny}\verb.^.\end{tiny}}%
\setbox\hltinyboxsinglequote=\hbox{\begin{tiny}\verb.'.\end{tiny}}%
\setbox\hltinyboxbacktick=\hbox{\begin{tiny}\verb.`.\end{tiny}}%
\setbox\hltinyboxhat=\hbox{\begin{tiny}\verb.^.\end{tiny}}%

\newsavebox{\hlscriptsizeboxclosebrace}%
\newsavebox{\hlscriptsizeboxopenbrace}%
\newsavebox{\hlscriptsizeboxbackslash}%
\newsavebox{\hlscriptsizeboxlessthan}%
\newsavebox{\hlscriptsizeboxgreaterthan}%
\newsavebox{\hlscriptsizeboxdollar}%
\newsavebox{\hlscriptsizeboxunderscore}%
\newsavebox{\hlscriptsizeboxand}%
\newsavebox{\hlscriptsizeboxhash}%
\newsavebox{\hlscriptsizeboxat}%
\newsavebox{\hlscriptsizeboxpercent}% 
\newsavebox{\hlscriptsizeboxhat}%
\newsavebox{\hlscriptsizeboxsinglequote}%
\newsavebox{\hlscriptsizeboxbacktick}%

\setbox\hlscriptsizeboxopenbrace=\hbox{\begin{scriptsize}\verb.{.\end{scriptsize}}%
\setbox\hlscriptsizeboxclosebrace=\hbox{\begin{scriptsize}\verb.}.\end{scriptsize}}%
\setbox\hlscriptsizeboxlessthan=\hbox{\begin{scriptsize}\verb.<.\end{scriptsize}}%
\setbox\hlscriptsizeboxdollar=\hbox{\begin{scriptsize}\verb.$.\end{scriptsize}}%
\setbox\hlscriptsizeboxunderscore=\hbox{\begin{scriptsize}\verb._.\end{scriptsize}}%
\setbox\hlscriptsizeboxand=\hbox{\begin{scriptsize}\verb.&.\end{scriptsize}}%
\setbox\hlscriptsizeboxhash=\hbox{\begin{scriptsize}\verb.#.\end{scriptsize}}%
\setbox\hlscriptsizeboxat=\hbox{\begin{scriptsize}\verb.@.\end{scriptsize}}%
\setbox\hlscriptsizeboxbackslash=\hbox{\begin{scriptsize}\verb.\.\end{scriptsize}}%
\setbox\hlscriptsizeboxgreaterthan=\hbox{\begin{scriptsize}\verb.>.\end{scriptsize}}%
\setbox\hlscriptsizeboxpercent=\hbox{\begin{scriptsize}\verb.%.\end{scriptsize}}%
\setbox\hlscriptsizeboxhat=\hbox{\begin{scriptsize}\verb.^.\end{scriptsize}}%
\setbox\hlscriptsizeboxsinglequote=\hbox{\begin{scriptsize}\verb.'.\end{scriptsize}}%
\setbox\hlscriptsizeboxbacktick=\hbox{\begin{scriptsize}\verb.`.\end{scriptsize}}%
\setbox\hlscriptsizeboxhat=\hbox{\begin{scriptsize}\verb.^.\end{scriptsize}}%

\newsavebox{\hlfootnotesizeboxclosebrace}%
\newsavebox{\hlfootnotesizeboxopenbrace}%
\newsavebox{\hlfootnotesizeboxbackslash}%
\newsavebox{\hlfootnotesizeboxlessthan}%
\newsavebox{\hlfootnotesizeboxgreaterthan}%
\newsavebox{\hlfootnotesizeboxdollar}%
\newsavebox{\hlfootnotesizeboxunderscore}%
\newsavebox{\hlfootnotesizeboxand}%
\newsavebox{\hlfootnotesizeboxhash}%
\newsavebox{\hlfootnotesizeboxat}%
\newsavebox{\hlfootnotesizeboxpercent}% 
\newsavebox{\hlfootnotesizeboxhat}%
\newsavebox{\hlfootnotesizeboxsinglequote}%
\newsavebox{\hlfootnotesizeboxbacktick}%

\setbox\hlfootnotesizeboxopenbrace=\hbox{\begin{footnotesize}\verb.{.\end{footnotesize}}%
\setbox\hlfootnotesizeboxclosebrace=\hbox{\begin{footnotesize}\verb.}.\end{footnotesize}}%
\setbox\hlfootnotesizeboxlessthan=\hbox{\begin{footnotesize}\verb.<.\end{footnotesize}}%
\setbox\hlfootnotesizeboxdollar=\hbox{\begin{footnotesize}\verb.$.\end{footnotesize}}%
\setbox\hlfootnotesizeboxunderscore=\hbox{\begin{footnotesize}\verb._.\end{footnotesize}}%
\setbox\hlfootnotesizeboxand=\hbox{\begin{footnotesize}\verb.&.\end{footnotesize}}%
\setbox\hlfootnotesizeboxhash=\hbox{\begin{footnotesize}\verb.#.\end{footnotesize}}%
\setbox\hlfootnotesizeboxat=\hbox{\begin{footnotesize}\verb.@.\end{footnotesize}}%
\setbox\hlfootnotesizeboxbackslash=\hbox{\begin{footnotesize}\verb.\.\end{footnotesize}}%
\setbox\hlfootnotesizeboxgreaterthan=\hbox{\begin{footnotesize}\verb.>.\end{footnotesize}}%
\setbox\hlfootnotesizeboxpercent=\hbox{\begin{footnotesize}\verb.%.\end{footnotesize}}%
\setbox\hlfootnotesizeboxhat=\hbox{\begin{footnotesize}\verb.^.\end{footnotesize}}%
\setbox\hlfootnotesizeboxsinglequote=\hbox{\begin{footnotesize}\verb.'.\end{footnotesize}}%
\setbox\hlfootnotesizeboxbacktick=\hbox{\begin{footnotesize}\verb.`.\end{footnotesize}}%
\setbox\hlfootnotesizeboxhat=\hbox{\begin{footnotesize}\verb.^.\end{footnotesize}}%

\newsavebox{\hlsmallboxclosebrace}%
\newsavebox{\hlsmallboxopenbrace}%
\newsavebox{\hlsmallboxbackslash}%
\newsavebox{\hlsmallboxlessthan}%
\newsavebox{\hlsmallboxgreaterthan}%
\newsavebox{\hlsmallboxdollar}%
\newsavebox{\hlsmallboxunderscore}%
\newsavebox{\hlsmallboxand}%
\newsavebox{\hlsmallboxhash}%
\newsavebox{\hlsmallboxat}%
\newsavebox{\hlsmallboxpercent}% 
\newsavebox{\hlsmallboxhat}%
\newsavebox{\hlsmallboxsinglequote}%
\newsavebox{\hlsmallboxbacktick}%

\setbox\hlsmallboxopenbrace=\hbox{\begin{small}\verb.{.\end{small}}%
\setbox\hlsmallboxclosebrace=\hbox{\begin{small}\verb.}.\end{small}}%
\setbox\hlsmallboxlessthan=\hbox{\begin{small}\verb.<.\end{small}}%
\setbox\hlsmallboxdollar=\hbox{\begin{small}\verb.$.\end{small}}%
\setbox\hlsmallboxunderscore=\hbox{\begin{small}\verb._.\end{small}}%
\setbox\hlsmallboxand=\hbox{\begin{small}\verb.&.\end{small}}%
\setbox\hlsmallboxhash=\hbox{\begin{small}\verb.#.\end{small}}%
\setbox\hlsmallboxat=\hbox{\begin{small}\verb.@.\end{small}}%
\setbox\hlsmallboxbackslash=\hbox{\begin{small}\verb.\.\end{small}}%
\setbox\hlsmallboxgreaterthan=\hbox{\begin{small}\verb.>.\end{small}}%
\setbox\hlsmallboxpercent=\hbox{\begin{small}\verb.%.\end{small}}%
\setbox\hlsmallboxhat=\hbox{\begin{small}\verb.^.\end{small}}%
\setbox\hlsmallboxsinglequote=\hbox{\begin{small}\verb.'.\end{small}}%
\setbox\hlsmallboxbacktick=\hbox{\begin{small}\verb.`.\end{small}}%
\setbox\hlsmallboxhat=\hbox{\begin{small}\verb.^.\end{small}}%

\newsavebox{\hllargeboxclosebrace}%
\newsavebox{\hllargeboxopenbrace}%
\newsavebox{\hllargeboxbackslash}%
\newsavebox{\hllargeboxlessthan}%
\newsavebox{\hllargeboxgreaterthan}%
\newsavebox{\hllargeboxdollar}%
\newsavebox{\hllargeboxunderscore}%
\newsavebox{\hllargeboxand}%
\newsavebox{\hllargeboxhash}%
\newsavebox{\hllargeboxat}%
\newsavebox{\hllargeboxpercent}% 
\newsavebox{\hllargeboxhat}%
\newsavebox{\hllargeboxsinglequote}%
\newsavebox{\hllargeboxbacktick}%

\setbox\hllargeboxopenbrace=\hbox{\begin{large}\verb.{.\end{large}}%
\setbox\hllargeboxclosebrace=\hbox{\begin{large}\verb.}.\end{large}}%
\setbox\hllargeboxlessthan=\hbox{\begin{large}\verb.<.\end{large}}%
\setbox\hllargeboxdollar=\hbox{\begin{large}\verb.$.\end{large}}%
\setbox\hllargeboxunderscore=\hbox{\begin{large}\verb._.\end{large}}%
\setbox\hllargeboxand=\hbox{\begin{large}\verb.&.\end{large}}%
\setbox\hllargeboxhash=\hbox{\begin{large}\verb.#.\end{large}}%
\setbox\hllargeboxat=\hbox{\begin{large}\verb.@.\end{large}}%
\setbox\hllargeboxbackslash=\hbox{\begin{large}\verb.\.\end{large}}%
\setbox\hllargeboxgreaterthan=\hbox{\begin{large}\verb.>.\end{large}}%
\setbox\hllargeboxpercent=\hbox{\begin{large}\verb.%.\end{large}}%
\setbox\hllargeboxhat=\hbox{\begin{large}\verb.^.\end{large}}%
\setbox\hllargeboxsinglequote=\hbox{\begin{large}\verb.'.\end{large}}%
\setbox\hllargeboxbacktick=\hbox{\begin{large}\verb.`.\end{large}}%
\setbox\hllargeboxhat=\hbox{\begin{large}\verb.^.\end{large}}%

\newsavebox{\hlLargeboxclosebrace}%
\newsavebox{\hlLargeboxopenbrace}%
\newsavebox{\hlLargeboxbackslash}%
\newsavebox{\hlLargeboxlessthan}%
\newsavebox{\hlLargeboxgreaterthan}%
\newsavebox{\hlLargeboxdollar}%
\newsavebox{\hlLargeboxunderscore}%
\newsavebox{\hlLargeboxand}%
\newsavebox{\hlLargeboxhash}%
\newsavebox{\hlLargeboxat}%
\newsavebox{\hlLargeboxpercent}% 
\newsavebox{\hlLargeboxhat}%
\newsavebox{\hlLargeboxsinglequote}%
\newsavebox{\hlLargeboxbacktick}%

\setbox\hlLargeboxopenbrace=\hbox{\begin{Large}\verb.{.\end{Large}}%
\setbox\hlLargeboxclosebrace=\hbox{\begin{Large}\verb.}.\end{Large}}%
\setbox\hlLargeboxlessthan=\hbox{\begin{Large}\verb.<.\end{Large}}%
\setbox\hlLargeboxdollar=\hbox{\begin{Large}\verb.$.\end{Large}}%
\setbox\hlLargeboxunderscore=\hbox{\begin{Large}\verb._.\end{Large}}%
\setbox\hlLargeboxand=\hbox{\begin{Large}\verb.&.\end{Large}}%
\setbox\hlLargeboxhash=\hbox{\begin{Large}\verb.#.\end{Large}}%
\setbox\hlLargeboxat=\hbox{\begin{Large}\verb.@.\end{Large}}%
\setbox\hlLargeboxbackslash=\hbox{\begin{Large}\verb.\.\end{Large}}%
\setbox\hlLargeboxgreaterthan=\hbox{\begin{Large}\verb.>.\end{Large}}%
\setbox\hlLargeboxpercent=\hbox{\begin{Large}\verb.%.\end{Large}}%
\setbox\hlLargeboxhat=\hbox{\begin{Large}\verb.^.\end{Large}}%
\setbox\hlLargeboxsinglequote=\hbox{\begin{Large}\verb.'.\end{Large}}%
\setbox\hlLargeboxbacktick=\hbox{\begin{Large}\verb.`.\end{Large}}%
\setbox\hlLargeboxhat=\hbox{\begin{Large}\verb.^.\end{Large}}%

\newsavebox{\hlLARGEboxclosebrace}%
\newsavebox{\hlLARGEboxopenbrace}%
\newsavebox{\hlLARGEboxbackslash}%
\newsavebox{\hlLARGEboxlessthan}%
\newsavebox{\hlLARGEboxgreaterthan}%
\newsavebox{\hlLARGEboxdollar}%
\newsavebox{\hlLARGEboxunderscore}%
\newsavebox{\hlLARGEboxand}%
\newsavebox{\hlLARGEboxhash}%
\newsavebox{\hlLARGEboxat}%
\newsavebox{\hlLARGEboxpercent}% 
\newsavebox{\hlLARGEboxhat}%
\newsavebox{\hlLARGEboxsinglequote}%
\newsavebox{\hlLARGEboxbacktick}%

\setbox\hlLARGEboxopenbrace=\hbox{\begin{LARGE}\verb.{.\end{LARGE}}%
\setbox\hlLARGEboxclosebrace=\hbox{\begin{LARGE}\verb.}.\end{LARGE}}%
\setbox\hlLARGEboxlessthan=\hbox{\begin{LARGE}\verb.<.\end{LARGE}}%
\setbox\hlLARGEboxdollar=\hbox{\begin{LARGE}\verb.$.\end{LARGE}}%
\setbox\hlLARGEboxunderscore=\hbox{\begin{LARGE}\verb._.\end{LARGE}}%
\setbox\hlLARGEboxand=\hbox{\begin{LARGE}\verb.&.\end{LARGE}}%
\setbox\hlLARGEboxhash=\hbox{\begin{LARGE}\verb.#.\end{LARGE}}%
\setbox\hlLARGEboxat=\hbox{\begin{LARGE}\verb.@.\end{LARGE}}%
\setbox\hlLARGEboxbackslash=\hbox{\begin{LARGE}\verb.\.\end{LARGE}}%
\setbox\hlLARGEboxgreaterthan=\hbox{\begin{LARGE}\verb.>.\end{LARGE}}%
\setbox\hlLARGEboxpercent=\hbox{\begin{LARGE}\verb.%.\end{LARGE}}%
\setbox\hlLARGEboxhat=\hbox{\begin{LARGE}\verb.^.\end{LARGE}}%
\setbox\hlLARGEboxsinglequote=\hbox{\begin{LARGE}\verb.'.\end{LARGE}}%
\setbox\hlLARGEboxbacktick=\hbox{\begin{LARGE}\verb.`.\end{LARGE}}%
\setbox\hlLARGEboxhat=\hbox{\begin{LARGE}\verb.^.\end{LARGE}}%

\newsavebox{\hlhugeboxclosebrace}%
\newsavebox{\hlhugeboxopenbrace}%
\newsavebox{\hlhugeboxbackslash}%
\newsavebox{\hlhugeboxlessthan}%
\newsavebox{\hlhugeboxgreaterthan}%
\newsavebox{\hlhugeboxdollar}%
\newsavebox{\hlhugeboxunderscore}%
\newsavebox{\hlhugeboxand}%
\newsavebox{\hlhugeboxhash}%
\newsavebox{\hlhugeboxat}%
\newsavebox{\hlhugeboxpercent}% 
\newsavebox{\hlhugeboxhat}%
\newsavebox{\hlhugeboxsinglequote}%
\newsavebox{\hlhugeboxbacktick}%

\setbox\hlhugeboxopenbrace=\hbox{\begin{huge}\verb.{.\end{huge}}%
\setbox\hlhugeboxclosebrace=\hbox{\begin{huge}\verb.}.\end{huge}}%
\setbox\hlhugeboxlessthan=\hbox{\begin{huge}\verb.<.\end{huge}}%
\setbox\hlhugeboxdollar=\hbox{\begin{huge}\verb.$.\end{huge}}%
\setbox\hlhugeboxunderscore=\hbox{\begin{huge}\verb._.\end{huge}}%
\setbox\hlhugeboxand=\hbox{\begin{huge}\verb.&.\end{huge}}%
\setbox\hlhugeboxhash=\hbox{\begin{huge}\verb.#.\end{huge}}%
\setbox\hlhugeboxat=\hbox{\begin{huge}\verb.@.\end{huge}}%
\setbox\hlhugeboxbackslash=\hbox{\begin{huge}\verb.\.\end{huge}}%
\setbox\hlhugeboxgreaterthan=\hbox{\begin{huge}\verb.>.\end{huge}}%
\setbox\hlhugeboxpercent=\hbox{\begin{huge}\verb.%.\end{huge}}%
\setbox\hlhugeboxhat=\hbox{\begin{huge}\verb.^.\end{huge}}%
\setbox\hlhugeboxsinglequote=\hbox{\begin{huge}\verb.'.\end{huge}}%
\setbox\hlhugeboxbacktick=\hbox{\begin{huge}\verb.`.\end{huge}}%
\setbox\hlhugeboxhat=\hbox{\begin{huge}\verb.^.\end{huge}}%

\newsavebox{\hlHugeboxclosebrace}%
\newsavebox{\hlHugeboxopenbrace}%
\newsavebox{\hlHugeboxbackslash}%
\newsavebox{\hlHugeboxlessthan}%
\newsavebox{\hlHugeboxgreaterthan}%
\newsavebox{\hlHugeboxdollar}%
\newsavebox{\hlHugeboxunderscore}%
\newsavebox{\hlHugeboxand}%
\newsavebox{\hlHugeboxhash}%
\newsavebox{\hlHugeboxat}%
\newsavebox{\hlHugeboxpercent}% 
\newsavebox{\hlHugeboxhat}%
\newsavebox{\hlHugeboxsinglequote}%
\newsavebox{\hlHugeboxbacktick}%

\setbox\hlHugeboxopenbrace=\hbox{\begin{Huge}\verb.{.\end{Huge}}%
\setbox\hlHugeboxclosebrace=\hbox{\begin{Huge}\verb.}.\end{Huge}}%
\setbox\hlHugeboxlessthan=\hbox{\begin{Huge}\verb.<.\end{Huge}}%
\setbox\hlHugeboxdollar=\hbox{\begin{Huge}\verb.$.\end{Huge}}%
\setbox\hlHugeboxunderscore=\hbox{\begin{Huge}\verb._.\end{Huge}}%
\setbox\hlHugeboxand=\hbox{\begin{Huge}\verb.&.\end{Huge}}%
\setbox\hlHugeboxhash=\hbox{\begin{Huge}\verb.#.\end{Huge}}%
\setbox\hlHugeboxat=\hbox{\begin{Huge}\verb.@.\end{Huge}}%
\setbox\hlHugeboxbackslash=\hbox{\begin{Huge}\verb.\.\end{Huge}}%
\setbox\hlHugeboxgreaterthan=\hbox{\begin{Huge}\verb.>.\end{Huge}}%
\setbox\hlHugeboxpercent=\hbox{\begin{Huge}\verb.%.\end{Huge}}%
\setbox\hlHugeboxhat=\hbox{\begin{Huge}\verb.^.\end{Huge}}%
\setbox\hlHugeboxsinglequote=\hbox{\begin{Huge}\verb.'.\end{Huge}}%
\setbox\hlHugeboxbacktick=\hbox{\begin{Huge}\verb.`.\end{Huge}}%
\setbox\hlHugeboxhat=\hbox{\begin{Huge}\verb.^.\end{Huge}}%

\def\urltilda{\kern -.15em\lower .7ex\hbox{\~{}}\kern .04em}%

\newcommand{\hlstd}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.16,0.16,1}{#1}}
\newcommand{\hlesc}[1]{\textcolor[rgb]{1,0,1}{#1}}
\newcommand{\hlstr}[1]{\textcolor[rgb]{1,0,0}{#1}}
\newcommand{\hldstr}[1]{\textcolor[rgb]{0.51,0.51,0}{#1}}
\newcommand{\hlslc}[1]{\textcolor[rgb]{0.51,0.51,0.51}{\it{#1}}}
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.51,0.51,0.51}{\it{#1}}}
\newcommand{\hldir}[1]{\textcolor[rgb]{0,0.51,0}{#1}}
\newcommand{\hlsym}[1]{\textcolor[rgb]{0,0,0}{#1}}
\newcommand{\hlline}[1]{\textcolor[rgb]{0.33,0.33,0.33}{#1}}
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0,0,0}{\bf{#1}}}
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.51,0,0}{#1}}
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0,0,0}{\bf{#1}}}
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0,0,0.51}{#1}}

\definecolor{fgcolor}{rgb}{0,0,0}
\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed}
\makeatother

\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage[pdftex, colorlinks]{hyperref}
\usepackage{amsmath, amsfonts}  % extended mathematics
\usepackage{booktabs} % book-quality tables
\usepackage{float}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}

%% Looks like a comment but it isn't! This is setting the default behavior for the Sweave chunk options, <<options >>= 
%

\begin{document}
\begin{frontmatter}
  \title{Limits to detection of early warning signals: Appendices}
  \author[davis]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  %\author[davis]{}
  \cortext[cor1]{Corresponding author.}
  \address[davis]{Center for Population Biology, University of California, Davis, United States}

  \begin{abstract}
  These appendices explore several concepts that do not get fully developed in the paper, while also providing an introduction to the early warning signals package we provide to support this paper.  This document is written in Sweave, so the code you see is the code that generates the figures and results shown.   
  \end{abstract}
 \end{frontmatter}

\section{Introduction}
\setkeys{Gin}{width=0.85\textwidth}

Early warning signals are most often presented in the literature as graph showing the monotonic increase of some statistic over time.  It is not necessarily obvious how best to quantify this pattern, nor does it seem common practice to do so.  Though these patterns often look pretty montonic, we must be careful not to forget that monotonic increases in a statistic will appear by chance at some rate.  In small data sets, the probabilty of a chance pattern is higher.  These basic statistical concerns seem largely absent from this literature, but if visual patterns speak more strongly than statistics, here's the problem in visual form:

Simulate from a system approaching a collapse~64 times,

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{populationdynamics}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{pars}{\ }\hlassignement{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlargument{Xo}{\ }\hlargument{=}{\ }\hlnumber{730}\hlkeyword{,}{\ }\hlargument{e}{\ }\hlargument{=}{\ }\hlnumber{0.5}\hlkeyword{,}{\ }\hlargument{a}{\ }\hlargument{=}{\ }\hlnumber{100}\hlkeyword{,}{\ }\hlargument{K}{\ }\hlargument{=}{\ }\hlnumber{1000}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{h}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{,}{\ }\hlargument{i}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{Da}{\ }\hlargument{=}{\ }\hlnumber{0.09}\hlkeyword{,}{\ }\hlargument{Dt}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{p}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{time}{\ }\hlassignement{=}{\ }\hlfunctioncall{seq}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{900}\hlkeyword{,}{\ }\hlargument{length}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{sn}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{saddle\usebox{\hlnormalsizeboxunderscore}node\usebox{\hlnormalsizeboxunderscore}ibm}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{reps}{\ }\hlargument{=}{\ }\hlsymbol{nreps}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

We do a little reformatting of the data, 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{X}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{time}{\ }\hlargument{=}{\ }\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{value}{\ }\hlargument{=}{\ }\hlsymbol{sn}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{x1}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{dat}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{X}\hlkeyword{,}{\ }\hlargument{id}{\ }\hlargument{=}{\ }\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{dat}\hlkeyword{)}\hlkeyword{[}\hlnumber{2}\hlkeyword{]}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlstring{"{}reps"{}}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

and compute the autocorrelation over a window in time equal half the length if the timeseries
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{acorr}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{summary\usebox{\hlnormalsizeboxunderscore}statistic}\hlkeyword{(}\hlsymbol{dat}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{windowsize}{\ }\hlargument{=}{\ }\hlfunctioncall{length}\hlkeyword{(}\hlsymbol{time}\hlkeyword{)}\hlkeyword{/}\hlnumber{2}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

\begin{figure}[H]
\begin{center}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{acorr}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{time}\hlkeyword{,}{\ }\hlsymbol{value}\hlkeyword{)}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{reps}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{opts}\hlkeyword{(}\hlargument{title}{\ }\hlargument{=}{\ }\hlstring{"{}Autocorrelation{\ }on{\ }replicates{\ }from{\ }a{\ }system{\ }approaching{\ }a{\ }crash"{}}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{crashacor}

\end{kframe}}
\end{knitrout}

\caption{The pattern of autocorrelation in replicate simulations from a simulation of a system which is approaching a saddle node bifurcation.}\label{crashacor}
\end{center}
\end{figure}

While we can see an average increasing trend, idenifying a particular squiggle as an early warning sign looks a lot more like judgement call then science.  Reality is unlikely to be any less murky, so the purpose this paper is to provide statistical guidance about how to start approaching this uncertainty.

We provide the code in this document to both illustrate the steps required and also that the commands can be copy-pasted into R using different choices for the warning indicator, the sampling of the simulation, etc.

\section{Model-based warning signals}

We load the package and the data set from the individual-based simulations.  To this data we fit both the LSN and OU models, 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{earlywarning}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{data}\hlkeyword{(}\hlsymbol{ibms}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{A}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{stability\usebox{\hlnormalsizeboxunderscore}model}\hlkeyword{(}\hlsymbol{ibm\usebox{\hlnormalsizeboxunderscore}critical}\hlkeyword{,}{\ }\hlstring{"{}OU"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{B}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{stability\usebox{\hlnormalsizeboxunderscore}model}\hlkeyword{(}\hlsymbol{ibm\usebox{\hlnormalsizeboxunderscore}critical}\hlkeyword{,}{\ }\hlstring{"{}LSN"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{observed}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlkeyword{-}\hlnumber{2}{\ }\hlkeyword{*}{\ }\hlkeyword{(}\hlfunctioncall{logLik}\hlkeyword{(}\hlsymbol{A}\hlkeyword{)}{\ }\hlkeyword{-}{\ }\hlfunctioncall{logLik}\hlkeyword{(}\hlsymbol{B}\hlkeyword{)}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

Comparing the models by the bootstrapping procedure is intensive.  `compare` simulates under each model, and then estimates both models on both simulated sets of data (for a total of 4 model fits).  As only the LSN fit is computationally intensive, this takes about twice the time as the above fitting step of model B.  We will need to repeat this many times to build up a bootstrap distribution. Consequently, it is convient to run this step in parallel over multiple processors or clusters of computers.

\subsection{Parallelization of the comparison}
Fortunately R has a diverse array of parallel tools.  Here we set up the flexible \texttt{snowfall} parallel envirnoment, which can run in serial, on a multicore chip, or over a cluster of machines.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{snowfall}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{sfInit}\hlkeyword{(}\hlargument{par}{\ }\hlargument{=}{\ }\hlsymbol{T}\hlkeyword{,}{\ }\hlargument{cpu}{\ }\hlargument{=}{\ }\hlnumber{16}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{sfLibrary}\hlkeyword{(}\hlsymbol{earlywarning}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{sfExportAll}\hlkeyword{(}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

Rather than build this choice into the earlywarning signals package, the user can choose whatever looping implementation they desire to compute replicates of the comparison.  The actual analysis takes just one line:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{reps}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{sfLapply}\hlkeyword{(}\hlnumber{1}\hlkeyword{:}\hlnumber{500}\hlkeyword{,}{\ }\hlkeyword{function}\hlkeyword{(}\hlformalargs{i}\hlkeyword{)}{\ }\hlfunctioncall{compare}\hlkeyword{(}\hlsymbol{A}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlsymbol{B}\hlkeyword{)}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

Our results are cached in the package, so anyone seeking to explore this example can run
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{data}\hlkeyword{(}\hlsymbol{manuscript\usebox{\hlnormalsizeboxunderscore}data}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

\subsection{Visualizing the distributions}

Once this step is finished, the remaining analysis is just a matter of shuffling and plotting the data.   We extract the likelihood ratios (deviances) from the null and test simulations with a simple function.  
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{lr}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{lik\usebox{\hlnormalsizeboxunderscore}ratios}\hlkeyword{(}\hlsymbol{reps}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

We can visualize this data in several ways: as two overlapping distributions,

\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{ggplot2}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{lr}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}density}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{value}\hlkeyword{,}{\ }\hlargument{fill}{\ }\hlargument{=}{\ }\hlsymbol{simulation}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{alpha}{\ }\hlargument{=}{\ }\hlnumber{0.7}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}vline}\hlkeyword{(}\hlargument{xintercept}{\ }\hlargument{=}{\ }\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{dists}

\end{kframe}}
\end{knitrout}

\end{figure}

or as beanplot, a modification of a boxplot showing the continuous distributions side-by-side:

\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{beanplot}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{beanplot}\hlkeyword{(}\hlsymbol{value}{\ }\hlkeyword{\urltilda{}}{\ }\hlsymbol{simulation}\hlkeyword{,}{\ }\hlsymbol{lr}\hlkeyword{,}{\ }\hlargument{what}{\ }\hlargument{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{1}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{0}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{abline}\hlkeyword{(}\hlargument{h}{\ }\hlargument{=}{\ }\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{beanplot}

\end{kframe}}
\end{knitrout}

\end{figure}

\subsection{Generating the ROC Curve}

We can make the ROC curve by calculating the true and false positive rates from the overlap of these distributions using the function \verb|roc_data|.  This gives us a data frame of threshold values for our statistic, and the corresponding false positive rate and true positive rate.  We can simply plot those rates against each other to create the ROC curve.

\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{roc}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{roc\usebox{\hlnormalsizeboxunderscore}data}\hlkeyword{(}\hlsymbol{lr}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{roc}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{False.positives}\hlkeyword{,}{\ }\hlsymbol{True.positives}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{lwd}{\ }\hlargument{=}{\ }\hlnumber{1}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{rocplot}

\end{kframe}}
\end{knitrout}

\end{figure}

\subsection{ Parameter distributions}

We can also look at the bootstraps of the parameters.  Another helper function will reformat this data from reps list.  The fit column uses a two-letter code to indicate first what model was used to simulate the data, and then what model was fit to the data.  For instance, AB means the data was simulated from model A (the null, OU model) but fit to B.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{pars}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{parameter\usebox{\hlnormalsizeboxunderscore}bootstraps}\hlkeyword{(}\hlsymbol{reps}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{head}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\begin{verbatim}
##     value parameter fit rep
## 1   2.440        Ro  AA   1
## 2 604.066     theta  AA   1
## 3 146.492     sigma  AA   1
## 4   1.782        Ro  AB   1
## 5 640.402     theta  AB   1
## 6 140.099     sigma  AB   1
\end{verbatim}
\end{kframe}}
\end{knitrout}

There are lots of options for visualizing this relatively high-dimensional data, which we can easily explore with a few commands from the \texttt{ggplot2} package. For instance, we can look at average and range of parameters estimated in the bootstraps of each model:

\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{Hmisc}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlfunctioncall{subset}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{fit}{\ }\hlkeyword{\usebox{\hlnormalsizeboxpercent}in\usebox{\hlnormalsizeboxpercent}}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlstring{"{}AA"{}}\hlkeyword{,}{\ }\hlstring{"{}BB"{}}\hlkeyword{)}\hlkeyword{)}\hlkeyword{,}{\ }\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{fit}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlsymbol{value}\hlkeyword{)}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{stat\usebox{\hlnormalsizeboxunderscore}summary}\hlkeyword{(}\hlargument{fun.y}{\ }\hlargument{=}{\ }\hlsymbol{mean}\hlkeyword{,}{\ }\hlargument{geom}{\ }\hlargument{=}{\ }\hlstring{"{}bar"{}}\hlkeyword{,}{\ }\hlargument{position}{\ }\hlargument{=}{\ }\hlstring{"{}dodge"{}}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlfunctioncall{stat\usebox{\hlnormalsizeboxunderscore}summary}\hlkeyword{(}\hlargument{fun.data}{\ }\hlargument{=}{\ }\hlsymbol{median\usebox{\hlnormalsizeboxunderscore}hilow}\hlkeyword{,}{\ }\hlargument{geom}{\ }\hlargument{=}{\ }\hlstring{"{}pointrange"{}}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }\hlargument{position}{\ }\hlargument{=}{\ }\hlfunctioncall{position\usebox{\hlnormalsizeboxunderscore}dodge}\hlkeyword{(}\hlargument{width}{\ }\hlargument{=}{\ }\hlnumber{0.9}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{conf.int}{\ }\hlargument{=}{\ }\hlnumber{0.95}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{parameter}\hlkeyword{,}{\ }\hlargument{scales}{\ }\hlargument{=}{\ }\hlstring{"{}free\usebox{\hlnormalsizeboxunderscore}y"{}}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{parplot}

\end{kframe}}
\end{knitrout}

\end{figure}

\subsection{Visualizing the simulation replicates}

The \texttt{reps} list also contains the simulation data from each replicate

\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{sims}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{get\usebox{\hlnormalsizeboxunderscore}replicates}\hlkeyword{(}\hlsymbol{reps}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\begin{verbatim}
## Error: argument is of length zero
\end{verbatim}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{sims}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{time}\hlkeyword{,}{\ }\hlsymbol{value}\hlkeyword{,}{\ }\hlargument{group}{\ }\hlargument{=}{\ }\hlsymbol{rep}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{alpha}{\ }\hlargument{=}{\ }\hlnumber{0.1}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{model}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\begin{verbatim}
## Error: object 'sims' not found
\end{verbatim}
\end{kframe}}
\end{knitrout}

\end{figure}

\section{Computing summary statistics with improved sampling}

Autocorrelation performs particularly poorly on the examples in the text, falling on approximately the 1:1 line in the ROC curves.  This example goes through the steps to demonstrate that in a sufficiently-frequently sampled timeseries, autocorrelation does contain *some* signal of early warning, it is simply a weaker signal than the corresponding analysis with variance, which in turn is weaker than the model-based approach. This example also illustrates the use of some of the summary statistics tools.

We begin by running the individual based simulation, supplied in our helper package, populationdynamics, with~1500 sample points, instead of the 40 shown in the text.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{populationdynamics}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{pars}{\ }\hlassignement{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlargument{Xo}{\ }\hlargument{=}{\ }\hlnumber{730}\hlkeyword{,}{\ }\hlargument{e}{\ }\hlargument{=}{\ }\hlnumber{0.5}\hlkeyword{,}{\ }\hlargument{a}{\ }\hlargument{=}{\ }\hlnumber{100}\hlkeyword{,}{\ }\hlargument{K}{\ }\hlargument{=}{\ }\hlnumber{1000}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{h}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{,}{\ }\hlargument{i}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{Da}{\ }\hlargument{=}{\ }\hlnumber{0.09}\hlkeyword{,}{\ }\hlargument{Dt}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{p}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{time}{\ }\hlassignement{=}{\ }\hlfunctioncall{seq}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{500}\hlkeyword{,}{\ }\hlargument{length}{\ }\hlargument{=}{\ }\hlnumber{1500}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{sn}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{saddle\usebox{\hlnormalsizeboxunderscore}node\usebox{\hlnormalsizeboxunderscore}ibm}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{time}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{X}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{time}{\ }\hlargument{=}{\ }\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{value}{\ }\hlargument{=}{\ }\hlsymbol{sn}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{x1}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

We compute the observed value of Kendall's tau for the autocorrelation computed in a moving window over the data.  (The warningtrend function is just a wrapper for the base function cor.test, which handles different window sizes.  The default uses half the length of the timeseries.)

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{observed}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{warningtrend}\hlkeyword{(}\hlsymbol{X}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{observed}\mbox{}
\normalfont
\end{flushleft}
\begin{verbatim}
##     tau 
## -0.1633 
\end{verbatim}
\end{kframe}}
\end{knitrout}

While this gives us single value which is useful for statistical comparisons that follow, recall that it is more common to simply plot the autocorrelation computed in this manner, and the increase is just the visual pattern.

\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{npts}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{length}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}value"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{autocorrelation}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}value"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{windowsize}{\ }\hlargument{=}{\ }\hlsymbol{npts}\hlkeyword{/}\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{autocorrelation}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlfunctioncall{rep}\hlkeyword{(}\hlnumber{NA}\hlkeyword{,}{\ }\hlsymbol{npts}\hlkeyword{/}\hlnumber{2}{\ }\hlkeyword{-}{\ }\hlnumber{1}\hlkeyword{)}\hlkeyword{,}{\ }\hlsymbol{autocorrelation}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{plot}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}time"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{,}{\ }\hlsymbol{autocorrelation}\hlkeyword{,}{\ }\hlargument{type}{\ }\hlargument{=}{\ }\hlstring{"{}l"{}}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{plot_autocorr}

\end{kframe}}
\end{knitrout}

\end{figure}

To bootstrap the estimate of tau, we need to be able to simulate under a null and test model.  We use our models of a stable system and a system approaching a collapse to do this.  We first estimate the model parameters from the data,

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{A}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{stability\usebox{\hlnormalsizeboxunderscore}model}\hlkeyword{(}\hlsymbol{X}\hlkeyword{,}{\ }\hlstring{"{}OU"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{B}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{stability\usebox{\hlnormalsizeboxunderscore}model}\hlkeyword{(}\hlsymbol{X}\hlkeyword{,}{\ }\hlstring{"{}LSN"{}}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

and then we can simulate some replicates

\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{reps}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlnumber{500}\hspace*{\fill}\\
\hlstd{}\hlsymbol{Asim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{simulate}\hlkeyword{(}\hlsymbol{A}\hlkeyword{,}{\ }\hlsymbol{reps}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{Bsim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{simulate}\hlkeyword{(}\hlsymbol{B}\hlkeyword{,}{\ }\hlsymbol{reps}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

We tidy up the data a bit; columns should be variables, not replicates.   
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{reshape2}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{Asim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{,}{\ }\hlargument{id}{\ }\hlargument{=}{\ }\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{Bsim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{,}{\ }\hlargument{id}{\ }\hlargument{=}{\ }\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{)}\hlkeyword{[}\hlnumber{2}\hlkeyword{]}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlstring{"{}reps"{}}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{)}\hlkeyword{[}\hlnumber{2}\hlkeyword{]}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlstring{"{}reps"{}}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

Now that we have replicates from each process, we can apply the windowed autocorrelation function to each replicate
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{plyr}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{require}\hlkeyword{(}\hlsymbol{doMC}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{registerDoMC}\hlkeyword{(}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{wsA}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{ddply}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{,}{\ }\hlstring{"{}reps"{}}\hlkeyword{,}{\ }\hlsymbol{warningtrend}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{.parallel}{\ }\hlargument{=}{\ }\hlnumber{TRUE}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlsymbol{wsB}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{ddply}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{,}{\ }\hlstring{"{}reps"{}}\hlkeyword{,}{\ }\hlsymbol{warningtrend}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{.parallel}{\ }\hlargument{=}{\ }\hlnumber{TRUE}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

And gather and plot the distributions,
\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{tidy}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{null}{\ }\hlargument{=}{\ }\hlsymbol{wsA}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{tau}\hlkeyword{,}{\ }\hlargument{test}{\ }\hlargument{=}{\ }\hlsymbol{wsB}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{tau}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{tidy}\hlkeyword{)}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlstring{"{}simulation"{}}\hlkeyword{,}{\ }\hlstring{"{}value"{}}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{beanplot}\hlkeyword{(}\hlsymbol{value}{\ }\hlkeyword{\urltilda{}}{\ }\hlsymbol{simulation}\hlkeyword{,}{\ }\hlsymbol{tidy}\hlkeyword{,}{\ }\hlargument{what}{\ }\hlargument{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlnumber{1}\hlkeyword{,}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{0}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{abline}\hlkeyword{(}\hlargument{h}{\ }\hlargument{=}{\ }\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{tidybean}

\end{kframe}}
\end{knitrout}

\end{figure}

and corrsponding ROC plot,
\begin{figure}[H]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlsymbol{roc}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{roc\usebox{\hlnormalsizeboxunderscore}data}\hlkeyword{(}\hlsymbol{tidy}\hlkeyword{)}\hspace*{\fill}\\
\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{roc}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{False.positives}\hlkeyword{,}{\ }\hlsymbol{True.positives}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
\hlstd{}{\ }{\ }{\ }{\ }\hlargument{lwd}{\ }\hlargument{=}{\ }\hlnumber{1}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}

\centering{}\includegraphics{anotherroc}

\end{kframe}}
\end{knitrout}

\end{figure}

Lastly we save the data from each of these examples, that it may be available to the users later.   
\begin{knitrout}
\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
\begin{flushleft}
\ttfamily\noindent
\hlfunctioncall{save}\hlkeyword{(}\hlargument{list}{\ }\hlargument{=}{\ }\hlfunctioncall{ls}\hlkeyword{(}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{file}{\ }\hlargument{=}{\ }\hlstring{"{}tutorial.rda"{}}\hlkeyword{)}\mbox{}
\normalfont
\end{flushleft}
\end{kframe}}
\end{knitrout}

%\section{Package Installation}

\section*{ }%bibliography
\bibliographystyle{elsarticle-harv}
%\bibliography{/home/cboettig/Documents/Mendeley/bib/library}
%\bibliography{../boettiger}

\end{document}

### 2012-02-16 appendix/appendices_knit_.Rnw
updating appendices examples

@@
 sims <- get_replicates(reps)
-ggplot2(sims) + geom_line(aes(time, value, group=rep), alpha=0.1) + facet_wrap(~model)
+ggplot(sims) + geom_line(aes(time, value, group=rep), alpha=0.1) + facet_wrap(~model)
 @
@@
 This example also illustrates the use of some of the summary statistics tools.
-
-We begin by running the individual based simulation, supplied in our helper package, populationdynamics, with 500 sample points, instead of the 40 shown in the text.
+<<npts, echo=FALSE>>=
+npts <- 1500
+@
+We begin by running the individual based simulation, supplied in our helper package, populationdynamics, with~\Sexpr{npts} sample points, instead of the 40 shown in the text.
 
@@
 pars = c(Xo = 730, e = 0.5, a = 100, K = 1000, h = 200, i = 0, Da = .09, Dt = 0, p = 2)
-time=seq(0, 500, length=500)
+time=seq(0, 500, length=1500)
 sn <- saddle_node_ibm(pars,time)
@@
 autocorrelation <- window_autocorr(X[["value"]], windowsize=npts/2)
-autocorrelation <- c(rep(NA, npts/2), autocorrelation)
-plot(X[[time]], autocorrelation, type="l")
+autocorrelation <- c(rep(NA, npts/2-1), autocorrelation)
+plot(X[["time"]], autocorrelation, type="l")
 @
@@
 <<moresims>>=
-reps <- 100
+reps <- 500
 Asim <- simulate(A, reps)
@@
 Bsim <- melt(Bsim, id="time")
-names(Asim)[2] <- "rep"
-names(Bsim)[2] <- "rep"
+names(Asim)[2] <- "reps"
+names(Bsim)[2] <- "reps"
 @
@@
 require(plyr)
-wsA <- ddply(Asim, "rep", warningtrend, window_autocorr)
-wsB <- ddply(Bsim, "rep", warningtrend, window_autocorr)
+require(doMC)
+registerDoMC()
+wsA <- ddply(Asim, "reps", warningtrend, window_autocorr, .parallel=TRUE)
+wsB <- ddply(Bsim, "reps", warningtrend, window_autocorr, .parallel=TRUE)
 @
@@
 
+Lastly we save the data from each of these examples, that it may be available to the users later.   
+<<saveall>>=
+save(list=ls(), file="tutorial.rda")
+@
+
 %\section{Package Installation}

### 2012-02-16 boettiger.tex
updating appendices examples

@@
   \author[esp]{Alan Hastings}
-  \cortext[cor1]{Corresponding author.}
+  %\cortext[cor1]{Corresponding author.}
   \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
@@
 through the lens of early warning indicators such as rising variance or increased return times.
-Without the benefit of hindsight of a collapse,
+Despite widespread appreciation of the difficulties and uncertainty involved in such forecasts,
+proposed methods hardly ever characterize their expected error rates. 
+Without the benefits of replicates, controls, or hindsight, 
 applications of these approaches must quantify how reliable different indicators are in avoiding false alarms,
 and how sensitive they are to missing subtle warning signs.
-We propose an approach which quantifies this trade-off between reliability and sensitivity,
-allows comparisons between different indicators,
-and estimates what data are necessary to achieve a desired error rate.
+We propose a model based approach in order to quantify this trade-off between reliability and sensitivity
+and allow comparisons between different indicators.
 We show these error rates can be quite severe for common indicators even under favorable assumptions,
-and illustrate how a likelihood-based indicator can improve this performance.
-% rewrite
-
+and also illustrate how a model-based indicator can improve this performance. 
+We demonstrate how the performance of an early warning indicator varies in different data sets, 
+and suggest that uncertainty quantification become a more central part of early warning predictions.  
   \end{abstract}
@@
 Many important ecosystems may currently be threatened with collapse, including corals~\citep{Bellwood2004}, fisheries~\citep{Berkes2006}⁠, lakes~\citep{Carpenter2011}, and semi-arid ecosystems~\citep{Kefi2007}⁠.
-Given the potential impact of  these shifts on the sustainable delivery of ecosystem services
+Given the potential impact of  these shifts on the sustainable delivery of ecosystem services~\citep{Folke2004}
 and the need for management to either avoid an undesirable shift or else to adapt to novel conditions,
@@
 to fit the model using either a time series or else independent estimates of parameters. 
-More generally, with a good model of the system, detail-oriented approaches could be useful~\citep{Lenton2009}.
+More generally, with a good model of the system, detail-oriented approaches could be useful~\citep{Lade2012}.  
+In this treatment we focus on the situation where these more detailed models are not available.
 
-However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009}
+However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
 that do not depend on estimating the parameters of a known model of a specific system. 
@@
 that look for generic signs of impending regime shifts.  
-Our goal here is to extend earlier work by providing much more precise estimates 
-of the ability of different statistics to signal impending regime shifts 
-and to develop new approaches that both are more efficient 
+Here we extend earlier work by providing estimates 
+of the ability of different potential indicators to accurately signal impending regime shifts,
+and develop new approaches that both are more efficient
 and also lay bare some of the important assumptions underlying attempts 
 to find general warning signs of regime shifts.  
+We distinguish this question from the extensive literature involving change-point analysis 
+for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodinonv2004, Lenton2009}, 
+which is of little use if the goal is the advanced detection of the shift.
 
-We begin our approach by first discussing the limitations of current approaches that rely on summary statistics
-and provide a clear description of assumptions through the introduction of a model based approach to detect early warning signals.
+We begin by discussing the limitations of current approaches that rely on summary statistics
+and provide a description of assumptions through the introduction of a model based approach to detect early warning signals.
 We then illustrate how stochastic differential equation (SDE) models can be used
-to reflect the uncertainty inherent in the detection of early warning signals. We caution against paradigms that are not useful for capturing uncertainty in
+to reflect the uncertainty inherent in the detection of early warning signals. 
+We caution against paradigms that are not useful for capturing uncertainty in
 a model-selection based approach, such as information criteria.
@@
 and different indicators have in detecting early warning signals and use this to explore a number of examples.
+This approach provides a visualization of the types of errors that arise and how one can trade off between them.
 
@@
 In addition to the assumption of a saddle node bifurcation, 
-the calculation of statistics that would be used to detect an impending transition is subject to several additional arbitrary choices.
+the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
 A basic difficulty arises from the need to assume a time-series is ergodic: 
@@
 while most of the literature simply forgoes quantifying the increase or estimating significance.
-While adequate in experimental systems that can compare patterns between controls and replicates~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}
+While adequate in experimental systems that can compare patterns between controls 
+and replicates~\citep[\emph{e.g.}][]{Drake2010, Carpenter2011}, 
 any real-world application of these approaches must be useful on a single time-series of observations.
 In these cases a quantitative definition of a statistically significant detection is essential.
+Without this, we have no assurance that a purported detection is not, in fact, a false positive. 
+By focusing primarily on examples known to be approaching a transition when testing warning signals,
+the probability of false positives has largely been overlooked.
 
@@
 
-\subsection{A simulation-based approach estimates the distributions}
+\subsection{Simulation-based comparisons} \label{simbased}
 We perform the identical analysis procedure described above on each of these three data sets.
 First, we estimate parameters for the null and test model to each data set by maximum likelihood.
-Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better --
-
+Comparing the likelihood of these fits directly gives us only a minimal indication of which model fits better. 
 To identify if these differences are significant,
 and by what probability they could arise as a false alarm or a missed event,
-we simulate 500 replicate time series from each estimated model,
-as visualized in the middle panel of each data set,
-Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation}.
+we simulate 500 replicate time series from each estimated model.
 
@@
 The extent to which these distributions overlap indicates our inability to distinguish between these scenarios.
-The tendency of the observed deviance (vertical line) to fall more clearly in the domain of one distribution or the other
+The tendency of the observed deviance to fall more clearly in the domain of one distribution or the other
 indicates the probability our observed data corresponds best with that model  --
@@
 While it trivial to assign a statistical significance to this observation based
-on how far into the tail of the (red) null distribution it falls,
+on how far into the tail of the null distribution it falls,
 for the reasons we discussed we prefer the more symmetric comparison of the probability that
@@
 We illustrate this approach on simulated data as well as several natural time-series that have been previously analyzed for early warning signals.
-All data and code for simulations and analysis are found in the accompanying R package, \verb|warningsignals|.
+All data and code for simulations and analysis are found in the accompanying R package, \verb|earlywarning|.
 
@@
 \subsection*{Analysis}
-The deviances $\delta$, Eqn~\eqref{delta} -- are 5.1, 6.0, 83.9 respectively for the simulation,
-the chemostat data and the glaciation data.
+The deviances $\delta$ observed are 5.1, 6.0, 83.9 for the simulation,
+the chemostat data and the glaciation data, respectively.
+Based on AIC score each is large enough to reject the null hypothesis of a stable model 
+with it's one extra parameter, but this does not give the full picture of the anticipated error rates.  
 The size of these differences reflects not only the magnitude of the difference in fit
@@
 
-As described above, we simulate 500 replicates under each model, 
+Our simulation approach can provide a better sense of the relative trade-off in error rates associated
+with these estimates.  
+As described above (Section~\ref{simbased}), we simulate 500 replicates under each model, 
 shown in the middle panels of Figures~\ref{fig:simulation},~\ref{fig:chemostat} and~\ref{fig:glaciation},
@@
 and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308. 
-Data and code are available at \href{https://github.com/cboettig/warningsignals}{ https://github.com/cboettig/warningsignals}.
+Data and code are available at \href{https://github.com/cboettig/earlywarning}{ https://github.com/cboettig/earlywarning}.
 

### 2012-02-16 appendix/appendices.tex
minor edits and touchups

@@
   \begin{abstract}
-  These appendices explore several concepts that do not get fully developed in the paper, while also providing an introduction to the early warning signals package we provide to support this paper.
+  These appendices explore several concepts that do not get fully developed in the paper, while also providing an introduction to the early warning signals package we provide (\href{https://github.com/cboettig/earlywarning}{https://github.com/cboettig/earlywarning}) to support this paper.
+We provide the code in this document to both illustrate the steps required and also that the commands can be copy-pasted into~\texttt{R} using different choices for the warning indicator, the sampling of the simulation, etc.
 This document is written in Sweave, so the code you see is the code that generates the figures and results shown.
@@
 
-\section{Introduction}
-\setkeys{Gin}{width=0.85\textwidth}
+\section{Introduction -- Visualizing early warning patterns}
+\setkeys{Gin}{width=0.95\textwidth}
 
@@
 
-Simulate from a system approaching a collapse~64 times,
+Simulate from a system approaching a collapse~64 times using our population dynamics package, available at \href{https://github.com/cboettig/populationdynamics}{https://github.com/cboettig/populationdynamics},
 
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{populationdynamics}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{pars}{\ }\hlassignement{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlargument{Xo}{\ }\hlargument{=}{\ }\hlnumber{730}\hlkeyword{,}{\ }\hlargument{e}{\ }\hlargument{=}{\ }\hlnumber{0.5}\hlkeyword{,}{\ }\hlargument{a}{\ }\hlargument{=}{\ }\hlnumber{100}\hlkeyword{,}{\ }\hlargument{K}{\ }\hlargument{=}{\ }\hlnumber{1000}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{h}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{,}{\ }\hlargument{i}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{Da}{\ }\hlargument{=}{\ }\hlnumber{0.09}\hlkeyword{,}{\ }\hlargument{Dt}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{p}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{time}{\ }\hlassignement{=}{\ }\hlfunctioncall{seq}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{900}\hlkeyword{,}{\ }\hlargument{length}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{sn}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{saddle\usebox{\hlnormalsizeboxunderscore}node\usebox{\hlnormalsizeboxunderscore}ibm}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{reps}{\ }\hlargument{=}{\ }\hlsymbol{nreps}\hlkeyword{)}\mbox{}
+\hlstd{}\hlsymbol{pars}{\ }\hlassignement{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlargument{Xo}{\ }\hlargument{=}{\ }\hlnumber{730}\hlkeyword{,}{\ }\hlargument{e}{\ }\hlargument{=}{\ }\hlnumber{0.5}\hlkeyword{,}{\ }\hlargument{a}{\ }\hlargument{=}{\ }\hlnumber{100}\hlkeyword{,}{\ }\hlargument{K}{\ }\hlargument{=}{\ }\hlnumber{1000}\hlkeyword{,}{\ }\hlargument{h}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{,}{\ }\hlargument{i}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{Da}{\ }\hlargument{=}{\ }\hlnumber{.09}\hlkeyword{,}{\ }\hlargument{Dt}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{p}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{time}\hlassignement{=}\hlfunctioncall{seq}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{900}\hlkeyword{,}{\ }\hlargument{length}\hlargument{=}\hlnumber{200}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{sn}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{saddle\usebox{\hlnormalsizeboxunderscore}node\usebox{\hlnormalsizeboxunderscore}ibm}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{reps}\hlargument{=}\hlsymbol{nreps}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \ttfamily\noindent
-\hlsymbol{X}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{time}{\ }\hlargument{=}{\ }\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{value}{\ }\hlargument{=}{\ }\hlsymbol{sn}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{x1}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{dat}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{X}\hlkeyword{,}{\ }\hlargument{id}{\ }\hlargument{=}{\ }\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
+\hlsymbol{X}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{time}\hlargument{=}\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{value}\hlargument{=}\hlsymbol{sn}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{x1}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{dat}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{X}\hlkeyword{,}{\ }\hlargument{id}\hlargument{=}\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
 \hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{dat}\hlkeyword{)}\hlkeyword{[}\hlnumber{2}\hlkeyword{]}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlstring{"{}reps"{}}\mbox{}
@@
 \ttfamily\noindent
-\hlsymbol{acorr}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{summary\usebox{\hlnormalsizeboxunderscore}statistic}\hlkeyword{(}\hlsymbol{dat}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{windowsize}{\ }\hlargument{=}{\ }\hlfunctioncall{length}\hlkeyword{(}\hlsymbol{time}\hlkeyword{)}\hlkeyword{/}\hlnumber{2}\hlkeyword{)}\mbox{}
+\hlsymbol{acorr}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{summary\usebox{\hlnormalsizeboxunderscore}statistic}\hlkeyword{(}\hlsymbol{dat}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}{\ }\hlargument{windowsize}\hlargument{=}\hlfunctioncall{length}\hlkeyword{(}\hlsymbol{time}\hlkeyword{)}\hlkeyword{/}\hlnumber{2}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \end{knitrout}
+
+\setkeys{Gin}{width=0.95\textwidth}
 
@@
 \ttfamily\noindent
-\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{acorr}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{time}\hlkeyword{,}{\ }\hlsymbol{value}\hlkeyword{)}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{reps}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{opts}\hlkeyword{(}\hlargument{title}{\ }\hlargument{=}{\ }\hlstring{"{}Autocorrelation{\ }on{\ }replicates{\ }from{\ }a{\ }system{\ }approaching{\ }a{\ }crash"{}}\hlkeyword{)}\mbox{}
+\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{acorr}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{time}\hlkeyword{,}{\ }\hlsymbol{value}\hlkeyword{)}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{reps}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
+\hlstd{}{\ }{\ }\hlfunctioncall{opts}\hlkeyword{(}\hlargument{title}\hlargument{=}\hlstring{"{}Autocorrelation{\ }on{\ }replicates{\ }from{\ }a{\ }system{\ }approaching{\ }a{\ }crash"{}}\hlkeyword{)}\mbox{}
 \normalfont
@@
 
-\caption{The pattern of autocorrelation in replicate simulations from a simulation of a system which is approaching a saddle node bifurcation.}\label{crashacor}
+\caption{The pattern of autocorrelation in replicate simulations from a simulation of a system which is approaching a saddle node bifurcation.
+While a defintitive trend is occassionally visible, clearly we will statistical methods to define what constitutes an "increase."  Our model-based toolset provides this.}\label{crashacor}
 \end{center}
 \end{figure}
+\setkeys{Gin}{width=0.5\textwidth}
 
@@
 Reality is unlikely to be any less murky, so the purpose this paper is to provide statistical guidance about how to start approaching this uncertainty.
-
-We provide the code in this document to both illustrate the steps required and also that the commands can be copy-pasted into R using different choices for the warning indicator, the sampling of the simulation, etc.
 
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{snowfall}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{sfInit}\hlkeyword{(}\hlargument{par}{\ }\hlargument{=}{\ }\hlsymbol{T}\hlkeyword{,}{\ }\hlargument{cpu}{\ }\hlargument{=}{\ }\hlnumber{16}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlfunctioncall{sfInit}\hlkeyword{(}\hlargument{par}\hlargument{=}\hlsymbol{T}\hlkeyword{,}{\ }\hlargument{cpu}\hlargument{=}\hlnumber{16}\hlkeyword{)}\hspace*{\fill}\\
 \hlstd{}\hlfunctioncall{sfLibrary}\hlkeyword{(}\hlsymbol{earlywarning}\hlkeyword{)}\hspace*{\fill}\\
@@
 \ttfamily\noindent
-\hlsymbol{reps}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{sfLapply}\hlkeyword{(}\hlnumber{1}\hlkeyword{:}\hlnumber{500}\hlkeyword{,}{\ }\hlkeyword{function}\hlkeyword{(}\hlformalargs{i}\hlkeyword{)}{\ }\hlfunctioncall{compare}\hlkeyword{(}\hlsymbol{A}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlsymbol{B}\hlkeyword{)}\hlkeyword{)}\mbox{}
+\hlsymbol{reps}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{sfLapply}\hlkeyword{(}\hlnumber{1}\hlkeyword{:}\hlnumber{500}\hlkeyword{,}{\ }\hlkeyword{function}\hlkeyword{(}\hlformalargs{i}\hlkeyword{)}{\ }\hlfunctioncall{compare}\hlkeyword{(}\hlsymbol{A}\hlkeyword{,}\hlsymbol{B}\hlkeyword{)}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{ggplot2}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{lr}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}density}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{value}\hlkeyword{,}{\ }\hlargument{fill}{\ }\hlargument{=}{\ }\hlsymbol{simulation}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{alpha}{\ }\hlargument{=}{\ }\hlnumber{0.7}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}vline}\hlkeyword{(}\hlargument{xintercept}{\ }\hlargument{=}{\ }\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\mbox{}
+\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{lr}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}density}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{value}\hlkeyword{,}{\ }\hlargument{fill}\hlargument{=}\hlsymbol{simulation}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{alpha}\hlargument{=}\hlnumber{.7}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
+\hlstd{}{\ }{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}vline}\hlkeyword{(}\hlargument{xintercept}\hlargument{=}\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}\hlargument{=}\hlnumber{2}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{beanplot}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{beanplot}\hlkeyword{(}\hlsymbol{value}{\ }\hlkeyword{\urltilda{}}{\ }\hlsymbol{simulation}\hlkeyword{,}{\ }\hlsymbol{lr}\hlkeyword{,}{\ }\hlargument{what}{\ }\hlargument{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{1}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{0}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{abline}\hlkeyword{(}\hlargument{h}{\ }\hlargument{=}{\ }\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\mbox{}
+\hlstd{}\hlfunctioncall{beanplot}\hlkeyword{(}\hlsymbol{value}{\ }\hlkeyword{\urltilda{}}{\ }\hlsymbol{simulation}\hlkeyword{,}{\ }\hlsymbol{lr}\hlkeyword{,}{\ }\hlargument{what}\hlargument{=}\hlfunctioncall{c}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}\hlnumber{1}\hlkeyword{,}\hlnumber{0}\hlkeyword{,}\hlnumber{0}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlfunctioncall{abline}\hlkeyword{(}\hlargument{h}\hlargument{=}\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}\hlargument{=}\hlnumber{2}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \hlsymbol{roc}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{roc\usebox{\hlnormalsizeboxunderscore}data}\hlkeyword{(}\hlsymbol{lr}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{roc}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{False.positives}\hlkeyword{,}{\ }\hlsymbol{True.positives}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{lwd}{\ }\hlargument{=}{\ }\hlnumber{1}\hlkeyword{)}\mbox{}
+\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{roc}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{False.positives}\hlkeyword{,}{\ }\hlsymbol{True.positives}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{lwd}\hlargument{=}\hlnumber{1}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{Hmisc}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlfunctioncall{subset}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{fit}{\ }\hlkeyword{\usebox{\hlnormalsizeboxpercent}in\usebox{\hlnormalsizeboxpercent}}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlstring{"{}AA"{}}\hlkeyword{,}{\ }\hlstring{"{}BB"{}}\hlkeyword{)}\hlkeyword{)}\hlkeyword{,}{\ }\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{fit}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlsymbol{value}\hlkeyword{)}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{stat\usebox{\hlnormalsizeboxunderscore}summary}\hlkeyword{(}\hlargument{fun.y}{\ }\hlargument{=}{\ }\hlsymbol{mean}\hlkeyword{,}{\ }\hlargument{geom}{\ }\hlargument{=}{\ }\hlstring{"{}bar"{}}\hlkeyword{,}{\ }\hlargument{position}{\ }\hlargument{=}{\ }\hlstring{"{}dodge"{}}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlfunctioncall{stat\usebox{\hlnormalsizeboxunderscore}summary}\hlkeyword{(}\hlargument{fun.data}{\ }\hlargument{=}{\ }\hlsymbol{median\usebox{\hlnormalsizeboxunderscore}hilow}\hlkeyword{,}{\ }\hlargument{geom}{\ }\hlargument{=}{\ }\hlstring{"{}pointrange"{}}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }\hlargument{position}{\ }\hlargument{=}{\ }\hlfunctioncall{position\usebox{\hlnormalsizeboxunderscore}dodge}\hlkeyword{(}\hlargument{width}{\ }\hlargument{=}{\ }\hlnumber{0.9}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{conf.int}{\ }\hlargument{=}{\ }\hlnumber{0.95}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{parameter}\hlkeyword{,}{\ }\hlargument{scales}{\ }\hlargument{=}{\ }\hlstring{"{}free\usebox{\hlnormalsizeboxunderscore}y"{}}\hlkeyword{)}\mbox{}
+\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlfunctioncall{subset}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{fit}{\ }\hlkeyword{\usebox{\hlnormalsizeboxpercent}in\usebox{\hlnormalsizeboxpercent}}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlstring{"{}AA"{}}\hlkeyword{,}{\ }\hlstring{"{}BB"{}}\hlkeyword{)}\hlkeyword{)}\hlkeyword{,}{\ }\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{fit}\hlkeyword{,}{\ }\hlsymbol{value}\hlkeyword{)}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
+\hlstd{}{\ }{\ }\hlfunctioncall{stat\usebox{\hlnormalsizeboxunderscore}summary}\hlkeyword{(}\hlargument{fun.y}{\ }\hlargument{=}{\ }\hlsymbol{mean}\hlkeyword{,}{\ }\hlargument{geom}{\ }\hlargument{=}{\ }\hlstring{"{}bar"{}}\hlkeyword{,}{\ }\hlargument{position}{\ }\hlargument{=}{\ }\hlstring{"{}dodge"{}}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
+\hlstd{}{\ }{\ }\hlfunctioncall{stat\usebox{\hlnormalsizeboxunderscore}summary}\hlkeyword{(}\hlargument{fun.data}{\ }\hlargument{=}{\ }\hlsymbol{median\usebox{\hlnormalsizeboxunderscore}hilow}\hlkeyword{,}{\ }\hlargument{geom}{\ }\hlargument{=}{\ }\hlstring{"{}pointrange"{}}\hlkeyword{,}\hspace*{\fill}\\
+\hlstd{}{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }{\ }\hlargument{position}{\ }\hlargument{=}{\ }\hlfunctioncall{position\usebox{\hlnormalsizeboxunderscore}dodge}\hlkeyword{(}\hlargument{width}{\ }\hlargument{=}{\ }\hlnumber{0.90}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{conf.int}{\ }\hlargument{=}{\ }\hlnumber{0.95}\hlkeyword{)}{\ }\hlkeyword{+}\hspace*{\fill}\\
+\hlstd{}{\ }{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{parameter}\hlkeyword{,}{\ }\hlargument{scales}\hlargument{=}\hlstring{"{}free\usebox{\hlnormalsizeboxunderscore}y"{}}\hlkeyword{)}\mbox{}
 \normalfont
@@
 
-\subsection{Visualizing the simulation replicates}
-
-The \texttt{reps} list also contains the simulation data from each replicate
-
-\begin{figure}[H]
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{.97, .97, .97}{\color{fgcolor}\begin{kframe}
-\begin{flushleft}
-\ttfamily\noindent
-\hlsymbol{sims}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{get\usebox{\hlnormalsizeboxunderscore}replicates}\hlkeyword{(}\hlsymbol{reps}\hlkeyword{)}\mbox{}
-\normalfont
-\end{flushleft}
-\begin{verbatim}
-## Error: argument is of length zero
-\end{verbatim}
-\begin{flushleft}
-\ttfamily\noindent
-\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{sims}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{time}\hlkeyword{,}{\ }\hlsymbol{value}\hlkeyword{,}{\ }\hlargument{group}{\ }\hlargument{=}{\ }\hlsymbol{rep}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{alpha}{\ }\hlargument{=}{\ }\hlnumber{0.1}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{facet\usebox{\hlnormalsizeboxunderscore}wrap}\hlkeyword{(}\hlkeyword{\urltilda{}}\hlsymbol{model}\hlkeyword{)}\mbox{}
-\normalfont
-\end{flushleft}
-\begin{verbatim}
-## Error: object 'sims' not found
-\end{verbatim}
-\end{kframe}}
-\end{knitrout}
-
-\end{figure}
+%\subsection{Visualizing the simulation replicates}
+%
+%The \texttt{reps} list also contains the simulation data from each replicate
+%\setkeys{Gin}{width=0.95\textwidth}
+%
+%\begin{figure}[H]
+%<<<repsplot, fig.width=7>>=
+%sims <- get_replicates(reps)
+%ggplot(sims) + geom_line(aes(time, value, group=rep), alpha=0.1) + facet_wrap(~model)
+%@
+%\end{figure}
+%\setkeys{Gin}{width=0.5\textwidth}
 
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{populationdynamics}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{pars}{\ }\hlassignement{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlargument{Xo}{\ }\hlargument{=}{\ }\hlnumber{730}\hlkeyword{,}{\ }\hlargument{e}{\ }\hlargument{=}{\ }\hlnumber{0.5}\hlkeyword{,}{\ }\hlargument{a}{\ }\hlargument{=}{\ }\hlnumber{100}\hlkeyword{,}{\ }\hlargument{K}{\ }\hlargument{=}{\ }\hlnumber{1000}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{h}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{,}{\ }\hlargument{i}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{Da}{\ }\hlargument{=}{\ }\hlnumber{0.09}\hlkeyword{,}{\ }\hlargument{Dt}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{p}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{time}{\ }\hlassignement{=}{\ }\hlfunctioncall{seq}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{500}\hlkeyword{,}{\ }\hlargument{length}{\ }\hlargument{=}{\ }\hlnumber{1500}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{sn}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{saddle\usebox{\hlnormalsizeboxunderscore}node\usebox{\hlnormalsizeboxunderscore}ibm}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}{\ }\hlsymbol{time}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{X}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{time}{\ }\hlargument{=}{\ }\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{value}{\ }\hlargument{=}{\ }\hlsymbol{sn}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{x1}\hlkeyword{)}\mbox{}
+\hlstd{}\hlsymbol{pars}{\ }\hlassignement{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlargument{Xo}{\ }\hlargument{=}{\ }\hlnumber{730}\hlkeyword{,}{\ }\hlargument{e}{\ }\hlargument{=}{\ }\hlnumber{0.5}\hlkeyword{,}{\ }\hlargument{a}{\ }\hlargument{=}{\ }\hlnumber{100}\hlkeyword{,}{\ }\hlargument{K}{\ }\hlargument{=}{\ }\hlnumber{1000}\hlkeyword{,}{\ }\hlargument{h}{\ }\hlargument{=}{\ }\hlnumber{200}\hlkeyword{,}{\ }\hlargument{i}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{Da}{\ }\hlargument{=}{\ }\hlnumber{.09}\hlkeyword{,}{\ }\hlargument{Dt}{\ }\hlargument{=}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlargument{p}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{time}\hlassignement{=}\hlfunctioncall{seq}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{500}\hlkeyword{,}{\ }\hlargument{length}\hlargument{=}\hlnumber{1500}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{sn}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{saddle\usebox{\hlnormalsizeboxunderscore}node\usebox{\hlnormalsizeboxunderscore}ibm}\hlkeyword{(}\hlsymbol{pars}\hlkeyword{,}\hlsymbol{time}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{X}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{time}\hlargument{=}\hlsymbol{time}\hlkeyword{,}{\ }\hlargument{value}\hlargument{=}\hlsymbol{sn}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{x1}\hlkeyword{)}\mbox{}
 \normalfont
@@
 ##    tau 
-## -0.1633 
+## 0.8628 
 \end{verbatim}
@@
 \hlsymbol{npts}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{length}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}value"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{autocorrelation}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}value"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{windowsize}{\ }\hlargument{=}{\ }\hlsymbol{npts}\hlkeyword{/}\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{autocorrelation}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlfunctioncall{rep}\hlkeyword{(}\hlnumber{NA}\hlkeyword{,}{\ }\hlsymbol{npts}\hlkeyword{/}\hlnumber{2}{\ }\hlkeyword{-}{\ }\hlnumber{1}\hlkeyword{)}\hlkeyword{,}{\ }\hlsymbol{autocorrelation}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{plot}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}time"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{,}{\ }\hlsymbol{autocorrelation}\hlkeyword{,}{\ }\hlargument{type}{\ }\hlargument{=}{\ }\hlstring{"{}l"{}}\hlkeyword{)}\mbox{}
+\hlstd{}\hlsymbol{autocorrelation}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}value"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{,}{\ }\hlargument{windowsize}\hlargument{=}\hlsymbol{npts}\hlkeyword{/}\hlnumber{2}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{autocorrelation}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlfunctioncall{rep}\hlkeyword{(}\hlnumber{NA}\hlkeyword{,}{\ }\hlsymbol{npts}\hlkeyword{/}\hlnumber{2}\hlkeyword{-}\hlnumber{1}\hlkeyword{)}\hlkeyword{,}{\ }\hlsymbol{autocorrelation}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlfunctioncall{plot}\hlkeyword{(}\hlsymbol{X}\hlkeyword{[[}\hlstring{"{}time"{}}\hlkeyword{]}\hlkeyword{]}\hlkeyword{,}{\ }\hlsymbol{autocorrelation}\hlkeyword{,}{\ }\hlargument{type}\hlargument{=}\hlstring{"{}l"{}}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \ttfamily\noindent
-\hlsymbol{reps}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlnumber{500}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{Asim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{simulate}\hlkeyword{(}\hlsymbol{A}\hlkeyword{,}{\ }\hlsymbol{reps}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{Bsim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{simulate}\hlkeyword{(}\hlsymbol{B}\hlkeyword{,}{\ }\hlsymbol{reps}\hlkeyword{)}\mbox{}
+\hlsymbol{nreps}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlnumber{500}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{Asim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{simulate}\hlkeyword{(}\hlsymbol{A}\hlkeyword{,}{\ }\hlsymbol{nreps}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{Bsim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{simulate}\hlkeyword{(}\hlsymbol{B}\hlkeyword{,}{\ }\hlsymbol{nreps}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \hlfunctioncall{require}\hlkeyword{(}\hlsymbol{reshape2}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{Asim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{,}{\ }\hlargument{id}{\ }\hlargument{=}{\ }\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{Bsim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{,}{\ }\hlargument{id}{\ }\hlargument{=}{\ }\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{Asim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{,}{\ }\hlargument{id}\hlargument{=}\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{Bsim}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{,}{\ }\hlargument{id}\hlargument{=}\hlstring{"{}time"{}}\hlkeyword{)}\hspace*{\fill}\\
 \hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{)}\hlkeyword{[}\hlnumber{2}\hlkeyword{]}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlstring{"{}reps"{}}\hspace*{\fill}\\
@@
 \hlstd{}\hlfunctioncall{registerDoMC}\hlkeyword{(}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{wsA}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{ddply}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{,}{\ }\hlstring{"{}reps"{}}\hlkeyword{,}{\ }\hlsymbol{warningtrend}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{.parallel}{\ }\hlargument{=}{\ }\hlnumber{TRUE}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlsymbol{wsB}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{ddply}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{,}{\ }\hlstring{"{}reps"{}}\hlkeyword{,}{\ }\hlsymbol{warningtrend}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{.parallel}{\ }\hlargument{=}{\ }\hlnumber{TRUE}\hlkeyword{)}\mbox{}
+\hlstd{}\hlsymbol{wsA}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{ddply}\hlkeyword{(}\hlsymbol{Asim}\hlkeyword{,}{\ }\hlstring{"{}reps"{}}\hlkeyword{,}{\ }\hlsymbol{warningtrend}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}{\ }\hlargument{.parallel}\hlargument{=}\hlnumber{TRUE}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlsymbol{wsB}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{ddply}\hlkeyword{(}\hlsymbol{Bsim}\hlkeyword{,}{\ }\hlstring{"{}reps"{}}\hlkeyword{,}{\ }\hlsymbol{warningtrend}\hlkeyword{,}{\ }\hlsymbol{window\usebox{\hlnormalsizeboxunderscore}autocorr}\hlkeyword{,}{\ }\hlargument{.parallel}\hlargument{=}\hlnumber{TRUE}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \ttfamily\noindent
-\hlsymbol{tidy}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{null}{\ }\hlargument{=}{\ }\hlsymbol{wsA}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{tau}\hlkeyword{,}{\ }\hlargument{test}{\ }\hlargument{=}{\ }\hlsymbol{wsB}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{tau}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
+\hlsymbol{tidy}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{melt}\hlkeyword{(}\hlfunctioncall{data.frame}\hlkeyword{(}\hlargument{null}\hlargument{=}\hlsymbol{wsA}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{tau}\hlkeyword{,}{\ }\hlargument{test}\hlargument{=}\hlsymbol{wsB}\hlkeyword{\usebox{\hlnormalsizeboxdollar}}\hlsymbol{tau}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
 \hlstd{}\hlfunctioncall{names}\hlkeyword{(}\hlsymbol{tidy}\hlkeyword{)}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlstring{"{}simulation"{}}\hlkeyword{,}{\ }\hlstring{"{}value"{}}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{beanplot}\hlkeyword{(}\hlsymbol{value}{\ }\hlkeyword{\urltilda{}}{\ }\hlsymbol{simulation}\hlkeyword{,}{\ }\hlsymbol{tidy}\hlkeyword{,}{\ }\hlargument{what}{\ }\hlargument{=}{\ }\hlfunctioncall{c}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlnumber{1}\hlkeyword{,}{\ }\hlnumber{0}\hlkeyword{,}{\ }\hlnumber{0}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{abline}\hlkeyword{(}\hlargument{h}{\ }\hlargument{=}{\ }\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}{\ }\hlargument{=}{\ }\hlnumber{2}\hlkeyword{)}\mbox{}
+\hlstd{}\hlfunctioncall{beanplot}\hlkeyword{(}\hlsymbol{value}{\ }\hlkeyword{\urltilda{}}{\ }\hlsymbol{simulation}\hlkeyword{,}{\ }\hlsymbol{tidy}\hlkeyword{,}{\ }\hlargument{what}\hlargument{=}\hlfunctioncall{c}\hlkeyword{(}\hlnumber{0}\hlkeyword{,}\hlnumber{1}\hlkeyword{,}\hlnumber{0}\hlkeyword{,}\hlnumber{0}\hlkeyword{)}\hlkeyword{)}\hspace*{\fill}\\
+\hlstd{}\hlfunctioncall{abline}\hlkeyword{(}\hlargument{h}\hlargument{=}\hlsymbol{observed}\hlkeyword{,}{\ }\hlargument{lty}\hlargument{=}\hlnumber{2}\hlkeyword{)}\mbox{}
 \normalfont
@@
 \hlsymbol{roc}{\ }\hlassignement{\usebox{\hlnormalsizeboxlessthan}-}{\ }\hlfunctioncall{roc\usebox{\hlnormalsizeboxunderscore}data}\hlkeyword{(}\hlsymbol{tidy}\hlkeyword{)}\hspace*{\fill}\\
-\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{roc}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{False.positives}\hlkeyword{,}{\ }\hlsymbol{True.positives}\hlkeyword{)}\hlkeyword{,}\hspace*{\fill}\\
-\hlstd{}{\ }{\ }{\ }{\ }\hlargument{lwd}{\ }\hlargument{=}{\ }\hlnumber{1}\hlkeyword{)}\mbox{}
+\hlstd{}\hlfunctioncall{ggplot}\hlkeyword{(}\hlsymbol{roc}\hlkeyword{)}{\ }\hlkeyword{+}{\ }\hlfunctioncall{geom\usebox{\hlnormalsizeboxunderscore}line}\hlkeyword{(}\hlfunctioncall{aes}\hlkeyword{(}\hlsymbol{False.positives}\hlkeyword{,}{\ }\hlsymbol{True.positives}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{lwd}\hlargument{=}\hlnumber{1}\hlkeyword{)}\mbox{}
 \normalfont
@@
 
-Lastly we save the data from each of these examples, that it may be available to the users later.   
+Lastly we save the data from each of these examples for future reference.   
 \begin{knitrout}
@@
 \ttfamily\noindent
-\hlfunctioncall{save}\hlkeyword{(}\hlargument{list}{\ }\hlargument{=}{\ }\hlfunctioncall{ls}\hlkeyword{(}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{file}{\ }\hlargument{=}{\ }\hlstring{"{}tutorial.rda"{}}\hlkeyword{)}\mbox{}
+\hlfunctioncall{save}\hlkeyword{(}\hlargument{list}\hlargument{=}\hlfunctioncall{ls}\hlkeyword{(}\hlkeyword{)}\hlkeyword{,}{\ }\hlargument{file}\hlargument{=}\hlstring{"{}tutorial.rda"{}}\hlkeyword{)}\mbox{}
 \normalfont

### 2012-02-16 appendix/appendices_knit_.Rnw
minor edits and touchups

@@
 %% Looks like a comment but it isn't! This is setting the default behavior for the Sweave chunk options, <<options >>= 
-% \SweaveOpts{fig.width=7, fig.height=6, fig.align=center, message=FALSE, warning=FALSE}
+% \SweaveOpts{fig.width=4, fig.height=4, fig.align=center, message=FALSE, warning=FALSE, tidy=FALSE, cache=TRUE}
 
@@
 
-Simulate from a system approaching a collapse~\Sexpr{nreps} times,
+Simulate from a system approaching a collapse~\Sexpr{nreps} times using our population dynamics package, available at \href{https://github.com/cboettig/populationdynamics}{https://github.com/cboettig/populationdynamics},
 
@@
 
+\setkeys{Gin}{width=0.95\textwidth}
+
 \begin{figure}[H]
 \begin{center}
-<<crashacor>>=
+<<crashacor, fig.width=7, fig.height=7>>=
 ggplot(acorr) + geom_line(aes(time, value)) + facet_wrap(~reps) + 
@@
 @
-\caption{The pattern of autocorrelation in replicate simulations from a simulation of a system which is approaching a saddle node bifurcation.}\label{crashacor}
+\caption{The pattern of autocorrelation in replicate simulations from a simulation of a system which is approaching a saddle node bifurcation.
+While a defintitive trend is occassionally visible, clearly we will statistical methods to define what constitutes an "increase."  Our model-based toolset provides this.}\label{crashacor}
 \end{center}
 \end{figure}
+\setkeys{Gin}{width=0.5\textwidth}
 
@@
 \begin{figure}[H]
-<<beanplot>>=
+<<beanplot, fig.width=6>>=
 require(beanplot)
@@
 
-\subsection{Visualizing the simulation replicates}
-
-The \texttt{reps} list also contains the simulation data from each replicate
-
-\begin{figure}[H]
-<<<repsplot, fig.width=10>>=
-sims <- get_replicates(reps)
-ggplot(sims) + geom_line(aes(time, value, group=rep), alpha=0.1) + facet_wrap(~model)
-@
-\end{figure}
+%\subsection{Visualizing the simulation replicates}
+%
+%The \texttt{reps} list also contains the simulation data from each replicate
+%\setkeys{Gin}{width=0.95\textwidth}
+%
+%\begin{figure}[H]
+%<<<repsplot, fig.width=7>>=
+%sims <- get_replicates(reps)
+%ggplot(sims) + geom_line(aes(time, value, group=rep), alpha=0.1) + facet_wrap(~model)
+%@
+%\end{figure}
+%\setkeys{Gin}{width=0.5\textwidth}
 
@@
 <<moresims>>=
-reps <- 500
-Asim <- simulate(A, reps)
-Bsim <- simulate(B, reps)
+nreps <- 500
+Asim <- simulate(A, nreps)
+Bsim <- simulate(B, nreps)
 @
@@
 \begin{figure}[H]
-<<tidybean>>=
+<<tidybean, fig.width=6>>=
 tidy <- melt(data.frame(null=wsA$tau, test=wsB$tau))
@@
 
-Lastly we save the data from each of these examples, that it may be available to the users later.   
+Lastly we save the data from each of these examples for future reference.   
 <<saveall>>=

### 2012-02-16 boettiger.tex
minor edits and touchups
spellcheck and ready to go

@@
 \usepackage[pdftex, colorlinks]{hyperref}
-\usepackage{float}
+%\usepackage{float}
+%\usepackage{endfloat}
 
@@
   \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
-  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}
+  \address[esp]{Department of Environmental Science and Policy, University of California, Davis, CA, 95616 United States}
 
@@
 We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.
-
-%%% hmm, what about all the red-herring topics, like change point analysis?
-Add a discussion such as Contamin & Ellison ~\citep{Contamin2009} have about forecasting shifts, not post-hoc identification of shifts.
 
@@
 reasonable null models are more readily available,
-such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedaticity.
+such as~\citet{Seekell2011}'s proposal to test for conditional heteroscedasticity.
 
@@
 
-where $P(n,t)$ is the probability of having $n$ individuals at time $t$, $b_n$ is the probability of a birth event occuring in a population of $n$ individuals, $d_n$ the probability of a death.
+where $P(n,t)$ is the probability of having $n$ individuals at time $t$, $b_n$ is the probability of a birth event occurring in a population of $n$ individuals, $d_n$ the probability of a death.
 $e, K, h$ and $a_t$ are parameters.
@@
 
-The use of box plots in Figure~\ref{fig:summary_box} provide a convient and familiar way 
+The use of box plots in Figure~\ref{fig:summary_box} provide a convenient and familiar way 
 to visualize the overlap between more than two distributions,
@@
 to the assumptions of the summary statistics approaches.  
-On the simulated data, the varianced based method approaches the 
+On the simulated data, the variance based method approaches the 
 true-positive rate of our likelihood method at higher levels of false positives, 
-but performs worse when the desired level of false postives is low.  
+but performs worse when the desired level of false positives is low.  
 The ROC curve helps us compare the performance of the different approaches at different tolerances.  
@@
 
- % past tense or not?
 We have used three examples to illustrate the performance of this approach in data from 
@@
 the uncertainty in the smaller simulated and experimental data forces a trade-off between errors.
-%Because such uncertainty may be more typical of early warning signal detection 
-%than the monotonically increasing patterns often used to illustrate the approach,
 
@@
 where a 5\% false positive rate often corresponds to only a 5\% true positive rate,
-performing no better than the flip of a coin. %%%% EEEK can we really say it that strongly?
+performing no better than the flip of a coin. 
 By estimating the ROC curve for a given set of data,
@@
 
-%%% Model adequacy
 In any application of early warning signals, it is essential to address the question of model adequacy.

### 2012-02-16 boettiger.tex
updated another citation

@@
 We distinguish this question from the extensive literature involving change-point analysis 
-for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodinonv2004, Lenton2009}, 
+for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodionov2004, Lenton2009}, 
 which is of little use if the goal is the advanced detection of the shift.

### 2012-02-23 boettiger.tex
submitted to JRSI
corrected transcritical bifurcation statement

@@
 and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
-However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
-they are characterized by a steady approach towards the alternative stable state, 
-such as a population size gradually approaching zero.  
+However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome.% --
+%they are characterized by a steady approach towards the alternative stable state, 
+%such as a population size gradually approaching zero.  
 While no approach will be applicable to all classes of sudden transitions,
@@
 \section{Acknowledgments}
-We would like to thank S. Schreiber, M. Holyoak, M. Baskett, and A. Perkins for comments on the manuscript. 
+We would like to thank S. Schreiber, M. Holyoak, M. Baskett, A. Perkins and N. Ross for comments on the manuscript. 
 This research was supported by funding from NSF Grant EF 0742674 

### 2012-04-02 fallacy.tex
fallacy note added
fallacy writing

@@
 
-\journal{\tiny Journal of the Royal Society Interface} 
+\journal{\tiny } 
 \begin{document}
@@
   from conditional probabilities is often known as the the Prosecutor's 
-  Fallacy.  As we pursue this important question of early warning signals, 
-  we must be careful to avoid this trap.
+  Fallacy.  We illustrate through simulation how examples of systems that have 
+  experienced transitions purely by chance show patterns associated with 
+  early warning signals more often than we might expect in systems that have not
+  been conditionally selected.  We further highlight how a model-based approach
+  is less subject to this bias.  In using historical examples of collapse to 
+  test and validate methods for early warning signals, greater care must be taken
+  to avoid this fallacy.
 
@@
  \end{frontmatter}
+
+ \section{Introduction}
+
+ \section{Methods and Results}
+ We simulated 1000 replicates of an individual-based birth-death process with an allee threshold.
+Above the allee threshold the population returns to a positive equilibrium size.
+Below the threshold the population decreases to zero.
+The simulation uses the Gillespie algorithm to provide an exact implementation of the master equation of birth-death process,
+
+\begin{align}
+  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
+    b_n &= \frac{e K n^2}{n^2 + h^2} \\
+    d_n &= e n + a
+\end{align}
+
+Though this system can be forced through a saddle-node bifurcation by
+increasing values of either $e$ or $a$, all parameters are held fixed.
+The simulation starts from the positive equilibrium population size.
+Though the chance of a transition across the allee threshold in any 
+given timestep are small, given enough time this system will eventually
+experience such a rare event driving the population extinct.  We evolved
+each replicate 50000 timesteps, sampling the system every 50 timesteps.  
+In this time window 155 of the 1000 replicates experience population collapse.  
+Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
+as an illustration.
+
+We selected replicates conditional on having collapsed in the simulations.
+We then selected a window around each system that ended just before the
+collapse, while the population values were still above the allee threshold.
+For each replicate, we calculated the most common warning signals, variance
+and autocorrelation~\citep[\emph{e.g.}][]{Carpenter2006,Dakos2008,Scheffer2009}, 
+around a moving window equal to half the length of that
+timeseries.
+
+To determine whether or not the variance or autocorrelation increased 
+over time in these 155 replicates, we computed Kendall's $\tau$ for each of the
+warning signal pattern~\citep{Dakos2008, Dakos2010}.  Positive values of $\tau$ 
+indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
+values observed across these replicates is shown in Figure~\ref{fig:indicator}.
+More positive values indicate stronger trends. Assessing the statistical significance
+of a given value requires an appropriate null model, which can be difficult 
+to do~\citep{Dakos2008}. As these examples have been selected from a system
+that is not approaching a bifurcation and therefore should not exhibit any 
+early warning sign, even the very strong values of $\tau$ should be considered 
+as arising from the null model.
+
+We compared this approach to a model-based approach
+
+  \begin{figure}[ht]
+    \begin{center}
+      \includegraphics{replicate_crashes}
+    \end{center}
+    \caption{Nine of the 155 replicates that experience a critical transition by chance during the course of the simulation.}
+    \label{fig:replicate_crashes}
+  \end{figure}
+
+  \begin{figure}[ht]
+    \begin{center}
+      \includegraphics{fallacy}
+    \end{center}
+    \caption{The distribution of three early warning indicators on replicates conditionally selected for having collapsed by chance in simulations.}
+    \label{fig:indicator}
+  \end{figure}
+
+ \section{Discussion}
+
  \section{References}%bibliography

### 2012-04-19 fallacy.tex
minor
updates to fallacy

@@
 %\documentclass[authoryear,5p]{elsarticle}
-\documentclass[authoryear,preprint,11pt]{elsarticle}
+\documentclass[authoryear,review,12pt]{elsarticle}
 \bibliographystyle{elsarticle-harv}
@@
 \linenumbers
+\usepackage{float}
 \usepackage{subfigure}
@@
 \definecolor{darkgreen}{rgb}{0,0.5,0}
-%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
 \usepackage[pdftex, colorlinks]{hyperref}
@@
 \newcommand{\V}{\mathrm{Var}}
+\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}
 
-\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}
+%% Redefines the elsarticle footer
+\makeatletter
+\def\ps@pprintTitle{
+\let\@oddhead\@empty
+\let\@evenhead\@empty
+\def\@oddfoot{\it \hfill\today}
+\let\@evenfoot\@oddfoot}
+\makeatother
 
@@
 \section{Introduction}
+In a criminal trial, the prosecution may introduce associative evidence such as 
+blood testing to establish a match between the perpetrator at the scene of the
+crime and a suspect on the defense stand.  Statistical testimony may accompany 
+such evidence reflecting the incidence rate of the matching characteristic -- 
+for instance, the frequency of individuals in the population with the matching
+blood type.  Such arguments may be misleading when they fail to reflect the 
+probability conditional on the manner in which the suspect was identified in 
+the first place.  If the suspect has been selected partly on the basis of belonging
+to demographic in which the characteristic is much more common than in the 
+population as a whole, these odds can be misleading.  Failing to recognize 
+these conditional probabilities leads to the ``Prosecutor's Fallacy'' -- the argument
+that the suspect is the perpetrator on the basis of the statistical probability 
+of a match without currectly identifying the baseline probability for such a match
+\citep{Thompson1987}.
+
+Statisical patterns associated with sudden, catastrophic transitions or ``tipping points''
+could play an important role as early warning signs or leading indicators of
+these shifts, which may give us precious time to react and avert them~\citep{Scheffer2009, Lenton2011}.
+Historical examples of sudden transitions taken from the paleoclimate record
+provide an important way to test and evaluate potential leading indicator methods,
+and have been widely used for this purpose~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010} 
+Similarly, it has been suggested that data gathered on ecological systems such as lakes
+that were monitored before the collapsed, or grasslands subjected to overgrazing,
+could contain data that could help reveal when similar systems are approaching 
+a tipping point~\citet{Inman2011}.  This is a silver lining to the clouds of past
+catastrophes -- we may be able to learn from our mistakes.
+
+In testing methods for early warning signals against historical examples of transitions,
+we must be careful to avoid falling into the Prosecutor's Fallacy.  Sudden transitions
+can occur for many reasons that do not involve the bifurcation proccesses on which
+most early warning indicators are predicated~\citep{Hastings2010}, including purely
+noise-induced transitions for which early warning signs will not exist~\citep{Ditlevsen2010, Lenton2011}.
+
+% Explain variance, autocorrelation, tau.
+
+\emph{Add more context \dots }
 
 \section{Methods and Results}
- We simulated 1000 replicates of an individual-based birth-death process with an allee threshold.
-Above the allee threshold the population returns to a positive equilibrium size.
-Below the threshold the population decreases to zero.
-The simulation uses the Gillespie algorithm to provide an exact implementation of the master equation of birth-death process,
+We simulated 1000 replicates of an individual-based birth-death process
+with an allee threshold.  Above the allee threshold the population returns
+to a positive equilibrium size.  Below the threshold the population decreases
+to zero. The simulation uses the Gillespie algorithm to provide an exact 
+implementation of the master equation of birth-death process,
 
@@
 Though this system can be forced through a saddle-node bifurcation by
-increasing values of either $e$ or $a$, all parameters are held fixed.
+increasing values of either $e$ or $a$, in these simulations all parameters
+are held constant and no early warning signal is anticipated.
 The simulation starts from the positive equilibrium population size.
@@
 
+  \begin{figure}[H]
+    \begin{center}
+      \includegraphics{replicate_crashes.pdf}
+    \end{center}
+    \caption{Example trajectories for nine of the 155 replicates that 
+             experience a critical transition by chance during the course
+             of the simulation.}
+    \label{fig:replicate_crashes}
+  \end{figure}
+
 We selected replicates conditional on having collapsed in the simulations.
@@
 and autocorrelation~\citep[\emph{e.g.}][]{Carpenter2006,Dakos2008,Scheffer2009}, 
-around a moving window equal to half the length of that
-timeseries.
+around a moving window equal to half the length of that timeseries.
 
@@
 that is not approaching a bifurcation and therefore should not exhibit any 
-early warning sign, even the very strong values of $\tau$ should be considered 
-as arising from the null model.
+early warning sign. %Even the very strong values of $\tau$ should be considered 
+%as arising from the null model.
 
-We compared this approach to a model-based approach
-
-  \begin{figure}[ht]
+\begin{figure}[H]
   \begin{center}
-      \includegraphics{replicate_crashes}
+    \includegraphics{indicators.pdf}
   \end{center}
-    \caption{Nine of the 155 replicates that experience a critical transition by chance during the course of the simulation.}
-    \label{fig:replicate_crashes}
-  \end{figure}
-
-  \begin{figure}[ht]
-    \begin{center}
-      \includegraphics{fallacy}
-    \end{center}
-    \caption{The distribution of three early warning indicators on replicates conditionally selected for having collapsed by chance in simulations.}
+  \caption{The distribution of two early warning indicators on replicates conditionally selected for having collapsed by chance in simulations.}
   \label{fig:indicator}
@@
 
+For each of these replicates we also take a model-based approach, estimating 
+parameters for an approximate linear model of the system approaching a
+saddle node bifurcation.
+
+\begin{equation}
+  dX = \sqrt{ r_t } (\phi(r_t) - X_t) dt +
+  \sigma\sqrt{\phi(r_t) } d B_t \label{LSN}
+\end{equation}
+
+where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilbirium postition.  
+This model emerges from the canonical form of the saddle-node bifurcation,
+
+\begin{equation}
+  \frac{dX}{dt} = r - x^2
+\end{equation}
+
+Where we introduce an explicit parameter $\theta$ to account for rescaling
+to the units of the data,
+
+\begin{equation}
+   \frac{dX}{dt} = r - (x-\theta)^2
+\end{equation}
+
+and then consider the corresponding stochastic process where the noise is
+proportional to the square root of the mean, as is the case for demographic
+noise processes,
+
+\begin{equation}
+  \frac{dX}{dt} = r - (x-\theta)^2 + \sqrt{ \beta (r + (x-\theta)^2 } dB_t
+\end{equation}
+
+$r$ remains the bifurcation parameter which we imagine changing with time,
+$r_t$, then linearizing the above model in neighboorhood of
+\(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\), we recover Equation~\eqref{LSN}.
+These models can be both be fit to the data by maximum likelihood estimation
+based on the equation for the evolution of their moments, as
+in~\citet{Boettiger2012b}.
+
+Though this approximation is more general than the original Markov process,
+it is also easier to estimate reliably from the data since it describes only 
+the process near the equilibrium where the data is avialable.
+
+We assume that $r_t$ is varying at constant rate, according to the model
+
+\begin{equation}
+ r_t = m t + r_0 
+\end{equation}
+
+and estimate $m$, $r_0$, $\theta$, and $\beta$ by maximum likelihood on each
+of the simulated datasets.  Estimates of $m < 0 $ are expected in systems 
+approaching a bifurcation, while for stable systems $m$ should be approximately zero.
+Of the 155 simulations, only six had estimates with machine precision different
+from zero, the greatest magnitude among these was $-7.1 \times 10^{-12}$,
+which would still correspond to a negilible change in $r_0$ over the 50,000 timesteps
+of the simulation.  Consequently, the model-based estimation shows no 
+evidence of bias on data that has been selected conditional on collapse.
+
  \section{Discussion}
+At first it seems tempting to argue that this bias is in fact valuable -- each
+of these systems did indeed collapse, so the fact that they show increased 
+probability to exhibit warning signals could be taken as a successful detection.
+Unfortunately, this reasoning is flawed. At the moment the forecast is made,
+these systems are no more likely to crash than their counterparts from the sample
+that did not collapse, just as a fair coin is still a fair coin after a string of
+consecutive ``heads.''  Taking a closer look at the patterns involved we can
+begin to understand why common indicators such as autocorrelation and variance
+can be misleading.
 
+Systems that crash by chance excursion beyond the tipping point
+
+Systems collapsing by chance do so from a string of unexpected events
+
+Discuss the pattern of a system crashing by chance -- why this looks like increased autocorrelation and increased variance.
+
+Discuss why this cannt be considered a successful detection.
+
+These are not more likely to crash than another system.
+
+Not a good test of the method's ability to detect bifurcations.
+
+ \section{Acknowledgements}
+This research was supported by funding from NSF Grant EF 0742674 to AH
+and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308 to CB. 
  \section{References}%bibliography
- \bibliography{boettiger}
+ \bibliography{fallacy}
  \end{document}

### 2012-04-21 ReplyToReviewers.md
corrections made based on reviews from Interface

Replies in _italics_

Referee: 1
==========

### Comments to the Author
I think that this is one of the most important papers yet written in what is emerging as an area of great interest.  My only concern is that it is written in a highly technical way that may not be accessible to the broad audience that should read it.   The idea of using a model-based approach and approaching the problem as one of model choice is very clever, but I had to read the section three times before I knew what that meant.   I would move some of that material up front to make the main points a little less drily than they are currently made. I think also some technical terms should be embellished with discussion and explanation up front.

_We thank the reviewer for this assessment, and we have done our best to address this important issue_

Tell the reader, for example, what a saddle-node bifurcation is.

_A definition now follows the first mention_

Explain how the coefficient of the noise term in the stochastic differential equation is chosen.

_Certainly this needed more explanation for any audience.  The justification now follows where this appears_

This part is written for experts, whereas for Interfaces it needs to be more pedagogical.

Why does the bifurcation parameter have a subscript t?

_to indicate it is changing in time, this has been stated when it first appears_

### Small points:  
- Line 25:  lead should be led
- Line 298:  Data are
- Line 309:  its

_Indeed, thanks for these corrections._

I like the examples.

Referee: 2
Comments to the Author
I am pleased to recommend this manuscript for publication in Proceedings of the Royal Society Interface. The study of early warning indicators for biological transitions is an extremely active area of research. This paper reports the first study (to my knowledge) of the general trade-off between reliability and sensitivity of various indicators. Its introduction of an approximate method-based approach is a major contribution. The authors may wish to consider the following minor issues:

1. Line 41: “powerful way” is gratuitous

_"powerful" has been deleted_

2. Lines 61-63: I think some transcritical bifurcations can be “blue-sky” transitions; the claim should be qualified

_We apoloigize that the jargon is may not be used consistently in the field.  Transcritical bifurcations swap the stability of two fixed points, the fixed points to not appear/dissappear into the "blue".  We have removed the term_

3. Lines 74-76: I think the question of whether a noise-induced transition can be indicated is open. Particularly, the argument of Ditlevsen is based on a local approximation that is not expected to hold just prior to the transition; the author's may wish to qualify this claim. Also, “noise-induced” transition is ambiguous, here referring to the stochastic fluctuation between attractors, but more commonly referring to critical values in the magnitude of the fluctuations, as in Horsthemke and Lefever (1984).

_Excellent point, this has been clarified.  In particular, we note that the patterns put forward as warning signals for saddle node bifurcations, such as increasing variance or increasing autocorrelation, should not be expected to hold if the transition between attractors is driven purely by chance deviation._

4. Line 97: It is not obvious to be that autocorrelation from interpolation to a grid necessarily introduces artificial correlations. Either a reference or numerical example should be provided or the idea should be stated in a more conjectural mood.

5. Line 248: “This introduces a nuisance parameter known as the statistical significance criterion” is overly prejudicial. Differences of statistical philosophy are fine, but I think this goes over the line.

_This has been rephrased (we meant only in the technical sense as a parameter you must choose arbitrarily rather than estimate from data, but the point is well taken)_

6. Line 301: Typo

_The missing noun (citation to Dakos et al 2008) has been inserted,  thanks for catching this._

### 2012-04-21 boettiger.tex
corrections made based on reviews from Interface

@@
 that do not depend on estimating the parameters of a known model of a specific system. 
-This has lead to a variety of approaches based on summary statistics
+This has led to a variety of approaches based on summary statistics
 ~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
@@
 Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
-as a powerful way to illustrate the sensitivity of different data sets
+as a way to illustrate the sensitivity of different data sets
 and different indicators have in detecting early warning signals and use this to explore a number of examples.
@@
 looking for patterns that are associated with this kind of transition.
+A saddle-node bifurcation is the result of a stable node and an unstable
+saddle-point colliding and annihilating, leaving the system to transition 
+to another attractor~\citep[see ][or other dynamical systems text 
+for an introduction to this and other bifurcations]{Guckenheimer1983}.
+
 Typical patterns include an increasing trend in a summary statistic such as
@@
 and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
-However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome.% --
+However, transcritical bifurcations involve a change in stability of a fixed point, rather
+than the sudden dissapearance of a fixed point that has made critical transitions so worrisome.% --
 %they are characterized by a steady approach towards the alternative stable state, 
@@
 it is certainly still useful to have an approach that detect transitions driven by
-saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.
+saddle-node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.
 
@@
 The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
-As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
+\citet{Livina2012} indicate that such noise induced transitions cannot be predicted through early warning signals -- 
+at least they are not expected to exhibit the same early warning patterns of increased variance and increased autocorrelation
+anticipated in the case of a saddle-node bifurcation.
 The third scenario is that the system does pass through a saddle-node bifurcation,
@@
 \subsubsection*{Arbitrary windows}
-In addition to the assumption of a saddle node bifurcation, 
+In addition to the assumption of a saddle-node bifurcation, 
 the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
@@
 we will write down a general model that is capable of approximating
-a wide class of models in which regime shifts are characterized by a saddle node bifurcation,
+a wide class of models in which regime shifts are characterized by a saddle-node bifurcation,
 and a second generic model that is capable of representing the behavior of such systems
@@
 \end{equation}
-where $x$ is the state variable and $r_t$ our bifurcation parameter,
-which may be slowly varying in time and
+where $x$ is the state variable and $r_t$ our bifurcation parameter.
+We have added a subscript $t$ to the bifurcation parameter as a reminder that
+it is the value which may be slowly varying in time and
 consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
@@
 This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
+Allowing the stochastic term to scale with the square root of $\phi$ follows from the assumption that
+of an internal-noise process, such as demographic stochasticity, that arises in deriving the SDE from
+a Markov process, see~\citet{Kampen2007a} or~\citet{Black2012}.  
 %It is also possible to derive this same expression from the master equation
@@
 It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
-This introduces a nuisance parameter known as the statistical significance criterion.
+This introduces the need for selecting a statistical significance criterion.
 In the hypothesis testing framework, a false positive is a Type I error,
@@
 as analyzed by~\citet{Dakos2008}.
-The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing
+The data are preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing
 to be as consistent as possible with the original analysis.
 We focus on the third glaciation event, consisting of 121 sample points.
-The match is not exact since estimates the de-trending window size manually,
+The match is not exact since~\citet{Dakos2008} estimates the de-trending window size manually,
 but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis.
@@
 Based on AIC score each is large enough to reject the null hypothesis of a stable model 
-with it's one extra parameter, but this does not give the full picture of the anticipated error rates.  
+with its one extra parameter, but this does not give the full picture of the anticipated error rates.  
 The size of these differences reflects not only the magnitude of the difference in fit

### 2012-04-23 ReplyToReviewers.md
minor edits

@@
-Replies in _italics_
+Dear Tim Holt,
+
+Thank you for handling the reviews of our manuscript.
+We are pleased to provide the updated version addressing the issues highlighted by the reviewers and hope that it is ready for publication.
+Please find our comments below addressing the points made by each referee.
+A pdf document highlighting all changes is also attached.
+(Our replies below in _italics_)
+
+Sincerely,
+
+Carl Boettiger
+
+Alan Hastings
 
@@
 
-_We thank the reviewer for this assessment, and we have done our best to address this important issue_
+_We thank the reviewer for this assessment, and we have done our best to address this important issue, clarifying definitions, expanding the explanation and making other small changes that should make this more accessible._
 
@@
 Its introduction of an approximate method-based approach is a major contribution.
+
 The authors may wish to consider the following minor issues:
@@
 
-_We apoloigize that the jargon is may not be used consistently in the field.
-Transcritical bifurcations swap the stability of two fixed points, the fixed points to not appear/dissappear into the "blue".
+_We apologize that the jargon is may not be used consistently in the field.
+Transcritical bifurcations swap the stability of two fixed points, the fixed points to not appear/disappear into the "blue".
 We have removed the term_
@@
 
-_The missing noun (citation to Dakos et al 2008) has been inserted,  thanks for catching this._
+_The missing noun (a citation to Dakos et al. 2008) has been inserted,  thanks for catching this._

### 2012-04-23 boettiger.tex
minor edits

@@
 
-However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
+Indeed, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
 that do not depend on estimating the parameters of a known model of a specific system. 
@@
 Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
-as a way to illustrate the sensitivity of different data sets
+as a way to illustrate the sensitivity different data sets
 and different indicators have in detecting early warning signals and use this to explore a number of examples.
-This approach provides a visualization of the types of errors that arise and how one can trade off between them.
+This approach provides a visualization of the types of errors that arise and how one can trade off between them,
+and is important for framing the problem as one focused on prediction.
 
@@
 looking for patterns that are associated with this kind of transition.
-A saddle-node bifurcation is the result of a stable node and an unstable
-saddle-point colliding and annihilating, leaving the system to transition 
-to another attractor~\citep[see ][or other dynamical systems text 
-for an introduction to this and other bifurcations]{Guckenheimer1983}.
+A saddle-node bifurcation occurs when a parameter changes and a stable equilibrium 
+(node) and an unstable equilibrium (saddle) coalesce and dissapear.  The system
+then moves to a more distant equilbrium.  \citet{Guckenheimer1983} or any other 
+textbook on dynamical systems will provide precise definitions and further explanation.
 
-Typical patterns include an increasing trend in a summary statistic such as
+Typical patterns used as warning signals include an increasing trend in a summary statistic such as
 variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
@@
 While attractive for their simplicity, such approaches must confront numerous challenges.
-In this paper we argue for a model-based approach,
+In this paper we argue for a model-based approach to warning signals,
 and describe how this can be done in a way that best addresses these difficulties.
@@
 While no approach will be applicable to all classes of sudden transitions,
-it is certainly still useful to have an approach that detect transitions driven by
+it is certainly still useful to have an approach that detects transitions driven by
 saddle-node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.
@@
 must further assume that a changing parameter has brought the system closer to the bifurcation.
-This assumption excludes at least three alternative explanations for the transition.
+This assumption excludes at least three alternative explanations for the transition in system behavior.
 The first possibility is that a large perturbation of the system state
-has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
+has moved the system into the alternative basin of attraction~\citep{Scheffer2001}.
 This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
-(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
+(An example might be a sudden dramatic increase in fishing effort that pushes a harvested population past a threshold.)
 The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
@@
 the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
-A basic difficulty arises from the need to assume a time-series is ergodic: 
+A basic difficulty arises from the need to assume a time-series is \emph{ergodic}: 
 that averaging over time is equivalent to averaging over replicate realizations,
@@
 As true replicates are seldom available in systems for which developing warning signals would be most desirable,
-these methods average across a single replicate using a moving window in time.
+typical methods average across a single replicate using a moving window in time.
 The selection of the size of this window and whether and by how much to overlap consecutive windows
@@
 much can be accomplished by framing the implicit assumptions into equations.
-\citet{Lade2012} introduces the idea of generalized models for early warning signals, and
+\citet{Lade2012} introduce the idea of generalized models for early warning signals, and
 \citet{Kuehn2011} presents normal forms for bifurcation processes that can give rise to critical transitions.
@@
 \subsection*{Models}
+In the neighborhood of a bifurcation a system can be transformed into its \emph{normal form}
+by a change of variables to facilitate analysis ~\citep{Guckenheimer1983}.
 The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
@@
 Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
-the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
+the system near the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
 We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:
@@
 a Markov process, see~\citet{Kampen2007a} or~\citet{Black2012}. 
+The square root could be removed for an external noise process, such as environmental noise.
+In practice it will be difficult to descriminate between the square root and linear scaling
+in these applications, since the average value of the state changes little before the bifurcation.
+
 %It is also possible to derive this same expression from the master equation
@@
 We assume a gradual, monotonic change which we approximate to first order:
+
 \begin{equation}
@@
 \end{equation}
+
 Detecting accelerating or otherwise nonlinear approaches to the bifurcation will generally require more power.
-When the system is stable, $r_t$ is constant ($m=0$) and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
+When the underlying system is not changing, $r_t$ is constant ($m=0$) and Equation~\eqref{LSN} will reduce to a simple Ornstein-Uhlenbeck process,
+
 \begin{equation}
@@
 \end{equation}
+
 This is the continuous time analog of the first-order autoregressive model considered as a null model elsewhere~\citep[\emph{e.g.}][]{Dakos2008, Guttal2008a}.
@@
 The probability \(P(X|M) \) of the data \(X\) given the model \(M\) is the product of the probability of observing each point in the time series given the previous point and the length of the interval,
+
 \begin{equation}
@@
 By estimating the stable and critical transition models from the data,
-and simulate 500 replicate data sets under each as in the analysis above,
+and simulating 500 replicate data sets under each as in the analysis above,
 we can then calculate the warning signals statistic over a sliding window
@@
 \section{Acknowledgments}
-We would like to thank S. Schreiber, M. Holyoak, M. Baskett, A. Perkins and N. Ross for comments on the manuscript. 
+We would like to thank S. Schreiber, M. Holyoak, M. Baskett, A. Perkins, N. 
+Ross, M. Holden, and two anonymous reviewers for comments on the manuscript. 
 This research was supported by funding from NSF Grant EF 0742674 

### 2012-04-23 fallacy.tex
minor edits

@@
 these shifts, which may give us precious time to react and avert them~\citep{Scheffer2009, Lenton2011}.
+
+% Explain Critical Slowing down, variance, autocorrelation, tau.
+
+% Saddle-node bifurcation
+\emph{Add more context \dots }
+
 Historical examples of sudden transitions taken from the paleoclimate record
@@
 noise-induced transitions for which early warning signs will not exist~\citep{Ditlevsen2010, Lenton2011}.
-
-% Explain variance, autocorrelation, tau.
-
-\emph{Add more context \dots }
 
@@
  \section{Discussion}
-At first it seems tempting to argue that this bias is in fact valuable -- each
+
+%We have shown ..
+
+It seems tempting to argue that this bias is in fact valuable -- each
 of these systems did indeed collapse, so the fact that they show increased 
@@
 
-Systems that crash by chance excursion beyond the tipping point
+% Delete this paragraph?
+The system simulated in our examples is governed by a birth-death process --
+a discrete Markov process that can in any instant only change by a single 
+birth or single death.  At each moment in time, there is some probability 
+that either event will occur next.  The system near its stable equilibrium
+can be well approximated as a Gaussian process, where consecutive random 
+draws buffet it above and below the stable point. While these random draws
+vary in magnitude, recall that this is simply an approximation to the 
+underlying process where events are all of the same magnitude but vary
+in their frequency following a state-dependent Poisson process.  This 
+means that what looks like a large random draw from the Gaussian distribution
+really corresponds to a series of more rapid events with some excess of 
+birth or deaths.
 
-Systems collapsing by chance do so from a string of unexpected events
+The farther the sytem gets from that stable point (while still 
+remaining in the local region), the more likely it is to draw a random step 
+that returns it towards the stable point, like a spring that is stretched 
+further and further, but there is some probability that it will move further
+still. Systems that manage to extend this random walk far enough to cross the
+tipping point must do so rather quickly -- with a string of events moving 
+closer and closer to the edge.  This pattern, clearly visible before the 
+crashes in each of the examples in Figure 1, produces a string of observations
+that appear more highly autocorrelated (if we are sampling the system
+frequently enought to catch the excursion at all) than we observe in the rest
+of the fluctuations around the equilbrium.  Yet this autocorrelation comes from
+a chance trajectory moving quickly \emph{away} from the stable state, not from
+the critical slowing down pattern in the return times to the stable state
+which precede a saddle-node bifurcation and motivate the early warning signal.
 
-Discuss the pattern of a system crashing by chance -- why this looks like increased autocorrelation and increased variance.
+This longer than expected excursion results in a higher than expected variance
+in that window as well. Both variance and autorcorrelation are calculated using
+a moving window over the time-series, which allows the method to pick out a 
+pattern of change as the window moves along the sequence. If this chance excursion 
+that precedes the crash happens to fill a significant part of the moving window,
+the resulting pattern will tend to show an increase in autocorrelation or variance.
+If the chance excursion is relatively rapid compared to the frequency at which
+the system is observed (spacing of the data) or the width of the moving window, 
+the excursion may not significantly alter the general pattern.  In this way, 
+some of the events in which a crash is observed will appear to present these
+statistical patterns of increased variance or autocorrelation without being
+harbingers of approaching critical transitions.
 
-Discuss why this cannt be considered a successful detection.
-
-These are not more likely to crash than another system.
-
-Not a good test of the method's ability to detect bifurcations.
+Because we have selected our data by conditioning
+on the fact that they have experienced such a crash, these excursions appear
+more frequently than we would expect by chance in the observations of a given
+stable system over some finite time window.  That biased selection is the
+heart of the Prosecutor's Fallacy.  In these cases, the presence of increasing
+autocorrelation or increasing variance does not represent the critical slowing
+down process.
 

### 2012-04-26 ReplyToReviewers.md
submission copies

@@
 
-### Comments to the Author
+Comments to the Author
+
 I think that this is one of the most important papers yet written in what is emerging as an area of great interest.
@@
 
-_We thank the reviewer for this assessment, and we have done our best to address this important issue, clarifying definitions, expanding the explanation and making other small changes that should make this more accessible._
+_We thank the reviewer for this assessment, and we have done our best to address this important issue, clarifying definitions, expanding the explanation and making other small changes that should make this more accessible.
+In an effort to make these technical methods both accessible and understandable to the broad audience of Interface, we  have also provided both step-by-step tutorials and software to implement this approach, available from the link specified at the end of the manuscript._
 
@@
 
-### Small points:  
+Small points:
+
 - Line 25:  lead should be led
@@
 Referee: 2
+==========
+
 Comments to the Author
+
 I am pleased to recommend this manuscript for publication in Proceedings of the Royal Society Interface.

### 2012-05-14 fallacy.tex
Alan's edits from 2012-05-08

@@
 \section{Introduction}
+
+\subsection*{Catastrophic Transitions}
+
+Catastrophic transitions or tipping points, where a complex system 
+shifts suddenly from one state to another, have been implicated in
+a wide array of ecological and global climate systems such as lake
+ecosystems~\citep{Carpenter2011}, coral reefs~\citep{Mumby2007},
+desertification~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and 
+tropical forests~\citep{Hirota2011}.  Recent research has begun to
+identify statistical patterns commonly associated with these sudden
+catastrophic transitions which could be used as an \emph{early warning sign}
+to identify an approaching tipping point, buying precious time 
+to react and avert them~\citep{Scheffer2009, Lenton2011}.
+
+An array of statistical patterns associated with tipping point phenomena has
+been suggested for the detection of early warning signals associated with
+such sudden transitions.  Two of the most commonly used are a pattern of
+increasing variance~\citep{Carpenter2006} and a pattern of increasing 
+autocorrelation~\citep{vanNes2007}, which have been tested in both experimental
+manipulation~\citep{Drake2010, Carpenter2011, Veraart2011} and historical 
+observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}
+
+\subsection*{Testing patterns on historical data}
+
+Historical examples of sudden transitions taken from the paleoclimate record
+provide an important way to test and evaluate potential leading indicator 
+methods, and have been widely used for this purpose 
+\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.
+Similarly, it has been suggested that data gathered from ecological systems such
+as lakes that were monitored before the collapsed, or grasslands subjected to
+overgrazing, could contain data that could help reveal when similar systems 
+are approaching a tipping point~\citet{Inman2011}.  This is a silver lining to
+the clouds of past catastrophes -- we may be able to learn from our mistakes.
+
+Yet in testing methods for early warning signals against historical examples of
+transitions, we must be careful to avoid statistical mistakes that arise when
+we select examples conditional on having exhibited a sudden transition.  Such
+sudden transitions can occur for many reasons that do not involve the 
+bifurcation processes on which most early warning indicators are predicated
+~\citep{Hastings2010}, including purely noise-induced transitions for which 
+early warning signs will not exist~\citep{Ditlevsen2010, Lenton2011}. Treating 
+such conditionally selected historical examples as if they were randomly 
+sampled representatives of systems that have experienced a saddle-node 
+bifurcation is an example of the statistical error commonly known as the
+\emph{Prosecutor's Fallacy.}
+
+\subsection*{The Prosecutor's Fallacy}
+
 In a criminal trial, the prosecution may introduce associative evidence such as 
@@
 probability conditional on the manner in which the suspect was identified in 
-the first place.  If the suspect has been selected partly on the basis of belonging
-to demographic in which the characteristic is much more common than in the 
-population as a whole, these odds can be misleading.  Failing to recognize 
-these conditional probabilities leads to the ``Prosecutor's Fallacy'' -- the argument
-that the suspect is the perpetrator on the basis of the statistical probability 
-of a match without currectly identifying the baseline probability for such a match
-\citep{Thompson1987}.
-
-Statisical patterns associated with sudden, catastrophic transitions or ``tipping points''
-could play an important role as early warning signs or leading indicators of
-these shifts, which may give us precious time to react and avert them~\citep{Scheffer2009, Lenton2011}.
-
-% Explain Critical Slowing down, variance, autocorrelation, tau.
-
-% Saddle-node bifurcation
-\emph{Add more context \dots }
-
-Historical examples of sudden transitions taken from the paleoclimate record
-provide an important way to test and evaluate potential leading indicator methods,
-and have been widely used for this purpose~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010} 
-Similarly, it has been suggested that data gathered on ecological systems such as lakes
-that were monitored before the collapsed, or grasslands subjected to overgrazing,
-could contain data that could help reveal when similar systems are approaching 
-a tipping point~\citet{Inman2011}.  This is a silver lining to the clouds of past
-catastrophes -- we may be able to learn from our mistakes.
-
-In testing methods for early warning signals against historical examples of transitions,
-we must be careful to avoid falling into the Prosecutor's Fallacy.  Sudden transitions
-can occur for many reasons that do not involve the bifurcation proccesses on which
-most early warning indicators are predicated~\citep{Hastings2010}, including purely
-noise-induced transitions for which early warning signs will not exist~\citep{Ditlevsen2010, Lenton2011}.
+the first place.  If the suspect has been selected partly on the basis of 
+belonging to demographic in which the characteristic is much more common than
+in the population as a whole, these odds can be misleading.  Failing to recognize 
+these conditional probabilities leads to the Prosecutor's Fallacy -- the 
+argument that the suspect is the perpetrator on the basis of the statistical 
+probability of a match without correctly identifying the baseline probability 
+for such a match \citep{Thompson1987}.
 
 \section{Methods and Results}
+We approach the problem by simulating a system that cannot undergo the saddle
+node bifurcation that underlies the usual early warning signal indicators.  
+This simulation approach is the only way to determine whether examining 
+historical events is a good way to test the utility of these indicators.
 We simulated 1000 replicates of an individual-based birth-death process
-with an allee threshold.  Above the allee threshold the population returns
+with an Allee threshold.  Above the Allee threshold the population returns
 to a positive equilibrium size.  Below the threshold the population decreases
@@
 The simulation starts from the positive equilibrium population size.
-Though the chance of a transition across the allee threshold in any 
-given timestep are small, given enough time this system will eventually
+Though the chance of a transition across the Allee threshold in any 
+given time step are small, given enough time this system will eventually
 experience such a rare event driving the population extinct.  We evolved
-each replicate 50000 timesteps, sampling the system every 50 timesteps.  
-In this time window 155 of the 1000 replicates experience population collapse.  
+each replicate 50000 time steps, sampling the system every 50 time steps.  
+In this time window 266 of the 1000 replicates experience population collapse.  
 Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
@@
     \end{center}
-    \caption{Example trajectories for nine of the 155 replicates that 
+    \caption{Example trajectories for nine of the 266 replicates that 
              experience a critical transition by chance during the course
@@
 We then selected a window around each system that ended just before the
-collapse, while the population values were still above the allee threshold.
+collapse, while the population values were still above the Allee threshold.
 For each replicate, we calculated the most common warning signals, variance
 and autocorrelation~\citep[\emph{e.g.}][]{Carpenter2006,Dakos2008,Scheffer2009}, 
-around a moving window equal to half the length of that timeseries.
+around a moving window equal to half the length of that time series.
 
 To determine whether or not the variance or autocorrelation increased 
-over time in these 155 replicates, we computed Kendall's $\tau$ for each of the
-warning signal pattern~\citep{Dakos2008, Dakos2010}.  Positive values of $\tau$ 
+over time in these 266 replicates, we computed Kendall's $\tau$ for each of the
+warning signal pattern~\citep{Dakos2008, Dakos2011}.  Positive values of $\tau$ 
 indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
@@
   \end{center}
-  \caption{The distribution of two early warning indicators on replicates conditionally selected for having collapsed by chance in simulations.}
+  \caption{The distribution of the correlation statistic $\tau$ for two early warning indicators (variance, autocorrelation) on replicates conditionally selected for having collapsed by chance in simulations.
+Positive values of $\tau$ correspond to a pattern of an indicator increasing with time; typically taken as evidence that a system is approaching a critical transition.
+In these simulations, the pattern arises instead from the Prosecutor's fallacy of conditional selection.}
   \label{fig:indicator}
@@
 
-where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilbirium postition.  
+where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilibrium position.  
 This model emerges from the canonical form of the saddle-node bifurcation,
@@
 $r$ remains the bifurcation parameter which we imagine changing with time,
-$r_t$, then linearizing the above model in neighboorhood of
+$r_t$, then linearizing the above model in neighborhood of
 \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\), we recover Equation~\eqref{LSN}.
@@
 it is also easier to estimate reliably from the data since it describes only 
-the process near the equilibrium where the data is avialable.
+the process near the equilibrium where the data is available.
 
@@
 approaching a bifurcation, while for stable systems $m$ should be approximately zero.
-Of the 155 simulations, only six had estimates with machine precision different
+Of the 266 simulations, only six had estimates with machine precision different
 from zero, the greatest magnitude among these was $-7.1 \times 10^{-12}$,
-which would still correspond to a negilible change in $r_0$ over the 50,000 timesteps
+which would still correspond to a negligible change in $r_0$ over the 50,000 time steps
 of the simulation.  Consequently, the model-based estimation shows no 
@@
 
-%We have shown ..
+We have shown that selecting datasets conditional on having experienced sudden 
+noise-induced transitions can lead to patterns that are more likely to resemble
+early warning signals typically associated with critical slowing down before 
+a bifurcation event.  Purely noise-induced transitions are distinctly separate 
+mechanism from critical slowing-down phenomena, in which the transition occurs
+as the result of a single or series of extremely unlikely events rather than
+a slow but regular degradation of the system stability.
 
-It seems tempting to argue that this bias is in fact valuable -- each
-of these systems did indeed collapse, so the fact that they show increased 
+\cb{The discussion that follows here is aimed at giving some intuition for our results, but comes off as a bit more technically focused than the section above.
+Should it be moved into an earlier section, or do we simply need a summarizing paragraph at the end?}
+
+It seems tempting to argue that the bias towards positive detection in historical
+examples is in fact valuable -- each of these systems did indeed collapse, so the fact that they show increased 
 probability to exhibit warning signals could be taken as a successful detection.
@@
 
-% Delete this paragraph?
+\cb{ Delete this paragraph?}
 The system simulated in our examples is governed by a birth-death process --
@@
 
-The farther the sytem gets from that stable point (while still 
+The farther the system gets from that stable point (while still 
 remaining in the local region), the more likely it is to draw a random step 
@@
 that appear more highly autocorrelated (if we are sampling the system
-frequently enought to catch the excursion at all) than we observe in the rest
-of the fluctuations around the equilbrium.  Yet this autocorrelation comes from
+frequently enough to catch the excursion at all) than we observe in the rest
+of the fluctuations around the equilibrium.  Yet this autocorrelation comes from
 a chance trajectory moving quickly \emph{away} from the stable state, not from
@@
 This longer than expected excursion results in a higher than expected variance
-in that window as well. Both variance and autorcorrelation are calculated using
+in that window as well. Both variance and autocorrelation are calculated using
 a moving window over the time-series, which allows the method to pick out a 
@@
 
+The problem we highlight ultimately stems from the difficulty of having only a
+single realization with which to examine a complex problem.  The only way to deal
+with this problem is through replication, as can be done in an experimental system
+in laboratory manipulations such as~\citet{Drake2010, Veraart2011} and at the 
+scale of whole lake ecosystems in~\citet{Carpenter2011}.  Historical data will 
+not suffice, more such replicated experiments are needed to test these patterns.
+
  \section{Acknowledgements}

### 2012-05-15 fallacy.tex
started fixing Alex edits

@@
   Early warning signals have been proposed to forecast the possibility of a 
-  critical transition, such as the eutriphication of a lake, desertification
+  critical transition, such as the eutrophication of a lake, desertification
   of savanna, the collapse of a coral reef, the end of a glacial period, or 
   the onset of epilepsy in the brain.  While such transitions may be difficult
-  to study in the laboratory setting and unethical to introduce on the a larger
+  to study in the laboratory setting and unethical to introduce on the larger
   natural scales where they may occur, it has often been observed that history
@@
   fact that we have observed a critical transition and those systems for which
-  we wish to forecast the approach of a transition. This difference arising
-  from conditional probabilities is often known as the the Prosecutor's 
+  we wish to forecast the approach of a transition. This difference arises
+  by conditionally selecting systems known to experience a transition of some 
+  sort and failing to account for this selection step -- a statistical error 
+  often known as the Prosecutor's
   Fallacy.  We illustrate through simulation how examples of systems that have 
@@
   test and validate methods for early warning signals, greater care must be taken
-  to avoid this fallacy.
+  to avoid this fallacy.  Groundbreaking experimental studies with replicates
+  avoid this pitfall, and more of such studies are needed to provide data
+  suitable for evaluating methods for early warning.
 
@@
 Similarly, it has been suggested that data gathered from ecological systems such
-as lakes that were monitored before the collapsed, or grasslands subjected to
+as lakes that were monitored before they collapsed, or grasslands subjected to
 overgrazing, could contain data that could help reveal when similar systems 
-are approaching a tipping point~\citet{Inman2011}.  This is a silver lining to
+are approaching a tipping point~\citep{Inman2011}.  This is a silver lining to
 the clouds of past catastrophes -- we may be able to learn from our mistakes.
@@
 bifurcation processes on which most early warning indicators are predicated
-~\citep{Hastings2010}, including purely noise-induced transitions for which 
-early warning signs will not exist~\citep{Ditlevsen2010, Lenton2011}. Treating 
-such conditionally selected historical examples as if they were randomly 
-sampled representatives of systems that have experienced a saddle-node 
-bifurcation is an example of the statistical error commonly known as the
+~\citep{Hastings2010}, including purely noise-induced transitions for proposed 
+early warning patterns are not anticipated~\citep{Ditlevsen2010, Lenton2011}. 
+Ignoring the potential biases introduced by conditionally selecting from 
+historical records is an example of the statistical error commonly known as the
 \emph{Prosecutor's Fallacy.}
@@
 for instance, the frequency of individuals in the population with the matching
-blood type.  Such arguments may be misleading when they fail to reflect the 
-probability conditional on the manner in which the suspect was identified in 
+blood type.  Such arguments may be misleading when they fail to account for  
+the manner in which the suspect was identified in 
 the first place.  If the suspect has been selected partly on the basis of 
-belonging to demographic in which the characteristic is much more common than
+belonging to a demographic in which the discriminating characteristic is much more common than
 in the population as a whole, these odds can be misleading.  Failing to recognize 
-these conditional probabilities leads to the Prosecutor's Fallacy -- the 
+these conditional probabilities leads to the Prosecutor's Fallacy: the 
 argument that the suspect is the perpetrator on the basis of the statistical 
@@
 for such a match \citep{Thompson1987}.
+
+% Consider an equation
 
@@
 historical events is a good way to test the utility of these indicators.
-We simulated 1000 replicates of an individual-based birth-death process
-with an Allee threshold.  Above the Allee threshold the population returns
-to a positive equilibrium size.  Below the threshold the population decreases
-to zero. The simulation uses the Gillespie algorithm to provide an exact 
-implementation of the master equation of birth-death process,
+We simulated 1000 replicates of a stochastic individual-based birth-death process
+with an Allee threshold~\citep{Courchamp2008}, which arises from positive fitness
+effects at low densities.  Above the Allee threshold the population returns
+to a positive equilibrium size, while below the threshold the population decreases
+to zero in expectation. The simulation uses the Gillespie algorithm to provide an exact 
+implementation of the master equation of the birth-death process,
 
@@
 
+a model with a linear death rate and density dependent birth rate that drives
+the Allee effect at low densities and limits growth at high densities.  The
+key feature of this model is the alternate stable states introduced by this effect;
+other functional forms for Eq.~\ref{master} could serve equally well for these
+simulations~\citep[see \emph{e.g.}][]{Scheffer2001}. 
 Though this system can be forced through a saddle-node bifurcation by
-increasing values of either $e$ or $a$, in these simulations all parameters
-are held constant and no early warning signal is anticipated.
+increasing the death rate, in these simulations all parameters
+are held constant and no bifurcation occurs.  Consequently we do not 
+anticipate an early warning signal of an approaching bifurcation.
 The simulation starts from the positive equilibrium population size.
@@
 experience such a rare event driving the population extinct.  We evolved
-each replicate 50000 time steps, sampling the system every 50 time steps.  
-In this time window 266 of the 1000 replicates experience population collapse.  
+each replicate 50,000 time steps, sampling the system every 50 time steps.  
+In this time window 266 of the 1,000 replicates experience population collapse.  
 Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
@@
 
-  \begin{figure}[H]
+  \begin{figure}
     \begin{center}
-      \includegraphics{replicate_crashes.pdf}
+      \includegraphics[width=6in]{replicate_crashes.pdf}
     \end{center}
@@
 
-We selected replicates conditional on having collapsed in the simulations.
+To test whether selecting systems that have experienced spontaneous transitions
+could bias the analysis towards false positive detection of early warning signals,
+(the Prosecutor's Fallacy) we selected replicates conditional on having collapsed
+in the simulations.
 We then selected a window around each system that ended just before the
 collapse, while the population values were still above the Allee threshold.
-For each replicate, we calculated the most common warning signals, variance
+For each replicate, we calculated the most common early warning indicators, variance
 and autocorrelation~\citep[\emph{e.g.}][]{Carpenter2006,Dakos2008,Scheffer2009}, 
@@
 
-To determine whether or not the variance or autocorrelation increased 
-over time in these 266 replicates, we computed Kendall's $\tau$ for each of the
-warning signal pattern~\citep{Dakos2008, Dakos2011}.  Positive values of $\tau$ 
+To test for the presence of a warning signal in these indicators we computed
+values of Kendall's $\tau$ for both indicators for each of the 266 replicates.
+Kendall's $tau$ takes values in $(-1, 1)$, and is frequently used to identify
+an increasing trend ($\tau > 0 )$ in early warning signals.~\citep{Dakos2008, Dakos2011}.
 indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
 values observed across these replicates is shown in Figure~\ref{fig:indicator}.
-More positive values indicate stronger trends. Assessing the statistical significance
-of a given value requires an appropriate null model, which can be difficult 
-to do~\citep{Dakos2008}. As these examples have been selected from a system
-that is not approaching a bifurcation and therefore should not exhibit any 
-early warning sign. %Even the very strong values of $\tau$ should be considered 
-%as arising from the null model.
+More positive values indicate stronger trends. The common statistical significance
+test used for Kendall's $tau$ assumes independence of the variables to achieve
+a normal distribution, this isn't satisfied for indicators coming from a time 
+series and computed with a sliding window.  For a more robust comparison, we 
+compare the distribution of this statistic from all the simulations to the 
+distribution conditioned on experiencing a chance transition to the alternative
+stable state, that is, from the replicates which are known to crash.  To avoid 
+an effect of sample size the timeseries are all chosen to be the same length.  
+As these examples have been selected from a system that is not approaching a 
+bifurcation, an indicator that does not suffer from the bias of Prosecutor's fallacy
+should have the same distribution in both cases.   Annotated, parallelized code to 
+replicate the analysis and produce the figures is provided in the appendix.
 
@@
   \begin{center}
-    \includegraphics{indicators.pdf}
+    \includegraphics[width=6in]{indicators.pdf}
   \end{center}
@@
 \end{equation}
-
 where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilibrium position.  
@@
 \end{equation}
-
 Where we introduce an explicit parameter $\theta$ to account for rescaling
@@
 \end{equation}
-
 and then consider the corresponding stochastic process where the noise is
@@
 \end{equation}
-
 $r$ remains the bifurcation parameter which we imagine changing with time,
@@
 These models can be both be fit to the data by maximum likelihood estimation
-based on the equation for the evolution of their moments, as
-in~\citet{Boettiger2012b}.
-
-Though this approximation is more general than the original Markov process,
-it is also easier to estimate reliably from the data since it describes only 
-the process near the equilibrium where the data is available.
-
-We assume that $r_t$ is varying at constant rate, according to the model
+based on the equation for the evolution of their moments, as explained
+in~\citet{Boettiger2012b}.  Though this approximation is more general than 
+the original Markov process (as other models containing a saddle-node bifurcation
+share the same approximation), it is also easier to estimate reliably from 
+the data since it describes only the process near the equilibrium where the
+data is available.  We assume that $r_t$ is varying at constant rate, 
+according to the model
 
@@
 \end{equation}
-
 and estimate $m$, $r_0$, $\theta$, and $\beta$ by maximum likelihood on each
@@
 approaching a bifurcation, while for stable systems $m$ should be approximately zero.
+
 Of the 266 simulations, only six had estimates with machine precision different
@@
 
-\cb{The discussion that follows here is aimed at giving some intuition for our results, but comes off as a bit more technically focused than the section above.
-Should it be moved into an earlier section, or do we simply need a summarizing paragraph at the end?}
-
 It seems tempting to argue that the bias towards positive detection in historical
@@
 
-\cb{ Delete this paragraph?}
-The system simulated in our examples is governed by a birth-death process --
-a discrete Markov process that can in any instant only change by a single 
-birth or single death.  At each moment in time, there is some probability 
-that either event will occur next.  The system near its stable equilibrium
-can be well approximated as a Gaussian process, where consecutive random 
-draws buffet it above and below the stable point. While these random draws
-vary in magnitude, recall that this is simply an approximation to the 
-underlying process where events are all of the same magnitude but vary
-in their frequency following a state-dependent Poisson process.  This 
-means that what looks like a large random draw from the Gaussian distribution
-really corresponds to a series of more rapid events with some excess of 
-birth or deaths.
-
-The farther the system gets from that stable point (while still 
+The farther the system gets from its stable point (while still 
 remaining in the local region), the more likely it is to draw a random step 
 that returns it towards the stable point, like a spring that is stretched 
-further and further, but there is some probability that it will move further
-still. Systems that manage to extend this random walk far enough to cross the
+further and further. Despite this, there is always some probability that it will 
+move further still, and with enough possible examples to choose from, we
+can be sure to see this happen occasionally.  
+Systems that manage to extend this random walk far enough to cross the
 tipping point must do so rather quickly -- with a string of events moving 
@@
 
+%% Comments on reproducibility?
+
  \section{Acknowledgements}

### 2012-05-16 fallacy.tex
fallacy updated

@@
 In this time window 266 of the 1,000 replicates experience population collapse.  
+To keep the examples of comperable sample size, we focus on a section of the data
+500 time points prior to the system approaching the transition.  
 Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
@@
              experience a critical transition by chance during the course
-             of the simulation.}
+             of the simulation, showing the window of 500 points prior to
+             crossing the threshold which is used in the analysis.  A wobble
+             at the unstable critical point is visible at the end of several 
+             trajectories.}
     \label{fig:replicate_crashes}
@@
 replicate the analysis and produce the figures is provided in the appendix. 
+(Currently at \href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).
 
@@
 \end{equation}
-Where we introduce an explicit parameter $\theta$ to account for rescaling
+where we introduce an explicit parameter $\theta$ to account for rescaling
 to the units of the data,
@@
 
-We have shown that selecting datasets conditional on having experienced sudden 
-noise-induced transitions can lead to patterns that are more likely to resemble
-early warning signals typically associated with critical slowing down before 
-a bifurcation event.  Purely noise-induced transitions are distinctly separate 
-mechanism from critical slowing-down phenomena, in which the transition occurs
-as the result of a single or series of extremely unlikely events rather than
-a slow but regular degradation of the system stability.
+Historical examples of systems that have undergone sudden transitions seem
+to offer an invaluable opportunity to test methods for the detection of
+early warning signals that could have forseen the collapse.  However,
+such data must be used with care, as the process of selecting systems
+already known to have collapsed opens up the possibility that these
+transitions were the result of rare chance events rather than examples
+of the more deterministic tipping point behaviour we wish to detect.
+When we think about forecasting potential collapse in a single system,
+these purely-chance transitions may be rare enough to ignore.  When we
+comb the historical record for any available data on sudden transitions,
+our odds of finding these chance transitions may be much higher.
 
-It seems tempting to argue that the bias towards positive detection in historical
-examples is in fact valuable -- each of these systems did indeed collapse, so the fact that they show increased 
-probability to exhibit warning signals could be taken as a successful detection.
-Unfortunately, this reasoning is flawed. At the moment the forecast is made,
-these systems are no more likely to crash than their counterparts from the sample
-that did not collapse, just as a fair coin is still a fair coin after a string of
-consecutive ``heads.''  Taking a closer look at the patterns involved we can
-begin to understand why common indicators such as autocorrelation and variance
-can be misleading.
+We have shown that systems which undergo such rare transitions look
+statistically different from their counterparts that do not, even though
+they are driven by the same stochastic process.  In particular, such
+conditionally selected examples are more likely to show signs associated
+with an early warning of an approaching tipping point, such as increasing
+variance or increasing autocorrelation, as measured by Kendall's $\tau$.
+This increases the risk of false positives -- cases in which a warning
+signal being tested appears to have successfully detected an oncoming
+tipping point, when in fact the example is falsely accused, coming
+instead from a stable system. Figure 2 shows the extent of this bias,
+where many of the chance crashes show values of $\tau$ that are 
+significantly larger than those observed in the otherwise identical
+replicates that did not experience a chance transition.
 
-The farther the system gets from its stable point (while still 
-remaining in the local region), the more likely it is to draw a random step 
-that returns it towards the stable point, like a spring that is stretched 
-further and further. Despite this, there is always some probability that it will 
-move further still, and with enough possible examples to choose from, we
-can be sure to see this happen occasionally.  
-Systems that manage to extend this random walk far enough to cross the
-tipping point must do so rather quickly -- with a string of events moving 
-closer and closer to the edge.  This pattern, clearly visible before the 
-crashes in each of the examples in Figure 1, produces a string of observations
-that appear more highly autocorrelated (if we are sampling the system
-frequently enough to catch the excursion at all) than we observe in the rest
-of the fluctuations around the equilibrium.  Yet this autocorrelation comes from
-a chance trajectory moving quickly \emph{away} from the stable state, not from
-the critical slowing down pattern in the return times to the stable state
-which precede a saddle-node bifurcation and motivate the early warning signal.
+\subsection{Chance transitions are false positives for early warning signals}
 
-This longer than expected excursion results in a higher than expected variance
-in that window as well. Both variance and autocorrelation are calculated using
-a moving window over the time-series, which allows the method to pick out a 
-pattern of change as the window moves along the sequence. If this chance excursion 
-that precedes the crash happens to fill a significant part of the moving window,
-the resulting pattern will tend to show an increase in autocorrelation or variance.
-If the chance excursion is relatively rapid compared to the frequency at which
-the system is observed (spacing of the data) or the width of the moving window, 
-the excursion may not significantly alter the general pattern.  In this way, 
-some of the events in which a crash is observed will appear to present these
-statistical patterns of increased variance or autocorrelation without being
-harbingers of approaching critical transitions.
+It seems tempting to argue that the bias towards positive detection
+in historical examples is valuable -- each of these systems did
+indeed collapse, so the fact that they show increased probability
+to exhibit warning signals could be taken as a successful detection.
+Unfortunately this is not the case. At the moment the forecast is made,
+these systems are no more likely to crash than their counterparts from
+the sample that did not collapse, just as a fair coin is still a fair
+coin after a string of consecutive ``heads.''  Taking a closer look at
+the patterns involved we can begin to understand why common indicators
+such as autocorrelation and variance can be misleading.
 
-Because we have selected our data by conditioning
-on the fact that they have experienced such a crash, these excursions appear
-more frequently than we would expect by chance in the observations of a given
-stable system over some finite time window.  That biased selection is the
-heart of the Prosecutor's Fallacy.  In these cases, the presence of increasing
-autocorrelation or increasing variance does not represent the critical slowing
-down process.
+The farther the system gets from its stable point (while still remaining
+in the local region), the more likely it is to draw a random step that
+returns it towards the stable point, like a spring that is stretched
+further and further. Despite this, there is always some probability that
+it will move further still, and with enough possible examples to choose
+from, we can be sure to see this happen occasionally.  Systems that manage
+to extend this random walk far enough to cross the tipping point must do
+so rather quickly -- with a string of events moving closer and closer to
+the edge.  This pattern, clearly visible before the crashes in each of
+the examples in Figure 1, produces a string of observations that appear
+more highly autocorrelated (if we are sampling the system frequently
+enough to catch the excursion at all) than we observe in the rest of the
+fluctuations around the equilibrium.  Yet this autocorrelation comes from
+a chance trajectory moving quickly \emph{away} from the stable state,
+not from the critical slowing down pattern in the return times to the
+stable state which precede a saddle-node bifurcation and motivate the
+early warning signal.
 
-The problem we highlight ultimately stems from the difficulty of having only a
-single realization with which to examine a complex problem.  The only way to deal
-with this problem is through replication, as can be done in an experimental system
-in laboratory manipulations such as~\citet{Drake2010, Veraart2011} and at the 
-scale of whole lake ecosystems in~\citet{Carpenter2011}.  Historical data will 
-not suffice, more such replicated experiments are needed to test these patterns.
+This longer than expected excursion results in a higher than expected
+variance in that window as well. Both variance and autocorrelation are
+calculated using a moving window over the time-series, which allows
+the method to pick out a pattern of change as the window moves along
+the sequence. If this chance excursion that precedes the crash happens
+to fill a significant part of the moving window, the resulting pattern
+will tend to show an increase in autocorrelation or variance.  If the
+chance excursion is relatively rapid compared to the frequency at which
+the system is observed (spacing of the data) or the width of the moving
+window, the excursion may not significantly alter the general pattern.
+In this way, some of the events in which a crash is observed will
+appear to present these statistical patterns of increased variance
+or autocorrelation without being harbingers of approaching critical
+transitions.
 
-%% Comments on reproducibility?
+\subsection{Comparing to the model-based method}
+
+In our experiment, the model-based estimate of early warning signals
+appears more robust than the summary statistics, producing the same
+estimates on both the conditionally selected replicates as on a random 
+sample of the replicates.  This is a consequence of the more rigid 
+specifications that come with a model-based approach -- the pattern
+expected is less general than any increase in variance or autocorrelation,
+but instead must be one that matches its approximation of the saddle-node
+bifurcation. This observation highlights the difference between the pattern
+driving the false positive trends in increasing variance and increasing
+autocorrelation and the pattern anticipated in the saddle-node model. This
+should not however be taken as evidence that the model-based approach is 
+immune to the bias of the Prosecutor's Fallacy.
+
+The problem we highlight ultimately stems from the difficulty of having
+only a single realization with which to examine a complex problem.
+The only way to deal with this problem is through replication, as
+can be done in an experimental system in laboratory manipulations
+such as~\citet{Drake2010, Veraart2011} and at the scale of whole lake
+ecosystems in~\citet{Carpenter2011}.  Historical data will not suffice,
+more such replicated experiments are needed to test these patterns.
 

### 2012-05-16 fallacy.tex
fallacy updated

@@
 \section{Methods and Results}
-We approach the problem by simulating a system that cannot undergo the saddle
-node bifurcation that underlies the usual early warning signal indicators.  
-This simulation approach is the only way to determine whether examining 
-historical events is a good way to test the utility of these indicators.
-We simulated 1000 replicates of a stochastic individual-based birth-death process
-with an Allee threshold~\citep{Courchamp2008}, which arises from positive fitness
-effects at low densities.  Above the Allee threshold the population returns
-to a positive equilibrium size, while below the threshold the population decreases
-to zero in expectation. The simulation uses the Gillespie algorithm to provide an exact 
-implementation of the master equation of the birth-death process,
+We approach the problem by simulating a system that cannot undergo the
+saddle node bifurcation that underlies the usual early warning signal
+indicators.  This simulation approach is the only way to determine whether
+examining historical events is a good way to test the utility of these
+indicators.  We simulated 1000 replicates of a stochastic individual-based
+birth-death process with an Allee threshold~\citep{Courchamp2008}, which
+arises from positive fitness effects at low densities.  Above the Allee
+threshold the population returns to a positive equilibrium size, while
+below the threshold the population decreases to zero in expectation. The
+simulation uses the Gillespie algorithm to provide an exact implementation
+of the master equation of the birth-death process,
 
@@
 \end{align}
-
-a model with a linear death rate and density dependent birth rate that drives
-the Allee effect at low densities and limits growth at high densities.  The
-key feature of this model is the alternate stable states introduced by this effect;
-other functional forms for Eq.~\ref{master} could serve equally well for these
-simulations~\citep[see \emph{e.g.}][]{Scheffer2001}. 
-Though this system can be forced through a saddle-node bifurcation by
-increasing the death rate, in these simulations all parameters
-are held constant and no bifurcation occurs.  Consequently we do not 
-anticipate an early warning signal of an approaching bifurcation.
-The simulation starts from the positive equilibrium population size.
-Though the chance of a transition across the Allee threshold in any 
-given time step are small, given enough time this system will eventually
-experience such a rare event driving the population extinct.  We evolved
-each replicate 50,000 time steps, sampling the system every 50 time steps.  
-In this time window 266 of the 1,000 replicates experience population collapse.  
-To keep the examples of comperable sample size, we focus on a section of the data
-500 time points prior to the system approaching the transition.  
-Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
-as an illustration.
+a model with a linear death rate and density dependent birth rate
+that drives the Allee effect at low densities and limits growth
+at high densities.  The key feature of this model is the alternate
+stable states introduced by this effect; other functional forms for
+Eq.~\ref{master} could serve equally well for these simulations~\citep[see
+\emph{e.g.}][]{Scheffer2001}.  Though this system can be forced
+through a saddle-node bifurcation by increasing the death rate, in
+these simulations all parameters are held constant and no bifurcation
+occurs.  Consequently we do not anticipate an early warning signal of
+an approaching bifurcation.  The simulation starts from the positive
+equilibrium population size.  Though the chance of a transition across
+the Allee threshold in any given time step are small, given enough
+time this system will eventually experience such a rare event driving
+the population extinct.  We evolved each replicate 50,000 time steps,
+sampling the system every 50 time steps.  In this time window 266 of the
+1,000 replicates experience population collapse.  To keep the examples of
+comparable sample size, we focus on a section of the data 500 time points
+prior to the system approaching the transition.  Nine of these replicates
+can be seen in Figure~\ref{fig:replicate_crashes}, as an illustration.
 
@@
 
-To test for the presence of a warning signal in these indicators we computed
-values of Kendall's $\tau$ for both indicators for each of the 266 replicates.
-Kendall's $tau$ takes values in $(-1, 1)$, and is frequently used to identify
-an increasing trend ($\tau > 0 )$ in early warning signals.~\citep{Dakos2008, Dakos2011}.
-indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
-values observed across these replicates is shown in Figure~\ref{fig:indicator}.
-More positive values indicate stronger trends. The common statistical significance
-test used for Kendall's $tau$ assumes independence of the variables to achieve
-a normal distribution, this isn't satisfied for indicators coming from a time 
-series and computed with a sliding window.  For a more robust comparison, we 
-compare the distribution of this statistic from all the simulations to the 
-distribution conditioned on experiencing a chance transition to the alternative
-stable state, that is, from the replicates which are known to crash.  To avoid 
-an effect of sample size the timeseries are all chosen to be the same length.  
-As these examples have been selected from a system that is not approaching a 
-bifurcation, an indicator that does not suffer from the bias of Prosecutor's fallacy
-should have the same distribution in both cases.   Annotated, parallelized code to 
-replicate the analysis and produce the figures is provided in the appendix. 
-(Currently at \href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).
+To test for the presence of a warning signal in these indicators we
+computed values of Kendall's $\tau$ for both indicators for each of
+the 266 replicates.  Kendall's $tau$ takes values in $(-1, 1)$, and is
+frequently used to identify an increasing trend ($\tau > 0 )$ in early
+warning signals.~\citep{Dakos2008, Dakos2011}.  The distribution
+of $\tau$ values observed across these replicates is shown in
+Figure~\ref{fig:indicator}.  More positive values indicate stronger
+trends. The common statistical significance test used for Kendall's $tau$
+assumes independence of the variables to achieve a normal distribution,
+this isn't satisfied for indicators coming from a time series and
+computed with a sliding window.  For a more robust comparison, we
+compare the distribution of this statistic from all the simulations to
+the distribution conditioned on experiencing a chance transition to the
+alternative stable state, that is, from the replicates which are known
+to crash.  To avoid an effect of sample size the time series are all chosen
+to be the same length.  As these examples have been selected from a system
+that is not approaching a bifurcation, an indicator that does not suffer
+from the bias of Prosecutor's fallacy should have the same distribution
+in both cases.   Annotated, parallelized code to replicate the analysis
+and produce the figures is provided in the appendix.  (Currently at
+\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).
 
-\begin{figure}[H]
+\begin{figure}
   \begin{center}
@@
   \end{center}
-  \caption{The distribution of the correlation statistic $\tau$ for two early warning indicators (variance, autocorrelation) on replicates conditionally selected for having collapsed by chance in simulations.
-Positive values of $\tau$ correspond to a pattern of an indicator increasing with time; typically taken as evidence that a system is approaching a critical transition.
-In these simulations, the pattern arises instead from the Prosecutor's fallacy of conditional selection.}
+  \caption{The distribution of the correlation statistic $\tau$ for two
+  early warning indicators (variance, autocorrelation) on replicates
+  conditionally selected for having collapsed by chance in simulations
+  is shown in gray bars.  Solid lines indicate the estimated density of
+  the statistic from a random sample of the simulations (not conditional
+  on observing a transition). Positive values of $\tau$ correspond to
+  a pattern of an indicator increasing with time; typically taken as
+  evidence that a system is approaching a critical transition.  In these
+  simulations, the pattern arises instead from the Prosecutor's fallacy
+  of conditional selection.}
   \label{fig:indicator}
@@
 to offer an invaluable opportunity to test methods for the detection of
-early warning signals that could have forseen the collapse.  However,
+early warning signals that could have foreseen the collapse.  However,
 such data must be used with care, as the process of selecting systems
@@
 but instead must be one that matches its approximation of the saddle-node
-bifurcation. This observation highlights the difference between the pattern
-driving the false positive trends in increasing variance and increasing
-autocorrelation and the pattern anticipated in the saddle-node model. This
-should not however be taken as evidence that the model-based approach is 
-immune to the bias of the Prosecutor's Fallacy.
+bifurcation. This observation highlights the difference between the
+pattern driving the false positive trends in increasing variance and
+increasing autocorrelation and the pattern anticipated in the saddle-node
+model. This should not however be taken as evidence that the model-based
+approach is immune to the bias of the Prosecutor's Fallacy.
 
@@
 more such replicated experiments are needed to test these patterns.
+Experimental procedures avoid the hazard of the Prosecutor's fallacy by
+generating a complete sample of replicates, rather than selecting a subset
+of cases some larger historical sample where the risk that if selecting a
+chance transition cannot be ignored.  Further, we note that the influence
+of the Prosecutors Fallacy need not be restricted to the context of early
+warning signals, but may arise any time we search historical data for
+examples containing some feature of interest and ignore the statistical
+effect that selection process introduces in our data.
 
@@
 This research was supported by funding from NSF Grant EF 0742674 to AH
-and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308 to CB. 
+and a Computational Sciences Graduate Fellowship from the Department of
+Energy grant DE-FG02-97ER25308 to CB.
+
  \section{References}%bibliography

### 2012-05-30 fallacy.tex
edits

@@
 
-  Early warning signals have been proposed to forecast the possibility of a 
-  critical transition, such as the eutrophication of a lake, desertification
-  of savanna, the collapse of a coral reef, the end of a glacial period, or 
-  the onset of epilepsy in the brain.  While such transitions may be difficult
-  to study in the laboratory setting and unethical to introduce on the larger
-  natural scales where they may occur, it has often been observed that history
-  has already performed the experiment for us, time and again.  Here we point 
-  out a critical difference between selecting systems for study based on the 
-  fact that we have observed a critical transition and those systems for which
-  we wish to forecast the approach of a transition. This difference arises
-  by conditionally selecting systems known to experience a transition of some 
-  sort and failing to account for this selection step -- a statistical error 
-  often known as the Prosecutor's
-  Fallacy.  We illustrate through simulation how examples of systems that have 
-  experienced transitions purely by chance show patterns associated with 
-  early warning signals more often than we might expect in systems that have not
-  been conditionally selected.  We further highlight how a model-based approach
-  is less subject to this bias.  In using historical examples of collapse to 
-  test and validate methods for early warning signals, greater care must be taken
-  to avoid this fallacy.  Groundbreaking experimental studies with replicates
-  avoid this pitfall, and more of such studies are needed to provide data
-  suitable for evaluating methods for early warning.
+  Early warning signals have been proposed to forecast the possibility
+  of a critical transition, such as the eutrophication of a lake,
+  the collapse of a coral reef, or the end
+  of a glacial period.  Because such transitions often occur on scales that
+  can be difficult to approach by experimental manipulation, research
+  has often relied on historical observations as a source of natural experiments.
+  Here we examine a critical difference between selecting systems for
+  study based on the fact that we have observed a critical transition
+  and those systems for which we wish to forecast the approach of a
+  transition. This difference arises by conditionally selecting systems
+  known to experience a transition of some sort and failing to account
+  for this bias this introduces -- a statistical error often known as the
+  Prosecutor's Fallacy.  By analyzing simulated systems that have
+  experienced transitions purely by chance, we reveal an elevated rate
+  of false postives in common warning signal statistics. We further
+  demonstrate a model-based approach which is less subject to this bias
+  than these more commonly used summary statistics.
+  In using historical examples of collapse to test and validate methods
+  for early warning signals, care must be taken to avoid this fallacy,
+  either through the use methods less sensitive to this effect or ideally
+  by greater reliance on Experimental studies with replicates avoid this 
+  pitfall entirely.  More such studies are needed to provide data suitable
+  for evaluating methods for early warning.
 
@@
 
-\subsection*{Catastrophic Transitions}
+%\subsection*{Catastrophic Transitions}
 
@@
 ecosystems~\citep{Carpenter2011}, coral reefs~\citep{Mumby2007},
-desertification~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and 
-tropical forests~\citep{Hirota2011}.  Recent research has begun to
-identify statistical patterns commonly associated with these sudden
-catastrophic transitions which could be used as an \emph{early warning sign}
-to identify an approaching tipping point, buying precious time 
-to react and avert them~\citep{Scheffer2009, Lenton2011}.
-
-An array of statistical patterns associated with tipping point phenomena has
-been suggested for the detection of early warning signals associated with
-such sudden transitions.  Two of the most commonly used are a pattern of
-increasing variance~\citep{Carpenter2006} and a pattern of increasing 
-autocorrelation~\citep{vanNes2007}, which have been tested in both experimental
-manipulation~\citep{Drake2010, Carpenter2011, Veraart2011} and historical 
-observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}
+savannah~\citep{Kefi2007}, fisheries~\citep{Berkes2006}, and tropical
+forests~\citep{Hirota2011}.  Recent research has begun to identify
+statistical patterns commonly associated with these sudden catastrophic
+transitions which could be used as an \emph{early warning sign} to
+identify an approaching tipping point, which might provide managers time
+to react to and avert an undesirable state shift~\citep{Scheffer2009,
+Lenton2011}.  An array of statistical patterns associated with tipping
+point phenomena has been suggested for the detection of early warning
+signals associated with such sudden transitions.  Two of the most commonly
+used are a pattern of increasing variance~\citep{Carpenter2006} and a
+pattern of increasing autocorrelation~\citep{vanNes2007}, which have
+been tested in both experimental manipulation~\citep{Drake2010,
+Carpenter2011, Veraart2011} and historical
+observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.
 
@@
 
-Historical examples of sudden transitions taken from the paleoclimate record
-provide an important way to test and evaluate potential leading indicator 
-methods, and have been widely used for this purpose 
+Historical examples of sudden transitions taken from the paleoclimate
+record provide an important way to test and evaluate potential
+leading indicator methods, and have been widely used for this purpose
 \citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.
-Similarly, it has been suggested that data gathered from ecological systems such
-as lakes that were monitored before they collapsed, or grasslands subjected to
-overgrazing, could contain data that could help reveal when similar systems 
-are approaching a tipping point~\citep{Inman2011}.  This is a silver lining to
-the clouds of past catastrophes -- we may be able to learn from our mistakes.
+Similarly, it has been suggested that data gathered from ecological
+systems such as lakes that were monitored before they experienced sudden
+eutriphication, or grasslands subjected to overgrazing, could contain
+data that could help reveal when similar systems are approaching a
+tipping point~\citep{Carpenter2011}.  This is a silver lining to the
+clouds of past catastrophes; we may be able to learn from such examples 
+how to detect and avert sudden shifts in the future.
 
-Yet in testing methods for early warning signals against historical examples of
-transitions, we must be careful to avoid statistical mistakes that arise when
-we select examples conditional on having exhibited a sudden transition.  Such
+However, testing methods for early warning signals against historical examples of
+transitions is suspetible to statistical mistakes that arise from selecting
+data conditional on that data having already exhibited a sudden transition.  
+A central tenent of early warning theory is that the system in question is slowly
+approaching a tipping point that lies some unknown distance away.  If nothing
+is done to remedy the situation, this slow change will inevitably carry the 
+system beyond the tipping point, which introduces a sudden, rapid transition into an 
+undesirable state~\citep{Scheffer2009}. This process can be described mathematically
+as a \emph{bifurcation}, in which a slowly changing parameter reaches a critical
+value that causes the system stability to change.
+
+Such
 sudden transitions can occur for many reasons that do not involve the 
@@
 historical records is an example of the statistical error commonly known as the
-\emph{Prosecutor's Fallacy.}
-
-\subsection*{The Prosecutor's Fallacy}
+Prosecutor's Fallacy.
 
@@
 crime and a suspect on the defense stand.  Statistical testimony may accompany 
-such evidence reflecting the incidence rate of the matching characteristic -- 
-for instance, the frequency of individuals in the population with the matching
-blood type.  Such arguments may be misleading when they fail to account for  
-the manner in which the suspect was identified in 
-the first place.  If the suspect has been selected partly on the basis of 
+such evidence reflecting the incidence rate of the matching characteristic,
+such as the frequency of individuals in the population with the matching
+blood type.  Such arguments are misleading when they fail to account for  
+the manner in which the suspect was identified in the first place.  If the suspect has been selected on the basis of 
 belonging to a demographic in which the discriminating characteristic is much more common than
-in the population as a whole, these odds can be misleading.  Failing to recognize 
+in the population as a whole, .
+
+Failing to recognize 
 these conditional probabilities leads to the Prosecutor's Fallacy: the 
@@
 \section{Methods and Results}
-We approach the problem by simulating a system that cannot undergo the
+We approach the problem by simulating a system that will not undergo the
 saddle node bifurcation that underlies the usual early warning signal
-indicators.  This simulation approach is the only way to determine whether
+indicators.  This simulation approach allows us to determine whether
 examining historical events is a good way to test the utility of these
@@
 computed values of Kendall's $\tau$ for both indicators for each of
-the 266 replicates.  Kendall's $tau$ takes values in $(-1, 1)$, and is
+the 266 replicates.  Kendall's $\tau$ takes values in $(-1, 1)$, and is
 frequently used to identify an increasing trend ($\tau > 0 )$ in early
@@
 Figure~\ref{fig:indicator}.  More positive values indicate stronger
-trends. The common statistical significance test used for Kendall's $tau$
+trends. The common statistical significance test used for Kendall's $\tau$
 assumes independence of the variables to achieve a normal distribution,
@@
 and a Computational Sciences Graduate Fellowship from the Department of
-Energy grant DE-FG02-97ER25308 to CB.
+Energy grant DE-FG02-97ER25308 to CB. The authors thank M. Baskett, 
+T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
+manuscript.
 

### 2012-06-01 fallacy.tex
fallacy explained more fully

@@
 
+\begin{quotation}
+\noindent \emph{Mathematics \dots while assisting the trier of fact in the search of truth, must not cast a spell over him.}
+-- California Supreme court, 1968.
+\end{quotation}
+
+\noindent In the case of \emph{People v. Collins} 1968, California Supreme
+Court considered the evidence of an expert witness described by the
+court as ``an instructor of mathematics at a state college'', which
+concluded that the probability that a randomly selected individual
+would match the description given by the victim would be less than 1 in
+12 million~\citep{PeopleCollins1968}.  The prosecution had produced an
+individual matching the prosecutor's detailed description, and convinced
+by the mathematics of the instructor, the lower courts had found him
+guilty.
+
+The prosecution has only observed that the probability of seeing the
+evidence ($E$) they produced given a random innocent individual ($I$),
+$P(E|I)$ is very small.  This is not a sufficient basis from which
+to conclude that the individual is indeed guilty, that is, that the
+probability the individual is innocent given the evidence $P(I|E)$
+is also very small. In a city with millions of people, there might
+be several individuals who match the description of the evidence.
+\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$, instead,
+these expressions are related by Bayes theorem,
+
+\begin{equation}
+  P(E|I) = P(I|E) \frac{P(E)}{P(I)}
+\end{equation} }
+
+With little other evidence supporting the conviction, The California
+Supreme Court reversed the decision, and the case became a widely
+recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
+While DNA evidence has brought a rise in the number of probability-based
+convictions in the courtroom, and a corresponding rise in the frequency
+with which suspects are falsely accused, trial juries are not the
+only ones who are susceptible to misinterpreting probabilistic evidence.
+In this paper we explore how a similar misconception can arise from the
+use of historical data to evaluate methods for detecting early warning
+signals of critical transitions.
+
 %\subsection*{Catastrophic Transitions}
@@
 identify an approaching tipping point, which might provide managers time
-to react to and avert an undesirable state shift~\citep{Scheffer2009,
-Lenton2011}.  An array of statistical patterns associated with tipping
+to react to and avert an undesirable state shift~\citep{Scheffer2009, Lenton2011}.
+An array of statistical patterns associated with tipping
 point phenomena has been suggested for the detection of early warning
@@
 
-However, testing methods for early warning signals against historical examples of
-transitions is suspetible to statistical mistakes that arise from selecting
-data conditional on that data having already exhibited a sudden transition.  
-A central tenent of early warning theory is that the system in question is slowly
-approaching a tipping point that lies some unknown distance away.  If nothing
-is done to remedy the situation, this slow change will inevitably carry the 
-system beyond the tipping point, which introduces a sudden, rapid transition into an 
-undesirable state~\citep{Scheffer2009}. This process can be described mathematically
-as a \emph{bifurcation}, in which a slowly changing parameter reaches a critical
-value that causes the system stability to change.
+However, testing methods for early warning signals against historical
+examples of transitions is suspetible to statistical mistakes that arise
+from selecting data conditional on that data having already exhibited
+a sudden transition.  A central tenent of early warning theory is that
+the system in question is slowly approaching a tipping point that lies
+some unknown distance away.  If nothing is done to remedy the situation,
+this slow change will inevitably carry the system beyond the tipping
+point, which introduces a sudden, rapid transition into an undesirable
+state~\citep{Scheffer2009}. This process can be described mathematically
+as a \emph{bifurcation}, in which a slowly changing parameter reaches
+a critical value that causes the system stability to change.
 
-Such
-sudden transitions can occur for many reasons that do not involve the 
-bifurcation processes on which most early warning indicators are predicated
-~\citep{Hastings2010}, including purely noise-induced transitions for proposed 
-early warning patterns are not anticipated~\citep{Ditlevsen2010, Lenton2011}. 
-Ignoring the potential biases introduced by conditionally selecting from 
-historical records is an example of the statistical error commonly known as the
-Prosecutor's Fallacy.
+Not all sudden transitions are caused by some guilty process slowly
+driving the system over a tipping point, that is, the kind of process
+that early warning signals are designed to detect.  Some systems may
+experience such transitions purely by chance, leaving a stable state on
+an extremely unlikely excursion that happens to stray to far from the
+stable attractor~\citep[\emph{e.g.} ][consider this possibility in 
+transitions that arise from analyzing historical climate record]{Ditlevsen2010, Lenton2011}. 
+Like the evidence presented before the California Supreme
+Court in 1968, the chance of observing such an ``innocent'' transition 
+a priori may be very small, but when selected from a historical record
+of many possible transitions, this possibility can no longer be ignored.
 
-In a criminal trial, the prosecution may introduce associative evidence such as 
-blood testing to establish a match between the perpetrator at the scene of the
-crime and a suspect on the defense stand.  Statistical testimony may accompany 
-such evidence reflecting the incidence rate of the matching characteristic,
-such as the frequency of individuals in the population with the matching
-blood type.  Such arguments are misleading when they fail to account for  
-the manner in which the suspect was identified in the first place.  If the suspect has been selected on the basis of 
-belonging to a demographic in which the discriminating characteristic is much more common than
-in the population as a whole, .
+In principle, as such chance transitions do not involve the shrinking
+eigenvalue that drives the critical slowing down and increasing variance
+phenomena upon which several early warning signal statistics are based,
+such data should not appear to display a warning signal. If this is the case,
+then we can use historical data with more confidence when seeking to 
+test early warning patterns, knowing that we will not mistake such examples
+as cases of the bifurcation process we seek to detect.
 
-Failing to recognize 
-these conditional probabilities leads to the Prosecutor's Fallacy: the 
-argument that the suspect is the perpetrator on the basis of the statistical 
-probability of a match without correctly identifying the baseline probability 
-for such a match \citep{Thompson1987}.
+Figure 1 shows a schematic illustrating critical transitions under
+each of these scenarios.  In the left panel, the system experiences 
+a bifurcation that forces it over the edge -- just the kind of data 
+we would want to test an early warning signal.  In the right panel,
+a similar-looking trajectory emerges from a simulation of a stable 
+system.  While the simulation of the bifurcation scenario shown on
+the left produces a similar transition every time, the transition 
+shown on the right is somewhat less likely, occurring in only 1\% of
+simulations.
+
+\begin{figure}
+  \begin{center}
+    \subfigure[\textbf{True criminal} A bifurcation
+    drives a transition into the undesirable state]{
+    \includegraphics[width=.4\textwidth]{bifurcation.pdf}
+    \label{fig:bifurcation} } \subfigure[\textbf{Innocent suspect} A
+    transition to the alternative stable state has occurred purely by
+    chance.]{ \includegraphics[width=.4\textwidth]{fallacytransition.pdf}
+    \label{fig:fallacy} }
+  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
+  state vs potential in two cases.  Potential is the negative integral
+  of the mean of dynamical equation.  The potential function gives an
+  intuitive picture of the stability of a system by imagining the curve as
+  a surface on which a ball is free to bounce across -- wells correspond
+  to stable points and peaks to unstable points.  Early warning signals
+  are aimed at detecting systems which are slowly moving towards a tipping
+  point or bifurcation, illustrated in the successively darker potential
+  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
+  a simulation under this process shows the state of the system as the
+  potential moves towards the bifurcation point. The state of the system
+  in the simulation is given by the horizontal axis, while the vertical
+  dimension is used to show time on the right-hand axis.  Some stable
+  systems, panel~\subref{fig:fallacy}, may experience chance transitions
+  to an alternate stable state without the presence of this slow change
+  driving the system over the edge.  An example of the trajectory of
+  such a chance transition is plotted over the potential.  While such
+  events may be unlikely over short timescales, selecting historical
+  examples of systems that have experienced sudden transitions increases
+  the probability of accidentally using chance events depicted in
+  panel~\subref{fig:fallacy} to test methods designed to detect the kind
+  of process depicted in panel~\subref{fig:bifurcation}.} \label{fig:1}
+\end{figure}
 
@@
 examining historical events is a good way to test the utility of these
-indicators.  We simulated 1000 replicates of a stochastic individual-based
+indicators.  We simulated 20,000 replicates of a stochastic individual-based
 birth-death process with an Allee threshold~\citep{Courchamp2008}, which
@@
 stable states introduced by this effect; other functional forms for
-Eq.~\ref{master} could serve equally well for these simulations~\citep[see
+Eq.~\eqref{master} could serve equally well for these simulations~\citep[see
 \emph{e.g.}][]{Scheffer2001}.  Though this system can be forced
@@
 comparable sample size, we focus on a section of the data 500 time points
-prior to the system approaching the transition.  Nine of these replicates
-can be seen in Figure~\ref{fig:replicate_crashes}, as an illustration.
-
-  \begin{figure}
-    \begin{center}
-      \includegraphics[width=6in]{replicate_crashes.pdf}
-    \end{center}
-    \caption{Example trajectories for nine of the 266 replicates that 
-             experience a critical transition by chance during the course
-             of the simulation, showing the window of 500 points prior to
-             crossing the threshold which is used in the analysis.  A wobble
-             at the unstable critical point is visible at the end of several 
-             trajectories.}
-    \label{fig:replicate_crashes}
-  \end{figure}
+prior to the system approaching the transition.
 

### 2012-06-05 fallacy.tex
eliminating words

@@
   the collapse of a coral reef, or the end
-  of a glacial period.  Because such transitions often occur on scales that
+  of a glacial period.  Because such transitions often unfold on temporal and spatial scales that
   can be difficult to approach by experimental manipulation, research
@@
   known to experience a transition of some sort and failing to account
-  for this bias this introduces -- a statistical error often known as the
+  for the bias this introduces -- a statistical error often known as the
   Prosecutor's Fallacy.  By analyzing simulated systems that have
@@
   of false postives in common warning signal statistics. We further
-  demonstrate a model-based approach which is less subject to this bias
+  demonstrate a model-based approach that is less subject to this bias
   than these more commonly used summary statistics.
@@
   for early warning signals, care must be taken to avoid this fallacy,
-  either through the use methods less sensitive to this effect or ideally
-  by greater reliance on Experimental studies with replicates avoid this 
+  either through the use of methods less sensitive to this effect or ideally
+  by greater reliance on experimental studies with replicates avoid this 
   pitfall entirely.  More such studies are needed to provide data suitable
@@
   P(E|I) = P(I|E) \frac{P(E)}{P(I)}
-\end{equation} }
+\end{equation}
 
-With little other evidence supporting the conviction, The California
+} With little other evidence supporting the conviction, The California
 Supreme Court reversed the decision, and the case became a widely
 recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
-While DNA evidence has brought a rise in the number of probability-based
-convictions in the courtroom, and a corresponding rise in the frequency
-with which suspects are falsely accused, trial juries are not the
-only ones who are susceptible to misinterpreting probabilistic evidence.
+Trial juries are not the
+only ones susceptible to misinterpreting probabilistic evidence.
 In this paper we explore how a similar misconception can arise from the
@@
 
-Not all sudden transitions are caused by some guilty process slowly
-driving the system over a tipping point, that is, the kind of process
+Not all sudden transitions are caused by some ``guilty'' process slowly
+driving the system over a tipping point -- the kind of process
 that early warning signals are designed to detect.  Some systems may
@@
 
-In principle, as such chance transitions do not involve the shrinking
-eigenvalue that drives the critical slowing down and increasing variance
-phenomena upon which several early warning signal statistics are based,
-such data should not appear to display a warning signal. If this is the case,
-then we can use historical data with more confidence when seeking to 
-test early warning patterns, knowing that we will not mistake such examples
-as cases of the bifurcation process we seek to detect.
-
 Figure 1 shows a schematic illustrating critical transitions under
-each of these scenarios.  In the left panel, the system experiences 
-a bifurcation that forces it over the edge -- just the kind of data 
-we would want to test an early warning signal.  In the right panel,
-a similar-looking trajectory emerges from a simulation of a stable 
-system.  While the simulation of the bifurcation scenario shown on
-the left produces a similar transition every time, the transition 
-shown on the right is somewhat less likely, occurring in only 1\% of
-simulations.
+each of these scenarios.  In the left panel, the system experiences a
+bifurcation  and should contain an early warning signal.  In the right
+panel, a similar-looking trajectory emerges from a simulation of a stable
+system which should not contain a warning signal.  While the simulation of
+the bifurcation scenario shown on the left produces a similar transition
+every time, the transition shown on the right is somewhat less likely,
+occurring in only 1\% of simulations.
 
@@
 \section{Methods and Results}
-We approach the problem by simulating a system that will not undergo the
-saddle node bifurcation that underlies the usual early warning signal
-indicators.  This simulation approach allows us to determine whether
+To investigate if early warning signals are vulnerable to this fallacy,
+we simulating a system that is not driven towards a bifurcation (such as
+in the right-hand panel).  This simulation approach allows us to determine whether
 examining historical events is a good way to test the utility of these
@@
 arises from positive fitness effects at low densities.  Above the Allee
-threshold the population returns to a positive equilibrium size, while
-below the threshold the population decreases to zero in expectation. The
+threshold the population returns to a positive equilibrium size, whereas
+below the threshold the population decreases to zero. The
 simulation uses the Gillespie algorithm to provide an exact implementation
@@
 \end{align}
-a model with a linear death rate and density dependent birth rate
-that drives the Allee effect at low densities and limits growth
-at high densities.  The key feature of this model is the alternate
-stable states introduced by this effect; other functional forms for
-Eq.~\eqref{master} could serve equally well for these simulations~\citep[see
-\emph{e.g.}][]{Scheffer2001}.  Though this system can be forced
-through a saddle-node bifurcation by increasing the death rate, in
-these simulations all parameters are held constant and no bifurcation
-occurs.  Consequently we do not anticipate an early warning signal of
-an approaching bifurcation.  The simulation starts from the positive
-equilibrium population size.  Though the chance of a transition across
-the Allee threshold in any given time step are small, given enough
-time this system will eventually experience such a rare event driving
-the population extinct.  We evolved each replicate 50,000 time steps,
-sampling the system every 50 time steps.  In this time window 266 of the
-1,000 replicates experience population collapse.  To keep the examples of
-comparable sample size, we focus on a section of the data 500 time points
-prior to the system approaching the transition.
 
-To test whether selecting systems that have experienced spontaneous transitions
-could bias the analysis towards false positive detection of early warning signals,
-(the Prosecutor's Fallacy) we selected replicates conditional on having collapsed
-in the simulations.
-We then selected a window around each system that ended just before the
-collapse, while the population values were still above the Allee threshold.
-For each replicate, we calculated the most common early warning indicators, variance
-and autocorrelation~\citep[\emph{e.g.}][]{Carpenter2006,Dakos2008,Scheffer2009}, 
+a model with a linear death rate and density-dependent birth
+rate that drives the Allee effect at low densities and limits
+growth at high densities.  The key feature of this model is the
+alternate stable states introduced by this effect; other functional
+forms for Eq.~\eqref{master} could serve equally well for these
+simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
+system can be forced through a bifurcation by increasing
+the death rate, in these simulations all parameters are held constant
+and no bifurcation occurs.  Consequently we do not anticipate an early
+warning signal of an approaching bifurcation.  The simulation starts
+from the positive equilibrium population size.  Though the chance of
+a transition across the Allee threshold in any given time step is 
+small, given enough time this system will eventually experience such a
+rare event driving the population extinct.  We ran each replicate over
+50,000 time unitss, sampling the system every 50 time units.  In this
+time window 266 of the 1,000 replicates experience population collapse.
+To keep the examples of comparable sample size, we focus on a section of
+the data 500 time points prior to the system approaching the transition.
+
+To test whether selecting systems that have experienced
+spontaneous transitions could bias the analysis towards false
+positive detection of early warning signals, (the Prosecutor's
+Fallacy) we selected replicates conditional on having collapsed
+in the simulations.  We then selected a window around each system
+that ended just before the collapse, while the population values
+were still above the Allee threshold.  For each replicate, we
+calculated the most common early warning indicators, variance and
+autocorrelation~\citep[\emph{e.g.}][]{Carpenter2006,Dakos2008,Scheffer2009},
 around a moving window equal to half the length of that time series.
@@
 frequently used to identify an increasing trend ($\tau > 0 )$ in early
-warning signals.~\citep{Dakos2008, Dakos2011}.  The distribution
+warning signals~\citep{Dakos2008, Dakos2011}.  The distribution
 of $\tau$ values observed across these replicates is shown in
-Figure~\ref{fig:indicator}.  More positive values indicate stronger
-trends. The common statistical significance test used for Kendall's $\tau$
-assumes independence of the variables to achieve a normal distribution,
-this isn't satisfied for indicators coming from a time series and
-computed with a sliding window.  For a more robust comparison, we
-compare the distribution of this statistic from all the simulations to
+Figure~\ref{fig:indicator}.   
+%The common statistical significance test used for Kendall's $\tau$
+%assumes independence of the variables to achieve a normal distribution,
+%this isn't satisfied for indicators coming from a time series and
+%computed with a sliding window.  For a more robust comparison, 
+We compare the distribution of $\tau$ from all the simulations to
 the distribution conditioned on experiencing a chance transition to the
-alternative stable state, that is, from the replicates which are known
-to crash.  To avoid an effect of sample size the time series are all chosen
-to be the same length.  As these examples have been selected from a system
-that is not approaching a bifurcation, an indicator that does not suffer
-from the bias of Prosecutor's fallacy should have the same distribution
-in both cases.   Annotated, parallelized code to replicate the analysis
+alternative stable state.  To avoid an effect of sample size the time series are all chosen
+to be the same length.  
+%As these examples have been selected from a system
+%that is not approaching a bifurcation, an indicator that does not suffer
+%from the bias of Prosecutor's fallacy should have the same distribution
+%in both cases.   
+Annotated, parallelized code to replicate the analysis
 and produce the figures is provided in the appendix.  (Currently at
@@
 
-For each of these replicates we also take a model-based approach, estimating 
-parameters for an approximate linear model of the system approaching a
-saddle node bifurcation.
-
-\begin{equation}
-  dX = \sqrt{ r_t } (\phi(r_t) - X_t) dt +
-  \sigma\sqrt{\phi(r_t) } d B_t \label{LSN}
-\end{equation}
-where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilibrium position.  
-This model emerges from the canonical form of the saddle-node bifurcation,
-
-\begin{equation}
-  \frac{dX}{dt} = r - x^2
-\end{equation}
-where we introduce an explicit parameter $\theta$ to account for rescaling
-to the units of the data,
-
-\begin{equation}
-   \frac{dX}{dt} = r - (x-\theta)^2
-\end{equation}
-and then consider the corresponding stochastic process where the noise is
-proportional to the square root of the mean, as is the case for demographic
-noise processes,
-
-\begin{equation}
-  \frac{dX}{dt} = r - (x-\theta)^2 + \sqrt{ \beta (r + (x-\theta)^2 } dB_t
-\end{equation}
-$r$ remains the bifurcation parameter which we imagine changing with time,
-$r_t$, then linearizing the above model in neighborhood of
-\(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\), we recover Equation~\eqref{LSN}.
-These models can be both be fit to the data by maximum likelihood estimation
-based on the equation for the evolution of their moments, as explained
-in~\citet{Boettiger2012b}.  Though this approximation is more general than 
-the original Markov process (as other models containing a saddle-node bifurcation
-share the same approximation), it is also easier to estimate reliably from 
-the data since it describes only the process near the equilibrium where the
-data is available.  We assume that $r_t$ is varying at constant rate, 
-according to the model
-
-\begin{equation}
- r_t = m t + r_0 
-\end{equation}
-and estimate $m$, $r_0$, $\theta$, and $\beta$ by maximum likelihood on each
-of the simulated datasets.  Estimates of $m < 0 $ are expected in systems 
-approaching a bifurcation, while for stable systems $m$ should be approximately zero.
-
-Of the 266 simulations, only six had estimates with machine precision different
-from zero, the greatest magnitude among these was $-7.1 \times 10^{-12}$,
-which would still correspond to a negligible change in $r_0$ over the 50,000 time steps
-of the simulation.  Consequently, the model-based estimation shows no 
-evidence of bias on data that has been selected conditional on collapse.
+For each of these replicates we also take a model-based
+approach, estimating parameters for an approximate linear
+model of the system approaching a saddle node bifurcation, as
+described~\citet{Boettiger2012b}.  In this model, the parameter $m$
+describes the approach towards the saddle-node bifurcation.  Estimates
+$m < 0 $ are expected in systems approaching a bifurcation, while for
+stable systems $m$ should be approximately zero. None of the estimates
+across the 266 simulations differed from zero in our study, hence the
+model-based estimation shows no evidence of bias on data that has been
+selected conditional on collapse.
 
@@
 
-Historical examples of systems that have undergone sudden transitions seem
-to offer an invaluable opportunity to test methods for the detection of
+Historical examples of systems that have undergone sudden transitions 
+offer a unique opportunity to test methods for the detection of
 early warning signals that could have foreseen the collapse.  However,

### 2012-06-15 fallacy.tex
british spelling

@@
-%\documentclass[authoryear,5p]{elsarticle}
-\documentclass[authoryear,review,12pt]{elsarticle}
+\documentclass[authoryear,5p]{elsarticle}
+%\documentclass[authoryear,review,12pt]{elsarticle}
 \bibliographystyle{elsarticle-harv}
@@
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 \usepackage{float}
@@
   for the bias this introduces -- a statistical error often known as the
-  Prosecutor's Fallacy.  By analyzing simulated systems that have
+  Prosecutor's Fallacy.  By analysing simulated systems that have
   experienced transitions purely by chance, we reveal an elevated rate
-  of false postives in common warning signal statistics. We further
+  of false positives in common warning signal statistics. We further
   demonstrate a model-based approach that is less subject to this bias
   than these more commonly used summary statistics.
-  In using historical examples of collapse to test and validate methods
-  for early warning signals, care must be taken to avoid this fallacy,
-  either through the use of methods less sensitive to this effect or ideally
-  by greater reliance on experimental studies with replicates avoid this 
-  pitfall entirely.  More such studies are needed to provide data suitable
-  for evaluating methods for early warning.
+  We note that experimental studies with replicates avoid this 
+  pitfall entirely.
 
@@
 individual matching the prosecutor's detailed description, and convinced
-by the mathematics of the instructor, the lower courts had found him
+by the mathematics, the lower courts had found him
 guilty.
@@
 evidence ($E$) they produced given a random innocent individual ($I$),
-$P(E|I)$ is very small.  This is not a sufficient basis from which
-to conclude that the individual is indeed guilty, that is, that the
+$P(E|I)$ is very small.  From this one cannot conclude that the individual is indeed guilty, that is, that the
 probability the individual is innocent given the evidence $P(I|E)$
@@
 
-} With little other evidence supporting the conviction, The California
+} The California
 Supreme Court reversed the decision, and the case became a widely
 recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
-Trial juries are not the
-only ones susceptible to misinterpreting probabilistic evidence.
-In this paper we explore how a similar misconception can arise from the
+Here we explore how a similar misconception can arise from the
 use of historical data to evaluate methods for detecting early warning
@@
 
-Historical examples of sudden transitions taken from the paleoclimate
+Historical examples of sudden transitions taken from the paleo-climate
 record provide an important way to test and evaluate potential
@@
 systems such as lakes that were monitored before they experienced sudden
-eutriphication, or grasslands subjected to overgrazing, could contain
+eutrophication, or grasslands subjected to overgrazing, could contain
 data that could help reveal when similar systems are approaching a
-tipping point~\citep{Carpenter2011}.  This is a silver lining to the
-clouds of past catastrophes; we may be able to learn from such examples
-how to detect and avert sudden shifts in the future.
+tipping point~\citep{Carpenter2011}.
 
 However, testing methods for early warning signals against historical
-examples of transitions is suspetible to statistical mistakes that arise
+examples of transitions is susceptible to statistical mistakes that arise
 from selecting data conditional on that data having already exhibited
-a sudden transition.  A central tenent of early warning theory is that
+a sudden transition.  A central tenant of early warning theory is that
 the system in question is slowly approaching a tipping point that lies
@@
   such a chance transition is plotted over the potential.  While such
-  events may be unlikely over short timescales, selecting historical
+  events may be unlikely over short time-scales, selecting historical
   examples of systems that have experienced sudden transitions increases
@@
 rare event driving the population extinct.  We ran each replicate over
-50,000 time unitss, sampling the system every 50 time units.  In this
+50,000 time units, sampling the system every 50 time units.  In this
 time window 266 of the 1,000 replicates experience population collapse.
@@
 %in both cases.   
-Annotated, parallelized code to replicate the analysis
+Annotated, parallelised code to replicate the analysis
 and produce the figures is provided in the appendix.  (Currently at
@@
 
-Historical examples of systems that have undergone sudden transitions 
-offer a unique opportunity to test methods for the detection of
-early warning signals that could have foreseen the collapse.  However,
-such data must be used with care, as the process of selecting systems
-already known to have collapsed opens up the possibility that these
-transitions were the result of rare chance events rather than examples
-of the more deterministic tipping point behaviour we wish to detect.
-When we think about forecasting potential collapse in a single system,
-these purely-chance transitions may be rare enough to ignore.  When we
-comb the historical record for any available data on sudden transitions,
-our odds of finding these chance transitions may be much higher.
-
-We have shown that systems which undergo such rare transitions look
+We have shown that systems which undergo rare sudden transitions look
 statistically different from their counterparts that do not, even though
@@
 signal being tested appears to have successfully detected an oncoming
-tipping point, when in fact the example is falsely accused, coming
+tipping point, when in fact the example comes
 instead from a stable system. Figure 2 shows the extent of this bias,
@@
 It seems tempting to argue that the bias towards positive detection
-in historical examples is valuable -- each of these systems did
-indeed collapse, so the fact that they show increased probability
-to exhibit warning signals could be taken as a successful detection.
+in historical examples is not problematic -- each of these systems did
+indeed collapse, so the increased probability
+of exhibiting warning signals could be taken as a successful detection.
 Unfortunately this is not the case. At the moment the forecast is made,
@@
 the sample that did not collapse, just as a fair coin is still a fair
-coin after a string of consecutive ``heads.''  Taking a closer look at
-the patterns involved we can begin to understand why common indicators
+coin after a string of consecutive ``heads.''  A closer look at
+the patterns involved shows why common indicators
 such as autocorrelation and variance can be misleading.
 
-The farther the system gets from its stable point (while still remaining
-in the local region), the more likely it is to draw a random step that
-returns it towards the stable point, like a spring that is stretched
-further and further. Despite this, there is always some probability that
-it will move further still, and with enough possible examples to choose
-from, we can be sure to see this happen occasionally.  Systems that manage
-to extend this random walk far enough to cross the tipping point must do
-so rather quickly -- with a string of events moving closer and closer to
-the edge.  This pattern, clearly visible before the crashes in each of
+As the system gets farther from its stable point, it it more likely to draw a random step that
+returns it towards the stable point. Despite this, there is always some probability that
+it will move further still, so systems that do cross the tipping point must do
+so rather quickly by a string of events.  This pattern, clearly visible before the crashes in each of
 the examples in Figure 1, produces a string of observations that appear
@@
 
-In our experiment, the model-based estimate of early warning signals
+In our numerical experiment, the model-based estimate of early warning signals
 appears more robust than the summary statistics, producing the same
@@
 only a single realization with which to examine a complex problem.
-The only way to deal with this problem is through replication, as
+The only way to deal with this problem embodied is through replication, as
 can be done in an experimental system in laboratory manipulations
 such as~\citet{Drake2010, Veraart2011} and at the scale of whole lake
-ecosystems in~\citet{Carpenter2011}.  Historical data will not suffice,
-more such replicated experiments are needed to test these patterns.
+ecosystems in~\citet{Carpenter2011}.  
 Experimental procedures avoid the hazard of the Prosecutor's fallacy by
 generating a complete sample of replicates, rather than selecting a subset
-of cases some larger historical sample where the risk that if selecting a
-chance transition cannot be ignored.  Further, we note that the influence
-of the Prosecutors Fallacy need not be restricted to the context of early
-warning signals, but may arise any time we search historical data for
-examples containing some feature of interest and ignore the statistical
-effect that selection process introduces in our data.
+of cases from some larger historical sample.  
+%We note as well that the Prosecutors Fallacy 
+%may arise in other contexts where we search historical data for
+%examples containing some feature of interest.
 

### 2012-06-19 fallacy.tex
finishing up corrections

@@
-\documentclass[authoryear,5p]{elsarticle}
-%\documentclass[authoryear,review,12pt]{elsarticle}
+%\documentclass[authoryear,5p]{elsarticle}
+\documentclass[authoryear,review,12pt]{elsarticle}
 \bibliographystyle{elsarticle-harv}
@@
 \usepackage{lineno}
-%\linenumbers
+\linenumbers
 \usepackage{float}
@@
 evidence ($E$) they produced given a random innocent individual ($I$),
-$P(E|I)$ is very small.  From this one cannot conclude that the individual is indeed guilty, that is, that the
-probability the individual is innocent given the evidence $P(I|E)$
-is also very small. In a city with millions of people, there might
-be several individuals who match the description of the evidence.
-\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$, instead,
-these expressions are related by Bayes theorem,
+$P(E|I)$ is very small.  From this one cannot conclude that the individual
+is indeed guilty, that is, that the probability the individual is innocent
+given the evidence $P(I|E)$ is also very small. In a city with millions
+of people, there might be several individuals who match the description
+of the evidence\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$,
+instead, these expressions are related by Bayes theorem,
 
@@
 
-} The California
-Supreme Court reversed the decision, and the case became a widely
-recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
-Here we explore how a similar misconception can arise from the
-use of historical data to evaluate methods for detecting early warning
-signals of critical transitions.
+}. The California Supreme Court reversed the decision, and
+the case became a widely recognized example of the Prosecutor's
+Fallacy~\citep{Thompson1987}.  Here we explore how a similar misconception
+can arise from the use of historical data to evaluate methods for
+detecting early warning signals of critical transitions.
 
@@
 been tested in both experimental manipulation~\citep{Drake2010,
-Carpenter2011, Veraart2011} and historical
+Carpenter2011, Veraart2011, Dai2012} and historical
 observations~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010}.
@@
   \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
-  state vs potential in two cases.  Potential is the negative integral
+  state versus potential in two cases.  Potential is the negative integral
   of the mean of dynamical equation.  The potential function gives an
@@
   to an alternate stable state without the presence of this slow change
-  driving the system over the edge.  An example of the trajectory of
-  such a chance transition is plotted over the potential.  While such
-  events may be unlikely over short time-scales, selecting historical
-  examples of systems that have experienced sudden transitions increases
-  the probability of accidentally using chance events depicted in
-  panel~\subref{fig:fallacy} to test methods designed to detect the kind
-  of process depicted in panel~\subref{fig:bifurcation}.} \label{fig:1}
+  in the shape of the potential well which tips the system off to the left.
+  An example of the trajectory of such a chance transition is
+  plotted over the potential.  While such events may be unlikely over
+  short time-scales, selecting historical examples of systems that
+  have experienced sudden transitions increases the probability of
+  accidentally using chance events depicted in panel~\subref{fig:fallacy}
+  to test methods designed to detect the kind of process depicted in
+  panel~\subref{fig:bifurcation}.} \label{fig:1}
 \end{figure}
@@
 To investigate if early warning signals are vulnerable to this fallacy,
-we simulating a system that is not driven towards a bifurcation (such as
-in the right-hand panel).  This simulation approach allows us to determine whether
-examining historical events is a good way to test the utility of these
+we simulating a system that is not driven towards a bifurcation such as
+in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
+examining historical events is a valid way to test the utility of these
 indicators.  We simulated 20,000 replicates of a stochastic individual-based
@@
 \begin{align}
-  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
-    b_n &= \frac{e K n^2}{n^2 + h^2} \\
-    d_n &= e n + a
+  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
+    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
+    d_n &= e n + a,
 \end{align}
@@
 rate that drives the Allee effect at low densities and limits
-growth at high densities.  The key feature of this model is the
+growth at high densities.  In this model $n$ indicates the discrete
+number of individuals in the population, $K$ indicates a carrying 
+capacity as set by a limiting resource, $e$ a per-capita death rate, 
+$a$ an additional mortality imposed on the population such as harvest,
+$h$ is a parameter controlling at what population size the addition of 
+more individuals switches from conferring a positive benefit on growth 
+from Allee interactions $n < h$ to a negative impact on growth due to 
+increased competition, $n > h$.  The key feature of this model is the
 alternate stable states introduced by this effect; other functional
@@
 and no bifurcation occurs.  Consequently we do not anticipate an early
-warning signal of an approaching bifurcation.  The simulation starts
-from the positive equilibrium population size.  Though the chance of
-a transition across the Allee threshold in any given time step is 
-small, given enough time this system will eventually experience such a
-rare event driving the population extinct.  We ran each replicate over
-50,000 time units, sampling the system every 50 time units.  In this
-time window 266 of the 1,000 replicates experience population collapse.
-To keep the examples of comparable sample size, we focus on a section of
-the data 500 time points prior to the system approaching the transition.
+warning signal of an approaching bifurcation.
+
+The simulation starts from the positive equilibrium population size.
+Though the chance of a transition across the Allee threshold in any
+given time step is small, given enough time this system will eventually
+experience such a rare event driving the population extinct.  We ran
+each replicate over 50,000 time units, sampling the system every 50
+time units.  In this time window 266 of the 1,000 replicates experience
+population collapse.  To keep the examples of comparable sample size,
+we focus on a section of the data 500 time points prior to the system
+approaching the transition.
 
@@
   conditionally selected for having collapsed by chance in simulations
-  is shown in gray bars.  Solid lines indicate the estimated density of
+  is shown in grey bars.  Solid lines indicate the estimated density of
   the statistic from a random sample of the simulations (not conditional
@@
 can be done in an experimental system in laboratory manipulations
-such as~\citet{Drake2010, Veraart2011} and at the scale of whole lake
+such as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole lake
 ecosystems in~\citet{Carpenter2011}.  

### 2012-06-20 fallacy.tex
formatting for PRSB submission

@@
 
+\usepackage{endfloat} % Move floats to the end of the document
+\renewcommand{\listoffigures}{} % but supress these lists
+\renewcommand{\listoftables}{} % supress these lists
+% compile without figures?
+\newif\ifhavefigures
+%\havefigurestrue
+\newif\ifnohavefigures % allows alternate condition
+\nohavefigurestrue     % turn on the alternate condition uncommenting this
+
 \journal{\tiny } 
@@
 Early warning signals have been proposed to forecast the possibility
-  of a critical transition, such as the eutrophication of a lake,
-  the collapse of a coral reef, or the end
-  of a glacial period.  Because such transitions often unfold on temporal and spatial scales that
-  can be difficult to approach by experimental manipulation, research
-  has often relied on historical observations as a source of natural experiments.
+of a critical transition, such as the eutrophication of a lake, the
+collapse of a coral reef, or the end of a glacial period.  Because such
+transitions often unfold on temporal and spatial scales that can be
+difficult to approach by experimental manipulation, research has often
+relied on historical observations as a source of natural experiments.
 Here we examine a critical difference between selecting systems for
@@
 known to experience a transition of some sort and failing to account
-  for the bias this introduces -- a statistical error often known as the
-  Prosecutor's Fallacy.  By analysing simulated systems that have
+for the bias this introduces -- a statistical error often known as
+the Prosecutor's Fallacy.  By analysing simulated systems that have
 experienced transitions purely by chance, we reveal an elevated rate
 of false positives in common warning signal statistics. We further
-  demonstrate a model-based approach that is less subject to this bias
-  than these more commonly used summary statistics.
-  We note that experimental studies with replicates avoid this 
-  pitfall entirely.
+demonstrate a model-based approach that is less subject to this
+bias than these more commonly used summary statistics.  We note that
+experimental studies with replicates avoid this pitfall entirely.
 
@@
       drives a transition into the undesirable state]{
-    \includegraphics[width=.4\textwidth]{bifurcation.pdf}
-    \label{fig:bifurcation} } \subfigure[\textbf{Innocent suspect} A
+      \ifhavefigures
+%        \includegraphics[width=.4\textwidth]{figure1a}
+      \fi
+      \ifnohavefigures
+        \hspace{.4\textwidth}
+      \fi
+      \label{fig:bifurcation}
+    } 
+    \subfigure[\textbf{Innocent suspect} A
     transition to the alternative stable state has occurred purely by
-    chance.]{ \includegraphics[width=.4\textwidth]{fallacytransition.pdf}
+    chance.]{
+    \ifhavefigures
+%      \includegraphics[width=.4\textwidth]{figure1b}
+    \fi
+    \ifnohavefigures
+      \hspace{.4\textwidth}
+    \fi
     \label{fig:fallacy} }
@@
   \begin{center}
-    \includegraphics[width=6in]{indicators.pdf}
+    \ifhavefigures
+%    \includegraphics[width=6in]{figure2}
+    \fi
   \end{center}
@@
  \section{References}%bibliography
- \bibliography{fallacy}
+
+ \begin{thebibliography}{24}
+\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
+\expandafter\ifx\csname url\endcsname\relax
+  \def\url#1{\texttt{#1}}\fi
+\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+
+\bibitem[{Berkes et~al.(2006)Berkes, Hughes, Steneck, Wilson, Bellwood, Crona,
+  Folke, Gunderson, Leslie, Norberg, Nystr\"{o}m, Olsson, Osterblom, Scheffer,
+  and Worm}]{Berkes2006}
+Berkes, F., Hughes, T.~P., Steneck, R.~S., Wilson, J.~A., Bellwood, D.~R.,
+  Crona, B., Folke, C., Gunderson, L.~H., Leslie, H.~M., Norberg, J.,
+  Nystr\"{o}m, M., Olsson, P., Osterblom, H., Scheffer, M., Worm, B., Mar.
+  2006. {Ecology. Globalization, roving bandits, and marine resources.} Science
+  (New York, N.Y.) 311~(5767), 1557--8.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16543444}
+
+\bibitem[{Boettiger and Hastings(2012)}]{Boettiger2012b}
+Boettiger, C., Hastings, A., 2012. {Quantifying Limits to Detection of Early
+  Warning for Critical Transitions}. Journal Of The Royal Society Interface.
+
+\bibitem[{{California Supreme Court}(1968)}]{PeopleCollins1968}
+{California Supreme Court}, 1968. People v. collins.
+
+\bibitem[{Carpenter(2011)}]{Carpenter2011}
+Carpenter, J., Feb. 2011. {May the Best Analyst Win}. Science 331~(6018),
+  698--699.
+\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.331.6018.698}
+
+\bibitem[{Carpenter and Brock(2006)}]{Carpenter2006}
+Carpenter, S.~R., Brock, W.~A., 2006. {Rising variance: a leading indicator of
+  ecological transition.} Ecology letters 9~(3), 311--8.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16958897}
+
+\bibitem[{Courchamp et~al.(2008)Courchamp, Berec, and
+  Gascoigne}]{Courchamp2008}
+Courchamp, F., Berec, L., Gascoigne, J., 2008. {Allee Effects in Ecology and
+  Conservation}. Oxford University Press, USA.
+\newline\urlprefix\url{http://www.amazon.com/Effects-Ecology-Conservation-Franck-Courchamp/dp/0198570309}
+
+\bibitem[{Dai et~al.(2012)Dai, Vorselen, Korolev, and Gore}]{Dai2012}
+Dai, L., Vorselen, D., Korolev, K.~S., Gore, J., May 2012. {Generic Indicators
+  for Loss of Resilience Before a Tipping Point Leading to Population
+  Collapse}. Science 336~(6085), 1175--1177.
+\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1219805}
+
+\bibitem[{Dakos et~al.(2011)Dakos, K\'{e}fi, Rietkerk, Nes, and
+  Scheffer}]{Dakos2011}
+Dakos, V., K\'{e}fi, S., Rietkerk, M., Nes, E. H.~V., Scheffer, M., 2011.
+  {Slowing Down in Spatially Patterned Ecosystems at the Brink of Collapse}.
+  The American Naturalist.
+
+\bibitem[{Dakos et~al.(2008)Dakos, Scheffer, van Nes, Brovkin, Petoukhov, and
+  Held}]{Dakos2008}
+Dakos, V., Scheffer, M., van Nes, E.~H., Brovkin, V., Petoukhov, V., Held, H.,
+  Sep. 2008. {Slowing down as an early warning signal for abrupt climate
+  change.} Proceedings of the National Academy of Sciences 105~(38), 14308--12.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/18787119}
+
+\bibitem[{Ditlevsen and Johnsen(2010)}]{Ditlevsen2010}
+Ditlevsen, P.~D., Johnsen, S.~J., Oct. 2010. {Tipping points: Early warning and
+  wishful thinking}. Geophysical Research Letters 37~(19), 2--5.
+\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2010/2010GL044486.shtml}
+
+\bibitem[{Drake and Griffen(2010)}]{Drake2010}
+Drake, J.~M., Griffen, B.~D., Sep. 2010. {Early warning signals of extinction
+  in deteriorating environments}. Nature 467~(7314), 456--459.
+\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature09389}
+
+\bibitem[{Guttal and Jayaprakash(2008)}]{Guttal2008}
+Guttal, V., Jayaprakash, C., Dec. 2008. {Spatial variance and spatial skewness:
+  leading indicators of regime shifts in spatial ecological systems}.
+  Theoretical Ecology 2~(1), 3--12.
+\newline\urlprefix\url{http://www.springerlink.com/index/10.1007/s12080-008-0033-1}
+
+\bibitem[{Hirota et~al.(2011)Hirota, Holmgren, {Van Nes}, and
+  Scheffer}]{Hirota2011}
+Hirota, M., Holmgren, M., {Van Nes}, E.~H., Scheffer, M., Oct. 2011. {Global
+  Resilience of Tropical Forest and Savanna to Critical Transitions}. Science
+  334~(6053), 232--235.
+\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1210657}
+
+\bibitem[{K\'{e}fi et~al.(2007)K\'{e}fi, Rietkerk, Alados, Pueyo, Papanastasis,
+  Elaich, and de~Ruiter}]{Kefi2007}
+K\'{e}fi, S., Rietkerk, M., Alados, C.~L., Pueyo, Y., Papanastasis, V.~P.,
+  Elaich, A., de~Ruiter, P.~C., Sep. 2007. {Spatial vegetation patterns and
+  imminent desertification in Mediterranean arid ecosystems.} Nature
+  449~(7159), 213--7.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17851524}
+
+\bibitem[{Lenton(2011)}]{Lenton2011}
+Lenton, T.~M., Jun. 2011. {Early warning of climate tipping points}. Nature
+  Climate Change 1~(4), 201--209.
+\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nclimate1143}
+
+\bibitem[{Lenton et~al.(2012)Lenton, Livina, and Dakos}]{Lenton2012}
+Lenton, T.~M., Livina, V., Dakos, V., 2012. {Early warning of climate tipping
+  points from critical slowing down: comparing methods to improve robustness}.
+  Philosophical Transactions of The Royal Society A (in press).
+\newline\urlprefix\url{http://sites.google.com/site/noraclpreprints/contributions/lenton-et-al/Early\_warning\_methods\_revised\_with\_figs\_and\_appendix2.pdf}
+
+\bibitem[{Livina and Lenton(2007)}]{Livina2007}
+Livina, V.~N., Lenton, T.~M., Feb. 2007. {A modified method for detecting
+  incipient bifurcations in a dynamical system}. Geophysical Research Letters
+  34~(3), 1--5.
+\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2007/2006GL028672.shtml}
+
+\bibitem[{Mumby et~al.(2007)Mumby, Hastings, and Edwards}]{Mumby2007}
+Mumby, P.~J., Hastings, A., Edwards, H.~J., Nov. 2007. {Thresholds and the
+  resilience of Caribbean coral reefs.} Nature 450~(7166), 98--101.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17972885}
+
+\bibitem[{Scheffer et~al.(2009)Scheffer, Bascompte, Brock, Brovkin, Carpenter,
+  Dakos, Held, van Nes, Rietkerk, and Sugihara}]{Scheffer2009}
+Scheffer, M., Bascompte, J., Brock, W.~A., Brovkin, V., Carpenter, S.~R.,
+  Dakos, V., Held, H., van Nes, E.~H., Rietkerk, M., Sugihara, G., 2009.
+  {Early-warning signals for critical transitions.} Nature 461~(7260), 53--9.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/19727193}
+
+\bibitem[{Scheffer et~al.(2001)Scheffer, Carpenter, Foley, Folke, and
+  Walker}]{Scheffer2001}
+Scheffer, M., Carpenter, S.~R., Foley, J.~A., Folke, C., Walker, B., Oct. 2001.
+  {Catastrophic shifts in ecosystems.} Nature 413~(6856), 591--6.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/11595939}
+
+\bibitem[{Thompson and Sieber(2010)}]{Thompson2010}
+Thompson, J. M.~T., Sieber, J., Dec. 2010. {Climate tipping as a noisy
+  bifurcation: a predictive technique}. IMA Journal of Applied Mathematics
+  76~(1), 27--46.
+\newline\urlprefix\url{http://imamat.oxfordjournals.org/cgi/doi/10.1093/imamat/hxq060}
+
+\bibitem[{Thompson and Schumann(1987)}]{Thompson1987}
+Thompson, W.~C., Schumann, E.~L., 1987. {Interpretation of statistical evidence
+  in criminal trials: The prosecutor's fallacy and the defense attorney's
+  fallacy.} Law and Human Behavior 11~(3), 167--187.
+\newline\urlprefix\url{http://psycnet.apa.org/journals/lhb/11/3/167/
+  http://doi.apa.org/getdoi.cfm?doi=10.1007/BF01044641}
+
+\bibitem[{van Nes and Scheffer(2007)}]{vanNes2007}
+van Nes, E.~H., Scheffer, M., Jun. 2007. {Slow recovery from perturbations as a
+  generic indicator of a nearby catastrophic shift.} The American naturalist
+  169~(6), 738--47.
+\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17479460}
+
+\bibitem[{Veraart et~al.(2011)Veraart, Faassen, Dakos, van Nes, L\"{u}rling,
+  and Scheffer}]{Veraart2011}
+Veraart, A.~J., Faassen, E.~J., Dakos, V., van Nes, E.~H., L\"{u}rling, M.,
+  Scheffer, M., Dec. 2011. {Recovery rates reflect distance to a tipping point
+  in a living system}. Nature, 2--5.
+\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature10723}
+
+\end{thebibliography}
+
  \end{document}

### 2012-08-22 prosecutor/fallacy_ah.tex
Cleanup

--- 2012-06-15 fallacy.tex
@@
   for the bias this introduces -- a statistical error often known as the
-  Prosecutor's Fallacy.  By analysing simulated systems that have
+  Prosecutor's Fallacy.  By analyzing simulated systems that have
   experienced transitions purely by chance, we reveal an elevated rate
-  of false positives in common warning signal statistics. We further
+  of false postives in common warning signal statistics. We further
   demonstrate a model-based approach that is less subject to this bias
@@
 
-Historical examples of sudden transitions taken from the paleo-climate
+Historical examples of sudden transitions taken from the paleoclimate
 record provide an important way to test and evaluate potential
@@
 systems such as lakes that were monitored before they experienced sudden
-eutrophication, or grasslands subjected to overgrazing, could contain
+eutriphication, or grasslands subjected to overgrazing, could contain
 data that could help reveal when similar systems are approaching a
@@
 However, testing methods for early warning signals against historical
-examples of transitions is susceptible to statistical mistakes that arise
+examples of transitions is suspetible to statistical mistakes that arise
 from selecting data conditional on that data having already exhibited
-a sudden transition.  A central tenant of early warning theory is that
+a sudden transition.  A central tenent of early warning theory is that
 the system in question is slowly approaching a tipping point that lies
@@
   such a chance transition is plotted over the potential.  While such
-  events may be unlikely over short time-scales, selecting historical
+  events may be unlikely over short timescales, selecting historical
   examples of systems that have experienced sudden transitions increases
@@
 rare event driving the population extinct.  We ran each replicate over
-50,000 time units, sampling the system every 50 time units.  In this
+50,000 time unitss, sampling the system every 50 time units.  In this
 time window 266 of the 1,000 replicates experience population collapse.
@@
 %in both cases.   
-Annotated, parallelised code to replicate the analysis
+Annotated, parallelized code to replicate the analysis
 and produce the figures is provided in the appendix.  (Currently at

### 2012-08-30 prosecutor/fallacy.tex
Cleanup
Alan's edits added

--- 2012-06-20 fallacy.tex
@@
 \linenumbers
-\usepackage{float}
+%\usepackage{float}
 \usepackage{subfigure}
-\usepackage[pdftex]{color}
-\definecolor{darkblue}{rgb}{0,0,0.5}
-\definecolor{darkgreen}{rgb}{0,0.5,0}
-\usepackage[pdftex, colorlinks]{hyperref}
+%\usepackage[pdftex]{color}
+%\definecolor{darkblue}{rgb}{0,0,0.5}
+%\definecolor{darkgreen}{rgb}{0,0.5,0}
+%\usepackage[pdftex, colorlinks]{hyperref}
 \textwidth 6.75in
@@
 
-\journal{\tiny } 
+\journal{\tiny Proceedings of the Royal Society B} 
 \begin{document}
@@
 of people, there might be several individuals who match the description
-of the evidence\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$,
+of the evidence.  Mathematically $P(E|I)$ need not equal $P(I|E)$,
 instead, these expressions are related by Bayes theorem,
@@
 \begin{equation}
-  P(E|I) = P(I|E) \frac{P(E)}{P(I)}
+  P(E|I) = P(I|E) \frac{P(E)}{P(I)},
 \end{equation}
 
-}. The California Supreme Court reversed the decision, and
+$P(E) \ll 1$ and $P(I)\approx 1$ , so $P(E|I) \approx P(I|E) P(E)$, and consequently we cannot conclude that $P(I|E) \ll 1$ from $P(E|I) \ll 1$.  
+Realizing this mistake, the California Supreme Court reversed the decision, and
 the case became a widely recognized example of the Prosecutor's
@@
 To investigate if early warning signals are vulnerable to this fallacy,
-we simulating a system that is not driven towards a bifurcation such as
+we simulate a system that is not driven towards a bifurcation such as
 in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
@@
 threshold the population returns to a positive equilibrium size, whereas
-below the threshold the population decreases to zero. The
-simulation uses the Gillespie algorithm to provide an exact implementation
-of the master equation of the birth-death process,
+below the threshold the population decreases to zero.
+The model can be represented as a continuous time birth-death process where births and deaths are Poisson events whcih depend on the current density,
 
 \begin{align}
-  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
-    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
-    d_n &= e n + a,
+    b(X) &= \frac{e K n^2}{n^2 + h^2}, \\
+    d(X) &= e n + a,
+
 \end{align}
+
+%The simulation uses the Gillespie algorithm to provide an exact implementation\footnote{ of the birth-death process,
+
+%\begin{align}
+%  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
+%    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
+%    d_n &= e n + a,
+%\end{align}
 
@@
 number of individuals in the population, $K$ indicates a carrying 
-capacity as set by a limiting resource, $e$ a per-capita death rate, 
+capacity as set by a limiting resource, $e$ a per-capita death rate (the $e$ scaling term in the birth equation allows the carrying capacity $K$ to correspond to a positive equilibrium point), 
 $a$ an additional mortality imposed on the population such as harvest,
@@
 computed values of Kendall's $\tau$ for both indicators for each of
-the 266 replicates.  Kendall's $\tau$ takes values in $(-1, 1)$, and is
-frequently used to identify an increasing trend ($\tau > 0 )$ in early
-warning signals~\citep{Dakos2008, Dakos2011}.  The distribution
+the 266 replicates.
+Kendall's $\tau$ is a non-parametric measure of rank correlation frequently used to identify an increasing trend ($\tau > 0 )$ in early warning signals~\citep{Dakos2008, Dakos2011}, defined as $\tau \def \tfrac{(\text{number of concordant pairs}) - (\text{number of discordant pairs})}{\frac{1}{2} n (n-1) }$ in $n$ observations.
+\footnote{A pair of observations $(x_i,y_i)$ and $(x_j, y_j)$ are concordant if $x_i > x_j$ and $y_i > y_j$ or $x_i < x_j$ and $y_i < y_j$ and discordant otherwise; equalties excepted.}  $\tau$ takes values in $(-1, 1)$.
+The distribution
 of $\tau$ values observed across these replicates is shown in
@@
 Annotated, parallelised code to replicate the analysis
-and produce the figures is provided in the appendix.  (Currently at
-\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).
+and produce the figures is provided in the appendix.
 
@@
 model of the system approaching a saddle node bifurcation, as
-described~\citet{Boettiger2012b}.  In this model, the parameter $m$
+described by~\citet{Boettiger2012b},
+
+\begin{equation}
+\ud X = \sqrt{ r_t } (\phi(r_t) - X_t)\ud t + \sigma\sqrt{\phi(r_t) } \ud B_t \label{LSN}
+\end{equation}
+
+In this model, the parameter $m$
 describes the approach towards the saddle-node bifurcation.  Estimates
@@
 
-We have shown that systems which undergo rare sudden transitions look
-statistically different from their counterparts that do not, even though
-they are driven by the same stochastic process.  In particular, such
-conditionally selected examples are more likely to show signs associated
-with an early warning of an approaching tipping point, such as increasing
-variance or increasing autocorrelation, as measured by Kendall's $\tau$.
-This increases the risk of false positives -- cases in which a warning
-signal being tested appears to have successfully detected an oncoming
-tipping point, when in fact the example comes
-instead from a stable system. Figure 2 shows the extent of this bias,
-where many of the chance crashes show values of $\tau$ that are 
-significantly larger than those observed in the otherwise identical
-replicates that did not experience a chance transition.
+The attempts to detect early warning signs for critical transitions
+are based on the concept of a deteriorating environment as embodied
+in a changing parameter~\cite{Scheffer2009}, which is a different kind
+of transition than one which is driven instead by stochasticity in an
+environment which is otherwise constant and exhibiting no directional
+change. When trying to use historical data to understand critical
+transitions we often do not know which category, changing environment
+or simply chance, an observed large change falls into.
+
+We have shown here that systems which undergo rare sudden transitions
+due to chance look statistically different from their counterparts that
+do not, even though they are driven by the same stochastic process.
+In particular, such conditionally selected examples are more likely to
+show signs associated with an early warning of an approaching tipping
+point, such as increasing variance or increasing autocorrelation,
+as measured by Kendall's $\tau$.  This increases the risk of false
+positives -- cases in which a warning signal being tested appears to
+have successfully detected an underlying change in the system leading
+to a tipping point, when in fact the example comes instead from a stable
+system with no underlying change in parameters. Figure 2 shows that many
+of the chance crashes show values of $\tau$ that are significantly larger
+than those observed in the otherwise identical replicates that did not
+experience a chance transition, thus ``detecting'' an underlying change
+in the system dynamics that is not in fact present.
 
@@
 
-It seems tempting to argue that the bias towards positive detection
-in historical examples is not problematic -- each of these systems did
-indeed collapse, so the increased probability
-of exhibiting warning signals could be taken as a successful detection.
-Unfortunately this is not the case. At the moment the forecast is made,
-these systems are no more likely to crash than their counterparts from
-the sample that did not collapse, just as a fair coin is still a fair
-coin after a string of consecutive ``heads.''  A closer look at
-the patterns involved shows why common indicators
-such as autocorrelation and variance can be misleading.
+It seems tempting to argue that this bias towards positive detection in
+historical examples is not problematic -- each of these systems did
+indeed collapse, so the increased probability of exhibiting warning
+signals could be taken as a successful detection.  Unfortunately this
+is not the case. At the moment the forecast is made, these systems are
+no more likely to crash than their counterparts from the sample that did
+not collapse, just as a fair coin is still a fair coin after a string of
+consecutive ``heads.''  A closer look at the patterns involved shows why
+common indicators such as autocorrelation and variance can be misleading.
 
-As the system gets farther from its stable point, it it more likely to draw a random step that
-returns it towards the stable point. Despite this, there is always some probability that
-it will move further still, so systems that do cross the tipping point must do
-so rather quickly by a string of events.  This pattern, clearly visible before the crashes in each of
-the examples in Figure 1, produces a string of observations that appear
-more highly autocorrelated (if we are sampling the system frequently
-enough to catch the excursion at all) than we observe in the rest of the
-fluctuations around the equilibrium.  Yet this autocorrelation comes from
-a chance trajectory moving quickly \emph{away} from the stable state,
-not from the critical slowing down pattern in the return times to the
-stable state which precede a saddle-node bifurcation and motivate the
-early warning signal.
+As the system gets farther from its stable point, it it more likely to
+draw a random step that returns it towards the stable point. Despite
+this, there is always some probability that it will move further still,
+so systems that do cross the tipping point must do so rather quickly by
+a string of events.  This pattern, clearly visible before the crashes
+in each of the examples in Figure 1, produces a string of observations
+that appear more highly autocorrelated (if we are sampling the system
+frequently enough to catch the excursion at all) than we observe in the
+rest of the fluctuations around the equilibrium.  Yet this autocorrelation
+comes from a chance trajectory moving quickly \emph{away} from the
+stable state, not from the critical slowing down pattern in the return
+times to the stable state which precede a saddle-node bifurcation and
+motivate the early warning signal.
 
@@
 
+\subsection{Importance of experimental approaches}
 The problem we highlight ultimately stems from the difficulty of having

### 2012-08-30 prosecutor/fallacy.tex
fixed equation formating

@@
 below the threshold the population decreases to zero.
-The model can be represented as a continuous time birth-death process where births and deaths are Poisson events whcih depend on the current density,
+The model can be represented as a continuous time birth-death process where births and deaths are Poisson events which depend on the current density with rates given by
 
 \begin{align}
-    b(X) &= \frac{e K n^2}{n^2 + h^2}, \\
-    d(X) &= e n + a,
-
+    b(n) &= \frac{e K n^2}{n^2 + h^2}, \label{master} \\
+    d(n) &= e n + a, 
 \end{align}

### 2012-09-03 prosecutor/fallacy.tex
updated discussion

@@
 \newif\ifhavefigures
-%\havefigurestrue
+\havefigurestrue
 \newif\ifnohavefigures % allows alternate condition
-\nohavefigurestrue     % turn on the alternate condition uncommenting this
+%\nohavefigurestrue     % turn on the alternate condition uncommenting this
 
@@
   \begin{center}
-    \subfigure[\textbf{True criminal} A bifurcation
-      drives a transition into the undesirable state]{
       \ifhavefigures
-%        \includegraphics[width=.4\textwidth]{figure1a}
+        \includegraphics[width=\textwidth]{Figure1}
       \fi
       \ifnohavefigures
-        \hspace{.4\textwidth}
+        \hspace{\textwidth}
       \fi
       \label{fig:bifurcation}
-    } 
-    \subfigure[\textbf{Innocent suspect} A
-    transition to the alternative stable state has occurred purely by
-    chance.]{
-    \ifhavefigures
-%      \includegraphics[width=.4\textwidth]{figure1b}
-    \fi
-    \ifnohavefigures
-      \hspace{.4\textwidth}
-    \fi
-    \label{fig:fallacy} }
-  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. Plots of system
-  state versus potential in two cases.  Potential is the negative integral
-  of the mean of dynamical equation.  The potential function gives an
-  intuitive picture of the stability of a system by imagining the curve as
-  a surface on which a ball is free to bounce across -- wells correspond
-  to stable points and peaks to unstable points.  Early warning signals
-  are aimed at detecting systems which are slowly moving towards a tipping
-  point or bifurcation, illustrated in the successively darker potential
-  curves in panel~\subref{fig:bifurcation}.  An example trajectory from
-  a simulation under this process shows the state of the system as the
-  potential moves towards the bifurcation point. The state of the system
-  in the simulation is given by the horizontal axis, while the vertical
-  dimension is used to show time on the right-hand axis.  Some stable
-  systems, panel~\subref{fig:fallacy}, may experience chance transitions
-  to an alternate stable state without the presence of this slow change
-  in the shape of the potential well which tips the system off to the left.
-  An example of the trajectory of such a chance transition is
-  plotted over the potential.  While such events may be unlikely over
-  short time-scales, selecting historical examples of systems that
-  have experienced sudden transitions increases the probability of
-  accidentally using chance events depicted in panel~\subref{fig:fallacy}
-  to test methods designed to detect the kind of process depicted in
-  panel~\subref{fig:bifurcation}.} \label{fig:1}
+  \end{center} \caption{\textbf{The Prosecutor's Fallacy}. 
+\textbf{(a)} Plot of the model functions shown in Eq~\eqref{master}
+with parameters $a=180$, $K=500$, $e=.5$, and $h=200$.  When the death
+rate is higher than the birth rate, the system dynamics drive the state
+(population size) to smaller values.  When birth rate is higher, the
+system moves right, as indicated by the arrows.\textbf{(b)} The potential
+energy is given by the negative integral of $b(n)-d(n)$, shown in the
+lower plot.   The potential function gives an intuitive picture of
+the stability of a system by imagining the curve as a surface on which
+a ball is free to bounce across -- wells correspond to stable points
+and peaks to unstable points. While most trajectories remain near the
+stable well, some transition out merely by chance.  An example of such
+a trajectory is shown in the top panel, in which time increases along
+the vertical axis. Though initially oscillating around the stable state,
+a chance excursion carries it beyond the threshold (vertical dotted line).
+Such chance trajectories can produce the statistical patterns as observed
+in true critical transitions seen in panel \textbf{(c)}: Early warning
+signals  are aimed at detecting systems which are slowly moving towards
+a tipping point or bifurcation, illustrated in the successive curves
+(detiorating and critical). Top panel: An example trajectory from a
+simulation under this process shows the state of the system as the
+potential moves towards the bifurcation point.
+} \label{fig:1}
 \end{figure}
@@
 we simulate a system that is not driven towards a bifurcation such as
-in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
+in Fig~ref{fig:1}(b).  This simulation approach allows us to determine whether
 examining historical events is a valid way to test the utility of these
@@
 \begin{align}
-    b(n) &= \frac{e K n^2}{n^2 + h^2}, \label{master} \\
+    b(n) &= \frac{ K n^2}{n^2 + h^2}, \label{master} \\
     d(n) &= e n + a, 
@@
 %  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
-%    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
+%    b_n &= \frac{K n^2}{n^2 + h^2}, \\
 %    d_n &= e n + a,
@@
 
-a model with a linear death rate and density-dependent birth
-rate that drives the Allee effect at low densities and limits
-growth at high densities.  In this model $n$ indicates the discrete
-number of individuals in the population, $K$ indicates a carrying 
-capacity as set by a limiting resource, $e$ a per-capita death rate (the $e$ scaling term in the birth equation allows the carrying capacity $K$ to correspond to a positive equilibrium point), 
-$a$ an additional mortality imposed on the population such as harvest,
-$h$ is a parameter controlling at what population size the addition of 
-more individuals switches from conferring a positive benefit on growth 
-from Allee interactions $n < h$ to a negative impact on growth due to 
-increased competition, $n > h$.  The key feature of this model is the
-alternate stable states introduced by this effect; other functional
-forms for Eq.~\eqref{master} could serve equally well for these
-simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
-system can be forced through a bifurcation by increasing
+a model with a linear death rate and density-dependent birth rate
+that drives the Allee effect at low densities and limits growth at high
+densities.  In this model $n$ indicates the discrete number of individuals
+in the population, $K$ indicates a carrying capacity as set by a limiting
+resource, $e$ a per-capita death rate (the $e$ scaling term in the birth
+equation allows the carrying capacity $K$ to correspond to a positive
+equilibrium point), $a$ an additional mortality imposed on the population
+such as harvest, $h$ is a parameter controlling at what population size
+the addition of more individuals switches from conferring a positive
+benefit on growth from Allee interactions $n < h$ to a negative impact
+on growth due to increased competition, $n > h$.  The key feature of
+this model is the alternate stable states introduced by this effect;
+other functional forms for Eq.~\eqref{master} could serve equally
+well for these simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.
+Though this system can be forced through a bifurcation by increasing
 the death rate, in these simulations all parameters are held constant
@@
 computed values of Kendall's $\tau$ for both indicators for each of
-the 266 replicates.
-Kendall's $\tau$ is a non-parametric measure of rank correlation frequently used to identify an increasing trend ($\tau > 0 )$ in early warning signals~\citep{Dakos2008, Dakos2011}, defined as $\tau \def \tfrac{(\text{number of concordant pairs}) - (\text{number of discordant pairs})}{\frac{1}{2} n (n-1) }$ in $n$ observations.
-\footnote{A pair of observations $(x_i,y_i)$ and $(x_j, y_j)$ are concordant if $x_i > x_j$ and $y_i > y_j$ or $x_i < x_j$ and $y_i < y_j$ and discordant otherwise; equalties excepted.}  $\tau$ takes values in $(-1, 1)$.
-The distribution
-of $\tau$ values observed across these replicates is shown in
-Figure~\ref{fig:indicator}.   
-%The common statistical significance test used for Kendall's $\tau$
-%assumes independence of the variables to achieve a normal distribution,
-%this isn't satisfied for indicators coming from a time series and
-%computed with a sliding window.  For a more robust comparison, 
-We compare the distribution of $\tau$ from all the simulations to
-the distribution conditioned on experiencing a chance transition to the
-alternative stable state.  To avoid an effect of sample size the time series are all chosen
-to be the same length.  
-%As these examples have been selected from a system
-%that is not approaching a bifurcation, an indicator that does not suffer
-%from the bias of Prosecutor's fallacy should have the same distribution
-%in both cases.   
-Annotated, parallelised code to replicate the analysis
-and produce the figures is provided in the appendix.
+the 266 replicates.  Kendall's $\tau$ is a non-parametric measure
+of rank correlation frequently used to identify an increasing trend
+($\tau > 0 )$ in early warning signals~\citep{Dakos2008, Dakos2011},
+defined as $\tau \def \tfrac{(\text{number of concordant pairs}) -
+(\text{number of discordant pairs})}{\frac{1}{2} n (n-1) }$ in $n$
+observations.  \footnote{A pair of observations $(x_i,y_i)$ and $(x_j,
+y_j)$ are concordant if $x_i > x_j$ and $y_i > y_j$ or $x_i < x_j$ and
+$y_i < y_j$ and discordant otherwise; equalities excepted.}  $\tau$ takes
+values in $(-1, 1)$.  The distribution of $\tau$ values observed across
+these replicates is shown in Figure~\ref{fig:indicator}.  We compare
+the distribution of $\tau$ from all the simulations to the distribution
+conditioned on experiencing a chance transition to the alternative
+stable state.  To avoid an effect of sample size the time series are
+all chosen to be the same length.
+
+To demonstrate the effect we observe is not unique to models with Allee
+effects, we provide an example of the effect arising in a discrete-time
+model with two non-zero stable states adapted from~\citep{May1977},
+\begin{equation}
+X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{s ^ Q + H ^ Q} \right). \label{May1977}
+\end{equation}
+Annotated, parallel code to replicate the analysis presented here is
+also provided in the appendix.
 
@@
     \ifhavefigures
-%    \includegraphics[width=6in]{figure2}
+    \includegraphics[width=6in]{figure2}
     \fi
@@
 
-It seems tempting to argue that this bias towards positive detection in
-historical examples is not problematic -- each of these systems did
+It seems tempting to argue that this bias towards positive detection
+in historical examples is not problematic -- each of these systems did
 indeed collapse, so the increased probability of exhibiting warning
-signals could be taken as a successful detection.  Unfortunately this
-is not the case. At the moment the forecast is made, these systems are
-no more likely to crash than their counterparts from the sample that did
-not collapse, just as a fair coin is still a fair coin after a string of
-consecutive ``heads.''  A closer look at the patterns involved shows why
+signals could be taken as a successful detection.  Unfortunately this is
+not the case. At the moment the forecast is made, these systems are not
+likely to transition, since they experience a strong pull towards the
+original stable state.  A closer look at the patterns involved shows why
 common indicators such as autocorrelation and variance can be misleading.
@@
 
+\subsection{ The truncation of observations }
+
+If we had a complete knowledge of the system dynamics, then we could
+eliminate the bias we observe here since the bias arises from the
+transient branch of the trajectory that crosses the threshold, and
+if the system were truncated at the minimum of the potential then the
+effect we emphasize here would not appear.  But, it is not possible to
+truncate the system in any practical application. The precise location of
+the minimum of the potential which is the location of the deterministic
+equilibrium is unknown. Moreover, under the hypothesis that the system is
+approaching a critical transition, the location of the minimum potential
+moves so it cannot easily be estimated by previous observations, (see
+Figure 1c where the equilibrium point moves in the direction of the
+transition). Thus it is neither practical nor desirable to suggest that
+historical time series can be used by following a simple truncation rule
+that avoids the branch of a trajectory crossing the threshold to another
+basin of attraction.  Exactly where a particular study will choose
+to truncate such a trajectory will necessarily be arbitrary without an
+underlying model of the process.  Frequently this is done by removing
+the very steep, monotonic branch of the trajectory expected once the
+system crosses the unstable threshold.  Such an approach corresponds
+with our choice of termination and produces the bias we discuss here.
+
+The examples of Figure 1, though only single replicates, may be useful
+in illustrating these issues.  Figure 1c, top panel shows a sample
+trajectory of a system with a parameter shift, while 1b shows a trajectory
+without a shift.  Both trajectories become more highly autocorrelated
+and higher variance near the end of the time series (time increases
+on the y axis in Figure 1).  The part of the time series following
+the critical transition shows a fast and monotonic trajectory to the
+unstable trajectory, and would usually be excluded by an analysis for
+warning signals in advance of the transition.  No such clear pattern
+exists prior to the transition in Figure 1b.   An alternative proposal
+to terminate the trajectory in panel B earlier would also risk decreasing
+the signal seen in panel c, and would be inconsistent with the application
+of warning signals in the forecasting context, where there would be no
+such truncation.
+
 \subsection{Comparing to the model-based method}
 
-In our numerical experiment, the model-based estimate of early warning signals
-appears more robust than the summary statistics, producing the same
-estimates on both the conditionally selected replicates as on a random
-sample of the replicates.  This is a consequence of the more rigid
-specifications that come with a model-based approach -- the pattern
+In our numerical experiment, the model-based estimate of early warning
+signals appears more robust than the summary statistics, producing
+the same estimates on both the conditionally selected replicates as on
+a random sample of the replicates.  This is a consequence of the more
+rigid specifications that come with a model-based approach -- the pattern
 expected is less general than any increase in variance or autocorrelation,
@@
 
-\subsection{Importance of experimental approaches}
-The problem we highlight ultimately stems from the difficulty of having
-only a single realization with which to examine a complex problem.
-The only way to deal with this problem embodied is through replication, as
-can be done in an experimental system in laboratory manipulations
-such as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole lake
-ecosystems in~\citet{Carpenter2011}.  
-Experimental procedures avoid the hazard of the Prosecutor's fallacy by
-generating a complete sample of replicates, rather than selecting a subset
-of cases from some larger historical sample.  
-%We note as well that the Prosecutors Fallacy 
-%may arise in other contexts where we search historical data for
-%examples containing some feature of interest.
+\subsection{Importance of experimental approaches} The problem we
+highlight ultimately stems from the difficulty of having only a single
+realization with which to examine a complex problem.  The only way
+to deal with this problem embodied is through replication, as can
+be done in an experimental system in laboratory manipulations such
+as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole
+lake ecosystems in~\citet{Carpenter2011}.  Experimental procedures avoid
+the hazard of the Prosecutor's fallacy by generating a complete sample
+of replicates, rather than selecting a subset of cases from some larger
+historical sample.
 
@@
 T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
-manuscript.
+manuscript, and also P. Ditlevsen and an anonymous reviewer for their 
+comments.
+
+\section{References}%bibliography
+\bibliography{fallacy}
+
+ \end{document}
 
@@
 \end{thebibliography}
-
- \end{document}

### 2012-09-04 prosecutor/fallacy.tex
reply letter done, waiting on final appendix runs & formating

@@
  \end{document}
-
- \section{References}%bibliography
-
- \begin{thebibliography}{24}
-\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
-\expandafter\ifx\csname url\endcsname\relax
-  \def\url#1{\texttt{#1}}\fi
-\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
-
-\bibitem[{Berkes et~al.(2006)Berkes, Hughes, Steneck, Wilson, Bellwood, Crona,
-  Folke, Gunderson, Leslie, Norberg, Nystr\"{o}m, Olsson, Osterblom, Scheffer,
-  and Worm}]{Berkes2006}
-Berkes, F., Hughes, T.~P., Steneck, R.~S., Wilson, J.~A., Bellwood, D.~R.,
-  Crona, B., Folke, C., Gunderson, L.~H., Leslie, H.~M., Norberg, J.,
-  Nystr\"{o}m, M., Olsson, P., Osterblom, H., Scheffer, M., Worm, B., Mar.
-  2006. {Ecology. Globalization, roving bandits, and marine resources.} Science
-  (New York, N.Y.) 311~(5767), 1557--8.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16543444}
-
-\bibitem[{Boettiger and Hastings(2012)}]{Boettiger2012b}
-Boettiger, C., Hastings, A., 2012. {Quantifying Limits to Detection of Early
-  Warning for Critical Transitions}. Journal Of The Royal Society Interface.
-
-\bibitem[{{California Supreme Court}(1968)}]{PeopleCollins1968}
-{California Supreme Court}, 1968. People v. collins.
-
-\bibitem[{Carpenter(2011)}]{Carpenter2011}
-Carpenter, J., Feb. 2011. {May the Best Analyst Win}. Science 331~(6018),
-  698--699.
-\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.331.6018.698}
-
-\bibitem[{Carpenter and Brock(2006)}]{Carpenter2006}
-Carpenter, S.~R., Brock, W.~A., 2006. {Rising variance: a leading indicator of
-  ecological transition.} Ecology letters 9~(3), 311--8.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16958897}
-
-\bibitem[{Courchamp et~al.(2008)Courchamp, Berec, and
-  Gascoigne}]{Courchamp2008}
-Courchamp, F., Berec, L., Gascoigne, J., 2008. {Allee Effects in Ecology and
-  Conservation}. Oxford University Press, USA.
-\newline\urlprefix\url{http://www.amazon.com/Effects-Ecology-Conservation-Franck-Courchamp/dp/0198570309}
-
-\bibitem[{Dai et~al.(2012)Dai, Vorselen, Korolev, and Gore}]{Dai2012}
-Dai, L., Vorselen, D., Korolev, K.~S., Gore, J., May 2012. {Generic Indicators
-  for Loss of Resilience Before a Tipping Point Leading to Population
-  Collapse}. Science 336~(6085), 1175--1177.
-\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1219805}
-
-\bibitem[{Dakos et~al.(2011)Dakos, K\'{e}fi, Rietkerk, Nes, and
-  Scheffer}]{Dakos2011}
-Dakos, V., K\'{e}fi, S., Rietkerk, M., Nes, E. H.~V., Scheffer, M., 2011.
-  {Slowing Down in Spatially Patterned Ecosystems at the Brink of Collapse}.
-  The American Naturalist.
-
-\bibitem[{Dakos et~al.(2008)Dakos, Scheffer, van Nes, Brovkin, Petoukhov, and
-  Held}]{Dakos2008}
-Dakos, V., Scheffer, M., van Nes, E.~H., Brovkin, V., Petoukhov, V., Held, H.,
-  Sep. 2008. {Slowing down as an early warning signal for abrupt climate
-  change.} Proceedings of the National Academy of Sciences 105~(38), 14308--12.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/18787119}
-
-\bibitem[{Ditlevsen and Johnsen(2010)}]{Ditlevsen2010}
-Ditlevsen, P.~D., Johnsen, S.~J., Oct. 2010. {Tipping points: Early warning and
-  wishful thinking}. Geophysical Research Letters 37~(19), 2--5.
-\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2010/2010GL044486.shtml}
-
-\bibitem[{Drake and Griffen(2010)}]{Drake2010}
-Drake, J.~M., Griffen, B.~D., Sep. 2010. {Early warning signals of extinction
-  in deteriorating environments}. Nature 467~(7314), 456--459.
-\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature09389}
-
-\bibitem[{Guttal and Jayaprakash(2008)}]{Guttal2008}
-Guttal, V., Jayaprakash, C., Dec. 2008. {Spatial variance and spatial skewness:
-  leading indicators of regime shifts in spatial ecological systems}.
-  Theoretical Ecology 2~(1), 3--12.
-\newline\urlprefix\url{http://www.springerlink.com/index/10.1007/s12080-008-0033-1}
-
-\bibitem[{Hirota et~al.(2011)Hirota, Holmgren, {Van Nes}, and
-  Scheffer}]{Hirota2011}
-Hirota, M., Holmgren, M., {Van Nes}, E.~H., Scheffer, M., Oct. 2011. {Global
-  Resilience of Tropical Forest and Savanna to Critical Transitions}. Science
-  334~(6053), 232--235.
-\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1210657}
-
-\bibitem[{K\'{e}fi et~al.(2007)K\'{e}fi, Rietkerk, Alados, Pueyo, Papanastasis,
-  Elaich, and de~Ruiter}]{Kefi2007}
-K\'{e}fi, S., Rietkerk, M., Alados, C.~L., Pueyo, Y., Papanastasis, V.~P.,
-  Elaich, A., de~Ruiter, P.~C., Sep. 2007. {Spatial vegetation patterns and
-  imminent desertification in Mediterranean arid ecosystems.} Nature
-  449~(7159), 213--7.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17851524}
-
-\bibitem[{Lenton(2011)}]{Lenton2011}
-Lenton, T.~M., Jun. 2011. {Early warning of climate tipping points}. Nature
-  Climate Change 1~(4), 201--209.
-\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nclimate1143}
-
-\bibitem[{Lenton et~al.(2012)Lenton, Livina, and Dakos}]{Lenton2012}
-Lenton, T.~M., Livina, V., Dakos, V., 2012. {Early warning of climate tipping
-  points from critical slowing down: comparing methods to improve robustness}.
-  Philosophical Transactions of The Royal Society A (in press).
-\newline\urlprefix\url{http://sites.google.com/site/noraclpreprints/contributions/lenton-et-al/Early\_warning\_methods\_revised\_with\_figs\_and\_appendix2.pdf}
-
-\bibitem[{Livina and Lenton(2007)}]{Livina2007}
-Livina, V.~N., Lenton, T.~M., Feb. 2007. {A modified method for detecting
-  incipient bifurcations in a dynamical system}. Geophysical Research Letters
-  34~(3), 1--5.
-\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2007/2006GL028672.shtml}
-
-\bibitem[{Mumby et~al.(2007)Mumby, Hastings, and Edwards}]{Mumby2007}
-Mumby, P.~J., Hastings, A., Edwards, H.~J., Nov. 2007. {Thresholds and the
-  resilience of Caribbean coral reefs.} Nature 450~(7166), 98--101.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17972885}
-
-\bibitem[{Scheffer et~al.(2009)Scheffer, Bascompte, Brock, Brovkin, Carpenter,
-  Dakos, Held, van Nes, Rietkerk, and Sugihara}]{Scheffer2009}
-Scheffer, M., Bascompte, J., Brock, W.~A., Brovkin, V., Carpenter, S.~R.,
-  Dakos, V., Held, H., van Nes, E.~H., Rietkerk, M., Sugihara, G., 2009.
-  {Early-warning signals for critical transitions.} Nature 461~(7260), 53--9.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/19727193}
-
-\bibitem[{Scheffer et~al.(2001)Scheffer, Carpenter, Foley, Folke, and
-  Walker}]{Scheffer2001}
-Scheffer, M., Carpenter, S.~R., Foley, J.~A., Folke, C., Walker, B., Oct. 2001.
-  {Catastrophic shifts in ecosystems.} Nature 413~(6856), 591--6.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/11595939}
-
-\bibitem[{Thompson and Sieber(2010)}]{Thompson2010}
-Thompson, J. M.~T., Sieber, J., Dec. 2010. {Climate tipping as a noisy
-  bifurcation: a predictive technique}. IMA Journal of Applied Mathematics
-  76~(1), 27--46.
-\newline\urlprefix\url{http://imamat.oxfordjournals.org/cgi/doi/10.1093/imamat/hxq060}
-
-\bibitem[{Thompson and Schumann(1987)}]{Thompson1987}
-Thompson, W.~C., Schumann, E.~L., 1987. {Interpretation of statistical evidence
-  in criminal trials: The prosecutor's fallacy and the defense attorney's
-  fallacy.} Law and Human Behavior 11~(3), 167--187.
-\newline\urlprefix\url{http://psycnet.apa.org/journals/lhb/11/3/167/
-  http://doi.apa.org/getdoi.cfm?doi=10.1007/BF01044641}
-
-\bibitem[{van Nes and Scheffer(2007)}]{vanNes2007}
-van Nes, E.~H., Scheffer, M., Jun. 2007. {Slow recovery from perturbations as a
-  generic indicator of a nearby catastrophic shift.} The American naturalist
-  169~(6), 738--47.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17479460}
-
-\bibitem[{Veraart et~al.(2011)Veraart, Faassen, Dakos, van Nes, L\"{u}rling,
-  and Scheffer}]{Veraart2011}
-Veraart, A.~J., Faassen, E.~J., Dakos, V., van Nes, E.~H., L\"{u}rling, M.,
-  Scheffer, M., Dec. 2011. {Recovery rates reflect distance to a tipping point
-  in a living system}. Nature, 2--5.
-\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature10723}
-
-\end{thebibliography}

### 2012-09-04 prosecutor/fallacy.tex
copy sent to alan

@@
 \makeatother
+
+\usepackage[hyphens]{url}
+\usepackage[pdftex]{color}
+\usepackage[breaklinks=true,linktocpage,colorlinks]{hyperref}
+\hypersetup{breaklinks=true, pdfborder={0 0 0}}
 
@@
 \begin{equation}
-X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{s ^ Q + H ^ Q} \right). \label{May1977}
+X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{X_t ^ Q + H ^ Q} \right). \label{May1977}
 \end{equation}
-Annotated, parallel code to replicate the analysis presented here is
-also provided in the appendix.
+which combines a logistic growth model with a saturating predator response (See~\citet{May1977} for detailed discussion), shown in Figure~\ref{fig:may}.
+Code to replicate the analysis can be found at \\
+\begin{sloppy}\href{https://github.com/cboettig/earlywarning/tree/prosecutor/}{https://github.com/cboettig/earlywarning/tree/prosecutor/}\end{sloppy}.
 
@@
   simulations, the pattern arises instead from the Prosecutor's fallacy
-  of conditional selection.}
+  of conditional selection. (the slight bimodality in the null distribution
+  arises from limited sample size).}
   \label{fig:indicator}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \ifhavefigures
+    \includegraphics[width=6in]{figure3}
+    \fi
+  \end{center}
+  \caption{The identical analysis from Figure~\ref{fig:indicator} is shown for the model in Eq~\eqref{May1977} using parameters $r = 0.75$, $K=10$, $a=1.7$, $Q=3$, and $H=1$.
+A similar statistical bias, particularly towards positive values of $tau$ occurs in this model as well.}
+  \label{fig:may}
 \end{figure}
@@
 
-\section{References}%bibliography
 \bibliography{fallacy}

### 2012-09-06 prosecutor/fallacy.tex
(re)submitted copies to PRSB

--- 2012-09-03 prosecutor/fallacy.tex
@@
 \linenumbers
-%\usepackage{float}
 \usepackage{subfigure}
-%\usepackage[pdftex]{color}
-%\definecolor{darkblue}{rgb}{0,0,0.5}
-%\definecolor{darkgreen}{rgb}{0,0.5,0}
-%\usepackage[pdftex, colorlinks]{hyperref}
 \textwidth 6.75in
@@
 \newif\ifnohavefigures % allows alternate condition
-%\nohavefigurestrue     % turn on the alternate condition uncommenting this
+\nohavefigurestrue     % turn on the alternate condition uncommenting this
 
@@
       \ifhavefigures
-        \includegraphics[width=\textwidth]{Figure1}
+%        \includegraphics[width=\textwidth]{Figure1}
       \fi
@@
 \begin{equation}
-X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{s ^ Q + H ^ Q} \right). \label{May1977}
+X_{t+1} = X_t \exp\left( r \left(1 - \frac{ X_t }{ K } \right) - \frac{ a * X_t ^ {Q - 1} }{X_t ^ Q + H ^ Q} \right). \label{May1977}
 \end{equation}
-Annotated, parallel code to replicate the analysis presented here is
-also provided in the appendix.
+which combines a logistic growth model with a saturating predator response (See~\citet{May1977} for detailed discussion), shown in Figure~\ref{fig:may}.
+Code to replicate the analysis can be found at \\
+\verb|https://github.com/cboettig/earlywarning/tree/prosecutor/|.
 
@@
     \ifhavefigures
-    \includegraphics[width=6in]{figure2}
+%    \includegraphics[width=6in]{Figure2}
+    \fi
+     \ifnohavefigures
+     \hspace{6in}
     \fi
@@
   \label{fig:indicator}
+\end{figure}
+
+\begin{figure}
+  \begin{center}
+    \ifhavefigures
+%    \includegraphics[width=6in]{Figure3}
+    \fi
+    \ifnohavefigures
+     \hspace{6in}
+    \fi
+  \end{center}
+  \caption{The identical analysis from Figure~\ref{fig:indicator} is shown for the model in Eq~\eqref{May1977} using parameters $r = 0.75$, $K=10$, $a=1.7$, $Q=3$, and $H=1$.
+A similar statistical bias, particularly towards positive values of $tau$ occurs in this model as well.}
+  \label{fig:may}
 \end{figure}
@@
 
-\section{References}%bibliography
-\bibliography{fallacy}
-
- \end{document}
-
- \section{References}%bibliography
-
- \begin{thebibliography}{24}
+\begin{thebibliography}{25}
 \expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
@@
   (New York, N.Y.) 311~(5767), 1557--8.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16543444}
 
 \bibitem[{Boettiger and Hastings(2012)}]{Boettiger2012b}
-Boettiger, C., Hastings, A., 2012. {Quantifying Limits to Detection of Early
-  Warning for Critical Transitions}. Journal Of The Royal Society Interface.
-
-\bibitem[{{California Supreme Court}(1968)}]{PeopleCollins1968}
-{California Supreme Court}, 1968. People v. collins.
+Boettiger, C., Hastings, A., May 2012. {Quantifying limits to detection of
+  early warning for critical transitions}. Journal of The Royal Society
+  Interface 9~(75), 2527 -- 2539.
 
@@
   698--699.
-\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.331.6018.698}
 
@@
   ecological transition.} Ecology letters 9~(3), 311--8.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/16958897}
 
@@
   Conservation}. Oxford University Press, USA.
-\newline\urlprefix\url{http://www.amazon.com/Effects-Ecology-Conservation-Franck-Courchamp/dp/0198570309}
 
@@
   Collapse}. Science 336~(6085), 1175--1177.
-\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1219805}
 
@@
   change.} Proceedings of the National Academy of Sciences 105~(38), 14308--12.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/18787119}
 
@@
   wishful thinking}. Geophysical Research Letters 37~(19), 2--5.
-\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2010/2010GL044486.shtml}
 
@@
   in deteriorating environments}. Nature 467~(7314), 456--459.
-\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature09389}
 
@@
   Theoretical Ecology 2~(1), 3--12.
-\newline\urlprefix\url{http://www.springerlink.com/index/10.1007/s12080-008-0033-1}
 
@@
   334~(6053), 232--235.
-\newline\urlprefix\url{http://www.sciencemag.org/cgi/doi/10.1126/science.1210657}
 
@@
   449~(7159), 213--7.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17851524}
 
@@
   Climate Change 1~(4), 201--209.
-\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nclimate1143}
 
@@
   Philosophical Transactions of The Royal Society A (in press).
-\newline\urlprefix\url{http://sites.google.com/site/noraclpreprints/contributions/lenton-et-al/Early\_warning\_methods\_revised\_with\_figs\_and\_appendix2.pdf}
 
@@
   34~(3), 1--5.
-\newline\urlprefix\url{http://www.agu.org/pubs/crossref/2007/2006GL028672.shtml}
+
+\bibitem[{May(1977)}]{May1977}
+May, R.~M., Oct. 1977. {Thresholds and breakpoints in ecosystems with a
+  multiplicity of stable states}. Nature 269~(5628), 471--477.
 
@@
   resilience of Caribbean coral reefs.} Nature 450~(7166), 98--101.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17972885}
 
@@
   {Early-warning signals for critical transitions.} Nature 461~(7260), 53--9.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/19727193}
 
@@
   {Catastrophic shifts in ecosystems.} Nature 413~(6856), 591--6.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/11595939}
+
+\bibitem[{{Supreme Court}(1968)}]{PeopleCollins1968}
+{Supreme Court}, C., 1968. {People v. Collins}.
 
@@
   76~(1), 27--46.
-\newline\urlprefix\url{http://imamat.oxfordjournals.org/cgi/doi/10.1093/imamat/hxq060}
 
@@
   fallacy.} Law and Human Behavior 11~(3), 167--187.
-\newline\urlprefix\url{http://psycnet.apa.org/journals/lhb/11/3/167/
-  http://doi.apa.org/getdoi.cfm?doi=10.1007/BF01044641}
 
@@
   169~(6), 738--47.
-\newline\urlprefix\url{http://www.ncbi.nlm.nih.gov/pubmed/17479460}
 
@@
   in a living system}. Nature, 2--5.
-\newline\urlprefix\url{http://www.nature.com/doifinder/10.1038/nature10723}
 
 \end{thebibliography}
+ \end{document}

### 2013-02-13 inst/examples/fallacy.Rmd
clean up

`ro warning=FALSE, message=FALSE, comment=NA, tidy=FALSE, cache=TRUE, verbose=TRUE, cache.path="fallacy/" or`

# Code for Prosecutors Fallacy 
<!--

`r require(knitcitations)`
This code is written in the `R` language for statistical computing.  
Population dynamics are simulated using the `populationdynamics` package
`r citep(citation("populationdynamics"))` for exact simulations of 
discrete birth-death processes in continuous time using the Gillespie
agorithm `r citep("10.1021/j100540a008")`.  Early warning signals
of variance and autocorrelation, as well as the model-based estimate
of `r citet("10.1098/rsif.2012.0125")` are estimated using the 
`earlywarning` package `r citep(citation("earlywarning")`.  These
packages can be installed from Github using the `devtools` R package

```r
library(devtools)
install_github("populationdynamics", "cboettig")
install_github("earlywarning", "cboettig")
```

For the individual-based simulation, the population dynamics are given by

<div>
\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
    d_n &= e n + a,
\end{align}
</div>

which is provided by the `saddle_node_ibm` model in `populationdynamics`.

-->

For each of the warning signal statistics in question, 
we need to generate the distibution over all replicates
and then over replicates which have been selected conditional 
on having experienced a crash.

We begin by running the simulation of the process for all replicates.

Load the required libraries

``` {r libraries}
library(populationdynamics)
library(earlywarning)
library(reshape2)		# data manipulation
library(data.table)	# data manipulation
library(ggplot2)		# graphics
library(snowfall)		# parallel
````

```{r plotting-theme}
theme_publish <- theme_set(theme_bw(12))
theme_publish <- 
  theme_update(legend.key=theme_blank(),
        panel.grid.major=theme_blank(),panel.grid.minor=theme_blank(),
        plot.background=theme_blank(), legend.title=theme_blank())
```

### Conditional distribution

Then we fix a set of paramaters we will use for the simulation function.  Since we will simulate 20,000 replicates with 50,000 pts a piece, we can save memory by performing the conditional selection on the ones that crash as we go along and disgard the others.  (We will create a null distribution in which we ignore this conditional selection later).

``` {r simdatf eval=FALSE}
select_crashes <- function(n){
	T<- 5000
	n_pts <- n
	pars = c(Xo = 500, e = 0.5, a = 180, K = 1000, h = 200,
    i = 0, Da = 0, Dt = 0, p = 2)
	sn <- saddle_node_ibm(pars, times=seq(0,T, length=n_pts), reps=1000)
	d <- dim(sn$x1)
	crashed <- which(sn$x1[d[1],]==0)
	sn$x1[,crashed] 
}
````

To take advantage of parallelization, we loop over this function a set number of times.  The `snowfall` library provides the parallelization
of the `lapply` loop.  A few extra commands format the data into a table
with columns of times, replicate id number, and population value at the
given time.

``` {r simdat}
sfInit(parallel=TRUE, cpu=12)
sfLibrary(populationdynamics)
sfExportAll()
examples <- sfLapply(1:20, function(i) select_crashes(50000))
dat <- melt(as.matrix(as.data.frame(examples, check.names=FALSE)))
names(dat) = c("time", "reps", "value")
levels(dat$reps) <- 1:length(levels(dat$reps)) # use numbers for reps
````

``` {r testing}
ggplot(subset(dat, reps %in% levels(dat$reps)[1:9])) + geom_line(aes(time, value)) + facet_wrap(~reps, scales="free")
````

Zoom in on the relevant area of data near the crash

``` {r subsetdata, dependson="simdat" }
require(plyr)
zoom <- ddply(dat, "reps", function(X){
    tip <- min(which(X$value<threshold))
#print(tip)
    index <- max(tip-200,1):tip
    data.frame(time=X$time[index], value=X$value[index])
    })
````

``` {r example-trajectories, dependson="subsetdata"}
ggplot(subset(zoom, reps %in% levels(zoom$reps)[1:9])) + geom_line(aes(time, value)) + facet_wrap(~reps, scales="free")
````

Compute model-based warning signals on all each of these.

``` {r plots_fallacy, dependson="subsetdata"}
dt <- data.table(subset(zoom, value>threshold))
var <- dt[, warningtrend(data.frame(time=time, value=value), window_var), by=reps]$V1
acor <- dt[, warningtrend(data.frame(time=time, value=value), window_autocorr), by=reps]$V1
dat <- melt(data.frame(Variance=var, Autocorrelation=acor))
````

### Null distribution

To compare against the expected distribution of these statistics, we create another set of simulations without conditioning on having experienced a chance transition, on which we perform the identical analysis.

``` {r simdatf_null, eval=FALSE}
threshold <- 250
select_crashes <- function(n){
	T<- 5000
	n_pts <- n
	pars = c(Xo = 500, e = 0.5, a = 180, K = 1000, h = 200,
    i = 0, Da = 0, Dt = 0, p = 2)
	sn <- saddle_node_ibm(pars, times=seq(0,T, length=n_pts), reps=500)
	d <- dim(sn$x1)
	sn$x1[1:501,]
}
````

``` {r simdat_null, dependson="simdatf_null"}
sfInit(parallel=TRUE, cpu=12)
sfLibrary(populationdynamics)
sfExportAll()
examples <-  sfLapply(1:24, function(i) select_crashes(50000))
nulldat <- melt(as.matrix(as.data.frame(examples, check.names=FALSE)))
nulldat <- melt(examples)
names(nulldat) = c("time", "reps", "value")
levels(nulldat$reps) <- 1:length(levels(dat$reps)) 
````

Zoom in on the relevant area of data near the crash

``` {r nullzoom, dependson="simdat_null"}
require(plyr)
nullzoom <- ddply(nulldat, "reps", function(X){
    data.frame(time=X$time, value=X$value)
    })
````

``` {r nullmelt, dependson="nullzoom"}
nulldt <- data.table(nullzoom)
nullvar <- nulldt[, warningtrend(data.frame(time=time, value=value), window_var), by=reps]$V1
nullacor <- nulldt[, warningtrend(data.frame(time=time, value=value), window_autocorr), by=reps]$V1
nulldat <- melt(data.frame(Variance=nullvar, Autocorrelation=nullacor))
````

``` {r figure2, dependson="nullmelt"}
ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.2, alpha=.5) +
 facet_wrap(~variable) + xlim(c(-1, 1)) + 
 geom_density(data=nulldat, aes(value), bw=0.2)
````

### 2013-02-13 inst/examples/fallacy.Rmd

@@
 # Code for Prosecutors Fallacy
-<!--
-
-`r require(knitcitations)`
-This code is written in the `R` language for statistical computing.  
-Population dynamics are simulated using the `populationdynamics` package
-`r citep(citation("populationdynamics"))` for exact simulations of 
-discrete birth-death processes in continuous time using the Gillespie
-agorithm `r citep("10.1021/j100540a008")`.  Early warning signals
-of variance and autocorrelation, as well as the model-based estimate
-of `r citet("10.1098/rsif.2012.0125")` are estimated using the 
-`earlywarning` package `r citep(citation("earlywarning")`.  These
-packages can be installed from Github using the `devtools` R package
-
-```r
-library(devtools)
-install_github("populationdynamics", "cboettig")
-install_github("earlywarning", "cboettig")
-```
 
@@
 which is provided by the `saddle_node_ibm` model in `populationdynamics`.
-
--->
 
@@
 
-``` {r simdatf eval=FALSE}
+``` {r simdatf}
+threshold <- 250
 select_crashes <- function(n){
@@
 
-``` {r simdat}
-sfInit(parallel=TRUE, cpu=12)
+``` {r parallel}
+sfInit(parallel=FALSE)
 sfLibrary(populationdynamics)
+```
+
+``` {r simdat, dependson="simdatf"}
 sfExportAll()
@@
 
-``` {r testing}
+``` {r testing, dependson="simdat"}
 ggplot(subset(dat, reps %in% levels(dat$reps)[1:9])) + geom_line(aes(time, value)) + facet_wrap(~reps, scales="free")
@@
 
-``` {r simdatf_null, eval=FALSE}
+``` {r simdatf_null}
 threshold <- 250
@@
 ``` {r simdat_null, dependson="simdatf_null"}
-sfInit(parallel=TRUE, cpu=12)
-sfLibrary(populationdynamics)
 sfExportAll()
-examples <-  sfLapply(1:24, function(i) select_crashes(50000))
+examples <-  sfLapply(1:10, function(i) select_crashes(50000))
 nulldat <- melt(as.matrix(as.data.frame(examples, check.names=FALSE)))
@@
 
-``` {r figure2, dependson="nullmelt"}
-ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.2, alpha=.5) +
+``` {r fig2, dependson="nullmelt"}
+ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
  facet_wrap(~variable) + xlim(c(-1, 1)) + 
- geom_density(data=nulldat, aes(value), bw=0.2)
+ geom_density(data=nulldat, aes(value), adjust=2) + xlab("Kendall's tau") + theme_bw()
+
 ````
+
+``` {r figure2, dev="CairoPS", fig.ext="eps", fig.width=6, fig.height=5, include=FALSE}
+ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
+ facet_wrap(~variable) + xlim(c(-1, 1)) + 
+ geom_density(data=nulldat, aes(value), adjust=2) + xlab("Kendall's tau") + theme_bw()
+
+````

### 2013-03-04 prosecutor/fallacy.tex
minor

@@
 \newif\ifnohavefigures % allows alternate condition
-\nohavefigurestrue     % turn on the alternate condition uncommenting this
+%\nohavefigurestrue     % turn on the alternate condition uncommenting this
 
@@
       \ifhavefigures
-%        \includegraphics[width=\textwidth]{Figure1}
+        \includegraphics[width=\textwidth]{Figure1}
       \fi
@@
 the vertical axis. Though initially oscillating around the stable state,
-a chance excursion carries it beyond the threshold (vertical dotted line).
+a chance excursion carries it beyond the Allee threshold (vertical dotted line).
 Such chance trajectories can produce the statistical patterns as observed
@@
 a tipping point or bifurcation, illustrated in the successive curves
-(detiorating and critical). Top panel: An example trajectory from a
+(deteriorating and critical). Top panel: An example trajectory from a
 simulation under this process shows the state of the system as the
-potential moves towards the bifurcation point.
+potential moves towards the bifurcation point. The original position of the Allee threshold
+is shown by the vertical dotted line (though it moves slightly as the parameter changes).  
 } \label{fig:1}
@@
     \ifhavefigures
-%    \includegraphics[width=6in]{Figure2}
+    \includegraphics[width=6in]{Figure2}
     \fi
@@
     \ifhavefigures
-%    \includegraphics[width=6in]{Figure3}
+    \includegraphics[width=6in]{Figure3}
     \fi
@@
 
- \section{Acknowledgements}
+ \section{Acknowledgments}
 This research was supported by funding from NSF Grant EF 0742674 to AH
-and a Computational Sciences Graduate Fellowship from the Department of
-Energy grant DE-FG02-97ER25308 to CB. The authors thank M. Baskett, 
-T.A. Perkins and N. Ross for helpful comments on earlier drafts of the
-manuscript, and also P. Ditlevsen and an anonymous reviewer for their 
-comments.
+and a Computational Sciences Graduate Fellowship from the Department
+of Energy grant DE-FG02-97ER25308 and NERSC Supercomputing grant
+DE-AC02-05CH11231 to CB. The authors thank M. Baskett, T.A. Perkins
+and N. Ross for helpful comments on earlier drafts of the manuscript,
+and also P. Ditlevsen and an anonymous reviewer for their comments.
 

### 2013-06-12 inst/examples/Figure2.Rmd

--- 2013-02-13 inst/examples/fallacy.Rmd
@@
-`ro warning=FALSE, message=FALSE, comment=NA, tidy=FALSE, cache=TRUE, verbose=TRUE, cache.path="fallacy/" or`
+`ro warning=FALSE, message=FALSE, comment=NA, tidy=FALSE, cache=TRUE, verbose=TRUE, cache.path="figure2/" or`
 
@@
     })
+save(zoom, file="zoom.rda")
 ````
@@
 ````
+
+``` {r save_final_data}
+write.csv(dat, file="Figure2_dat.csv")
+write.csv(nulldat, file="Figure2_nulldat.csv")
+```

### 2013-06-12 inst/doc/CommentReply.tex
submitted version of Figure (pdf)

\documentclass{article} 
\usepackage{hyperref}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.

\usepackage{setspace, lineno}
\doublespacing
\linenumbers

\title{No early warning signals for stochastic transitions: insights from large
       deviation theory}
\author{%
  Carl Boettiger\thanks{Center for Stock Assessment Research, Department of Applied Math and Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA 95064, USA} \and Alan Hastings\thanks{Department of Environmental Science and Policy, University of California Davis, 1 Shields Avenue, Davis, CA 95616 USA}}
\begin{document}

\maketitle

In Boettiger \& Hastings {[}1{]} we demonstrated that conditioning on
observing a purely stochastic transition from one stable basin to
another could generate time-series trajectories that could be mistaken
for an early warning signal of a critical transition (such as might be
due to a fold bifurcation {[}2{]}), when instead the shift is merely due
to chance. While the goal was to highlight a potential danger in mining
historical records for patterns showing sudden shifts when seeking to
test early warning techniques, Drake {[}3{]} draws attention to a
potentially more interesting consequence of our analysis. Drake argues
that the bias observed could be used to forecast purely stochastic
transitions -- a task previously thought to be impossible {[}4{]}. We
feel this interpretation is too generous. The pattern Drake points to
arises in any large deviation, regardless of whether a system is or is
not at elevated risk for a transition. We illustrate this pattern in
systems with and without bistability, demonstrating that early warning
signals do not exist for purely stochastic transitions.

Here we provide a numerical demonstration that the pattern in question
for consideration of an early warning signal appears not only before
purely stochastic transitions (as seen in Reference 1) but during any
large deviation. As large deviations can occur even in stochastic
systems that have only a single stable point, these patterns cannot be
considered indicators of stochastic transitions. We demonstrate this in
two scenarios: first using the Allee model of alternative stable states
considered in Reference 1, Eqn 2.1 - 2.2 and Figure 2, and then in a
simple Ornstein-Uhlenbeck (OU) model which has only a single stable
state. Rather than condition on a stochastic transition having occurred
(as in Reference 1), we now condition on having merely observed a
sufficiently large deviation. It does not matter precisely what
``large'' deviation is considered, only that the larger the deviation
the more replicates or longer simulation times will be needed to
sufficiently populate the sample. We pick values such that we get a
sample of a few hundred large deviation events in a sample of 20,000
replicates.

The OU model is defined by a stochastic differential equation in which
there is only a single optimum whose strength is proportional to the
displacement,

\[ dX_t = - \alpha X_t dt + \sigma dB_t, \]

where the state $X_t$ oscillates around a stable point (at zero in these
arbitrary units), driven by Brownian noise $dB_t$ of intensity $\sigma$
and restorative force $\alpha$.

The analysis for each model proceeds exactly as in Reference 1: For each
model we generate 20,000 replicate time series. We condition upon only
those experiencing a deviation of size $L$ ($X \leq 250$ in the Allee
model and $X \leq -4$ in the OU model). For the sequence of observations
immediately leading up to the large deviation we compute the warning
signals of variance and autocorrelation over a sliding window of half
the length of the time-series, and we summarise the increasing or
decreasing trend observed in the variance and autocorrelation using
Kendall's $\tau$ rank correlation coefficient (all following the method
for early warning indicators outlined in Reference 5). We repeat this
analysis on the entire set of time-series under each model to obtain
null distributions for $\tau$ statistic.

\begin{figure}[htbp]
\centering
\includegraphics{Figure1.pdf}
\caption{Figure 1. Histogram shows the frequency the correlation
statistic $\tau$ observed for each warning signal (variance,
autocorrelation coefficient) on the large deviation samples from each
model. Background distribution of all samples show by smooth line
(kernel density estimate). More positive values of tau are supposed to
indicate a rising indicator which can be a signal of an approaching
transition {[}2{]}. The OU model uses $\alpha = 5$, $\sigma=3.5$,
$t \in (0, 10)$, 2000 replicates, 20,000 sample points each.
Conditionally selected trajectories experiencing a deviation of at least
-4, and analyzed the 1,500 data points prior to the threshold to
determine a warning signal (following Reference 5). (Code at:
https://raw.github.com/cboettig/earlywarning/resubmission/inst/doc/Figure1.Rmd,
data at:
https://raw.github.com/cboettig/earlywarning/resubmission/inst/doc/Figure1.csv}
\end{figure}

We find (Figure 1) that $\tau$ is significantly skewed towards positive
values when conditioning on large deviations in both models. This
demonstrates that it is the presence of the large deviation, not the
presence of the stochastic transition we condition on in Reference 1,
that is responsible for this pattern (just as we claimed without example
then).

Observing the bias shown in the figures here depends on having a rapid
enough sample frequency to capture the escape trajectory and a long
enough trajectory for the statistic to demonstrate an increase over
time. Since large deviations due to stochastic forces alone must be
fast, so must the accompanying warning signal and management response
(which will show up on the time scale of the perturbation). Note that
fast relative to the system dynamics may or may not be fast relative to
the timescale of management (just as with bifurcation-driven warning
signals, Reference 6). The wider null distribution in the OU model
results from the sample window being shorter relative to the system
timescale.

One might consider this a corollary of the Prosecutor's Fallacy we
originally presented, which demonstrated that examples of sudden
transitions historically selected from the literature could be mistaken
for positive evidence of early warning signals when they were in fact
due to purely stochastic transitions. Here we have seen how any large
deviation could be similarly misleading, whether or not it results in a
stochastic transition to an alternative stable state. From a classical
result of the large deviation theory one can gain considerable intuition
about why these chance deviations show much higher variance and
autocorrelation than expected from the stationary distribution of a
stable point. Though large deviations are rare -- the time we must wait
to observe a deviation of size $L$ in the system above scales as
$\exp\left(L^2/\sigma^2\right)$ (the familiar Arrhenius relationship),
when these deviations occur they occur very rapidly. The expected time
for an excursion to a distant point $L$ that does not again cross the
stable point before reaching $L$ scales as $\log(L/\sigma)$, just as a
trajectory returning down the gradient of the attractor from $L$ to the
stable point (proofs in Reference 7 or Reference 8). While most
trajectories in the stationary distribution take steps in each direction
with equal probability, these large deviations moving rapidly to the
boundary will consequently show the greater autocorrelation. In
achieving a much greater deviation than typically observed, these
trajectories will also show an increase in variance, as observed. That
such trajectories appear to be pulled in the direction of their escape
rather than climbing away against a restorative force has led to
confusion before. Reference 8 argues how this shows how a ``punctuated
equilibrium'' pattern of stasis followed by rapid change could arise
entirely from small steps, and Reference 9 empirically demonstrates this
phenomenon in the trajectories of local population extinctions.

In conclusion, we heartily agree with the need for a decision-theoretic
approach to early warning signal questions {[}10{]}. Central to a
decision-theoretic approach is enumerating alternative scenarios that
are possible given the observed data. We have highlighted how purely
stochastic transitions and large deviations are such possibilities. The
challenge of sufficient or unique early warning indicators is not
limited to stochastic shifts, but includes the more typical critical
transitions. For instance, rising variance or autocorrelation patterns
typical of fold bifurcations can be observed in more benign bifurcations
or smooth transitions {[}11{]}. Early warning signals may offer a
promising technique that will one day allow us to avoid seemingly
unpredictable catastrophes -- but we must not lose sight of just how
difficult are the challenges involved. A key step here and for early
warning indicators more generally is to understand these other
circumstances in which they can arise, that we may then develop ways to
eliminate those possibilities. Though we may never be able to detect
purely stochastic transitions, perhaps these approaches in this
discussion may lead to more unique and sufficient indicators for true
critical transitions.

The authors acknowledge the generous support of NSF grant to AH and
helpful comments from TA Perkins, as well as the reviews from an
anonymous reviewer and P. Ditlevsen.

1 Boettiger, C. \& Hastings, A. 2012 Early warning signals and the
prosecutor's fallacy. \emph{Proceedings of the Royal Society B:
Biological Sciences} (doi:10.1098/rspb.2012.2085)

2 Scheffer, M. et al. 2009 Early-warning signals for critical
transitions. \emph{Nature} \textbf{461}, 53--9.

3 Drake, J. M. 2013 Early warning signals of stochastic switching.
\emph{Proceedings of The Royal Society B} \textbf{in press}.

4 Ditlevsen, P. D. \& Johnsen, S. J. 2010 Tipping points: Early warning
and wishful thinking. \emph{Geophysical Research Letters} \textbf{37},
2--5. (doi:10.1029/2010GL044486)

5 Dakos, V., Scheffer, M., van Nes, E. H., Brovkin, V., Petoukhov, V. \&
Held, H. 2008 Slowing down as an early warning signal for abrupt climate
change. \emph{Proceedings of the National Academy of Sciences}
\textbf{105}, 14308--12. (doi:10.1073/pnas.0802430105)

6 Hughes, T. P., Linares, C., Dakos, V., van de Leemput, I. a \& van
Nes, E. H. 2013 Living dangerously on borrowed time during slow,
unrecognized regime shifts. \emph{Trends in ecology \& evolution}
\textbf{28}, 149--55. (doi:10.1016/j.tree.2012.08.022)

7 Ludwig, D. 1981 Escape from Domains of Attraction for Sytstems
Perturbed by Noise. In \emph{Nonlinear Phenomena in Physics and Biology}
(eds R. H. Enns B. L. Jones R. M. Miura \& S. S. Rangnekar), pp.
549--566. Boston, MA: Springer New York.

8 Lande, R. 1985 Expected time for random genetic drift of a population
between stable phenotypic states. \emph{Proceedings of the National
Academy of Sciences} \textbf{82}, 7641--7645.

9 Drake, J. M. \& Griffen, B. D. 2009 Speed of expansion and extinction
in experimental populations. \emph{Ecology letters} \textbf{12}, 772--8.
(doi:10.1111/j.1461-0248.2009.01325.x)

10 Boettiger, C. \& Hastings, A. 2012 Quantifying limits to detection of
early warning for critical transitions. \emph{Journal of The Royal
Society Interface} \textbf{9}, 2527--2539. (doi:10.1098/rsif.2012.0125)

11 Kéfi, S., Dakos, V., Scheffer, M., Van Nes, E. H. \& Rietkerk, M.
2012 Early warning signals also precede non-catastrophic transitions.
\emph{Oikos} (doi:10.1111/j.1600-0706.2012.20838.x)

\end{document}

### 2013-06-13 inst/doc/CommentReply.tex
line wrap, add dryad DOI

@@
 data at:
-https://raw.github.com/cboettig/earlywarning/resubmission/inst/doc/Figure1.csv}
+https://raw.github.com/cboettig/earlywarning/resubmission/inst/doc/Figure1.csv
+(these links should be replaced with the appropriate citation to Dryad,
+where these files are now deposited under provisional DOI:
+10.5061/dryad.1dj62)}
 \end{figure}
@@
 
-The authors acknowledge the generous support of NSF grant to AH and
-helpful comments from TA Perkins, as well as the reviews from an
-anonymous reviewer and P. Ditlevsen.
+The authors acknowledge the generous support of NSF grant EF 0742674 to
+AH and helpful comments from TA Perkins, an anonymous reviewer and
+reviewer P. Ditlevsen.
 

### 2013-07-11 inst/doc/sig-tests.Rmd
significance tests

```{r}
opts_chunk$set(comment=NA)
download.file("https://raw.github.com/cboettig/earlywarning/resubmission/inst/doc/Figure1.csv", dest="Figure1.csv", method="curl")
dat <- read.csv("Figure1.csv")
```

There should be some clever way in plyr or data.table to do these combinations automatically, but anyway:

Focusing on the OU model, we see that autocorrelation is not sufficiently different:

```{r}
x <- subset(dat, variable=="Autocorrelation" & data=="null" & model == "OU")$value
y <- subset(dat, variable=="Autocorrelation" & data=="conditional" & model == "OU")$value
wilcox.test(x,y, alternative="greater")
ks.test(x,y, alternative="greater")
```

Variances are more different:

```{r}
x <- subset(dat, variable=="Variance" & data=="null" & model == "OU")$value
y <- subset(dat, variable=="Variance" & data=="conditional" & model == "OU")$value
wilcox.test(x,y, alternative="greater")
ks.test(x,y, alternative="greater")
```

Clearer for Allee case:

```{r}
x <- subset(dat, variable=="Variance" & data=="null" & model == "Allee")$value
y <- subset(dat, variable=="Variance" & data=="conditional" & model == "Allee")$value
wilcox.test(x,y, alternative="greater")
ks.test(x,y, alternative="greater")
```

We opt not to include a comparison statistic since it is unclear to me what is most appropriate.  Wilcox test is a bit of a strange choice given the skew of the conditional distribution.  Kolmogorov-Smirnov is slightly more natural, but really hard to defend either one in this context, as distributions of Kendall's $\tau$ statistics is already a somewhat curious distribution - ideally we should consider how it should actually be distributed given the dynamical process as the null (which is what we generate numerically).

### 2013-07-11 inst/doc/sig-tests.Rmd
significance tests (use correct one-sided test)

@@
 y <- subset(dat, variable=="Autocorrelation" & data=="conditional" & model == "OU")$value
-wilcox.test(x,y, alternative="greater")
-ks.test(x,y, alternative="greater")
+wilcox.test(x,y, alternative="less") # x - y < 0 since y bigger than x
+ks.test(x,y, alternative="greater") # CDF x > then CDF y since more weight to the left
 ```
@@
 y <- subset(dat, variable=="Variance" & data=="conditional" & model == "OU")$value
-wilcox.test(x,y, alternative="greater")
-ks.test(x,y, alternative="greater")
+wilcox.test(x,y, alternative="less") # x - y < 0 since y bigger than x
+ks.test(x,y, alternative="greater") # CDF x > then CDF y since more weight to the left
 ```
@@
 y <- subset(dat, variable=="Variance" & data=="conditional" & model == "Allee")$value
-wilcox.test(x,y, alternative="greater")
-ks.test(x,y, alternative="greater")
+wilcox.test(x,y, alternative="less") # x - y < 0 since y bigger than x
+ks.test(x,y, alternative="greater") # CDF x > then CDF y since more weight to the left
 ```

### 2014-11-20 R/LSN.R
use earlywarning package

@@ -1,64 +1,0 @@
-
-###########  Linearized Saddle-Node (LSN) Model ##################
-#' Linearized Saddle Node model
-#'
-#' Estimate the mean and variance of the linearized saddle
-#' node process (in which the spring constant changes over time)
-#' @param Xo initial condition
-#' @param to initial time
-#' @param t1 final time
-#' @param pars numeric of parameters named Ro, theta, and sigma
-#' @return a list with values Ex, the expected x value, and 
-#' Vx, the expected variance
-#' @keywords internal
-#' @import deSolve
-LSN <- function(Xo, to, t1, pars){
-
-  R <- function(t, pars){pars[1] + pars[2]*t }
-
-  check <- any(R(t1,pars) < 0)
-	if(is.na(check) | check | pars[3] < 0){
-		Ex <- Xo
-		Vx <- rep(Inf,length(Xo))
-	} else {
-		moments <- function(t,y,p){ 
-			sqrtR <- sqrt(R(t,pars)) 
-			yd1 <- 2*sqrtR*(sqrtR+pars[3] - y[1]) 
-			yd2 <- -2*sqrtR*y[2] + p[4]^2*(sqrtR+pars[3])
-			list(c(yd1=yd1, yd2=yd2))
-		}
-		jacfn <- function(t,y,p){
-			sqrtR <- sqrt(R(t,pars)) 
-			c(
-			-2*sqrtR, 0,
-			0, -2*sqrtR
-		)}
-## The apply calls needed to work with vector inputs as Xo (whole timeseries)
-		times <- matrix(c(to, t1), nrow=length(to))
-		out <- lapply(1:length(Xo), function(i){
-			lsoda(y=c(xhat=Xo[i], sigma2=0), times=times[i,], func=moments, 
-			      parms=pars, jacfunc=jacfn) 
-		})
-		Ex <- sapply(1:length(Xo), function(i) out[[i]][2,2]) # times are in rows, cols are time, par1, par2
-		Vx <- sapply(1:length(Xo), function(i) out[[i]][2,3])
-	}
-# Handle badly defined parameters by creating very low probability returns,
-# needed particularly for the L-BFGS-B bounded method, which can't handle Infs 
-# and does rather poorly... Worth investigating a better way to handle this.  
-# Note errors can apply to some of the timeseries, possibly by estimates of m 
[…]

### 2014-11-20 R/bootstrap_trend.R

@@ -1,32 +1,0 @@
-#' parameteric bootstrap the correlation coefficient under both hypotheses
-#' 
-#' @param X data sampling time and sampled value as columns. If only
-#' one column is provided, assumes that the time is uniform
-#' @param indicator a function that returns the warning indicator over time,
-#' the correlation of the resulting series with time is returned parameter
-#' @param ... additional arguments to \link{warningtrend}
-#' @param reps number of replicate simulations to use.
-#' @return  data.frame with a column indicating the model used (null or test)
-#' and a column with the requested correlation coefficient, see \link{cor.test}
-#' @seealso \link{warningtrend}
-#' @examples
-#' data(ibm)
-#' bootstrap_trend(ibm_critical, window_var, method="pearson", rep=10)
-#' @export
-bootstrap_trend <- function(X, indicator, ..., reps=100){
-  A <- stability_model(X, "OU")
-  B <- stability_model(X, "LSN")
-  Asim <- simulate(A, reps)
-  Bsim <- simulate(B, reps)
-
-  Asim <- melt(Asim, id="time")
-  Bsim <- melt(Bsim, id="time")
-  names(Asim)[2] <- "rep"
-  names(Bsim)[2] <- "rep"
-
-  wsA <- ddply(Asim, "rep", warningtrend, indicator, ...)
-  wsB <- ddply(Bsim, "rep", warningtrend, indicator, ...)
-  tidy <- melt(data.frame(null=wsA[[2]], test=wsB[[2]]))
-  names(tidy) <- c("simulation", "value")
-  tidy
-}

### 2014-11-20 R/compare.R

@@ -1,34 +1,0 @@
-#' Bootstrap parameters and likelihood ratios 
-#'
-#' Simulate under both models and fit results to simulation
-#' @param A a fitted model with update and simulate methods
-#' @param B another model capable of fitting the same data object
-#' @return a list with: 
-#' @return AA the model A fitted to simulations from A 
-#' @return BB model B fitted to simulations from B
-#' @return AB the model A fitted to simulations from B 
-#' @return BA model B fitted to simulations from A
-#' @return likelihood estimates, 
-#' @return parameter values
-#' @details will repeat simulation and fitting if one of the model updates
-#' fails to converge on that data
-#' @export
-compare <- function(A, B){
-  done <- 0
-  while(!done){
-    Asim <- simulate(A)
-    AfitA <- update(A, X = Asim)
-    BfitA <- update(B, X = Asim)
-    done <- converge(AfitA) && converge(BfitA)
-  }
-  done <- 0
-  while(!done){
-    Bsim <- simulate(B)
-    BfitB <- update(B, X = Bsim)
-    AfitB <- update(A, X = Bsim)
-    done <- converge(AfitB) && converge(BfitB)
-  }
-  list(AA = AfitA, BB = BfitB, AB = AfitB, BA = BfitA)
-}
-
-

### 2014-11-20 R/constOU.R

@@ -1,25 +1,0 @@
-
-#' OU model for stable system
-#' 
-#' Estimate the appropriate mean and variance of the OU process
-#' @param Xo initial condition
-#' @param to initial time
-#' @param t1 final time
-#' @param pars numeric of parameters named Ro, theta, and sigma
-#' @return a list with values Ex, the expected x value, and 
-#' Vx, the expected variance
-#' @keywords internal
-constOU <- function(Xo, to, t1, pars){
-  Dt <- t1 - to
-  Ex <- pars["theta"]*(1 - exp(-pars["Ro"] * Dt)) + Xo *
-        exp(-pars["Ro"] * Dt) 
-  Vx <- 0.5 * pars["sigma"] ^ 2 * 
-       (1 - exp(-2 * pars["Ro"] * Dt)) / pars["Ro"]
-  if(pars['Ro'] < 0 ) Vx <- rep(Inf, length(Xo)) 
-  if(pars['sigma'] < 0 ) Vx <- rep(Inf, length(Xo)) 
-  return(list(Ex = Ex, Vx = Vx))
-}
-
-
-
-

### 2014-11-20 R/gauss.R

@@ -1,148 +1,0 @@
-
-# gauss class has:
-# data (data.frame), 
-# parameters (named vector numeric), 
-# model-name (character), 
-# log-likelihood (scalar numeric)
-# convergence (logical)
-# 
-
-#' converge generic test
-#' 
-#' @param object a class of model fit
-#' @param ... additional arguments 
-#' @return convergence status (logical)
-#' @export 
-converge <- function(object, ...) UseMethod("converge")
-
-#' Test if a gaussian model fit has converged
-#' 
-#' @param object a class of model fit
-#' @param ... additional arguments 
-#' @return convergence status (logical)
-#' @S3method converge gauss
-#' @method converge gauss
-converge.gauss <- function(object, ...){
-  object$convergence
-}
-
-#' extract the fitted model parameters
-#' 
-#' @param object a gauss class object
-#' @param ... aditional arguments (currently ignored)
-#' @return a (named) vector of the MLE parameters
-#' @S3method fitted gauss
-#' @method fitted gauss
-fitted.gauss <- function(object, ...){
-  object$pars
-}
-
-
-
-#' update routine for gaussian model fits
-#' 
-#' @param object any gauss-class object
-#' @param ... additional arguments, see details.  Must at least include X=
-#' @return an updated gauss-class object depending on other parameters given
-#' @details data has to be passed in as X= via the ..., see example.
-#' any commands to the optim routine used are also passed in in this manner.
-#' one can also add the option store_data = FALSE to avoid returning the 
-#' input data, see \code{\link{stability_model}} for details.  
-#' @S3method update gauss
-#' @method update gauss
-update.gauss <- function(object, ...){
-  stability_model(..., model = object$model, p = object$pars)
-}
-
-#' simulate method
-#' 
-#' @param object any gauss-class object
-#' @param ... additional arguments (currently ignored)
-#' @return a data frame with the time simulated values
-#' @S3method simulate gauss
-#' @method simulate gauss
-simulate.gauss <- function(object, nsim = 1, seed = NULL, ...){
-  if(object$model == "LSN") 
-    setmodel <- LSN
-  else if(object$model == "OU") 
-    setmodel <- constOU
-   time <- object$X[,1]
-   N <- length(time)
-   value <- sapply(1:nsim, function(j){
-     X <- numeric(N - 1)
-     X[1] <- object$X[1,2]
-     for(i in 1:(N - 1) ){
[…]

### 2014-11-20 R/get_replicates.R

@@ -1,26 +1,0 @@
-#' extract the replicates simulations by model  
-#' @param reps a list of outputs from compare()
-#' @return a data.frame listing the "time", "rep" id number,
-#' "value" observed at that time, and the "model" producing the 
-#' simulated data.  
-#' @seealso \link{compare}, \link{roc_data}
-#' @import reshape2
-#' @export 
-get_replicates <- function(reps){
-  ous <- sapply(reps, function(rep){
-          if(rep$model=="OU")
-            rep$X[[2]] 
-  })
-  lsns <- sapply(reps, function(rep){
-          if(rep$model=="LSN")
-            rep$X[[2]] 
-  })
-  dat <- melt(list(OU=ous, LSN=lsns))
-  names(dat) <- c("time", "rep", "value", "model")
-  dat$time <- reps[[1]][dat$time, 1]
-  dat
-}
-
-
-
-

### 2014-11-20 R/indicators.R

@@ -1,114 +1,0 @@
-
-
-## Potential indicator functions
-
-#' Compute variance over a sliding window
-#' @param X a numeric containing evenly sampled values from a time series
-#' @param windowsize the size of the sliding time window (in # of pts) used to 
-#' compute the statistic.  
-#' @return a numeric of size length(X)-windowsize of values of the statistic 
-#' calculated in the sliding window.  
-#' @seealso \link{warningtrend}
-#' @export
-window_var <- function(X, windowsize=length(X)/2){
-	sapply(0:(length(X)-windowsize), function(i){
-		var(X[(i+1):(i+windowsize)]) 
-	})
-}
-
-#' Compute the coefficient of variation (var/mean) in a sliding window
-#' @param X a numeric containing evenly sampled values from a time series
-#' @param windowsize the size of the sliding time window (in # of pts) used to 
-#' compute the statistic.  
-#' @return a numeric of size length(X)-windowsize of values of the statistic 
-#' calculated in the sliding window.  
-#' @seealso \link{warningtrend}
-#' @export
-window_cv <- function(X, windowsize=length(X)/2){
-	sapply(0:(length(X)-windowsize), function(i){
-			var(X[(i+1):(i+windowsize)]) / mean(X[(i+1):(i+windowsize)]) 
-		})
-}
-
-#' Compute mean over a sliding window
-#' @param X a numeric containing evenly sampled values from a time series
-#' @param windowsize the size of the sliding time window (in # of pts) used to 
-#' compute the statistic.  
-#' @return a numeric of size length(X)-windowsize of values of the statistic 
-#' calculated in the sliding window.  
-#' @seealso \link{warningtrend}
-#' @export
-window_mean <- function(X, windowsize=length(X)/2){
-	sapply(0:(length(X)-windowsize), function(i){
-		mean(X[(i+1):(i+windowsize)]) 
-	})
-}
-
-#' Compute skew in a sliding window
-#' @param X a numeric containing evenly sampled values from a time series
-#' @param windowsize the size of the sliding time window (in # of pts) used to 
-#' compute the statistic.  
-#' @return a numeric of size length(X)-windowsize of values of the statistic 
-#' calculated in the sliding window.  
-#' @seealso \link{warningtrend}
-#' @import psych
-#' @export
-window_skew <- function(X, windowsize=length(X)/2){
[…]

### 2014-11-20 R/lik_ratios.R

@@ -1,22 +1,0 @@
-#' extract the likelihood ratios under both null and test simulations
-#' @param reps a list of outputs from compare()
-#' @return a data.frame of columns "simulation" (null or test), 
-#' "value" (deviance between models on that simulation), 
-#' and "rep", a replicate id number.  
-#' @seealso \link{compare}, \link{roc_data}
-#' @import reshape2
-#' @export 
-lik_ratios <- function(reps){
-  dat <- sapply(reps, function(rep){
-          null <- -2*(logLik(rep$AA) - logLik(rep$BA))
-          test <- -2*(logLik(rep$AB) - logLik(rep$BB))
-          c(null=null, test=test)
-  })
-  dat <- melt(dat)
-  names(dat) <- c("simulation", "rep", "value") 
-  dat
-}
-
-
-
-

### 2014-11-20 R/parameter_bootstraps.R

@@ -1,32 +1,0 @@
-#' extract a data.frame of parameters from the bootstrapped replicate comparisons 
-#' @param reps a list of outputs from compare()
-#' @return a data.frame of columns "value" (parameter value), 
-#' "parameter" (name of the parameter), "fit", a double-letter code indicating
-#' first which model created the simulation and second which model was fit to the data,
-#' "rep", a replicate id number.  
-#' @details Consider ggplot2 tools to visualize this data, or use the subset() function 
-#' on this data set to look at a particular parameter from a particular model fit.
-#' see examples for illustrations.  
-#' @seealso \link{compare}, \link{roc_data}, \link{lik_ratios}, \link{subset}
-#' @import reshape2
-#' @export
-#' @examples
-#' data("example_analysis")
-#' pars <- parameter_bootstraps(reps)
-#' subset(pars, fit=="BB" && parameter=="m")
-#' require(ggplot2)
-#' require(Hmisc)
-#' ggplot(subset(pars, fit %in% c("AA", "BB")), aes(parameter, value, fill = fit)) +
-#'   stat_summary(fun.y = mean, geom = "bar", position = "dodge") +
-#'   stat_summary(fun.data = median_hilow, geom = "pointrange", aes(color = fit), 
-#'               position = position_dodge(width = 0.90), conf.int = 0.95)
-
-parameter_bootstraps <- function(reps){
-  dat <- lapply(reps, function(rep){
-          list(AA=as.list(fitted(rep$AA)), AB=as.list(fitted(rep$AB)), 
-               BA=as.list(fitted(rep$BA)), BB=as.list(fitted(rep$BB)))
-  })
-  dat <- melt(dat)
-  names(dat) <- c("value", "parameter", "fit", "rep") 
-  dat
-}

### 2014-11-20 R/roc_data.R

@@ -1,59 +1,0 @@
-#' Compute data to draw an ROC curve
-#'
-#' Grab the data for the ROC curve from the null & test distributions
-#' @param dat a data.frame with columns "simulation", indicating "null" or "test",
-#' "value", with the value of the test statistic, and "rep" indicating the replicate number
-#' Or set dat=NULL and specify null_dist and test_dist directly, see below.  
-#' @param pts (optional) number of output points desired, defaults to 50, 
-#' should not be more than the number of points in each distribution
-#' @param null_dist a distribution (vector of samples) of the test statistic 
-#' under the null hypothesis, if not given as a data frame
-#' @param test_dist the values under the test hypothesis
-#' @return a data frame with thresholds, false positive and 
-#' corresponding true positive rates
-#' @export
-roc_data <- function(dat, pts=50, null_dist=NULL, test_dist=NULL, silent=TRUE){
-  # Creates the ROC curve from the output of the bootstrap function
-
-  if(!is.null(dat)){
-    null_dist <- subset(dat, simulation=="null")$value
-    test_dist <- subset(dat, simulation=="test")$value
-  }
-
-  n_null <- length(null_dist)
-  n_test <- length(test_dist)
-
-
-  lower <- min(null_dist, test_dist)
-  upper <- max(null_dist, test_dist)
-  threshold <- seq(lower, upper, length=pts)
-
-  # Calculate the ROC curve
-  roc <- sapply(threshold, 
-                function(thresh){
-                  c(sum(null_dist > thresh)/n_null,
-                  sum(test_dist > thresh)/n_test)
-                })
-  roc <- t(roc)
-
-  # Caculate the AUC
-  f<-approx(roc[,1], roc[,2], n=200) 
-  delta <- f$x[2] - f$x[1]
-  area  <- sum(f$y*delta)
-
-  # display some summary information
-  if(!silent){
-    message(paste("Area Under Curve = ", area))
-    id <- which(roc[,1]<=.05)[1]
-    message(paste("True positive rate = ", roc[id,2], "at false positive rate of", roc[id,1]))
-  }
-  out <- cbind(threshold,roc)
-  out <- as.data.frame(out)
-  names(out) <- c("Threshold", "False.positives", "True.positives")
-  out
-}
-
-
-
-
-

### 2014-11-20 R/stability_model.R

@@ -1,77 +1,0 @@
-#' @title stability model fitting
-#'
-#' @param X data sampling time and sampled value as columns. If only
-#' one column is provided, assumes that the time is uniform
-#' @param model the model used
-#' @param p initial parameter guesses, if NULL, will guess from data
-#' @param ... additional parameters passed to the optim routine
-#' @param store_data if FALSE will not return an object with 
-#' a copy of the original data
-#' @return a gauss model fit object (list), containing
-#' the data (if requested), the model type used, the maximum-likelihood
-#' estimated parameters, the log-likelihood at those values, 
-#' and a convergence indicator
-#' @details depends on the gauss class of functions and the model 
-#' definitions LSN and constOU.  
-#' @import deSolve
-#' @export
-stability_model <- function(X, model=c("LSN", "OU"), p = NULL, ..., 
-                            store_data=TRUE){
-  model <- match.arg(model)
-  # reformat time series objects into proper data frames
-  if(is(X, "ts"))
-    X <- data.frame(as.numeric(time(X)), X@.Data)
-  # if time values are not provided
-  else if(is.null(dim(X)))
-    X <- data.frame(1:length(X), X)
-  
-
-  if(!is.null(p)){ ## got everything? then rock & roll
-    f1 <- switch(model, 
-                LSN = f_closure(X, LSN),
-                OU = f_closure(X, constOU))
-    o <- optim(p, f1, ...)
-
-  } else if(is.null(p)){ ## oh, need p? try:
-    p <- c(Ro=1/max(time(X[,1])), theta=mean(X[,2]), sigma=sd(X[,2]))
-    f2 <- f_closure(X, constOU)
-    o <- optim(p, f2, ...)
-
-    # if model is "OU", we're done.  otherwise:
-    if(model=="LSN"){
-      f3 <- f_closure(X, LSN) # switch to the LSN model
-      p_est <- o$par  # & use the OU estimated pars as starting guess
-      # but rescale them to the new definitions:
-      Ro <- as.numeric(p_est[1]^2)
-      theta <- as.numeric(p_est[2]+p_est[1])
-      sigma <- as.numeric(abs(p_est[3]/sqrt(2*p_est[1]+ p_est[2])))
-      p <- c(Ro=Ro, m=0, theta=theta, sigma=sigma)
-
-      ## now fit the LSN model
-      o <- optim(p, f3, ...)
-    }
-  }
[…]

### 2014-11-20 R/summary_statistic.R

@@ -1,28 +1,0 @@
-
-#' apply a summary statistic over a set of replicate data
-#' 
-#' @param dat a data frame with columns "time", "reps", and "value"
-#' observed in a timeseries of replicates.   
-#' @param indicator a function taking X, such as \link{window_var}
-#' @param ... additional optional arguments to the indicator function,
-#' such as windowsize
-#' @return a data frame with columns "time", "reps", and "value",
-#' where values are the sliding-window computed summary statistics produced
-#' by indicator.  
-#' @details helper function just helps simple indicator functions be 
-#' applied to data formated with replicate observations and still
-#' return nicely formatted data 
-#' @import plyr
-#' @import reshape2
-#' @export
-summary_statistic <- function(dat, indicator, ...){
-  tmp <- ddply(dat, "reps", function(X) 
-    indicator(X$value, ...))
-  stat <- melt(t(tmp))
-  stat <- stat[-1,]
-  names(stat) <- c("time", "reps", "value")
-  stat$time <- as.numeric(gsub("\\D", "", stat$time))
-  class(stat$value) <- "numeric"
-  stat
-}
-

### 2014-11-20 R/warningtrend.R

@@ -1,40 +1,0 @@
-#' calculate the correlation coefficient of some summary statistic over time
-#'
-#' @param X the data, either a time-series, a numeric, or a matrix
-#' with times in the first column and values in the second column. 
-#' @param indicator a function of a numeric string of values.  Additional
-#' arguments such as windowsize or explicit time vector can be passed in
-#' as optional arguments.  
-#' @param ... additional arguments to the indicator function
-#' @param method the correlation test performed, see ?cor.test for details.
-#' @seealso \link{cor.test} 
-#' @details supported indicators in the package so far include:
-#' \link{window_var}, \link{window_skew}, \link{window_autocorr}.  
-#' 
-#' Note that this function does not return the p-value
-#' given by the test, as this is not meaningful in this context,
-#' since the assumptions are not met by early warning signals data. 
-#' @export
-warningtrend <- function(X, indicator, windowsize=NULL, ...,
-                        method=c("kendall", "pearson", "spearman"))
-{
-  # Handle alternative data formats 
-  if(is.ts(X))
-    X <- data.frame(as.numeric(time(X)), X@.Data)
-  else if(is.null(dim(X)))
-    X <- data.frame(1:length(X), X)
-
-  npts <- length(X[["time"]])
-  if(is.null(windowsize))
-    windowsize <- npts/2
-
-  # compute the indicator with the function specified
-  Y <- indicator(X[["value"]], c(windowsize, ...)) 
-  # apply the correclation test indicated
-  method=match.arg(method)
-  correlation <- cor.test(X[windowsize:npts,1], Y, method=method)
-  correlation$estimate
-}
-
-
-

### 2014-11-20 comment-reply-code/beer.Rmd
rename

``` {r libraries}
rm(list=ls())
library(earlywarning)
library(reshape2)		# data manipulation
library(data.table)	# data manipulation
library(ggplot2)		# graphics
opts_chunk$set(warning = FALSE, message = FALSE, comment = NA, tidy = FALSE, 
    verbose = TRUE, cache = FALSE, fig.path="beer_")

````

### Conditional distribution

``` {r simdatf}
threshold <- -4
require(sde)
M <- 2000   # replicates
N <- 20000 # sample points
d <- expression(-5 * x)
s <- expression(3.5)
X <- sde.sim(X0=0, drift=d, T=10, sigma=s, N = N, M = M)
timeseries <- matrix(X@.Data, ncol=M)
# Get the ids of samples that have a point less than the theshold
w <- sapply(data.frame(timeseries), function(x) any(x < threshold))
# extract that subset by id 
W <- timeseries[,w]
sample <- 1500 # sample length
dev <- sapply(as.data.frame(W), which.min)
drop <- which(dev - sample < 1)
if(length(drop) > 0){
W <- W[,-drop]
dev <- dev[-drop]
}
D <- rbind((dev - sample), dev)
# extract just that range
M <- as.matrix(W)
dat <- sapply(1:length(dev), function(i) M[D[1,i]:D[2,i], i])
dat <- as.data.frame(dat)
dat <- as.data.frame(cbind(time = 1:dim(dat)[1], dat))
df <- melt(dat, id="time")
names(df) = c("time", "reps", "value")
levels(df$reps) <- 1:length(levels(df$reps)) # use numbers for reps instead of V1, V2, etc
zoom <- df
````

``` {r example-trajectories}
ggplot(subset(zoom, reps %in% levels(zoom$reps)[1:9])) + geom_line(aes(time, value)) + facet_wrap(~reps, scales="free")
````

Compute model-based warning signals on all each of these.

``` {r plots_fallacy}
dt <- data.table(zoom)
var <- dt[, warningtrend(data.frame(time=time, value=value), window_var), by=reps]$V1
acor <- dt[, warningtrend(data.frame(time=time, value=value), window_autocorr), by=reps]$V1
dat <- melt(data.frame(Variance=var, Autocorrelation=acor))
````

### Null distribution

To compare against the expected distribution of these statistics, we create another set of simulations without conditioning on having experienced a chance transition, on which we perform the identical analysis.

``` {r simdatf_null}
null <- timeseries[1000:1201,]
null <- as.data.frame(cbind(time = 1:dim(null)[1], null))
ndf <- melt(null, id="time")
names(ndf) = c("time", "reps", "value")
levels(ndf$reps) <- 1:length(levels(ndf$reps)) # use numbers for reps instead of V1, V2, etc
nulldt <- data.table(ndf)
nullvar <- nulldt[, warningtrend(data.frame(time=time, value=value), window_var), by=reps]$V1
nullacor <- nulldt[, warningtrend(data.frame(time=time, value=value), window_autocorr), by=reps]$V1
nulldat <- melt(data.frame(Variance=nullvar, Autocorrelation=nullacor))
````

``` {r fig}
ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
 facet_wrap(~variable) + xlim(c(-1, 1)) + 
 geom_density(data=nulldat, aes(value), adjust=2) + xlab("Kendall's tau") + theme_bw()
````

``` {r beer, dev="CairoPS", fig.ext="eps", fig.width=6, fig.height=5, include=FALSE}
ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
 facet_wrap(~variable) + xlim(c(-1, 1)) + 
 geom_density(data=nulldat, aes(value), adjust=2) + xlab("Kendall's tau") + theme_bw()

````

``` {r save_final_data}
write.csv(dat, file="ou_dat.csv")
write.csv(nulldat, file="ou_nulldat.csv")
```

### 2014-11-20 comment-reply-code/comment.Rmd
rename

--- 2013-02-13 inst/examples/fallacy.Rmd
@@
-`ro warning=FALSE, message=FALSE, comment=NA, tidy=FALSE, cache=TRUE, verbose=TRUE, cache.path="fallacy/" or`
-
-# Code for Prosecutors Fallacy
+``` {r libraries}
+rm(list=ls())
+library(earlywarning)
+library(reshape2)		# data manipulation
+library(data.table)	# data manipulation
+library(ggplot2)		# graphics
+opts_chunk$set(warning=FALSE, message=FALSE, comment=NA, tidy=FALSE, verbose=TRUE, cache=FALSE)
+````
 
@@
 
-``` {r libraries}
+``` {r libs}
 library(populationdynamics)
-library(earlywarning)
-library(reshape2)		# data manipulation
-library(data.table)	# data manipulation
-library(ggplot2)		# graphics
 library(snowfall)   # parallel
@@
 Then we fix a set of paramaters we will use for the simulation function.
-Since we will simulate 20,000 replicates with 50,000 pts a piece, we can save memory by performing the conditional selection on the ones that crash as we go along and disgard the others.
+Since we will simulate 20,000 replicates with 50,000 pts a piece, we can save memory by performing the conditional selection on the ones that cross a threshold we go along and disgard the others.
 (We will create a null distribution in which we ignore this conditional selection later).
@@
   d <- dim(sn$x1)
-	crashed <- which(sn$x1[d[1],]==0)
+  crashed <- which(sn$x1[d[1],] <= threshold)
+# crashed <- which(sn$x1[d[1],]==0)
   sn$x1[,crashed] 
@@
 
-To take advantage of parallelization, we loop over this function a set number of times.  The `snowfall` library provides the parallelization
-of the `lapply` loop.  A few extra commands format the data into a table
+ A few extra commands format the data into a table
 with columns of times, replicate id number, and population value at the
@@
     tip <- min(which(X$value<threshold))
-#print(tip)
     index <- max(tip-200,1):tip
@@
 
-``` {r fig2, dependson="nullmelt"}
+``` {r fig, dependson="nullmelt"}
 ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
@@
 
-``` {r figure2, dev="CairoPS", fig.ext="eps", fig.width=6, fig.height=5, include=FALSE}
+``` {r save_final_data}
+write.csv(dat, file="allee.csv")
+write.csv(nulldat, file="allee_nulldat.csv")
+```
+
+``` {r comment, dev="CairoPS", fig.ext="eps", fig.width=6, fig.height=5, include=FALSE}
 ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +

### 2014-11-20 doc/no-signals-for-stochastic-transitions/sig-tests.Rmd
rename

--- 2013-07-11 inst/doc/sig-tests.Rmd
@@
 
-There should be some clever way in plyr or data.table to do these combinations automatically, but anyway:
+(there should be some clever way in plyr or data.table to do these combinations automatically, but anyway)
 
-Focusing on the OU model, we see that autocorrelation is not sufficiently different:
+Focusing on the OU model, we do see some 'significant' difference in autocorrelation distributions:
 

### 2014-11-20 prosecutor-fallacy-code/warning_signal_slides.Rmd
rename

```{r }
#opts_knit$set(upload.fun = socialR::flickr.url)

require(ggplot2)
require(reshape2)
theme_publish <- theme_set(theme_bw(14))

lower <- 0
upper <- 700
x <- seq(lower, upper, length.out=100)
e <- 0.5 
K <- 1000
h <- 200

a <- seq(120, 220, by=20)

d0 <- sapply(a, function(a){
b <- function(n)  (0.5 * K * n ^ 2) / (n ^ 2 + h ^ 2)
d <- function(n) e * n + a
f <- function(n) b(n) - d(n)
bx <- sapply(x, b) 
dx <- sapply(x, d) 
growth <- bx-dx
})

d1 <- melt(data.frame(State=x, d0), id="State")
ggplot(d1) + 
  geom_line(aes(State, value, lty=variable), lwd=1) + 
  ylab("growth rate")

```

```{r }
a <- 180
u1 <- -1 * sapply(x, function(x) integrate(f, lower, x)$value)
a <- 190
u2 <- -1 * sapply(x, function(x) integrate(f, lower, x)$value)
a <- 200
u3 <- -1 * sapply(x, function(x) integrate(f, lower, x)$value)
```

```{r chance_crash_data} 
load("data/zoom.rda")
dat <- subset(zoom, reps==2)
#dat <- subset(dat, value > 250)
a <- dat$time - min(dat$time)
a <- a/max(a)*(34000-19000) + 19000
d2 <- data.frame(x = dat$value, value = a)

```

```{r Figure1b, tidy=FALSE}
df <- data.frame(x=x, value=u1)
df$variable = "stable potential"
d2$variable = "stable potential"
df$panel <- "Potential Energy"
d2$panel <- "Time of Sample Trajectory"
d <- rbind(df, d2)
p1 <- ggplot(data = d, mapping = aes(x = x, y = value)) + 
  facet_grid(panel ~ ., scale = "free") +
  layer(data = d2, geom = "point", size=1) +
  layer(data = df,  mapping = aes(lty=variable),  geom = c("line")) +
  geom_vline(aes(xintercept=250), lty=4) +
  coord_cartesian(ylim=c(18000, 35000)) + 
  xlab("State") + ylab("arbitrary units") +
  opts(title = "(b) Trajectory of a chance \n transition from a stable system")

write.csv(d, "Figure1_panelb.csv")
```

simulate an example with a bifurcation for comparison.

```{r crash_sim, cache=TRUE}
  library(populationdynamics)
	T<- 5000
	n_pts <- 50000
	pars = c(Xo = 500, e = 0.5, a = 180, K = 1000, h = 200,
    i = 0, Da = .5, Dt = 0, p = 2)
	sn <- saddle_node_ibm(pars, times=seq(0,T, length=n_pts), reps=1)
  X <- sn$x1
  X <- X[X > 0]
  times <- 1:length(X)*T/n_pts
  a <- times - min(times) 
  a <- a/max(a)*(34000-19000) + 19000
  d3 <- data.frame(x=X, value=a)
````

```{r Figure1c, tidy=FALSE}
df <- data.frame(x=x, "stable potential"=u1, "deteriorating"=u2, "critical"=u3)
d1 <- melt(df, id="x")
d1$panel <- "Potential Energy"
d3$panel <- "Time of Sample Trajectory"
d3$variable <- d3$x # match col numbers to rbind
d <- rbind(d1, d3)
p2 <- ggplot(data = d, mapping = aes(x = x, y = value)) + 
  facet_grid(panel ~ ., scale = "free") +
  layer(data = d3, geom = "point", size=1) +
  layer(data = d1,  mapping = aes(lty=variable), geom = c("line")) +
  geom_vline(aes(xintercept=250), lty=4) +
  coord_cartesian(ylim=c(18000, 35000)) + xlab("State") + ylab("arbitrary units") +
  opts(title="(c) Trajectory during an \n underlying parameter shift")
write.csv(d, "Figure1_panelc.csv")
```

```{r Figure1_upload, fig.width=15}
require(grid)
pushViewport(viewport(layout = grid.layout(1, 3)))
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)
print(p0, vp = vplayout(1, 1))
print(p1, vp = vplayout(1, 2))
print(p2, vp = vplayout(1, 3))
```

```{r Figure1_orig, fig.width=15, dev="Cairo_pdf"}
require(grid)
pushViewport(viewport(layout = grid.layout(1, 3)))
vplayout <- function(x, y) viewport(layout.pos.row = x, layout.pos.col = y)
print(p0, vp = vplayout(1, 1))
print(p1, vp = vplayout(1, 2))
print(p2, vp = vplayout(1, 3))
```

### 2014-11-20 prosecutor-fallacy-code/README.Rmd
add some notes to the README

## Reproducing the full analysis

All simulation data and analyses can be generated and reproduced following the code
provided in the R scripts, Figure1.Rmd, Figure2.Rmd, and Figure3.Rmd.  These files
are dynamic documents, and can be run directly in R using the `knitr` package command,
such as:

```r
library(knitr)
knit("Figure2.Rmd")
```

Note that `Figure1.Rmd` uses data generated the `Figure2.Rmd` script, so run that one first!

which will generate a file Figure1.md, a plain text (markdown) file of the results
with links to the figures. After running `knit`, all variables are available at the R command line for further exploration as well.

## Reproducing the figures

Raw data for the results, Figures 2 and 3, is also provided as csv files for convience.  
Each figure includes two files, one for the null data and one for the conditional data.
Column values are: `X`, the index, `variable`, the indicator (variance or autocorrelation),
and `value`, the value of Kendall's tau.  The figures can be recreated in R from the
csv files using the code below.

``` {r Figure2, fig.width=6, fig.height=5}
library(ggplot2)
dat <- read.csv("Figure2_dat.csv")
nulldat <- read.csv("Figure2_nulldat.csv")
ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
 facet_wrap(~variable) + xlim(c(-1, 1)) + 
 geom_density(data=nulldat, aes(value), adjust=3) + xlab("Kendall's tau") + theme_bw()

```

``` {r Figure3, fig.width=6, fig.height=5}
library(ggplot2)
dat <- read.csv("Figure3_dat.csv")
nulldat <- read.csv("Figure3_nulldat.csv")
ggplot(dat) + geom_histogram(aes(value, y=..density..), binwidth=0.3, alpha=.5) +
 facet_wrap(~variable) + xlim(c(-1, 1)) + 
 geom_density(data=nulldat, aes(value), adjust=3) + xlab("Kendall's tau") + theme_bw()

```
